---
title: "06_CellTypesIslet_cells_Islet"
author: "Nicolas Damond, Nathan Steenbuck"
date: "Created: 24 May, 2022; Compiled: `r format(Sys.time(), '%d %b, %Y')`"
output: workflowr::wflow_html
editor_options:
  chunk_output_type: inline
---

```{r setup, include=FALSE}
if (rstudioapi::isAvailable()) {
  script_name <- basename(rstudioapi::getSourceEditorContext()$path)
} else {
  script_name <- knitr::current_input()
}

cur_user <- Sys.info()[["user"]]
script_name <- "06_CellTypesIslet_cells_Islet.Rmd"

if (cur_user == "ubuntu") {
    source(file.path("/", "mnt", "central_nas", "projects", 
    "type1_diabetes", "nathan", "T1D_Vol", "T1D_analysis", 
    "analysis", "helpers.R"))
  #n_cores <- future::availableCores() - 1
  n_cores <- 4
  future::plan(future::multicore(workers = n_cores))
  paths <- getPaths(script_name)
  knitr::opts_knit$set(root_dir = paths$home_data)
  do_print <- FALSE
} else {
  source(file.path("/", "home", "nsteen", "scripts", "T1D_analysis", "analysis", "helpers.R"))
  n_cores <- 8
  future::plan(future::multicore(workers = n_cores))
  paths <- getPaths(script_name)
  paths$folder_script <- paths$cluster_folder_script
  paths$folder_in <- paths$cluster_folder_in
  paths$folder_out <- paths$cluster_home 
  knitr::opts_knit$set(root_dir = paths$cluster_home)
  do_print <- TRUE
}

seed <- 123456
set.seed(seed)
options <- furrr::furrr_options(seed = seed)
paths$prev <- paste("05_CellCategories", paths$object_type, paths$panel_type, sep = "_")
knitr::opts_chunk$set(echo = TRUE)
```



# **Goal**

The goal of this script is to attribute consensus cell types based on the clustering results obtained at the previous step.  
Also, potential islet cells cells outside of the islet-mask are attributed to islet cell types using a random forest classifier 
trained on the islet cells inside the islet mask.  
Finally, segmentation scores are calculated for each islet cell type based on the expression of cell type-specific hormones versus non-specific hormones.

# **Settings**  

## Load packages

```{r packages, results='hide'}
suppressPackageStartupMessages(c(
  library(data.table),
  library(dplyr),
  library(SpatialExperiment),
  library(parallel)
))
```

## Paths and settings

```{r settings}
# Paths
if (!dir.exists(paths$folder_script)) dir.create(paths$folder_script)
plotsave_param$path <- paths$folder_script
plotsave_param_large$path <- paths$folder_script

# Misc settings
today <- gsub("-", "", Sys.Date())
```

##  Read in the data

Load the SpatialExperiment (SPE) object saved at the previous step.

```{r load-data}
fn_spe <- file.path(paths$folder_out, paths$prev, paste0(paths$object_type, "_", paths$panel_type, ".rds"))
spe <- readRDS(fn_spe)
print(spe)
```

## Methods and assays

Select channels (`channels_clust`), assays (`assay_sel`) and reduced dimensions (`dimred_sel`) to use.

```{r select-methods-assays}
methods_sel <- c("Pheno1", "Leiden1")

#assay_sel <- c("fastMNN", "scaled")
assay_sel <- c("fastMNN_case_id", "scaled")
#names(assay_sel) <- c("exprs", "scaled")
names(assay_sel) <- c("exprs", "exprs", "scaled")
writeLines(c("Assays:", assay_sel[assay_sel %in% assayNames(spe)]))

dimred_sel <- c("UMAP")
writeLines(c("\nReduced dimensions:", dimred_sel))

channels <- rownames(spe)[!(grepl("DNA|H3", rownames(spe)))]
```

## Subset SPE

```{r subset-islet-cells}
# Extract the metadata (to avoid conflicts when merging the SCEs)
meta <- metadata(spe)
spatial_coords <- spatialCoords(spe)
colpairs <- colPairs(spe)

# Split the SPE object (cells outside vs inside islets)
sce_isl <- spe[, spe$distance_to_islet >= 0]
sce_exo <- spe[, !(colnames(spe) %in% colnames(sce_isl))]

# Convert to SingleCellExperiment
sce_isl <- as(sce_isl, "SingleCellExperiment")
sce_exo <- as(sce_exo, "SingleCellExperiment")
remove(spe)
```



# **Consensus cell types**

Attribute consensus cell types for islet cells based on the clustering performed with the previous script.

## Packages

```{r packages2, results='hide'}
suppressPackageStartupMessages(c(
  library(ggplot2),
  library(ggalluvial),
  library(heatmaply),
  library(scuttle),
  library(ranger),
  library(cytomapper)
))
```


## Cell type concordance

Compare concordance of cell types attributed using different methods.

```{r celltypes-concordance}
celltype_cols <- colnames(colData(sce_isl))[grepl(paste0(
  "CellType_", methods_sel, collapse = "|"), colnames(colData(sce_isl)))]

names(celltype_cols) <- celltype_cols
celltype_cols <- celltype_cols[!celltype_cols %in% c("CellType_Pheno1_fastMNN_case_id")]

celltypes <- as.data.table(colData(sce_isl)[
  , c("cell_id", "case_id", "donor_type", "image_number", "cell_number", celltype_cols)])

ct_combination <- t(combn(unique(celltype_cols), 2))

for (i in (seq_along(ct_combination[, 1]))) {
  cat("\n", ct_combination[i, 1], " vs ", ct_combination[i, 2], "\n")
  
  ct_concord <- nrow(
    celltypes[get(ct_combination[i, 1]) == get(ct_combination[i, 2]), ]
  )
  ct_discord <- nrow(
    celltypes[get(ct_combination[i, 1]) != get(ct_combination[i, 2]), ]
  )
  cat("Number of cells with concordant cell types: ", ct_concord,
      "\nNumber of cells with discordant cell types: ", ct_discord,
      "\nPercentage of cells with concordant cell types:",
      100 * ct_concord / (ct_discord + ct_concord), "%", "\n")
}
```
The same assays have a high concordance.
Scaled assay: Pheno vs Leiden (95 % concordance).

Different assays have a lower concordance.
Pheno: scaled vs fastMNN: 83 %
Across clust/assay have similar lower concordance:
Pheno-fastMNN vs Leiden-scaled: 84 %

## Attribute consensus cell types

### Majority vote

Cells for which 3 out of 4 methods agree on the cell type are attributed this cell type, other cells as considered as "Ambiguous".

```{r consensus-celltype}
# Majority vote
func_majority_vote <- function(x) {
  ux <- unique(x)
  ux[which.max(tabulate(match(x, ux)))]
}

celltypes[, cell_type := apply(celltypes, 1, func_majority_vote)]
celltypes[, Paste := do.call(paste0, .SD),
          .SDcols = grep("CellType_", colnames(celltypes))]
## Count the number of times the majority-voted cell type appears in the Paste column
celltypes[, count := stringr::str_count(Paste, cell_type)]
celltypes[count < 2, cell_type := "Ambiguous"]
#celltypes[, `:=`(Paste = NULL, count = NULL)]

cat("Percentage of cells with an ambiguous cell type: ",
    100 * nrow(celltypes[cell_type == "Ambiguous"]) / nrow(celltypes),
    "%")

## Frequency of nr. of counts per cell type.
celltypes |> 
  as_tibble() |>
  group_by(cell_type)  |> 
  mutate(n_cell = n())  |> 
  ungroup() |> 
  group_by(cell_type, count, n_cell) |>
  # freq. per cell_type.
  summarise(n = n() / dplyr::first(n_cell)) |>
  print(n = "all")
```

Note: beta, alpha-cells are pretty robust. (~ 80, 70% counts = 5)
Gamma, delta, epsilon: less robust (45 % counts = 5 (or 4))
Non-islet: problematic! (only 20 % counts = 5)

### Rescue rare cells

Epsilon or gamma cells are sometimes too rare for the above methods to work. 
Usually they are best identified using scaled counts, so if these rare cells 
cannot be identified using different counts transformations, cell types are 
attributed to them based on scaled counts only.

```{r rescue-rare-populations}
# # Select cell type columns based on scaled counts
# cn_rare_celltypes <- colnames(colData(sce_isl))[grepl(paste0(
#   paste("cell_type", methods_sel, "scaled", sep = "_"), collapse = "|"),
#   colnames(colData(sce_isl)))]
# 
# # Attribute cell types
# celltypes[get(cn_rare_celltypes[1]) == "Epsilon" &
#             get(cn_rare_celltypes[2]) == "Epsilon", cell_type := "Epsilon"]
# 
# celltypes[get(cn_rare_celltypes[1]) == "Gamma" &
#             get(cn_rare_celltypes[2]) == "Gamma", cell_type := "Gamma"]
```

### Compare cell type attributions

```{r celltypes-alluvial, warning=FALSE}
celltype_cols <- c("cell_type", colnames(colData(sce_isl))[grepl(paste0(
  "CellType_", methods_sel, collapse = "|"), colnames(colData(sce_isl)))])

names(celltype_cols) <- celltype_cols
celltype_cols <- celltype_cols[!celltype_cols %in% c("CellType_Pheno1_fastMNN_case_id")]

p <- celltypes[, .(number = .N), by = celltype_cols] |>
  ggplot(aes(axis1 = get(celltype_cols[1]),
             axis2 = get(celltype_cols[2]),
             axis3 = get(celltype_cols[3]),
             axis4 = get(celltype_cols[4]),
             #axis5 = get(celltype_cols[3]),
             #axis6 = get(celltype_cols[4]),
             y = number)) +
  ggalluvial::geom_stratum() +
  geom_text(stat = "stratum", infer.label = TRUE) +
  ggalluvial::geom_alluvium(aes(fill = get(celltype_cols[1]))) +
  scale_x_discrete(limits = gsub("cell_type_", "", celltype_cols)) +
  scale_fill_manual(values = palettes$colors[c(TRUE, FALSE)],
                    name = "Cell Type") +
  mytheme$standard() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
if (do_print) {
  print(p)
}

fn <- paste0(paste(today, "CellTypes", "alluvial", sep = "_"), ".png")
do.call(ggsave, c(list(fn, p), plotsave_param))
```

### Number of cells by cell type

```{r cells-per-celltype}
tibble::as_tibble(celltypes) |>
  dplyr::select(cell_type) |>
  dplyr::group_by(cell_type) |>
  add_count(name = "Number of cells") |>
  distinct()
```


## Add celltyes to islet SPE

```{r add-celltypes-spe-islet}
celltypes <- celltypes[order(match(celltypes$cell_id, colnames(sce_isl))), ]
colData(sce_isl)$cell_type <- celltypes$cell_type
```



# **Islet cells outside islet masks**

Some islet cells may found themselves outside the islet masks generated during preprocessing. 
In that case, their distance to islet is smaller than 0 and they are not considered for clustering in the previous script.
Here, we use a random forest classifier to attribute islet cell types to cells with a distance to islet < 0. 
Training is performed on the `sce_isl` object and cell types are predicted for cells in the `sce_exo` object. 
As above, final cell types are attributed based on a majority vote.

## Train classifier

```{r random-forest-train-predict}
channels_clust <- rownames(rowData(sce_isl)[rowData(sce_isl)$clustering == 1, ])
n_features <- length(channels_clust)

assay_sel <- c("fastMNN_case_id", "scaled")
names(assay_sel) <- c("exprs", "scaled")

# Vector to map cell types to integers
celltype_map <- c("Alpha" = 1L, "Beta" = 2L, "Delta" = 3L, "Gamma" = 4L,
                  "Epsilon" = 5L, "NonIslet" = 6L, "Ambiguous" = 7L)
celltype_map_inv <- setNames(names(celltype_map), celltype_map)

purrr::walk(assay_sel, \(cur_assay) {
  purrr::walk(methods_sel, \(cur_method) {
    clust_name <- paste("CellType", cur_method, cur_assay, sep = "_")
    print(clust_name)

    if (paste(cur_method, cur_assay, sep = "_") != "Pheno1_fastMNN_case_id") {  

      # Define training (`sce_isl`) and test data (`sce_exo`)
      #training_data <- cbind(data.frame(t(assay(sce_isl, cur_assay))),
      #                       cluster = sce_isl[[clust_name]])
      if (cur_assay %in% assayNames(sce_isl)) {
        training_data <- dplyr::bind_cols(t(assay(sce_isl, cur_assay)),
                                            cluster = sce_isl[[clust_name]])
        test_data <- dplyr::bind_cols(t(assay(sce_exo, cur_assay)),
                                        cluster = sce_exo[[clust_name]])
      } else {
        training_data <- dplyr::bind_cols(reducedDim(sce_isl, cur_assay),
                                          cluster = sce_isl[[clust_name]])
        test_data <- dplyr::bind_cols(reducedDim(sce_exo, cur_assay),
                                      cluster = sce_exo[[clust_name]])
      }

      training_data$cluster <- celltype_map[training_data$cluster]
      test_data$cluster <- celltype_map[test_data$cluster]

      # Train
      ranger_model <- ranger(
        formula = cluster ~ .,
        data = training_data,
        mtry = floor(n_features / 3),
        num.trees = 10 * n_features,
        classification = TRUE,
        num.threads = n_cores,
        seed = seed)

      # Predict
      rf_predict <- predict(ranger_model, test_data)
      sce_exo[[clust_name]] <<- rf_predict$predictions
      sce_exo[[clust_name]] <<- celltype_map_inv[sce_exo[[clust_name]]]
    }
  })
})
remove(training_data, test_data, ranger_model, rf_predict)
```

```{r save-sce-exo}
fn_spe_exo <- file.path(paths$folder_script, paste0(paths$object_type, "_", paths$panel_type, "_exo.rds"))
saveRDS(sce_exo, fn_spe_exo)
```

## Cell type concordance

Compare concordance of cell types attributed using different methods.

```{r celltypes-concordance-exo}
celltype_cols <- colnames(colData(sce_exo))[grepl(paste0(
  "CellType_", methods_sel, collapse = "|"), colnames(colData(sce_exo)))]

celltypes <- as.data.table(colData(sce_exo)[
  , c("cell_id", "case_id", "image_number", "cell_number", celltype_cols)])

ct_combination <- t(combn(unique(celltype_cols), 2))


for (i in seq_along(ct_combination[, 1])) {
  cat("\n", ct_combination[i, 1], " vs ", ct_combination[i, 2], "\n")
  
  ct_concord <- nrow(
    celltypes[get(ct_combination[i, 1]) == get(ct_combination[i, 2]), ]
  )
  ct_discord <- nrow(
    celltypes[get(ct_combination[i, 1]) != get(ct_combination[i, 2]), ]
  )
  cat("Number of cells with concordant cell types: ", ct_concord,
      "\nNumber of cells with discordant cell types: ", ct_discord,
      "\nPercentage of cells with concordant cell types:",
      100 * ct_concord / (ct_discord + ct_concord), "%", "\n")
}
```

## Attribute consensus cell types

### Majority vote

Cells for which 3 out of 4 methods agree on the cell type are attributed this cell type, other cells as considered as "Ambiguous".

```{r consensus-celltype-exo}
# Majority vote
## Count the number of times the majority-voted cell type appears in the Paste column
celltypes[, cell_type := apply(celltypes, 1, func_majority_vote)]
celltypes[, Paste := do.call(paste0, .SD),
          .SDcols = grep("CellType_", colnames(celltypes))]
celltypes[, count := stringr::str_count(Paste, cell_type)]
celltypes[count < 2, cell_type := "Ambiguous"]
#celltypes[, `:=`(Paste = NULL, count = NULL)]

cat("Percentage of cells with an ambiguous cell type: ",
    100 * nrow(celltypes[cell_type == "Ambiguous"]) / nrow(celltypes),
    "%")
```


### Compare cell type attributions

```{r celltypes-alluvial-exo, warning=FALSE}
celltype_cols <- c("cell_type", colnames(colData(sce_exo))[grepl(paste0(
  "CellType_", methods_sel, collapse = "|"), colnames(colData(sce_exo)))])

names(celltype_cols) <- celltype_cols
celltype_cols <- celltype_cols[!celltype_cols %in% c("CellType_Pheno1_fastMNN_case_id")]

p <- celltypes[, .(number = .N), by = celltype_cols] |>
  ggplot(aes(axis1 = get(celltype_cols[1]),
             axis2 = get(celltype_cols[2]),
             axis3 = get(celltype_cols[3]),
             axis4 = get(celltype_cols[4]),
             #axis5 = get(celltype_cols[4]),
             #axis6 = get(celltype_cols[6]),
             y = number)) +
  ggalluvial::geom_stratum() +
  ggplot2::geom_text(stat = "stratum", infer.label = TRUE) +
  ggalluvial::geom_alluvium(aes(fill = get(celltype_cols[1]))) +
  ggplot2::scale_x_discrete(limits = gsub("cell_type_", "", celltype_cols)) +
  ggplot2::scale_fill_manual(values = palettes$colors[c(TRUE, FALSE)],
                             name = "Cell Type") +
  mytheme$standard() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
if (do_print) {
  print(p)
}
fn <- paste0(paste(today, "CellTypes", "alluvial", "exo", sep = "_"), ".png")
do.call(ggsave, c(list(fn, p), plotsave_param))
```

### Number of cells by cell type

```{r cells-per-celltype-exo}
tibble::as_tibble(celltypes) |>
  dplyr::select(cell_type) |>
  dplyr::group_by(cell_type) |>
  dplyr::add_count(name = "Number of cells") |>
  dplyr::distinct()
```


## Add consensus cell types for all cells

### Add celltyes to exo SPE

```{r add-celltypes-spe-exo}
celltypes <- celltypes[order(match(celltypes$cell_id, colnames(sce_exo))), ]
colData(sce_exo)$cell_type <- celltypes$cell_type
```

### Merge objects

Merge the "islet" and "non-islet" SCE objects, convert to SPE.

```{r merge-spe}
# Merge and re-order
spe <- cbind(sce_isl, sce_exo)
spe <- spe[, order(colData(spe)$case_id,
                   colData(spe)$image_number,
                   colData(spe)$cell_number)]

# Fix sample_id and convert to SPE
spe$sample_id <- spe$image_id
spe <- as(spe, "SpatialExperiment")

# Add the saved metadata, spatial coordinates and colPairs
metadata(spe) <- meta
spatialCoords(spe) <- spatial_coords
colPairs(spe) <- colpairs
remove(sce_isl, sce_exo)
```

### Temp save SPE

```{r save-spe-1}
fn_spe <- file.path(paths$folder_script, paste0(paths$object_type, "_", paths$panel_type, ".rds"))
saveRDS(spe, fn_spe)
# spe <- readRDS(fn_spe)
```

### Add cell categories

```{r add-cellcat}
islet_celltypes <- c("Alpha", "Beta", "Gamma", "Delta", "Epsilon")
spe$cell_category <- ifelse(spe$cell_type %in% islet_celltypes,
                            "islet", "other")
```



# **Plot cell types**

## Reduced dimensions

Plot consensus cell types on reduced dimensions

```{r consensus-dimred}
ct <- c("beta", "alpha", "gamma", "delta", "epsilon", "ambiguous", "non-islet")
names(ct) <- c("Beta", "Alpha", "Gamma", "Delta", "Epsilon",
               "Ambiguous", "NonIslet")
palette_ct <- c(palettes$colors[1:(length(ct) - 2)], "grey40", "grey")
names(palette_ct) <- names(ct)


cur_dat <- scuttle::makePerCellDF(spe[, colnames(spe) %in% metadata(spe)$subset],
                                  use_dimred = TRUE) |>
  dplyr::mutate(case_id = factor(case_id, levels = metadata(spe)$cases),
                donor_type = factor(donor_type, levels = metadata(spe)$stages),
                cell_type = factor(cell_type, levels = names(ct))) |>
  dplyr::arrange(cell_type, case_id, donor_type)

purrr::walk(assay_sel, \(cur_assay) {
  purrr::walk(dimred_sel, \(cur_dimred) {
    dimred_name <- paste(cur_dimred, cur_assay, sep = "_")
    
    p <- plot_dim_red(cur_dat, dimred_name, "cell_type",
                      sample = TRUE, size = 0.1, alpha = 1,
                      palette = palette_ct)
    if (do_print) {
      print(p)
    }
    fn <- paste0(paste(today, "CellType", "Consensus", dimred_name,
                       sep = "_"), ".png")
    do.call(ggsave, c(list(fn, p), plotsave_param))
  })
})
remove(cur_dat)
```

## Heatmap

```{r consensus-heatmap}
hmlist <- list()
purrr::walk(assay_sel, \(cur_assay) {
  clust_name <- paste("cell_type", cur_assay, sep = "_")
  
  # Summarize the data
  hm <- summarize_heatmap(spe,
                          expr_values = cur_assay,
                          cluster_by = "cell_type",
                          channels = channels)
  
  # Display the heatmap
  fn <- paste0(paste(today, "CellType", "Consensus", cur_assay,
                     "Heatmap", sep = "_"), ".html")
  
  hmlist[[i]] <- heatmaply(
    heatmaply::normalize(hm), main = clust_name,
    file = file.path(paths$folder_script, fn))
})
htmltools::tagList(setNames(hmlist, NULL))
```


## Correlation between channels

```{r celltype-correlation}
# Define the correlation method
cor_method <- "pearson"

# Function to calculate p values
cor_test_p <- function(x) {
  FUN <- function(x, y) cor.test(x, y, method = cor_method)[["p.value"]]
  z <- outer(
    colnames(x), 
    colnames(x), 
    Vectorize(function(i, j) FUN(x[, i], x[, j]))
  )
  dimnames(z) <- list(colnames(x), colnames(x))
  z
}

hmlist <- list()
purrr::walk(assay_sel, \(cur_assay) {
  # Summarize the data
  hm <- summarize_heatmap(spe,
                          expr_values = cur_assay,
                          cluster_by = "cell_type",
                          channels = channels)
  
  # Run the correlation test
  r <- cor(hm, method = cor_method)
  p <- cor_test_p(hm)
  diag(p) <- NA
  
  # Plot and save the heatmap
  fn <- paste0(paste(today, "CellTypes", "Consensus", "Correlation",
                     cur_assay, sep = "_"), ".html")
  
  hmlist[[i]] <- heatmaply_cor(
    r, node_type = "scatter",
    point_size_mat = -log10(p), point_size_name = "-log10(p-value)",
    label_names = c("x", "y", "Correlation"),
    main = paste("Cell type -", cur_assay),
    file = file.path(paths$folder_script, fn))
})
htmltools::tagList(setNames(hmlist, NULL))
```

## Images

Load random images and masks

```{r consensus-viz-load}
nb_images <- 14
image_extension <- ".tiff"

# Folders
folder_images <- file.path(paths$folder_in, "img", paths$panel_type)
folder_masks <- file.path(paths$folder_in, "masks_cells",
                          paths$panel_type, "whole-cell")

# Select random images and the associated masks
set.seed(seed)
image_list <- list.files(folder_images)
image_list <- mask_list <- sort(c(sample(unique(image_list), nb_images))) #"6447_Islet_ROI_055.tiff", "6437_Islet_ROI_054.tiff", "6437_Islet_ROI_055.tiff", "6126_Islet_ROI_066.tiff", 
writeLines(c("Selected images:", image_list))

# Load images and masks
images <- yoloader(
  x = spe,
  image_dir = folder_images,
  image_names = image_list,
  type = "stacks"
)

masks <- yoloader(
  x = spe,
  image_dir = folder_masks,
  image_names = mask_list,
  as.is = TRUE,
  type = "masks"
)

if (length(setdiff(mcols(images)$ImageName,
                   mcols(masks)$ImageName)) != 0) {
  print("mcols(images)$ImageName and mcols(masks)$ImageName should have the same values")
}

# Subset the SPE object
spe_sub <- spe[, spe$image_fullname %in% image_list]
spe_sub$ImageName <- gsub(image_extension, "", spe_sub$image_fullname)
```

Plot the images

```{r viz-celltype-plot}
fn <- list(filename = file.path(
  plotsave_param$path,
  paste0(paste(today, "CellType", "Cytomapper", sep = "_"), ".png")))

color_key <- list(c(
  Alpha = "red", Beta = "blue",  Gamma = "yellow", Delta = "green",
  Epsilon = "orange", Ambiguous = "white", NonIslet = "grey30"))
names(color_key) <- "cell_type"
    
image_names <- c(unique(colData(spe_sub)[, c("ImageName", "case_id")])$case_id)

cytomapper::plotPixels(images, object = spe_sub, masks,
                       cell_id = "cell_number", img_id = "ImageName",
                       colour_by = c("ProINS", "ProGCG", "SST", "PPY", "INS", "GHRL"),
                       outline_by = "cell_type",
                       # image_title = list(text = image_names),
                       colour = color_key,
                       bcg = list(ProINS = c(0, 3, 1), ProGCG = c(0, 7, 1), SST = c(0, 5, 1),
                                  PPY = c(0, 3, 1), INS = c(0, 3, 1), GHRL = c(0, 10, 1)),
                       image_title = NULL, scale = TRUE, save_plot = fn)
```

### Plot images

Images are saved to disk but not shown.
FIXME: Comment out when running on the cluster.
```{r viz-cluster-plot-interactive}
#channels_view <- channels_clust
#sub_images <- cytomapper::getChannels(images, channels_view)
# Use cytoviewer with images, masks and object
#library(cytoviewer)
#app <- cytoviewer(image = sub_images, 
#                  mask = masks, 
#                  object = spe_sub[channels_view, ], 
#                  img_id = "ImageName", 
#                  cell_id = "cell_number")

#if (interactive()) {
#  shiny::runApp(app)
#}
```

# **Segmentation score**

For each cell type, we measure the expression level of the cell type-specific hormone (e.g. glucagon in alpha cells) 
versus the non-specific hormones (e.g. insulin and somatostatin in alpha cells). 
The ratio of specific vs non-specific hormone expression is then calculated, which gives a score for each islet cell type.
The final score is obtained by weighting the scores of each cell type by the fraction of this cell type in the total number of islet cells.  

## Parameters

```{r segscore-parameters}
# Cell types and channels
ct <- c("beta", "alpha", "gamma", "delta", "nonislet", "epsilon")
names(ct) <- c("Beta", "Alpha", "Gamma", "Delta", "NonIslet", "Epsilon")
hormone_channels <- c("ProINS", "ProGCG", "PPY", "SST", "CAV1", "GHRL")

#spe_isl <- spe[, spe$cell_category == "islet"]
# spe_exo <- spe[, spe$cell_category == "other"]
spe_isl <- spe[, spe$distance_to_islet > 0]

celltype_cols <- c("cell_type", colnames(colData(spe_isl))[grepl(paste0(
  "CellType_", methods_sel, collapse = "|"), colnames(colData(spe_isl)))])


# Plotting theme
theme_segm <- list(
  geom_jitter(size = 0.1, aes(color = as.factor(channel))),
  geom_violin(scale = "width", draw_quantiles = 0.5, alpha = 0.5),
  xlab(NULL),
  ylab("Scaled counts"),
  theme(
    plot.title = element_text(size = 10),
    legend.position = "none",
    axis.title.y = element_text(size = 8),
    axis.text.x = element_text(size = 6),
    axis.text.y = element_text(size = 6))
)

# Plotting function
plotSegScore <- function(x) {
  segscore[segscore$CT == names(ct)[x] &
             segscore$channel %in% hormone_channels, ] |>
    ggplot(aes(x = channel, y = counts)) +
    ggtitle(paste(names(ct)[x], "Cell Score:",
                  round(score.table[CT == names(ct)[x]]$score, digits = 3))) +
    mytheme$standard() +
    theme_segm
}
```


## Calculate segmentation score

Calculate segmentation scores for each islet cell type and plot the results

```{r segscore-calculate, warning=FALSE}
# Prepare the data
dat <- scuttle::makePerCellDF(spe_isl,
                              use.dimred = FALSE,
                              assay.type = "scaled",
                              features = hormone_channels)

dat <- dat[, c(hormone_channels, celltype_cols)] |>
  tidyr::pivot_longer(all_of(hormone_channels),
                      names_to = "channel", values_to = "counts")

for (k in seq_along(celltype_cols)) {
  segscore <- dat |>
    dplyr::mutate(CT = get(celltype_cols[k])) |>
    dplyr::select(CT, channel, counts) |>
    dplyr::filter(CT %in% names(ct)) |>
    dplyr::mutate(CT = factor(CT, levels = names(ct))) |>
    dplyr::mutate(channel = factor(channel, levels = hormone_channels)) |>
    dplyr::arrange(CT, channel)
  
  # Number of cells per cell type
  nb.cells <- segscore |> 
    dplyr::group_by(CT) |> 
    dplyr::summarise(n = n()) |>
    dplyr::mutate(tot = sum(n)) |>
    dplyr::mutate(fract = n / tot) |>
    as.data.table()
  
  # Calculate segmentation score
  score.table <- segscore |>
    dplyr::group_by(CT, channel) |>
    dplyr::summarise(expr = mean(counts)) |>
    tidyr::spread(key = channel, value = expr) |>
    as.data.table()
  
  score.table[, ":="(mean = 0, score = 0, w_score = 0)] 

  ## Calculate the mean expression level of cell type-specific hormones
  for (i in seq_len(nrow(score.table))) {
    for (j in 1:(ncol(score.table) - 2)) {
      if (i == j) (score.table[[i, "mean"]] <- score.table[[i, (j + 1)]])
    }
  }
  
  ## Calculate the segmentation score for each cell type
  for (i in seq_len(nrow(score.table))) {
    for (j in 1:(ncol(score.table) - 4)) {
      score.table[[i, "score"]] <-
        ifelse(i == j,
              score.table[[i, "score"]] + (score.table[[i, (j + 1)]] / score.table[[i, "mean"]]),
              score.table[[i, "score"]] - (score.table[[i, (j + 1)]] / score.table[[i, "mean"]])
        )
    }
  }

  
  ## Calculate weighted score
  score.table <- merge(score.table, nb.cells, by = "CT")
  ## Calculate the segmentation score for each cell type
  #if (celltype_cols[k] != "CellType_Pheno1_fastMNN_case_id") {
  #  for (i in seq_len(nrow(score.table))) {
  #    for (j in 1:(ncol(score.table) - 7)) {
  #      score.table[[i, "w_score"]] <- 
  #              score.table[[i, "w_score"]] + ((score.table[[i, "mean"]] * score.table[[i, "fract"]]) / (score.table[[i, (j + 1)]] * score.table[[j, "fract"]]) 
  #        )
  #    }
  #  }
  #}

  score.table[, weighted.score := score * fract]
  
  final.score <- sum(score.table$weighted.score)
  
  cat("\n", celltype_cols[k], "\n")
  print(score.table[, .(CT, n, score, weighted.score)]) #, w_score
  
  # Plot the results
  p <- list()
  p <- lapply(seq_along(ct), plotSegScore)
  
  g <- gridExtra::grid.arrange(grobs = p, nrow = 2, top = paste(
    "Segmentation score:", celltype_cols[k], round(final.score, digits = 3)
  ))
  
  fn <- paste0(paste(today, "SegScore", celltype_cols[k], sep = "_"), ".png")
  do.call(ggsave, c(list(fn, g), plotsave_param_large))
}
```



# **Save the updated SPE object**

Contains consensus islet cell types.

```{r save-spe}
print(spe)
fn_spe <- file.path(paths$folder_script, paste0(paths$object_type, "_", paths$panel_type, ".rds"))
saveRDS(spe, fn_spe)
```
