---
title: "06_CellTypesImmune_cells_Immune"
author: "Nicolas Damond, Nathan Steenbuck"
date: "Created: 08 June, 2022; Compiled: `r format(Sys.time(), '%d %b, %Y')`"
output: workflowr::wflow_html
editor_options:
  chunk_output_type: inline
---
```{r setup, include=FALSE}
if (rstudioapi::isAvailable()) {
  script_name <- basename(rstudioapi::getSourceEditorContext()$path)
} else {
  script_name <- knitr::current_input()
}

cur_user <- Sys.info()[["user"]]
script_name <- "06_CellTypesImmune_cells_Immune.Rmd"

if (cur_user == "ubuntu") {
  source(file.path("/", "mnt", "central_nas", "projects", 
    "type1_diabetes", "nathan", "T1D_Vol", "T1D_analysis", 
    "analysis", "helpers.R"))
  #n_cores <- future::availableCores() - 1
  n_cores <- 4
  future::plan(future::multicore(workers = n_cores))
  paths <- getPaths(script_name)
  knitr::opts_knit$set(root_dir = paths$home_data)
  do_print <- FALSE
} else {
  source(file.path("/", "home", "nsteen", "scripts", "T1D_analysis", "analysis", "helpers.R"))
  n_cores <- 8
  future::plan(future::multicore(workers = n_cores))
  paths <- getPaths(script_name)
  paths$folder_script <- paths$cluster_folder_script
  paths$folder_in <- paths$cluster_folder_in
  paths$folder_out <- paths$cluster_home 
  knitr::opts_knit$set(root_dir = paths$cluster_home)
  do_print <- TRUE
}

seed <- 123456
set.seed(seed)
options <- furrr::furrr_options(seed = seed)
paths$prev <- paste("05_CellCategories", paths$object_type, paths$panel_type, sep = "_")
knitr::opts_chunk$set(echo = TRUE)
```



# **Goal**

In scripts 05 to 08, cell types are attributed to all cells in the dataset
in an iterative way:

- `05_CellCategories_cells_Immune`: Attribution of main cell categories 
(immune, islet, exocrine, ...).
- `06_CellTypesImmune_cells_Immune`(this script): Attribution of main cell 
types to immune cells (Lymphocyte, Myeloid, Neutrophil).
- `07_ClustersImmune_cells_Immune`: Further separation of immune cell types 
into clusters.
- `08_CellTypesNonImmune_cells_Immune`: Attribution of cell types
to non-immune cells (Beta, Alpha, Exocrine, ...).
- `08_CellTypesIslet_cells_Immune`: Re-attribution of islet cells
to cell types based on clustering results from the previous step.
- `08_CellTypesIslet_cells_ImmuneII`: Attribution of rare & specific immune cell clusters
like T-reg and T-EMRAs.

Specifically, the goal of the current script is to attribute immune cell types 
to cells that have been attributed the `immune` category in the previous 
script.

This is performed in several steps:

1. Unsupervised clustering using two approaches (*i)* PhenoGraph and (*ii)* 
Shared-nearest neighbors (SNN) + Leiden community detection on two different 
count transformations (`exprs` and `scaled`). Only immune cell markers are used
for clustering.

2. For each of the four clustering results (two approaches X two count 
transformations), clusters are manually merged into cell categories based on
marker expression.

3. Consensus cell categories are attributed by majority vote: if the cell 
type is the same for three or more of the four approaches, this category is 
attributed to the cell. If not, the cell is categorized as `ambiguous`.

Concretely, immune cells are divided in the following "cell types": 
`Lymphocyte`, `Myeloid`, `Neutrophil`, and `Other`. These cell types are stored
in the `colData(spe)$cell_type` column.

In the next script, Lymphocytes will be further separated into more specific
cell types (`T_CD4`, `B`, `NK`, ...).


# **Settings**  

## Load packages

```{r packages, results='hide'}
suppressPackageStartupMessages(c(
  library(data.table),
  library(dplyr),
  library(SpatialExperiment),
  library(parallel),
  library(purrr),
  library(furrr)
))
```

## Paths and settings

```{r settings}
# Paths
if (!dir.exists(paths$folder_script)) dir.create(paths$folder_script)
plotsave_param$path <- paths$folder_script
plotsave_param_large$path <- paths$folder_script

# Misc settings
today <- gsub("-", "", Sys.Date())
```

##  Read in the data

Load the SpatialExperiment (SPE) object saved at the previous step.

```{r load-data}
fn_spe <- file.path(paths$folder_out, paths$prev, paste0(paths$object_type, "_", paths$panel_type, ".rds"))
spe <- readRDS(fn_spe)
print(spe)
```


## Methods and assays

Select channels (`channels_clust`), assays (`assay_sel`) and clustering methods (`methods_sel`) to use.

```{r select-methods-assays}
methods_sel <- c("Pheno2", "Leiden2")

assay_sel <- c("fastMNN_case_id", "scaled")
names(assay_sel) <- c("exprs", "scaled")
writeLines(c("Assays:", assay_sel[assay_sel %in% assayNames(spe)]))

dimred_sel <- c("UMAP") # , "tSNE")
writeLines(c("\nReduced dimensions:", dimred_sel))

channels <- rownames(spe)[!(grepl("DNA|H3", rownames(spe)))]
cat(c("\nChannels:", channels[channels %in% rownames(spe)]))
cat(c("\nNumber of channels:", length(channels)))
all_channels <- rownames(spe)
```

### Select the channels for clustering

Only immune cell markers are used for clustering.

```{r select-channels-clustering}
channels_clust <- c("MPO", "CD16", "CD11b", "HLA_DR", "CS", "CD57", "CD163",
                    "CD20", "CD45RA", "CD56", "CD54", "CD3e", "CD11c", "ATP5A",
                    "CD73", "CD7", "GranB", "LDHA", "TMEM173", "CD45RO", "HK1",
                    "CD27", "CD8a", "PD1", "FoxP3", "Arg1", "CD103", "CD4",
                    "CD204", "TIM3", "CD303", "CD206", "CD66b", "CD15")
cat(c("\nChannels used for unsupervised clustering:",
      channels_clust[channels_clust %in% rownames(spe)]))
```

## Extract immune cells

The clustering is performed only on immune cells (`sce_imm`).

```{r spe-extract}
# Extract the metadata (to avoid conflicts when merging the SCEs)
meta <- metadata(spe)
spatial_coords <- spatialCoords(spe)
colpairs <- colPairs(spe)
```

```{r extract-immune-cells}
# Subset the SPE (immune cells only)
sce_imm <- spe[channels_clust, spe$cell_category == "immune"]

# Convert to SingleCellExperiment
sce_imm <- as(sce_imm, "SingleCellExperiment")
```



# **Dimensionality reduction**

Dimensionality reduction is performed on immune cells only.


## Packages

Change the path to the FIt-SNE source file below if needed.  

```{r packages2, results='hide'}
suppressPackageStartupMessages(c(
  library(ggplot2),
  library(uwot),
  library(scuttle),
  library(viridis)
))
#source(file.path("", "home", "ubuntu", "Git", "FIt-SNE", "fast_tsne.R"),
#       chdir = TRUE)
```

## Run dimensionality reduction

### Remove existing dimreds

```{r remove-reduced-dims}
reducedDims(sce_imm) <- NULL
```

### Run UMAP

UMAP for every selected assay, only if there is no object with the same name in `reducedDimNames(sce_imm)`.  

```{r run-UMAP}
purrr::walk(assay_sel, \(cur_assay) {
  dimred_name <- paste("UMAP", "immune", cur_assay, sep = "_")
  print(dimred_name)
  
  # Run UMAP
  if ((!dimred_name %in% reducedDimNames(sce_imm)) & ("UMAP" %in% dimred_sel)) {
    cur_umap <- uwot::umap(t(assay(sce_imm, cur_assay)),
                             n_threads = n_cores)
    colnames(cur_umap) <- c("UMAP1", "UMAP2")
    rownames(cur_umap) <- colnames(assay(sce_imm, cur_assay))
    reducedDim(sce_imm, dimred_name) <<- cur_umap
  }
})
```

### Subset SCE

```{r subset-sce-imm}
## Select 90.000 cells for Leiden2-clustering
# Cells per case
nb_cells <- 1000

# Subset the SPE object (nb_cells per case)
set.seed(seed)
cell_subset <- as_tibble(cbind(rn = rownames(colData(sce_imm)),
                                case_id = colData(sce_imm)$case_id)) |> 
  group_by(case_id) |>
  sample_n(nb_cells, replace = TRUE) |>
  pull(rn) |>
  unique()

# Keep the subset cell ids in SPE metadata
meta[["subset_immune"]] <- sort(as.vector(cell_subset))
```

```{r subset-sce-imm2}
sce_sub <- sce_imm[channels_clust, colnames(sce_imm) %in% meta$subset_immune]
sce_not <- sce_imm[channels_clust, !(colnames(sce_imm) %in% colnames(sce_sub))]
```


## Plot marker expression

```{r plot-dimred-channels, fig.height=10, fig.width=15}
purrr::walk(assay_sel, \(cur_assay) {
  purrr::walk(dimred_sel, \(cur_dimred) {
    dimred_name <- paste(cur_dimred, "immune", cur_assay, sep = "_")

    # Prepare the data
    cur_dat <- scuttle::makePerCellDF(sce_sub,
                                      features = channels_clust,
                                      exprs_values = cur_assay,
                                      use_dimred = TRUE) |>
      dplyr::select(c(all_of(channels_clust), 
      paste0(dimred_name, ".1"), paste0(dimred_name, ".2")))

    cur_dat <- cur_dat |>
      tidyr::pivot_longer(cols = all_of(channels_clust),
                          names_to = "channel",
                          values_to = cur_assay)
    
    # Plot marker expression
    p <- plot_dim_red_channels(cur_dat, dimred_name, cur_assay,
                               channels) # , force.points = TRUE)

    fn <- paste0(paste(today, dimred_name, "Channels", sep = "_"), ".png")
    do.call(ggsave, c(list(fn, p), plotsave_param_large))
    if (do_print) print(p)
  })
})
```



# **Clustering**

Unsupervised clustering is performed by generating a nearest-neighbor graph followed by community detection (using either Louvain or Leiden algorithms).

## Settings

### Load packages

```{r packages3, results='hide'}
suppressPackageStartupMessages(c(
  library(scater),
  library(scran),
  library(igraph),
  library(Rphenoannoy),
  library(clustree),
  library(patchwork),
  library(ranger),
  library(foreach),
  library(doParallel),
  library(BiocParallel)
))
```


## **Clustering: PhenoGraph**

Unsupervised clustering is performed with the [PhenoGraph](https://doi.org/10.1016/j.cell.2015.05.047) algorithm. 
This method works by generating a nearest-neighbor (kNN) graph of phenotpyic similarities followed by Louvain community dectection. 
For speed, we run PhenoGraph using the [Rphenoannoy implementation](https://github.com/stuchly/Rphenoannoy).

```{r phenograph-run}
clust_method <- c("Pheno2")

# Number of nearest-neighbors
k <- 25

purrr::walk(assay_sel, \(cur_assay) {
  clust_name <- paste(clust_method, cur_assay, sep = "_")
  writeLines(c("\n", clust_name))

  # Run Phenograph
  set.seed(seed)
  cur_pheno_annoy <- Rphenoannoy::Rphenoannoy(t(assay(sce_imm, cur_assay)), k = k)

  cur_pheno <- DataFrame(cur_pheno_annoy[[2]]$membership)
  colnames(cur_pheno) <- clust_name
  rownames(cur_pheno) <- colnames(assay(sce_imm, cur_assay))

  # Add Phenograph clusters to the colData of the SCE object
  colData(sce_imm)[, clust_name] <<- cur_pheno
  remove(cur_pheno)
})
```

```{r save_pheno}
fn_spe <- file.path(paths$folder_script, paste0(paths$object_type, "_", paths$panel_type, "temp_Pheno.rds"))
saveRDS(sce_imm, fn_spe)
# sce_imm_test <- readRDS(fn_spe)
```
## **Clustering: SNN + Leiden**

Here, we first construct a shared nearest-neighbor graph and identify clusters using the Leiden community detection algorithm.

### Run clustering

```{r SNN-leiden-run}
clust_method <- c("Leiden2")
do_leiden <- FALSE

# Number of nearest-neighbors for SNN graph
k <- 15

# Resolutions for Leiden algorithm
resolutions <- c(0.000005, 0.0000075, 0.00001, 0.000025,
                  0.00005,  0.000075,  0.0001,  0.00025,
                   0.0005,   0.00075)


purrr::walk(assay_sel, \(cur_assay) {
  clust_name <- paste(clust_method, cur_assay, sep = "_")
  writeLines(c("\n", clust_name))
  
  # if (!clust_name %in% colnames(colData(sce_imm))) {
  if (!clust_name %in% colnames(colData(sce_sub))) {
    do_leiden <- TRUE
    
    # Build the SNN graph
    set.seed(seed)
     cur_graph <- scran::buildSNNGraph(sce_sub, type = "jaccard", k = k,
    #cur_graph <- scran::buildSNNGraph(sce_imm, type = "jaccard", k = k,
                                       assay.type = cur_assay, 
                                       BPPARAM = MulticoreParam(workers = n_cores))
    
    # Leiden clustering.
    set.seed(seed)
    clusters <- furrr::future_map(resolutions, \(cur_resolution) {
      cur_leiden <- igraph::cluster_leiden(
        cur_graph,
        objective_function = "CPM", # could change for additional scalability.
        resolution_parameter = cur_resolution)

      cur_leiden <- cur_leiden$membership
    }, .options = options)

    # Add clusters to colData
    for (k in seq_along(resolutions)) {
      res_name <- paste(clust_method, cur_assay, k, sep = "_")
      colData(sce_sub)[, res_name] <<- clusters[[k]]
      #colData(sce_imm)[, res_name] <<- clusters[[k]]
    }
    
    # Plot cluster tree
    p <- clustree::clustree(sce_sub, prefix = paste0(clust_method, "_", cur_assay, "_"))
    # p <- clustree::clustree(sce_imm, prefix = paste0(clust_method, "_", cur_assay, "_"))

    if (do_print) print(p)
    
    # Save
    fn <- paste0(paste(today, clust_method, cur_assay, "ClusterTree", sep = "_"), ".png")
    do.call(ggsave, c(list(fn, p), plotsave_param))
    remove(cur_graph, clusters)
  }
})
```
```{r save_leiden}
fn_spe <- file.path(paths$folder_script, paste0(paths$object_type, "_", paths$panel_type, "temp_leiden.rds"))
saveRDS(sce_imm, fn_spe)
```

### Optimal cluster resolution

Based on the trees above, select the optimal cluster resolution. Columns corresponding to other resolutions are removed from `colData(sce_sub)`.

```{r SNN-leiden-select}
do_leiden <- TRUE

if (do_leiden) {
  #optimal_res <- c("fastMNN" = 7, "scaled" = 6)
  optimal_res <- c("fastMNN_case_id" = 8, "scaled" = 8)

  purrr::walk(assay_sel, \(cur_assay) {
    clust_name <- paste(clust_method, cur_assay, sep = "_")

    optimal_col <- paste(clust_method, cur_assay, optimal_res[cur_assay], sep = "_")
    colData(sce_sub)[, clust_name] <<- colData(sce_sub)[, optimal_col]
    colData(sce_not)[, clust_name] <<- 0

    other_cols <- colnames(colData(sce_sub))[grep(paste0(clust_name, "_"), colnames(colData(sce_sub)))]
    colData(sce_sub)[, other_cols] <<- NULL
  })
}
gc()
```

```{r save_leiden2}
fn_spe <- file.path(paths$folder_script, paste0(paths$object_type, "_", paths$panel_type, "sce_sub_temp.rds"))
saveRDS(sce_sub, fn_spe)
```

## **Train classifier**

Clustering is performed on a cell subset. A random forest classifier is used to attribute clusters to the other cells.

### Random forest

Train a random forest classifier to attribute clusters to cells that were not subsetted.

```{r random-forest-train-predict}

if (do_leiden) {
  n_features <- length(channels_clust)

  purrr::walk(assay_sel, \(cur_assay) {  
    #purrr::walk(methods_sel, \(cur_method) {
    cur_method <- "Leiden2"
      # cur_assay <- assay_sel[1]; cur_method <- methods_sel[1]
    clust_name <- paste(cur_method, cur_assay, sep = "_")
    print(clust_name)

    # Define training (`sce_sub`) and test data (`sce_not`)
    training_data <- dplyr::bind_cols(t(assay(sce_sub, cur_assay)),
                                        cluster_x = sce_sub[[clust_name]])
    df_data <- t(assay(sce_not, cur_assay))

    # Changed to dplyr for scalability.
    test_data <- dplyr::bind_cols(df_data, cluster_x = sce_not[[clust_name]])

    # Train
    ranger_model <- ranger::ranger(
      formula = cluster_x ~ .,
      data = training_data,
      importance = "impurity",
      mtry = floor(n_features / 3),
      num.trees = 10 * n_features,
      classification = TRUE,
      num.threads = n_cores,
      seed = seed)

    # Predict
    rf_predict <- predict(ranger_model, test_data, seed = seed)
    sce_not[[clust_name]] <<- rf_predict$predictions
  })
} 
```

### Original vs. predicted

Compare predicted clusters and original clusters on train set (cells on which the clustering algorithms were applied).

```{r random-forest-compare, fig.width=15, fig.height=10}
dat_sub <- scuttle::makePerCellDF(sce_sub, use_dimred = TRUE) |>
  dplyr::mutate(case_id = factor(case_id, levels = meta$cases),
                donor_type = factor(donor_type, levels = meta$stages)) |>
  dplyr::arrange(case_id, donor_type)

dat_not <- scuttle::makePerCellDF(sce_not, use_dimred = TRUE) |>
  dplyr::mutate(case_id = factor(case_id, levels = meta$cases),
                donor_type = factor(donor_type, levels = meta$stages)) |>
  dplyr::arrange(case_id, donor_type)
dat_not <- dat_not[sample(rownames(dat_not), nrow(dat_sub)), ]

purrr::walk(assay_sel, \(cur_assay) {
  #purrr::walk(methods_sel, \(cur_method) {
  cur_method <- "Leiden2"
    dimred_name <- paste("UMAP", "immune", cur_assay, sep = "_")
    clust_name <- paste(cur_method, cur_assay, sep = "_")

    # Plot all clusters
    p_sub <- plot_dim_red(dat_sub, dimred_name, clust_name,
                          sample = TRUE, size = 0.1, alpha = 1) +
      ggtitle(paste(clust_name, "Original", sep = " - "))

    p_not <- plot_dim_red(dat_not, dimred_name, clust_name,
                          sample = TRUE, size = 0.1, alpha = 1) +
      ggtitle(paste(clust_name, "Predicted", sep = " - "))

    p <- p_sub + p_not
    # print(p)

    fn <- paste0(paste(today, clust_name, "RFcomparison", sep = "_"), ".png")
    do.call(ggsave, c(list(fn, p), plotsave_param))
})

dat_sub <- makePerCellDF(sce_sub, use_dimred = TRUE) |>
  mutate(case_id = factor(case_id, levels = meta$cases),
         donor_type = factor(donor_type, levels = meta$stages)) |>
  arrange(case_id, donor_type)

dat_not <- makePerCellDF(sce_not, use_dimred = TRUE) |>
  mutate(case_id = factor(case_id, levels = meta$cases),
         donor_type = factor(donor_type, levels = meta$stages)) |>
  arrange(case_id, donor_type)
dat_not <- dat_not[sample(rownames(dat_not), nrow(dat_sub)), ]

purrr::walk(assay_sel, \(cur_assay) {
  # purrr::walk(methods_sel, \(cur_method) {
  cur_method <- "Leiden2"
    dimred_name <- paste("UMAP", "immune", cur_assay, sep = "_")
    clust_name <- paste(cur_method, cur_assay, sep = "_")

    # Plot all clusters
    p_sub <- plot_dim_red(dat_sub, dimred_name, clust_name,
                          sample = TRUE, size = 0.1, alpha = 1) +
      ggtitle(paste(clust_name, "Original", sep = " - "))

    p_not <- plot_dim_red(dat_not, dimred_name, clust_name,
                          sample = TRUE, size = 0.1, alpha = 1) +
      ggtitle(paste(clust_name, "Predicted", sep = " - "))

    p <- p_sub + p_not
    # print(p)

    fn <- paste0(paste(today, clust_name, "RFcomparison", sep = "_"), ".png")
    do.call(ggsave, c(list(fn, p), plotsave_param))
})
```

### Merge test and train datasets

Merge datasets and add clusters to the SCE object containing immune cells (`sce_imm`).

```{r random-forest-merge}
# Cluster columns
# cluster_cols <- c(outer(methods_sel, assay_sel, FUN = paste, sep = "_"))
cluster_cols <- c("Leiden2_fastMNN_case_id", "Leiden2_scaled")
# Merge datasets
clusters <- rbind(colData(sce_sub)[, cluster_cols],
                  colData(sce_not)[, cluster_cols])
clusters <- clusters[order(match(rownames(clusters), colnames(sce_imm))), ]

# Add to the SCE object
colData(sce_imm)[, cluster_cols] <- clusters
# remove(sce_sub, sce_not)
```

```{r save_rf2}
metadata(sce_imm)$all_channels <- all_channels
fn_spe <- file.path(paths$folder_script, paste0(paths$object_type, "_", paths$panel_type, "rf.rds"))
saveRDS(sce_imm, fn_spe)
# sce_imm <- readRDS(fn_spe)
```
# **Visualize clusters**

## Load packages

```{r packages4, results='hide'}
suppressPackageStartupMessages(c(
  library(heatmaply),
  library(htmltools),
  library(cytomapper),
  library(ggalluvial)
))
```


## Visualize clusters on reduced dimensions

```{r dimred-clusters}
cur_dat <- makePerCellDF(sce_imm, use_dimred = TRUE) |>
  mutate(case_id = factor(case_id, levels = meta$cases),
         donor_type = factor(donor_type, levels = meta$stages)) |>
  arrange(case_id, donor_type)

purrr::walk(assay_sel, \(cur_assay) {
  purrr::walk(dimred_sel, \(cur_dimred) {
    purrr::walk(methods_sel, \(cur_method) {
      # cur_assay <- "scaled"; cur_method <- "Pheno2"; cur_dimred <- "UMAP"
      dimred_name <- paste(cur_dimred, "immune", cur_assay, sep = "_")
      clust_name <- paste(cur_method, cur_assay, sep = "_")
      
      # Plot all clusters
      p <- plot_dim_red(cur_dat, dimred_name, clust_name,
                        sample = TRUE, size = 0.1, alpha = 1)
      if (do_print) print(p)
      
      fn <- paste0(paste(today, "Clusters", clust_name, cur_dimred,
                         sep = "_"), ".png")
      do.call(ggsave, c(list(fn, p), plotsave_param))
    })
  })
})
```


## Distribution of marker expression by cluster

```{r violin-clusters-markers, message=FALSE, warning=FALSE, fig.width=15, fig.height=10}
purrr::walk(assay_sel, \(cur_assay) {
  purrr::walk(methods_sel, \(cur_method) {
    clust_name <- paste(cur_method, cur_assay, sep = "_")
    sce_imm[[clust_name]] <- as.factor(sce_imm[[clust_name]])
    cur_dat <- makePerCellDF(sce_imm, assay.type = cur_assay,
                             features = channels_clust, use.dimred = FALSE)
    
    cur_dat <- melt.data.table(
      as.data.table(cur_dat),
      id.vars = c("image_number", "cell_number", clust_name),
      measure.vars = channels_clust,
      variable.name = "channel",
      value.name = cur_assay
    )
    cur_dat[, MeanExprs := mean(get(cur_assay)),
            by = c("channel", clust_name)]
    
    p <- plot_violin(cur_dat, x = "channel", y = cur_assay,
                     fill_by = "MeanExprs", facet_by = clust_name,
                     title = clust_name) +
      theme(axis.text.x = element_text(size = 8, angle = 90))
    if (do_print) suppressMessages(print(p))
    fn <- paste0(paste(today, "Clusters", clust_name, "Violins",
                       sep = "_"), ".png")
    do.call(ggsave, c(list(fn, p), plotsave_param_large))
  })
})
```


## Heatmap of marker expression by cluster

```{r heatmap-clusters-markers}
hmlist <- list()
n <- 1
purrr::walk(assay_sel, \(cur_assay) {
  purrr::walk(methods_sel, \(cur_method) {
    clust_name <- paste(cur_method, cur_assay, sep = "_")
    
    # Summarize the data
    hm <- summarize_heatmap(sce_imm,
                            expr_values = cur_assay,
                            cluster_by = clust_name,
                            channels = channels_clust)
    
    # Display the heatmap
    fn <- paste0(paste(today, "Clusters", clust_name, "Heatmap",
                       sep = "_"), ".html")
    hmlist[[n]] <- heatmaply::heatmaply(
      heatmaply::normalize(hm), main = clust_name,
      file = file.path(paths$folder_script, fn))
    n <- n + 1

    # remove all clusters with less than 10 cells and plot again.
    clust_freq <- table(colData(sce_imm)[[clust_name]])
    keep_clusts <- names(clust_freq)[clust_freq >= 110]
    sce_isl_sub <- sce_imm[, colData(sce_imm)[[clust_name]] %in% keep_clusts]

    hm <- summarize_heatmap(sce_isl_sub,
                            expr_values = cur_assay,
                            cluster_by = clust_name,
                            channels = channels_clust)
    # Display the reduced heatmap -> this improves the visualization.
    # Otherwise heatmap is scaled by expression of outliers.
    fn <- paste0(paste(today, "Clusters", clust_name, "Heatmap_reduced",
                       sep = "_"), ".html")
    hmlist[[n]] <- heatmaply(
       heatmaply::normalize(hm), main = clust_name, 
       file = file.path(paths$folder_script, fn))
    n <- n + 1
  })
})
htmltools::tagList(setNames(hmlist, NULL))
```

## Barplot with cluster frequencies
Visualize the number of cells per cluster.
```{r viz-cluster-barplot}
cur_dat <- scuttle::makePerCellDF(sce_imm, use_dimred = TRUE) |>
  dplyr::mutate(case_id = factor(case_id, levels = meta$cases),
                donor_type = factor(donor_type, levels = meta$stages)) |>
  dplyr::arrange(case_id, donor_type) |> 
  tibble::as_tibble()

purrr::walk(methods_sel, \(cur_method) {
  purrr::walk(assay_sel, \(cur_assay) {
    clust_name <- paste(cur_method, cur_assay, sep = "_")
    title <- paste0("Cluster Abundances: ", clust_name)
    df <- cur_dat |> 
      dplyr::select(all_of(clust_name))  |> 
      dplyr::group_by(across(all_of(clust_name))) |> 
      dplyr::summarise(n = n())

    p <- df |> 
      ggplot2::ggplot(ggplot2::aes(x = factor(get(clust_name)),
                                   y = n, fill = factor(get(clust_name)))) +
      mytheme$standard() +
      ggplot2::theme(
        axis.text.x = ggplot2::element_text(angle = 45, hjust = 1),
        panel.background = ggplot2::element_rect(colour = "grey20")
      ) + 
      ggplot2::geom_bar(stat = "identity") +
      ggtitle(title) + 
      ggplot2::scale_fill_manual(values = palettes$colors[seq_along(unique(cur_dat[[clust_name]]))]) + 
      xlab("Cluster") +
      ylab("Number of cells")
    print(p)
    fn <- paste0(paste(today, "Clusters", clust_name, "Barplot",
                       sep = "_"), ".png")
    do.call(ggsave, c(list(fn, p), plotsave_param_large))
  })
})
```

```{r check_enrichment}
df <- as_tibble(colData(sce_imm)) |> select(case_id, donor_type, image_id, Pheno2_scaled)

df |> 
  dplyr::add_count(Pheno2_scaled) |> 
  dplyr::count(case_id, Pheno2_scaled, n) |> 
  mutate(fraction = nn / n) |> 
  filter(n > 30) |> 
  arrange(desc(fraction)) |> 
  print(n = 30)

df |> dplyr::count(image_id, Pheno2_scaled) |> filter(Pheno2_scaled == 10) |> arrange(desc(n))
```
## Visualize clusters with cytomapper

### Select cluster(s) and assay to show

```{r viz-clust-select}
# Lymphocytes: viz_clust <- c(3, 9, 17, 24, 29) #c(3, 17, 21) NULL
viz_clust <- NULL # c(25, 27, 1, 23, 2, 31, 30)
viz_method <- "Pheno2"
viz_assay <- "scaled"
# all_channels <- metadata(sce_imm)$all_channels
```

### Load images and masks

```{r viz-cluster-load}
if (!is.null(viz_clust)) {
  nb_images <- 14
  image_extension <- ".tiff"
  clust_name <- paste(viz_method, viz_assay, sep = "_")
  
  # Subset the SCE
  sce_viz <- sce_imm[, colData(sce_imm)[[clust_name]] %in% viz_clust]

  set.seed(seed)
  image_sub <- sort(sample(
    unique(sce_viz$image_fullname),
    min(length(unique(sce_viz$image_fullname)), nb_images)))

  # Folders
  folder_images <- file.path(paths$folder_in, "img", paths$panel_type)
  folder_masks <- file.path(paths$folder_in, "masks_cells",
                            paths$panel_type, "whole-cell")
  
  # Load images and masks
  #images <- yoloader(
  #  x = sce_viz,
  #  image_dir = folder_images,
  #  image_names = image_sub,
  #  type = "stacks"
  #)
  # 
  # 27: "6347_Immune_ROI_021.tiff", "6422_Immune_ROI_039.tiff", "6424_Immune_ROI_064.tiff"
  # 22: "6228_Immune_ROI_029.tiff", "8011_Immune_ROI_017.tiff", "6514_Immune_ROI_005.tiff"
  # 29: "6225_Immune_ROI_053.tiff", "6225_Immune_ROI_043.tiff", "6225_Immune_ROI_057.tiff"
  # 28: "6328_Immune_ROI_076.tiff", "6328_Immune_ROI_078.tiff", "6328_Immune_ROI_001.tiff"
  # 23: "6533_Immune_ROI_051.tiff", "6209_Immune_ROI_057.tiff", "6506_Immune_ROI_060.tiff"
  image_sub <- unique(c(image_sub, "6347_Immune_ROI_021.tiff", "6422_Immune_ROI_039.tiff", "6424_Immune_ROI_064.tiff", "6228_Immune_ROI_029.tiff", 
                 "8011_Immune_ROI_017.tiff", "6514_Immune_ROI_005.tiff", "6225_Immune_ROI_053.tiff", "6225_Immune_ROI_043.tiff", "6225_Immune_ROI_057.tiff",
                 "6328_Immune_ROI_076.tiff", "6328_Immune_ROI_078.tiff", "6328_Immune_ROI_001.tiff",
                 "6533_Immune_ROI_051.tiff", "6209_Immune_ROI_057.tiff", "6506_Immune_ROI_060.tiff", 
                 "6310_Immune_ROI_006.tiff", # "6043_Immune_ROI_014.tiff",  
                 "6228_Immune_ROI_082.tiff", "6247_Immune_ROI_072.tiff"))                             
                 # )

  images <- cytomapper::loadImages(file.path(folder_images, image_sub))
  channelNames(images) <- all_channels
  images <- getChannels(images, channels_clust)
  # Add image names to metadata
  mcols(images)$ImageName <- names(images)
  mcols(images)$ImageName <- paste0(mcols(images)$ImageName)

  masks <- yoloader(
    x = sce_viz,
    image_dir = folder_masks,
    image_names = image_sub,
    as.is = TRUE,
    type = "masks"
  )
  
  sce_viz <- sce_viz[, sce_viz$image_fullname %in% image_sub]
  sce_viz$ImageName <- gsub(image_extension, "", sce_viz$image_fullname)
}
```

### Plot images

Images are saved to disk but not shown.

```{r viz-cluster-plot}
if (!is.null(viz_clust)) {
  fn <- list(filename = file.path(
    plotsave_param$path,
    paste0(paste(today, "Cytomapper", clust_name,
                 paste(viz_clust, collapse = "-"), sep = "_"), ".png")))

  plotPixels(images, object = sce_viz, masks,
             cell_id = "cell_number", img_id = "ImageName",
             colour_by = c("CD3e", "PD1", "MPO", "HLA_DR", "CD163", "CD206"),
             outline_by = c(clust_name),
             bcg = list(CD3e = c(0, 5, 1), PD1 = c(0, 5, 1), MPO = c(0, 5, 1),
                        HLA_DR = c(0, 5, 1), CD163 = c(0, 5, 1), CD206 = c(0, 5, 1)),
             image_title = NULL, scale = TRUE, save_plot = fn)
}
```

### Cytoviewer app.
```{r cytoviewer}
if (!is.null(viz_clust)) {
  library(cytoviewer)
  sce_viz[[clust_name]] <- as.factor(sce_viz[[clust_name]])
  app <- cytoviewer(image = images, 
                    mask = masks, 
                    object = sce_viz,
                    img_id = "ImageName", 
                    cell_id = "cell_number")

  if (interactive()) {
    shiny::runApp(app)
  }
}
```

### Remove generated files

```{r viz-cluster-remove}
if (!is.null(viz_clust)) {
  remove(sce_viz, images, masks)
}
```


## Plot specific cluster(s)

When many clusters are present, they may not be easy fo visualize on reduced dimension maps. Here, a subset of clusters can be selected for plotting.

### Select cluster(s) and assay to show

```{r dimred-cluster-select}
viz_clust <- NULL # c(1, 4, 10) # NULL
viz_method <- "Leiden2"
viz_assay <- "fastMNN_case_id"
```

```{r dimred-cluster-plot}
if (!is.null(viz_clust)) {
  # Prepare the data
  cur_dat <- makePerCellDF(
    sce_imm[, colnames(sce_imm) %in% meta[["subset"]]],
    use.dimred = TRUE) |>
    mutate(case_id = factor(case_id, levels = meta$cases),
           donor_type = factor(donor_type, levels = meta$stages)) |>
    arrange(case_id, donor_type)
  
  # Clusters and palette
  clust_name <- paste(viz_method, viz_assay, sep = "_")
  clust_nb <-  length(unique(cur_dat[[clust_name]]))
  viz_pal <- palettes$colors[1:clust_nb]
  names(viz_pal) <- 1:clust_nb
  viz_pal[!names(viz_pal) %in% viz_clust] <- "lightgrey"
  
  # Plot
  purrr::walk(dimred_sel, \(cur_dimred) {
    dimred_name <- paste(cur_dimred, "immune", viz_assay, sep = "_")
    
    p <- plot_dim_red(cur_dat, dimred_name, clust_name, palette = viz_pal,
                      sample = TRUE, size = 0.1, alpha = 1)
    
    fn <- paste0(paste(today, paste0("Cluster", paste(viz_clust, collapse = "-")),
           clust_name, cur_dimred, sep = "_"), ".png")
    do.call(ggsave, c(list(fn, p), plotsave_param))
  })
}
```



# **Attribute cell types**

**Note: this section requires manual intervention**  

Here, clusters obtained at the previous steps are manually merged into meaningful cell types. Cluster numbers are attributed to the different cell types based on the plots and heatmaps above. Consequently, this attribution has to be adapted to the clustering results.  

## Cell type attribution

Cell type attribution has to be performed once per clustering method and per selected assay.

### Phenograph x scaled counts

```{r celltypes-pheno-scaled}
clust_method <- "Pheno2"
clust_assay <- "scaled"

if (!clust_assay %in% assay_sel)
  stop("The selected assay is not in the assays selected for clustering")

clust_name <- paste(clust_method, clust_assay, sep = "_")
print(clust_name)

# Define which clusters correspond to which cell types
#clust_lympho  <- c(1, 5, 7, 17, 18)
#clust_myelo <- c(2, 3, 4, 8, 11, 12, 13, 14, 15, 16)
#clust_neutr <- c(6, 9, 10)
#clust_other <- c()
clust_lympho <- c(6, 13, 14, 20, 23, 29)
clust_neutr <- c(5, 22, 24)
clust_myelo <- c(1, 3, 4, 7, 8, 9, 10, 11, 12, 15, 16, 17, 18, 19, 21, 25, 26, 27, 30, 31)
clust_other <- c(2, 28, 32, 33)

all_clust <- sort(c(clust_lympho, clust_myelo, clust_neutr, clust_other))

if ((!length(unique(all_clust)) ==
     length(unique(colData(sce_imm)[, clust_name]))) ||
    any(duplicated(all_clust))) {
  stop("Recheck cluster attribution")
}

# Add cell types to the SCE object
colData(sce_imm)[colData(sce_imm)[, clust_name] %in% clust_lympho,
                 paste("CellType", clust_name, sep = "_")] <- "Lymphocyte"
colData(sce_imm)[colData(sce_imm)[, clust_name] %in% clust_myelo,
                 paste("CellType", clust_name, sep = "_")] <- "Myeloid"
colData(sce_imm)[colData(sce_imm)[, clust_name] %in% clust_neutr,
                 paste("CellType", clust_name, sep = "_")] <- "Neutrophil"
colData(sce_imm)[colData(sce_imm)[, clust_name] %in% clust_other,
                 paste("CellType", clust_name, sep = "_")] <- "Other"
```

### Phenograph x fastMNN counts

```{r celltypes-pheno-fastMNN}
clust_method <- "Pheno2"
clust_assay <- "fastMNN_case_id"

if (!clust_assay %in% assay_sel)
  stop("The selected assay is not in the assays selected for clustering")

clust_name <- paste(clust_method, clust_assay, sep = "_")
print(clust_name)

# Define which clusters correspond to which cell types
#clust_lympho  <- c(1, 7, 10, 14)
#clust_myelo <- c(2, 3, 4, 5, 6, 9, 13, 15)
#clust_other <- c()
##clust_neutr <- c(8, 11, 12)

clust_lympho <- c(6, 9, 15, 21, 22, 23, 24)
clust_neutr <- c(5, 11, 20)
clust_myelo <- c(1:4, 7, 8, 10, 12:14, 16:19, 26)
clust_other <- c(25, 27)
all_clust <- sort(c(clust_lympho, clust_myelo, clust_neutr, clust_other))

if ((!length(unique(all_clust)) ==
     length(unique(colData(sce_imm)[, clust_name]))) ||
    any(duplicated(all_clust))) {
  stop("Recheck cluster attribution")
}

# Add cell types to the SCE object
colData(sce_imm)[colData(sce_imm)[, clust_name] %in% clust_lympho,
                 paste("CellType", clust_name, sep = "_")] <- "Lymphocyte"
colData(sce_imm)[colData(sce_imm)[, clust_name] %in% clust_myelo,
                 paste("CellType", clust_name, sep = "_")] <- "Myeloid"
colData(sce_imm)[colData(sce_imm)[, clust_name] %in% clust_neutr,
                 paste("CellType", clust_name, sep = "_")] <- "Neutrophil"
colData(sce_imm)[colData(sce_imm)[, clust_name] %in% clust_other,
                 paste("CellType", clust_name, sep = "_")] <- "Other"
```

### Leiden x scaled counts

```{r celltypes-leiden-scaled}
clust_method <- "Leiden2"
clust_assay <- "scaled"

if (!clust_assay %in% assay_sel)
  stop("The selected assay is not in the assays selected for clustering")

clust_name <- paste(clust_method, clust_assay, sep = "_")
print(clust_name)

# Define which clusters correspond to which cell types
# clust_lympho  <- c(1, 7, 10)
#clust_myelo <- c(3, 4, 6, 8, 9)
#clust_neutr <- c(5, 11)
#clust_other <- c(2, 12)

clust_lympho <- c(3, 9, 17, 24, 29)
clust_myelo <- c(1, 4:8, 11:14, 18:23, 26:28,
                 30, 31, 32, 33, 35:37)
clust_neutr <- c(2, 25, 34)
clust_other <- c(10, 15, 16)


all_clust <- sort(c(clust_lympho, clust_myelo, clust_neutr, clust_other))

if ((!length(unique(all_clust)) ==
     length(unique(colData(sce_imm)[, clust_name]))) ||
    any(duplicated(all_clust))) {
  stop("Recheck cluster attribution")
}

# Add cell types to the SCE object
colData(sce_imm)[colData(sce_imm)[, clust_name] %in% clust_lympho,
                 paste("CellType", clust_name, sep = "_")] <- "Lymphocyte"
colData(sce_imm)[colData(sce_imm)[, clust_name] %in% clust_myelo,
                 paste("CellType", clust_name, sep = "_")] <- "Myeloid"
colData(sce_imm)[colData(sce_imm)[, clust_name] %in% clust_neutr,
                 paste("CellType", clust_name, sep = "_")] <- "Neutrophil"
colData(sce_imm)[colData(sce_imm)[, clust_name] %in% clust_other,
                 paste("CellType", clust_name, sep = "_")] <- "Other"
```

### Leiden x fastMNN counts

```{r celltypes-leiden-fastMNN}
clust_method <- "Leiden2"
clust_assay <- "fastMNN_case_id"

if (!clust_assay %in% assay_sel)
  stop("The selected assay is not in the assays selected for clustering")

clust_name <- paste(clust_method, clust_assay, sep = "_")
print(clust_name)

# Define which clusters correspond to which cell types
#clust_lympho  <- c(1, 2, 8)
#clust_myelo <- c(3, 4, 5, 6)
#clust_neutr <- c(7)
#clust_other <- c(9)
clust_neutr <- c(5, 16, 21)
clust_lympho <- c(8, 9, 15, 20, 22, 27)
clust_other <- c()
clust_myelo <- c(1, 2, 3, 4, 6, 7, 10, 11, 12, 13, 14, 17, 18, 19, 23, 24, 25, 26, 28:41)


all_clust <- sort(c(clust_lympho, clust_myelo, clust_neutr, clust_other))

if ((!length(unique(all_clust)) ==
     length(unique(colData(sce_imm)[, clust_name]))) ||
    any(duplicated(all_clust))) {
  stop("Recheck cluster attribution")
}

# Add cell types to the SCE object
colData(sce_imm)[colData(sce_imm)[, clust_name] %in% clust_lympho,
                 paste("CellType", clust_name, sep = "_")] <- "Lymphocyte"
colData(sce_imm)[colData(sce_imm)[, clust_name] %in% clust_myelo,
                 paste("CellType", clust_name, sep = "_")] <- "Myeloid"
colData(sce_imm)[colData(sce_imm)[, clust_name] %in% clust_neutr,
                 paste("CellType", clust_name, sep = "_")] <- "Neutrophil"
colData(sce_imm)[colData(sce_imm)[, clust_name] %in% clust_other,
                 paste("CellType", clust_name, sep = "_")] <- "Other"
```


## Visualize attributed cell types

### Plot cell types on reduced dimensions

```{r palette$celltypes}
palettes$celltype <- (c(
  Myeloid = palettes$colors[1], Lymphocyte = palettes$colors[2],
  Neutrophil = palettes$colors[3], Other = palettes$colors[8]))
```


```{r celltypes-dimred}
# Prepare the data
cur_dat <- makePerCellDF(
  sce_imm[, colnames(sce_imm) %in% meta[["subset_immune"]]],
  use_dimred = TRUE) |>
  mutate(case_id = factor(case_id, levels = meta$cases),
         donor_type = factor(donor_type, levels = meta$stages)) |>
  arrange(case_id, donor_type)

# Plot
purrr::walk(assay_sel, \(cur_assay) {
  purrr::walk(dimred_sel, \(cur_dimred) {
    purrr::walk(methods_sel, \(cur_method) {
      # cur_assay <- assay_sel[2];  cur_method <- methods_sel[2]; cur_dimred <- dimred_sel[1]
      dimred_name <- paste(cur_dimred, "immune", cur_assay, sep = "_")
      clust_name <- paste("CellType", cur_method, cur_assay,
                          sep = "_")
      
      p <- plot_dim_red(cur_dat, dimred_name, clust_name,
                        palette = palettes$celltype,
                        sample = TRUE, size = 0.1, alpha = 1)
      if (do_print) print(p)
      
      fn <- paste0(paste(today, clust_name, cur_dimred,
                         sep = "_"), ".png")
      do.call(ggsave, c(list(fn, p), plotsave_param))
    })
  })
})
```

### Cell types heatmap

```{r celltypes-heatmap}
hmlist <- list()
n <- 1
purrr::walk2(assay_sel, names(assay_sel), \(cur_assay, name_cur_assay) {
  purrr::walk(methods_sel, \(cur_method) {
    clust_name <- paste("CellType", cur_method, cur_assay,
                        sep = "_")
  
    # Summarize the data
    hm <- summarize_heatmap(sce_imm,
                            expr_values = name_cur_assay,
                            cluster_by = clust_name,
                            channels = channels_clust)
    
    # Display the heatmap
    fn <- paste0(paste(today, clust_name, "Heatmap",
                       sep = "_"), ".html")
    
    hmlist[[n]] <- heatmaply(
      heatmaply::normalize(hm), main = clust_name,
      file = file.path(paths$folder_script, fn))
    n <- n + 1
  })
})
htmltools::tagList(setNames(hmlist, NULL))
```

### Visualize celltypes with cytomapper

Load the masks

```{r viz-celltype-load}
nb_images <- 14
image_extension <- ".tiff"

set.seed(seed)
image_sub <- sort(sample(
  unique(sce_imm$image_fullname),
  min(length(unique(sce_imm$image_fullname)), nb_images)))

# Subset the SCE to subsampled images
sce_sub <- sce_imm[, sce_imm$image_fullname %in% image_sub]
sce_sub$ImageName <- gsub(image_extension, "", sce_sub$image_fullname)

# Folders
folder_images <- file.path(paths$folder_in, "img", paths$panel_type)
folder_masks <- file.path(paths$folder_in, "masks_cells",
                          paths$panel_type, "whole-cell")

# Load masks
masks <- yoloader(
  x = sce_sub,
  image_dir = folder_masks,
  image_names = image_sub,
  as.is = TRUE,
  type = "masks"
)
```

Plot the images

```{r viz-celltype-plot}
purrr::walk(assay_sel, \(cur_assay) {
  purrr::walk(methods_sel, \(cur_method) {
    celltype_viz <- paste("CellType", cur_method, cur_assay,
                          sep = "_")
    
    colour_key <- list(celltype_viz = palettes$celltype)

    fn <- list(filename = file.path(
      plotsave_param$path,
      paste0(paste(today, "Cytomapper", celltype_viz, sep = "_"), ".png")))

    plotCells(masks, object = sce_sub,
              cell_id = "cell_number", img_id = "ImageName",
              colour_by = celltype_viz, colour = colour_key, save_plot = fn)
  })
})
```

# **Consensus cell types**

Attribute consensus cell types for immune cells based on the clustering 
performed above.


## Cell type concordance

Compare concordance of cell types attributed using different methods.

```{r celltypes-concordance}
celltype_cols <- colnames(colData(sce_imm))[grepl(paste0(
  "CellType_", methods_sel, collapse = "|"), colnames(colData(sce_imm)))]

celltypes <- as.data.table(colData(sce_imm)[
  , c("cell_id", "case_id", "image_number", "cell_number", celltype_cols)])

ct_combination <- t(combn(unique(celltype_cols), 2))

for (i in 1:nrow(ct_combination)) {
  cat("\n", ct_combination[i, 1], " vs ", ct_combination[i, 2], "\n")
  
  ct_concord <- nrow(
    celltypes[get(ct_combination[i, 1]) == get(ct_combination[i, 2]), ]
  )
  ct_discord <- nrow(
    celltypes[get(ct_combination[i, 1]) != get(ct_combination[i, 2]), ]
  )
  cat("Number of cells with concordant cell types: ", ct_concord,
      "\nNumber of cells with discordant cell types: ", ct_discord,
      "\nPercentage of cells with concordant cell types:",
      100 * ct_concord / (ct_discord + ct_concord), "%", "\n")
}
```
Pheno2-scaled - Leiden2-fastMNN: 96.0 %
Pheno2-scaled - Pheno2-fastMNN: 95.6 %
Pheno2-scaled - Leiden2-scaled: 97.4 %
Leiden2-fastMNN - Pheno2-fastMNN: 97.4 %
Leiden2-fastMNN - Leiden2-scaled: 96.7 %
Pheno2-fastMNN - Leiden2-scaled: 95.7 %



## Attribute consensus cell types

### Majority vote

Cells for which 3 out of 4 methods agree on the cell type are attributed this cell type, other cells as considered as "Ambiguous".

```{r consensus-celltype}
#column_to_remove <- grep("Pheno1_fastMNN", names(celltypes))

# Remove the column by setting it to NULL
#celltypes[, (column_to_remove) := NULL]

# Majority vote
Mode <- function(x) {
  ux <- unique(x)
  ux[which.max(tabulate(match(x, ux)))]
}

celltypes[, cell_type := apply(celltypes, 1, Mode)]
celltypes[, Paste := do.call(paste0, .SD),
          .SDcols = grep("CellType_", colnames(celltypes))]
celltypes[, count := stringr::str_count(Paste, cell_type)]
celltypes[count < 3, cell_type := "Ambiguous"]
celltypes[, `:=`(Paste = NULL, count = NULL)]

cat("Percentage of cells with an ambiguous cell type: ",
    100 * nrow(celltypes[cell_type == "Ambiguous"]) / nrow(celltypes),
    "%")
```

### Compare cell type attributions

```{r celltypes-alluvial, warning=FALSE}
celltype_cols <- c("cell_type", colnames(colData(sce_imm))[grepl(paste0(
  "CellType_", methods_sel, collapse = "|"), colnames(colData(sce_imm)))])

p <- celltypes[, .(number = .N), by = celltype_cols] |>
  ggplot(aes(axis1 = get(celltype_cols[1]),
             axis2 = get(celltype_cols[2]),
             axis3 = get(celltype_cols[3]),
             axis4 = get(celltype_cols[4]),
             axis5 = get(celltype_cols[5]),
             y = number)) +
  geom_stratum() +
  geom_text(stat = "stratum", infer.label = TRUE) +
  geom_alluvium(aes(fill = get(celltype_cols[1]))) +
  scale_x_discrete(limits = gsub("cell_type_", "", celltype_cols)) +
  scale_fill_manual(values = palettes$colors[c(TRUE, FALSE)],
                    name = "Cell Type") +
  mytheme$standard() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
if (do_print) print(p)

fn <- paste0(paste(today, "CellTypes", "alluvial", sep = "_"), ".png")
do.call(ggsave, c(list(fn, p), plotsave_param))
```

### Number of cells by cell type

```{r cells-per-celltype}
as_tibble(celltypes) |>
  select(cell_type) |>
  group_by(cell_type) |>
  add_count(name = "Number of cells") |>
  distinct()
```
170k Lymphocytes!

## Add cell types to the immune SCE object

```{r add-immune-celltypes-sce}
celltypes <- celltypes[order(match(celltypes$cell_id, colnames(sce_imm))), ]
colData(sce_imm)$cell_type <- celltypes$cell_type
```



# **Plot cell types**

## Reduced dimensions

Plot consensus cell types on reduced dimensions

```{r consensus-dimred}
ct <- c(names(palettes$celltype), "Ambiguous")
palette_ct <- c(palettes$celltype[1:(length(ct) - 2)], "grey40", "grey")
names(palette_ct) <- names(ct) <- ct

cur_dat <- makePerCellDF(
  sce_imm[, colnames(sce_imm) %in% metadata(sce_imm)$subset],
  use_dimred = TRUE) |>
  mutate(case_id = factor(case_id, levels = metadata(sce_imm)$cases),
         donor_type = factor(donor_type, levels = metadata(sce_imm)$stages),
         cell_type = factor(cell_type, levels = names(ct))) |>
  arrange(cell_type, case_id, donor_type)

purrr::walk(assay_sel, \(cur_assay) {
  purrr::walk(dimred_sel, \(cur_dimred) {
    dimred_name <- paste(cur_dimred, "immune", cur_assay, sep = "_")
    
    p <- plot_dim_red(cur_dat, dimred_name, "cell_type",
                      sample = TRUE, size = 0.1, alpha = 1,
                      palette = palette_ct)
    if (do_print) print(p)
    
    fn <- paste0(paste(today, "CellType", "Consensus", dimred_name,
                       sep = "_"), ".png")
    do.call(ggsave, c(list(fn, p), plotsave_param))
  })
})
```

## Heatmap

```{r consensus-heatmap}
hmlist <- list()
purrr::walk2(assay_sel, names(assay_sel), \(cur_assay, name_cur_assay) {
  clust_name <- paste("CellType", cur_assay, sep = "_")
  
  # Summarize the data
  hm <- summarize_heatmap(sce_imm,
                          expr_values = name_cur_assay,
                          cluster_by = "cell_type",
                          channels = channels_clust)
  
  # Display the heatmap
  fn <- paste0(paste(today, "CellType", "Consensus", cur_assay,
                     "Heatmap", sep = "_"), ".html")
  
  hmlist[[i]] <- heatmaply(
    heatmaply::normalize(hm), main = clust_name,
    file = file.path(paths$folder_script, fn))
})
htmltools::tagList(setNames(hmlist, NULL))
```


## Correlation between channels

```{r celltype-correlation}
# Define the correlation method
cor_method <- "pearson"

# Function to calculate p values
cor_test_p <- function(x) {
  FUN <- function(x, y) cor.test(x, y, method = cor_method)[["p.value"]]
  z <- outer(
    colnames(x), 
    colnames(x), 
    Vectorize(function(i, j) FUN(x[, i], x[, j]))
  )
  dimnames(z) <- list(colnames(x), colnames(x))
  z
}

hmlist <- list()
purrr::walk(assay_sel, \(cur_assay) {
  
  # Summarize the data
  hm <- summarize_heatmap(sce_imm,
                          expr_values = cur_assay,
                          cluster_by = "cell_type",
                          channels = channels_clust)
  
  # Run the correlation test
  r <- cor(hm, method = cor_method)
  p <- cor_test_p(hm)
  diag(p) <- NA
  
  # Plot and save the heatmap
  fn <- paste0(paste(today, "CellTypes", "Consensus", "Correlation",
                     cur_assay, sep = "_"), ".html")
  
  hmlist[[i]] <- heatmaply_cor(
    r, node_type = "scatter",
    point_size_mat = -log10(p), point_size_name = "-log10(p-value)",
    label_names = c("x", "y", "Correlation"),
    main = paste("Cell type -", cur_assay),
    file = file.path(paths$folder_script, fn))
})
htmltools::tagList(setNames(hmlist, NULL))
```



# **Finalize and save **

## Merge objects

Add immune cell types to the main SPE

```{r add-immune-celltypes}
# Add a new "cell_type" column to the SPE object
spe$cell_type <- "0"

# Make sure row order is preserved
row_order <- rownames(colData(spe)[spe$cell_category == "immune", ])
sce_imm <- sce_imm[, order(match(colnames(sce_imm), row_order))]

# Add cell types
colData(spe)[row_order, ]$cell_type <- sce_imm$cell_type
```


## Save the updated SCE object

Contains all attributed cell types and cell categories

```{r save-spe}
print(spe)
fn_spe <- file.path(paths$folder_script, paste0(paths$object_type, "_", paths$panel_type, ".rds"))
saveRDS(spe, fn_spe)
```
