---
title: "08_CellTypesNonImmune_cells_Immune"
author: "Nicolas Damond, Nathan Steenbuck"
date: "Created: 01 Sep, 2022; Compiled: `r format(Sys.time(), '%d %b, %Y')`"
output: workflowr::wflow_html
editor_options:
  chunk_output_type: inline
---

```{r setup, include=FALSE}
if (rstudioapi::isAvailable()) {
  script_name <- basename(rstudioapi::getSourceEditorContext()$path)
} else {
  script_name <- knitr::current_input()
}

cur_user <- Sys.info()[["user"]]
script_name <- "08_CellTypesNonImmune_cells_Immune.Rmd"

if (cur_user == "ubuntu") {
  source(file.path("/", "mnt", "central_nas", "projects", 
    "type1_diabetes", "nathan", "T1D_Vol", "T1D_analysis", 
    "analysis", "helpers.R"))
  #n_cores <- future::availableCores() - 1
  n_cores <- 4
  future::plan(future::multicore(workers = n_cores))
  paths <- getPaths(script_name)
  knitr::opts_knit$set(root_dir = paths$home_data)
} else {
  source(file.path("/", "home", "nsteen", "scripts", "T1D_analysis", "analysis", "helpers.R"))
  n_cores <- 8
  future::plan(future::multicore(workers = n_cores))
  paths <- getPaths(script_name)
  paths$folder_script <- paths$cluster_folder_script
  paths$folder_in <- paths$cluster_folder_in
  paths$folder_out <- paths$cluster_home 
  knitr::opts_knit$set(root_dir = paths$cluster_home)
}

seed <- 123456
set.seed(seed)
options <- furrr::furrr_options(seed = seed)
paths$prev <- paste("07_ClustersImmune", paths$object_type, paths$panel_type, sep = "_")
knitr::opts_chunk$set(echo = TRUE)
do_print <- TRUE
```



# **Goal**

In scripts 05 to 08, cell types are attributed to all cells in the dataset
in an iterative way:

- `05_CellCategories_cells_Immune`: Attribution of main cell categories 
(immune, islet, exocrine, ...).
- `06_CellTypesImmune_cells_Immune`: Attribution of main cell 
types to immune cells (Lymphocyte, Myeloid, Neutrophil).
- `07_ClustersImmune_cells_Immune`: Further separation of immune cell types 
into clusters.
- `08_CellTypesNonImmune_cells_Immune`(this script): Attribution of cell types
to non-immune cells (Beta, Alpha, Exocrine, ...).
- `08_CellTypesIslet_cells_Immune`: Re-attribution of islet cells
to cell types based on clustering results from the previous step.
- `08_CellTypesIslet_cells_ImmuneII`: Attribution of rare & specific immune cell clusters
like T-reg and T-EMRAs.

Specifically, the goal of the current script is to attribute cell types 
to non-immune cells in the dataset. These cells were identified in script 06
as main cell categories (`islet`, `exocrine`, ...), but were not further 
clustered.

Cell type attribution is performed in a similar way as in previous scripts.

1. Unsupervised clustering using two approaches (i) PhenoGraph and (ii) Shared-
nearest neighbors (SNN) + Leiden community detection on two different count 
transformations (`exprs` and `scaled`). Only immune cell markers are used for
clustering.

2. For each of the four clustering results (two approaches X two count 
transformations), clusters are manually merged into cell categories based on
marker expression.

3. Consensus cell categories are attributed by majority vote: if the cell 
type is the same for three or more of the four approaches, this category is 
attributed to the cell. If not, the cell is categorized as `ambiguous`.

As an immune panel was used to stain the sections, the number of markers 
available to precisely identify non-immune cells is limited. The cell type 
resolution is therefore limited. Yet, markers of important cell types, such as 
beta cells, have been included, allowing to reliably identify them.



# **Settings**  

## Load packages

```{r packages, results='hide'}
suppressPackageStartupMessages(c(
  library(data.table),
  library(dplyr),
  library(SpatialExperiment),
  library(parallel)
))
```

## Paths and settings

```{r settings}
# Paths
if (!dir.exists(paths$folder_script)) dir.create(paths$folder_script)
plotsave_param$path <- paths$folder_script
plotsave_param_large$path <- paths$folder_script

# Misc settings
today <- gsub("-", "", Sys.Date())
```

##  Read in the data

Load the SpatialExperiment (SPE) object saved at the previous step.

```{r load-data}
fn_spe <- file.path(paths$folder_out, paths$prev, paste0(paths$object_type, "_", paths$panel_type, ".rds"))
spe <- readRDS(fn_spe)
print(spe)
```

### Channels and assays

Select channels (`channels_clust`), assays (`assay_sel`) and clustering methods (`methods_sel`) to use.

```{r select-methods-assays}
methods_sel <- c("Pheno5", "Leiden5")

assay_sel <- c("fastMNN_case_id", "scaled")
names(assay_sel) <- c("exprs", "scaled")
writeLines(c("Assays:", assay_sel[assay_sel %in% assayNames(spe)]))

dimred_sel <- c("UMAP") # , "tSNE")
writeLines(c("\nReduced dimensions:", dimred_sel))

channels <- rownames(spe)[!(grepl("DNA|H3", rownames(spe)))]
cat(c("\nChannels:", channels[channels %in% rownames(spe)]))
cat(c("\nNumber of channels:", length(channels)))
```

### Select the channels for clustering

Exclusive immune cell markers are not used for clustering.

```{r select-channels-clustering}
# Consider de-selecting: CS, CD57, CD56, CD54, ATP5A, CD73, LDHA, TMEM173, HK1
clust_islet <- c("SST", "INS", "GCG", "PDX1", "NKX6_1", "SYP")
clust_exocrine <- c("CAV1", "VIM", "PDX1", "SMA")
channel_other <- c("CD54", "TMEM173", "CD56", "CD57", "HLA_DR")
channel_met <- c("HK1", "ATP5A", "LDHA", "CS")
channel_func <- c("CD73")

channels_clust <- c(clust_islet, clust_exocrine, channel_other)
cat(c("\nChannels used for unsupervised clustering:",
      channels_clust[channels_clust %in% rownames(spe)]))
```


### Extract non-immune cells

The clustering is performed only on non-immune cells (`sce_imm`).

```{r extract-immune-cells}
# Extract the metadata (to avoid conflicts when merging the SCEs)
meta <- metadata(spe)
spatial_coords <- spatialCoords(spe)
colpairs <- colPairs(spe)

# Subset the SPE (non-immune cells only)
sce_exo <- spe[channels_clust, !spe$cell_category %in% c("immune")]

# Convert to SingleCellExperiment
sce_exo <- as(sce_exo, "SingleCellExperiment")
```

### Subset SCE

Clustering and cell type attribution will only be performed on a subset of the
dataset, and then extended to the rest of the cells using random forest.

```{r subset-sce-imm}
sce_sub <- sce_exo[channels_clust, colnames(sce_exo) %in% meta$subset]
sce_exo <- sce_exo[channels_clust, !(colnames(sce_exo) %in% colnames(sce_sub))]
```



# **Clustering**

Unsupervised clustering is performed by generating a nearest-neighbor graph 
followed by community detection (using either Louvain or Leiden algorithms).


## Settings

### Load packages

```{r packages2, results='hide'}
suppressPackageStartupMessages(c(
  library(ggplot2),
  library(viridis),
  library(scater),
  library(scuttle),
  library(scran),
  library(igraph),
  library(Rphenoannoy),
  library(clustree),
  library(patchwork),
  library(ranger),
  library(foreach),
  library(doParallel),
  library(BiocParallel)
))
```


## **Clustering: PhenoGraph**

Unsupervised clustering is performed with the [PhenoGraph](https://doi.org/10.1016/j.cell.2015.05.047) algorithm. 
This method works by generating a nearest-neighbor (kNN) graph of phenotpyic similarities followed by Louvain community dectection. 
For speed, we run PhenoGraph using the [Rphenoannoy implementation](https://github.com/stuchly/Rphenoannoy).

```{r phenograph-run}
clust_method <- c("Pheno5")

# Number of nearest-neighbors
k <- 15

purrr::walk(assay_sel, \(cur_assay) {
  clust_name <- paste(clust_method, cur_assay, sep = "_")
  writeLines(c("\n", clust_name))

  # Run Phenograph
  if (!clust_name %in% colnames(colData(sce_sub))) {
    set.seed(seed)
    cur_pheno <- Rphenoannoy::Rphenoannoy(t(assay(sce_sub, cur_assay)), k = k)

    cur_pheno <- DataFrame(cur_pheno[[2]]$membership)
    colnames(cur_pheno) <- clust_name
    rownames(cur_pheno) <- colnames(assay(sce_sub, cur_assay))

    # Add Phenograph clusters to the colData of the SCE object
    colData(sce_sub)[, clust_name] <<- cur_pheno
    remove(cur_pheno)
  }
})
```


## **Clustering: SNN + Leiden**

Here, we first construct a shared nearest-neighbor graph and identify clusters using the Leiden community detection algorithm.

### Run clustering

# Iterate over selected Assays.

```{r SNN-leiden-run}
clust_method <- c("Leiden5")
do_leiden <- FALSE

# Number of nearest-neighbors for SNN graph
k <- 15

# Resolutions for Leiden algorithm
resolutions <- c(0.000005, 0.0000075, 0.00001, 0.000025,
                  0.00005,  0.000075,  0.0001,  0.00025,
                   0.0005,   0.00075)


purrr::walk(assay_sel, \(cur_assay) {
  clust_name <- paste(clust_method, cur_assay, sep = "_")
  writeLines(c("\n", clust_name))
  
  if (!clust_name %in% colnames(colData(sce_sub))) {
    do_leiden <- TRUE
    
    # Build the SNN graph
    set.seed(seed)
    cur_graph <- scran::buildSNNGraph(sce_sub,
                                      type = "jaccard",
                                      k = k,
                                      assay.type = cur_assay,
                                      BPPARAM = BiocParallel::MulticoreParam(workers = n_cores))
    
    # Leiden clustering
    set.seed(seed)
    clusters <- furrr::future_map(resolutions, \(cur_resolution) {
      cur_leiden <- igraph::cluster_leiden(
        cur_graph,
        objective_function = "CPM", # could change for additional scalability.
        resolution_parameter = cur_resolution)

      cur_leiden <- cur_leiden$membership
    }, .options = options)
    
    # Add clusters to colData
    for (k in seq_along(resolutions)) {
      res_name <- paste(clust_method, cur_assay, k, sep = "_")
      colData(sce_sub)[, res_name] <<- clusters[[k]]
    }
    
    # Plot cluster tree
    p <- clustree::clustree(sce_sub, prefix = paste0(clust_method, "_", cur_assay, "_"))
    if (do_print) print(p)
    
    fn <- paste0(paste(today, clust_method, cur_assay, "ClusterTree", sep = "_"), ".png")
    do.call(ggsave, c(list(fn, p), plotsave_param))
    remove(cur_graph, clusters)
  }
})
```

### Optimal cluster resolution

Based on the trees above, select the optimal cluster resolution. Columns corresponding to other resolutions are removed from `colData(sce_sub)`.

```{r SNN-leiden-select}
do_leiden <- TRUE
if (isTRUE(do_leiden)) {
  optimal_res <- c("fastMNN_case_id" = 7, "scaled" = 6)
  
  purrr::walk(assay_sel, \(cur_assay) {
    clust_name <- paste(clust_method, cur_assay, sep = "_")
    
    optimal_col <- paste(clust_method, cur_assay, optimal_res[cur_assay], sep = "_")
    colData(sce_sub)[, clust_name] <<- colData(sce_sub)[, optimal_col]
    colData(sce_exo)[, clust_name] <<- 0

    other_cols <- colnames(colData(sce_sub))[grep(paste0(clust_name, "_"), colnames(colData(sce_sub)))]
    colData(sce_sub)[, other_cols] <<- NULL
  })
}
```
```{r save-sce-sub}
fn_sce_sub <- file.path(paths$folder_script, paste0("sce_sub", paths$object_type, paths$panel_type, ".rds"))
saveRDS(sce_sub, fn_sce_sub)
```

## Extend clusters to all cells

Cell types were attributed to a subset of the dataset. Here, we train a random
forest classifier to extend these cell types to the rest of the dataset.

### Random forest

Train a random forest classifier to attribute clusters to cells that were not subsetted.

```{r random-forest-train-predict-clusters}
n_features <- length(channels_clust)

purrr::walk(assay_sel, \(cur_assay) {
  purrr::walk(methods_sel, \(cur_method) {
    clust_name <- paste(cur_method, cur_assay, sep = "_")
    print(clust_name)
    
    
    # Define training (`sce_sub`) and test data (`sce_exo`)
    training_data <- dplyr::bind_cols(t(assay(sce_sub, cur_assay)),
                                        cluster_x = sce_sub[[clust_name]])
    df_data <- t(assay(sce_exo, cur_assay))

    # Changed to dplyr for scalability.
    test_data <- dplyr::bind_cols(df_data, cluster_x = sce_exo[[clust_name]])
    
    # Train
    ranger_model <- ranger(
      formula = cluster_x ~ .,
      data = training_data,
      importance = "impurity",
      mtry = floor(n_features / 3),
      num.trees = 10 * n_features,
      classification = TRUE,
      num.threads = n_cores,
      seed = seed)
    
    # Predict
    rf_predict <- predict(ranger_model, test_data, seed = seed)
    sce_exo[[clust_name]] <<- rf_predict$predictions
  })
})
```

### Original vs. predicted

Compare predicted clusters and original clusters on train set (cells on which the clustering algorithms were applied).

```{r random-forest-compare-clusters, fig.width=15, fig.height=10}
dat_sub <- scuttle::makePerCellDF(sce_sub, use_dimred = TRUE) |>
  mutate(case_id = factor(case_id, levels = meta$cases),
         donor_type = factor(donor_type, levels = meta$stages)) |>
  arrange(case_id, donor_type)

dat_exo <- scuttle::makePerCellDF(sce_exo, use_dimred = TRUE) |>
  mutate(case_id = factor(case_id, levels = meta$cases),
         donor_type = factor(donor_type, levels = meta$stages)) |>
  arrange(case_id, donor_type)
dat_exo <- dat_exo[sample(rownames(dat_exo), nrow(dat_sub)), ]

purrr::walk(assay_sel, \(cur_assay) {
  purrr::walk(methods_sel, \(cur_method) {
    dimred_name <- paste("UMAP", cur_assay, sep = "_")
    clust_name <- paste(cur_method, cur_assay, sep = "_")

    # Plot all clusters
    p_sub <- plot_dim_red(dat_sub, dimred_name, clust_name,
                          sample = TRUE, size = 0.1, alpha = 1) +
      ggtitle(paste(clust_name, "Original", sep = " - "))

    p_not <- plot_dim_red(dat_exo, dimred_name, clust_name,
                          sample = TRUE, size = 0.1, alpha = 1) +
      ggtitle(paste(clust_name, "Predicted", sep = " - "))

    p <- p_sub + p_not
    if (do_print) print(p)

    fn <- paste0(paste(today, clust_name, "RFcomparison", sep = "_"), ".png")
    do.call(ggsave, c(list(fn, p), plotsave_param))
  })
})
remove(dat_sub, dat_exo, p, p_sub, p_not)
```

### Merge SCE objects

The `sce_exo` and `sce_sub` again.

```{r merge-sce-objects}
sce_exo <- cbind(sce_exo, sce_sub)
```

```{r save-sce-exo}
fn_sce_exo <- file.path(paths$folder_script, paste0("sce_exo", paths$object_type, paths$panel_type, ".rds"))
saveRDS(sce_exo, fn_sce_exo)
# sce_exo <- readRDS(fn_sce_exo)
# sce_sub <- sce_exo[channels_clust, colnames(sce_exo) %in% metadata(sce_exo)$subset]
# meta <- metadata(sce_exo)
```

# **Visualize clusters**

## Load packages

```{r packages4, results='hide'}
suppressPackageStartupMessages(c(
  library(heatmaply),
  library(htmltools),
  library(cytomapper),
  library(ggalluvial)
))
```


## Visualize clusters on reduced dimensions

### Plot marker expression

```{r plot-dimred-channels, fig.height=10, fig.width=15}
purrr::walk(assay_sel, \(cur_assay) {
  purrr::walk(dimred_sel, \(cur_dimred) {
    dimred_name <- paste(cur_dimred, cur_assay, sep = "_")

    # Prepare the data
    cur_dat <- scuttle::makePerCellDF(sce_sub,
                                      features = channels_clust,
                                      exprs_values = cur_assay) |>
      select(c(all_of(channels_clust),
               paste0(dimred_name, ".1"), paste0(dimred_name, ".2")))

    cur_dat <- cur_dat |>
      tidyr::pivot_longer(cols = all_of(channels_clust),
                          names_to = "channel",
                          values_to = cur_assay)
    
    # Plot marker expression
    p <- plot_dim_red_channels(cur_dat, dimred_name, cur_assay,
                               channels_clust, force_points = TRUE)

    fn <- paste0(paste(today, dimred_name, "Channels", sep = "_"), ".png")
    do.call(ggsave, c(list(fn, p), plotsave_param_large))
    if (do_print) print(p)
  })
})
```

### Plot clusters

```{r dimred-clusters}
cur_dat <- makePerCellDF(sce_sub, use_dimred = TRUE) |>
  mutate(case_id = factor(case_id, levels = meta$cases),
         donor_type = factor(donor_type, levels = meta$stages)) |>
  arrange(case_id, donor_type)

purrr::walk(assay_sel, \(cur_assay) {
  purrr::walk(dimred_sel, \(cur_dimred) {
    purrr::walk(methods_sel, \(cur_method) {
      dimred_name <- paste(cur_dimred, cur_assay, sep = "_")
      clust_name <- paste(cur_method, cur_assay, sep = "_")
      
      # Plot all clusters
      p <- plot_dim_red(cur_dat, dimred_name, clust_name,
                        sample = TRUE, size = 0.1, alpha = 1)
      if (do_print) print(p)
      
      fn <- paste0(paste(today, "Clusters", clust_name, cur_dimred,
                         sep = "_"), ".png")
      do.call(ggsave, c(list(fn, p), plotsave_param))
    })
  })
})
# remove(cur_dat, p)
```


## Distribution of marker expression by cluster

```{r violin-clusters-markers, message=FALSE, warning=FALSE, fig.width=15, fig.height=10}
purrr::walk(assay_sel, \(cur_assay) {
  purrr::walk(methods_sel, \(cur_method) {
    clust_name <- paste(cur_method, cur_assay, sep = "_")
    sce_sub[[clust_name]] <- as.factor(sce_sub[[clust_name]])
    cur_dat <- makePerCellDF(sce_sub, assay.type = cur_assay,
                             features = channels_clust, use.dimred = FALSE)
    
    cur_dat <- melt.data.table(
      as.data.table(cur_dat),
      id.vars = c("image_number", "cell_number", clust_name),
      measure.vars = channels_clust,
      variable.name = "channel",
      value.name = cur_assay
    )
    cur_dat[, MeanExprs := mean(get(cur_assay)),
            by = c("channel", clust_name)]
    
    p <- plot_violin(cur_dat, x = "channel", y = cur_assay,
                     fill_by = "MeanExprs", facet_by = clust_name,
                     title = clust_name) +
      theme(axis.text.x = element_text(size = 8, angle = 90))
    suppressMessages(if (do_print) print(p))
    fn <- paste0(paste(today, "Clusters", clust_name, "Violins",
                       sep = "_"), ".png")
    do.call(ggsave, c(list(fn, p), plotsave_param_large))
  })
})
```

## Heatmap of marker expression by cluster

```{r heatmap-clusters-markers}
hmlist <- list()
n <- 1
purrr::walk(assay_sel, \(cur_assay) {
  purrr::walk(methods_sel, \(cur_method) {
    clust_name <- paste(cur_method, cur_assay, sep = "_")
    
    # Summarize the data
    hm <- summarize_heatmap(sce_sub,
                            expr_values = cur_assay,
                            cluster_by = clust_name,
                            channels = channels_clust)
    
    # Display the heatmap
    fn <- paste0(paste(today, "Clusters", clust_name, "Heatmap",
                       sep = "_"), ".html")
    hmlist[[n]] <- heatmaply(
      heatmaply::normalize(hm), main = clust_name,
      file = file.path(paths$folder_script, fn))
    n <- n + 1

    # remove all clusters with less than 10 cells and plot again.
    clust_freq <- table(colData(sce_sub)[[clust_name]])
    keep_clusts <- names(clust_freq)[clust_freq >= 110]
    sce_isl_sub <- sce_sub[, colData(sce_sub)[[clust_name]] %in% keep_clusts]

    hm <- summarize_heatmap(sce_isl_sub,
                            expr_values = cur_assay,
                            cluster_by = clust_name,
                            channels = channels_clust)
    # Display the reduced heatmap -> this improves the visualization.
    # Otherwise heatmap is scaled by expression of outliers.
    fn <- paste0(paste(today, "Clusters", clust_name, "Heatmap_reduced",
                       sep = "_"), ".html")
    hmlist[[n]] <- heatmaply(
       heatmaply::normalize(hm), main = clust_name, 
       file = file.path(paths$folder_script, fn))
    n <- n + 1
  })
})
htmltools::tagList(setNames(hmlist, NULL))
```


## Barplot with cluster frequencies
Visualize the number of cells per cluster.
```{r viz-cluster-barplot}
cur_dat <- scuttle::makePerCellDF(sce_sub, use_dimred = TRUE) |>
  dplyr::mutate(case_id = factor(case_id, levels = meta$cases),
                donor_type = factor(donor_type, levels = meta$stages)) |>
  dplyr::arrange(case_id, donor_type) |> 
  tibble::as_tibble()

purrr::walk(methods_sel, \(cur_method) {
  purrr::walk(assay_sel, \(cur_assay) {
    clust_name <- paste(cur_method, cur_assay, sep = "_")
    title <- paste0("Cluster Abundances: ", clust_name)
    df <- cur_dat |> 
      dplyr::select(all_of(clust_name))  |> 
      dplyr::group_by(across(all_of(clust_name))) |> 
      dplyr::summarise(n = n())

    p <- df |> 
      ggplot2::ggplot(ggplot2::aes(x = factor(get(clust_name)),
                                   y = n, fill = factor(get(clust_name)))) +
      mytheme$standard() +
      ggplot2::theme(
        axis.text.x = ggplot2::element_text(angle = 45, hjust = 1),
        panel.background = ggplot2::element_rect(colour = "grey20")
      ) + 
      ggplot2::geom_bar(stat = "identity") +
      ggtitle(title) + 
      ggplot2::scale_fill_manual(values = palettes$colors[seq_along(unique(cur_dat[[clust_name]]))]) + 
      xlab("Cluster") +
      ylab("Number of cells")
    if (do_print) print(p)
    fn <- paste0(paste(today, "Clusters", clust_name, "Barplot",
                       sep = "_"), ".png")
    do.call(ggsave, c(list(fn, p), plotsave_param_large))
  })
})
```

## Visualize clusters with cytomapper

### Select cluster(s) and assay to show

```{r viz-clust-select}
viz_clust <- NULL # c(7, 16, 20) # c(8, 12) # NULL #c(3, 17, 21) NULL
viz_method <- "Pheno5"
viz_assay <- "scaled"
```

### Load images and masks
```{r sce_sub_viz}
if (!is.null(viz_clust)) {
  sce_exo2 <- spe[, spe$cell_category != "immune"]
  combinations <- expand.grid(methods_sel, assay_sel)
  combined_vector <- apply(combinations, 1, paste, collapse = "_")
  # sce_exo2[[clust_name]] <- sce_exo[[clust_name]]
  clusters_oi <- colData(sce_exo)[, combined_vector]
  cols_oi <- colnames(sce_exo)
  rm(spe, sce_exo); gc()
  sce_exo2 <- sce_exo2[, cols_oi]
  colData(sce_exo2)[, combined_vector] <- clusters_oi
}
```

```{r viz-cluster-load}
if (!is.null(viz_clust)) {
  nb_images <- 14
  image_extension <- ".tiff"
  clust_name <- paste(viz_method, viz_assay, sep = "_")
  
  # Subset the SCE
  sce_viz <- sce_exo2[, colData(sce_exo2)[[clust_name]] %in% viz_clust]

  set.seed(seed)
  image_sub <- sort(sample(
    unique(sce_viz$image_fullname),
    min(length(unique(sce_viz$image_fullname)), nb_images)))

  image_sub <- unique(c(image_sub, "6197_Immune_ROI_083.tiff", # "6422_Immune_ROI_039.tiff", 
                "6034_Immune_ROI_050.tiff", "6494_Immune_ROI_008.tiff", 
                "6197_Immune_ROI_072.tiff", "6514_Immune_ROI_005.tiff", "6225_Immune_ROI_053.tiff", "6225_Immune_ROI_043.tiff", "6225_Immune_ROI_057.tiff",
                "6314_Immune_ROI_008.tiff", "6328_Immune_ROI_078.tiff", # "6328_Immune_ROI_001.tiff",
                "6034_Immune_ROI_034.tiff", "6209_Immune_ROI_057.tiff", "6550_Immune_ROI_072.tiff", 
                "6310_Immune_ROI_006.tiff", # "6043_Immune_ROI_014.tiff",  
                "6512_Immune_ROI_026.tiff", "6458_Immune_ROI_077.tiff"))

  # Folders
  folder_images <- file.path(paths$folder_in, "img", paths$panel_type)
  folder_masks <- file.path(paths$folder_in, "masks_cells",
                            paths$panel_type, "whole-cell")
  
  # Load images and masks
  images <- yoloader(
    x = sce_viz,
    image_dir = folder_images,
    image_names = image_sub,
    type = "stacks"
  )

  masks <- yoloader(
    x = sce_viz,
    image_dir = folder_masks,
    image_names = image_sub,
    as.is = TRUE,
    type = "masks"
  )
  
  sce_viz <- sce_viz[, sce_viz$image_fullname %in% image_sub]
  sce_viz$ImageName <- gsub(image_extension, "", sce_viz$image_fullname)
}
```

### Plot images

Images are saved to disk but not shown.

```{r viz-cluster-plot}
if (!is.null(viz_clust)) {
  fn <- list(filename = file.path(
    plotsave_param$path,
    paste0(paste(today, "Cytomapper", clust_name,
                 paste(viz_clust, collapse = "-"), sep = "_"), ".png")))

  plotPixels(images, object = sce_viz, masks,
             cell_id = "cell_number", img_id = "ImageName",
             colour_by = c("CD8a", "CD3e", "MPO", "CD163", "CD204", "CD7"),
             outline_by = c(clust_name),
             bcg = list(CD8a = c(0, 5, 1), CD3e = c(0, 5, 1), MPO = c(0, 7, 1),
                        CD163 = c(0, 5, 1), CD204 = c(0, 5, 1), CD7 = c(0, 5, 1)),
             image_title = NULL, scale = TRUE, save_plot = fn)
}
```

```{r cytoviewer}
if (!is.null(viz_clust)) {
  library(cytoviewer)
  sce_viz[[clust_name]] <- as.factor(sce_viz[[clust_name]])
  app <- cytoviewer(image = images, 
                    mask = masks, 
                    object = sce_viz,
                    img_id = "ImageName", 
                    cell_id = "cell_number")

  if (interactive()) {
    shiny::runApp(app)
  }
}
```

### Remove generated files

```{r viz-cluster-remove}
if (!is.null(viz_clust)) {
  remove(sce_viz, images, masks)
}
```


## Plot specific cluster(s)

When many clusters are present, they may not be easy fo visualize on reduced dimension maps. Here, a subset of clusters can be selected for plotting.

### Select cluster(s) and assay to show

```{r dimred-cluster-select}
viz_clust <- NULL # c(7, 19, 20)
viz_method <- "Leiden5"
viz_assay <- "fastMNN_case_id"
```

```{r dimred-cluster-plot}
if (!is.null(viz_clust)) {
  # Prepare the data
  cur_dat <- scuttle::makePerCellDF(
    sce_exo2[, colnames(sce_exo2) %in% meta[["subset"]]],
    use.dimred = TRUE) |>
    mutate(case_id = factor(case_id, levels = meta$cases),
           donor_type = factor(donor_type, levels = meta$stages)) |>
    arrange(case_id, donor_type)
  
  # Clusters and palette
  clust_name <- paste(viz_method, viz_assay, sep = "_")
  clust_nb <-  length(unique(cur_dat[[clust_name]]))
  viz_pal <- palettes$colors[1:clust_nb]
  names(viz_pal) <- 1:clust_nb
  viz_pal[!names(viz_pal) %in% viz_clust] <- "lightgrey"
  
  # Plot
  purrr::walk(dimred_sel, \(cur_dimred) {
    dimred_name <- paste(cur_dimred, viz_assay, sep = "_")
    
    p <- plot_dim_red(cur_dat, dimred_name, clust_name, palette = viz_pal,
                      sample = TRUE, size = 0.1, alpha = 1)
    if (do_print) print(p)
    
    fn <- paste0(
       paste(today, paste0("Cluster", paste(viz_clust, collapse = "-")),
             clust_name, cur_dimred, sep = "_"), ".png")
    do.call(ggsave, c(list(fn, p), plotsave_param))
  })
}
```



# **Attribute cell types**

**Note: this section requires manual intervention**  

Here, clusters obtained at the previous steps are manually merged into 
meaningful cell types. Cluster numbers are attributed to the different cell 
types based on the plots and heatmaps above. Consequently, this attribution has
to be adapted to the clustering results.  

## Cell type attribution

Cell type attribution has to be performed once per clustering method and per 
selected assay.

### Phenograph x scaled counts

```{r celltypes-pheno-scaled}
clust_methods <- c("Pheno5")
clust_assay <- c("scaled")

if (!clust_assay %in% assay_sel)
  stop("The selected assay is not in the assays selected for clustering")
if (length(clust_assay) != 1)
  stop("Select only one assay")

clust_name <- paste(clust_methods, clust_assay, sep = "_")
print(clust_name)

# Define which clusters correspond to which cell types
#clust_beta <- c(11)
#clust_alpha  <- c(16, 18)
#clust_delta <- c(21)
#clust_islet_other <- c()
#clust_ductal <- c(4, 6, 7, 14, 20, 30)
#clust_acinar <- c(1, 3, 10, 23, 25, 26, 28, 29, 31, 32)
#clust_endothelial <- c(12)
#clust_mesenchymal <- c(19, 24, 27)
#clust_other <- c(2, 5, 8, 9, 13, 15, 17, 22)
clust_beta <- c(19)
clust_alpha <- c(6)
clust_delta <- c(22)
clust_islet_other <- c()
clust_ductal <- c(1, 8, 11, 13, 14, 18, 21, 28, 29, 31)
clust_acinar <- c(2:5, 7, 9, 10, 16, 17, 20)
clust_endothelial <- c(12, 15, 23)
clust_mesenchymal <- c(24)
clust_smooth_muscle <- c(30)
clust_nerves <- c(26)
clust_other <- c(25, 27)

all_clust <- sort(c(clust_beta, clust_alpha, clust_delta, clust_islet_other,
                    clust_ductal, clust_acinar, clust_endothelial,
                    clust_mesenchymal, clust_smooth_muscle, clust_nerves, clust_other))

if ((!length(unique(all_clust)) ==
     length(unique(colData(sce_exo)[, clust_name]))) ||
    any(duplicated(all_clust))) {
  stop("Recheck cluster attribution")
}

# Add cell types to the SCE object
colData(sce_exo)[colData(sce_exo)[, clust_name] %in% clust_beta,
             paste("CellType", clust_name, sep = "_")] <- "Beta"
colData(sce_exo)[colData(sce_exo)[, clust_name] %in% clust_alpha,
             paste("CellType", clust_name, sep = "_")] <- "Alpha"
colData(sce_exo)[colData(sce_exo)[, clust_name] %in% clust_delta,
             paste("CellType", clust_name, sep = "_")] <- "Delta"
colData(sce_exo)[colData(sce_exo)[, clust_name] %in% clust_islet_other,
             paste("CellType", clust_name, sep = "_")] <- "IsletOther"
colData(sce_exo)[colData(sce_exo)[, clust_name] %in% clust_ductal,
             paste("CellType", clust_name, sep = "_")] <- "Ductal"
colData(sce_exo)[colData(sce_exo)[, clust_name] %in% clust_acinar,
             paste("CellType", clust_name, sep = "_")] <- "Acinar"
colData(sce_exo)[colData(sce_exo)[, clust_name] %in% clust_endothelial,
             paste("CellType", clust_name, sep = "_")] <- "Endothelial"
colData(sce_exo)[colData(sce_exo)[, clust_name] %in% clust_mesenchymal,
             paste("CellType", clust_name, sep = "_")] <- "Mesenchymal"
colData(sce_exo)[colData(sce_exo)[, clust_name] %in% clust_smooth_muscle,
             paste("CellType", clust_name, sep = "_")] <- "SmoothMuscle"
colData(sce_exo)[colData(sce_exo)[, clust_name] %in% clust_nerves,
              paste("CellType", clust_name, sep = "_")] <- "Nerves"
colData(sce_exo)[colData(sce_exo)[, clust_name] %in% clust_other,
             paste("CellType", clust_name, sep = "_")] <- "Other"
```

### Phenograph x fastMNN counts

```{r celltypes-pheno-fastMNN}
clust_methods <- c("Pheno5")
clust_assay <- c("fastMNN_case_id")

if (!clust_assay %in% assay_sel)
  stop("The selected assay is not in the assays selected for clustering")
if (length(clust_assay) != 1)
  stop("Select only one assay")

clust_name <- paste(clust_methods, clust_assay, sep = "_")
print(clust_name)

# Define which clusters correspond to which cell types
#clust_beta <- c(13)
#clust_alpha  <- c(12, 25)
#clust_delta <- c(24)
#clust_islet_other <- c(20)
#clust_ductal <- c(1, 4, 6, 8, 9)
#clust_acinar <- c(2, 3, 7, 10, 11, 19, 27)
#clust_endothelial <- c(26)
#clust_mesenchymal <- c(5, 15, 18, 21)
#clust_other <- c(14, 16, 17, 22, 23)
clust_beta <- c(18)
clust_alpha <- c(6)
clust_delta <- c(20)
clust_islet_other <- c()
clust_ductal <- c(1, 3, 4, 9, 16, 23)
clust_acinar <- c(2, 5, 7, 8, 10:14, 24) 
clust_mesenchymal <- c(19)
clust_smooth_muscle <- c(22)
clust_endothelial <- c(15, 21)
clust_nerves <- c()
clust_other <- c(17)

all_clust <- sort(c(clust_beta, clust_alpha, clust_delta, clust_islet_other,
                    clust_ductal, clust_acinar, clust_endothelial, 
                    clust_mesenchymal, clust_smooth_muscle,
                    clust_nerves, clust_other))

if ((!length(unique(all_clust)) ==
     length(unique(colData(sce_exo)[, clust_name]))) ||
    any(duplicated(all_clust))) {
  stop("Recheck cluster attribution")
}

# Add cell types to the SCE object
colData(sce_exo)[colData(sce_exo)[, clust_name] %in% clust_beta,
                 paste("CellType", clust_name, sep = "_")] <- "Beta"
colData(sce_exo)[colData(sce_exo)[, clust_name] %in% clust_alpha,
                 paste("CellType", clust_name, sep = "_")] <- "Alpha"
colData(sce_exo)[colData(sce_exo)[, clust_name] %in% clust_delta,
                 paste("CellType", clust_name, sep = "_")] <- "Delta"
colData(sce_exo)[colData(sce_exo)[, clust_name] %in% clust_islet_other,
                 paste("CellType", clust_name, sep = "_")] <- "IsletOther"
colData(sce_exo)[colData(sce_exo)[, clust_name] %in% clust_ductal,
                 paste("CellType", clust_name, sep = "_")] <- "Ductal"
colData(sce_exo)[colData(sce_exo)[, clust_name] %in% clust_acinar,
                 paste("CellType", clust_name, sep = "_")] <- "Acinar"
colData(sce_exo)[colData(sce_exo)[, clust_name] %in% clust_endothelial,
                 paste("CellType", clust_name, sep = "_")] <- "Endothelial"
colData(sce_exo)[colData(sce_exo)[, clust_name] %in% clust_mesenchymal,
                 paste("CellType", clust_name, sep = "_")] <- "Mesenchymal"
colData(sce_exo)[colData(sce_exo)[, clust_name] %in% clust_smooth_muscle,
                 paste("CellType", clust_name, sep = "_")] <- "SmoothMuscle"
colData(sce_exo)[colData(sce_exo)[, clust_name] %in% clust_nerves,
                 paste("CellType", clust_name, sep = "_")] <- "Nerves"
colData(sce_exo)[colData(sce_exo)[, clust_name] %in% clust_other,
                 paste("CellType", clust_name, sep = "_")] <- "Other"
```

### Leiden x scaled counts

```{r celltypes-leiden-scaled}
clust_methods <- c("Leiden5")
clust_assay <- c("scaled")

if (!clust_assay %in% assay_sel)
  stop("The selected assay is not in the assays selected for clustering")
if (length(clust_assay) != 1)
  stop("Select only one assay")

clust_name <- paste(clust_methods, clust_assay, sep = "_")
print(clust_name)

# Define which clusters correspond to which cell types
#clust_beta <- c(11)
#clust_alpha  <- c(17)
#clust_delta <- c(22)
#clust_ductal <- c(5, 8, 9, 10, 12, 18)
#clust_islet_other <- c()
#clust_acinar <- c(4, 6, 14, 15, 20, 21, 24)
#clust_endothelial <- c(13)
#clust_mesenchymal <- c(3, 16)
#clust_other <- c(1, 2, 7, 19, 23)
clust_beta <- c(14)
clust_alpha <- c(13)
clust_delta <- c(20)
clust_islet_other <- c()
clust_ductal <- c(1, 2, 12, 17, 25, 26)
clust_acinar <- c(3, 5:11, 23, 24)
clust_endothelial <- c(19, 21)
clust_mesenchymal <- c(4, 18)
clust_smooth_muscle <- c(16)
clust_nerves <- c(22)
clust_other <- c(15)

all_clust <- sort(c(clust_beta, clust_alpha, clust_delta, clust_islet_other,
                    clust_ductal, clust_acinar, clust_endothelial,
                    clust_mesenchymal, clust_smooth_muscle, clust_nerves, clust_other))

if ((!length(unique(all_clust)) ==
     length(unique(colData(sce_exo)[, clust_name]))) ||
    any(duplicated(all_clust))) {
  stop("Recheck cluster attribution")
}

# Add cell types to the SCE object
colData(sce_exo)[colData(sce_exo)[, clust_name] %in% clust_beta,
                 paste("CellType", clust_name, sep = "_")] <- "Beta"
colData(sce_exo)[colData(sce_exo)[, clust_name] %in% clust_alpha,
                 paste("CellType", clust_name, sep = "_")] <- "Alpha"
colData(sce_exo)[colData(sce_exo)[, clust_name] %in% clust_delta,
                 paste("CellType", clust_name, sep = "_")] <- "Delta"
colData(sce_exo)[colData(sce_exo)[, clust_name] %in% clust_islet_other,
                 paste("CellType", clust_name, sep = "_")] <- "IsletOther"
colData(sce_exo)[colData(sce_exo)[, clust_name] %in% clust_ductal,
                 paste("CellType", clust_name, sep = "_")] <- "Ductal"
colData(sce_exo)[colData(sce_exo)[, clust_name] %in% clust_acinar,
                 paste("CellType", clust_name, sep = "_")] <- "Acinar"
colData(sce_exo)[colData(sce_exo)[, clust_name] %in% clust_endothelial,
                 paste("CellType", clust_name, sep = "_")] <- "Endothelial"
colData(sce_exo)[colData(sce_exo)[, clust_name] %in% clust_mesenchymal,
                 paste("CellType", clust_name, sep = "_")] <- "Mesenchymal"
colData(sce_exo)[colData(sce_exo)[, clust_name] %in% clust_smooth_muscle,
                 paste("CellType", clust_name, sep = "_")] <- "SmoothMuscle"
colData(sce_exo)[colData(sce_exo)[, clust_name] %in% clust_nerves,
                 paste("CellType", clust_name, sep = "_")] <- "Nerves"
colData(sce_exo)[colData(sce_exo)[, clust_name] %in% clust_other,
                 paste("CellType", clust_name, sep = "_")] <- "Other"
```

### Leiden1 x fastMNN counts

```{r celltypes-leiden-fastMNN}
clust_methods <- c("Leiden5")
clust_assay <- c("fastMNN_case_id")

if (!clust_assay %in% assay_sel)
  stop("The selected assay is not in the assays selected for clustering")
if (length(clust_assay) != 1)
  stop("Select only one assay")

clust_name <- paste(clust_methods, clust_assay, sep = "_")
print(clust_name)

# Define which clusters correspond to which cell types
#clust_beta <- c(13)
#clust_alpha  <- c(11)
#clust_islet_other <- c()
#clust_delta <- c(14)
#clust_ductal <- c(1, 7, 9, 20, 21)
#clust_acinar <- c(2, 5, 6, 12, 19)
#clust_endothelial <- c(18)
#clust_mesenchymal <- c(3, 15)
#clust_other <- c(4, 8, 10, 16, 17, 22)
clust_beta <- c(19)
clust_alpha <- c(7)
clust_delta <- c(20)
clust_ductal <- c(1, 2, 13, 16, 22)
clust_endothelial <- c(15, 21)
clust_mesenchymal <- c()
clust_smooth_muscle <- c(10)
clust_acinar <- c(3:6, 8, 9, 11, 12, 14, 17, 18)
clust_nerves <- c()
clust_other <- c(23:25)

all_clust <- sort(c(clust_beta, clust_alpha, clust_delta, clust_islet_other,
                    clust_ductal, clust_acinar, clust_endothelial,
                    clust_mesenchymal, clust_smooth_muscle, clust_nerves, clust_other))

if ((!length(unique(all_clust)) ==
     length(unique(colData(sce_exo)[, clust_name]))) ||
    any(duplicated(all_clust))) {
  stop("Recheck cluster attribution")
}

# Add cell types to the SCE object
colData(sce_exo)[colData(sce_exo)[, clust_name] %in% clust_beta,
                 paste("CellType", clust_name, sep = "_")] <- "Beta"
colData(sce_exo)[colData(sce_exo)[, clust_name] %in% clust_alpha,
                 paste("CellType", clust_name, sep = "_")] <- "Alpha"
colData(sce_exo)[colData(sce_exo)[, clust_name] %in% clust_delta,
                 paste("CellType", clust_name, sep = "_")] <- "Delta"
colData(sce_exo)[colData(sce_exo)[, clust_name] %in% clust_islet_other,
                 paste("CellType", clust_name, sep = "_")] <- "IsletOther"
colData(sce_exo)[colData(sce_exo)[, clust_name] %in% clust_ductal,
                 paste("CellType", clust_name, sep = "_")] <- "Ductal"
colData(sce_exo)[colData(sce_exo)[, clust_name] %in% clust_acinar,
                 paste("CellType", clust_name, sep = "_")] <- "Acinar"
colData(sce_exo)[colData(sce_exo)[, clust_name] %in% clust_endothelial,
                 paste("CellType", clust_name, sep = "_")] <- "Endothelial"
colData(sce_exo)[colData(sce_exo)[, clust_name] %in% clust_mesenchymal,
                 paste("CellType", clust_name, sep = "_")] <- "Mesenchymal"
colData(sce_exo)[colData(sce_exo)[, clust_name] %in% clust_smooth_muscle,
                 paste("CellType", clust_name, sep = "_")] <- "SmoothMuscle"
colData(sce_exo)[colData(sce_exo)[, clust_name] %in% clust_nerves,
                 paste("CellType", clust_name, sep = "_")] <- "Nerves"
colData(sce_exo)[colData(sce_exo)[, clust_name] %in% clust_other,
                 paste("CellType", clust_name, sep = "_")] <- "Other"
```


## Visualize attributed cell types

### Plot cell types on reduced dimensions

```{r palette$celltypes}
palettes$celltype <- (c(
  Beta = palettes$colors[1], Alpha = palettes$colors[2],
  Delta = palettes$colors[3], IsletOther = palettes$colors[4],
  Ductal = palettes$colors[5], Acinar = palettes$colors[6],
  Endothelial = palettes$colors[7], Mesenchymal = palettes$colors[8],
  SmoothMuscle = palettes$colors[9], Nerves = palettes$colors[10],
  Other = "lightgrey")
)
```


```{r celltypes-dimred}
# Subset again (to include newly-attributed cell types)
sce_sub <- sce_exo[channels_clust, colnames(sce_exo) %in% meta$subset]

# Prepare the data
cur_dat <- makePerCellDF(
  sce_sub[, colnames(sce_sub) %in% meta[["subset"]]],
  use_dimred = TRUE) |>
  mutate(case_id = factor(case_id, levels = meta$cases),
         donor_type = factor(donor_type, levels = meta$stages)) |>
  arrange(case_id, donor_type)

# Plot
purrr::walk(assay_sel, \(cur_assay) {
  purrr::walk(dimred_sel, \(cur_dimred) {
    purrr::walk(methods_sel, \(cur_method) {
      dimred_name <- paste(cur_dimred, cur_assay, sep = "_")
      clust_name <- paste("CellType", cur_method, cur_assay,
                          sep = "_")
      
      p <- plot_dim_red(cur_dat, dimred_name, clust_name,
                        palette = palettes$celltype,
                        sample = TRUE, size = 0.1, alpha = 1)
      if (do_print) print(p)
      
      fn <- paste0(paste(today, clust_name, cur_dimred,
                         sep = "_"), ".png")
      do.call(ggsave, c(list(fn, p), plotsave_param))
    })
  })
})
```

### Cell types heatmap

```{r celltypes-heatmap}
hmlist <- list()
n <- 1
purrr::walk2(assay_sel, names(assay_sel), \(cur_assay, name_cur_assay) {
  purrr::walk(methods_sel, \(cur_method) {
    clust_name <- paste("CellType", cur_method, cur_assay,
                        sep = "_")
  
    # Summarize the data
    hm <- summarize_heatmap(sce_sub,
                            expr_values = name_cur_assay,
                            cluster_by = clust_name,
                            channels = channels_clust)
    
    # Display the heatmap
    fn <- paste0(paste(today, clust_name, "Heatmap",
                       sep = "_"), ".html")
    
    hmlist[[n]] <- heatmaply(
      heatmaply::normalize(hm), main = clust_name,
      file = file.path(paths$folder_script, fn))
    n <- n + 1
  })
})
htmltools::tagList(setNames(hmlist, NULL))
```

### Visualize celltypes with cytomapper

Load the masks

```{r viz-celltype-load}
nb_images <- 14
image_extension <- ".tiff"

set.seed(seed)
image_sub <- sort(sample(
  unique(sce_sub$image_fullname),
  min(length(unique(sce_sub$image_fullname)), nb_images)))

# Subset the SCE to subsampled images
sce_viz <- sce_exo[, sce_exo$image_fullname %in% image_sub]
sce_viz$ImageName <- gsub(image_extension, "", sce_viz$image_fullname)

# Folders
folder_images <- file.path(paths$folder_in, "img", paths$panel_type)
folder_masks <- file.path(paths$folder_in, "masks_cells",
                          paths$panel_type, "whole-cell")

# Load masks
masks <- yoloader(
  x = sce_viz,
  image_dir = folder_masks,
  image_names = image_sub,
  as.is = TRUE,
  type = "masks"
)
```

Plot the images

```{r viz-celltype-plot}
purrr::walk(assay_sel, \(cur_assay) {
  purrr::walk(methods_sel, \(cur_method) {
    celltype_viz <- paste("CellType", cur_method, cur_assay,
                          sep = "_")
    
    colour_key <- list(celltype_viz = palettes$celltype)

    fn <- list(filename = file.path(plotsave_param$path,
      paste0(paste(today, "Cytomapper", celltype_viz, sep = "_"), ".png")))

    cytomapper::plotCells(masks, object = sce_viz, 
                          cell_id = "cell_number", img_id = "ImageName",
                          colour_by = celltype_viz, colour = colour_key, 
                          save_plot = fn)
  })
})
```



# **Consensus cell categories**

Attribute consensus cell categories based on the clustering performed in the previous script.

## Load packages

```{r packages5, results='hide'}
suppressPackageStartupMessages(c(
  library(ggalluvial)
))
```

## Cell category concordance

Compare concordance of cell categories attributed using different methods.

```{r cellcats-concordance}
celltype_cols <- colnames(colData(sce_exo))[grepl(paste0(
  "CellType_", methods_sel, collapse = "|"), colnames(colData(sce_exo)))]

cell_types <- as.data.table(colData(sce_exo)[
  , c("cell_id", "case_id", "image_number", "cell_number", celltype_cols)])

ct_combination <- t(combn(unique(celltype_cols), 2))

for (i in seq_len(nrow(ct_combination))) {
  cat("\n", ct_combination[i, 1], " vs ", ct_combination[i, 2], "\n")
  
  ct_concord <- nrow(cell_types[
    get(ct_combination[i, 1]) == get(ct_combination[i, 2]), ]
  )
  ct_discord <- nrow(cell_types[
    get(ct_combination[i, 1]) != get(ct_combination[i, 2]), ]
  )
  cat("Number of cells with concordant cell types: ", ct_concord,
      "\nNumber of cells with discordant cell types: ", ct_discord,
      "\nPercentage of cells with concordant cell types:",
      100 * ct_concord / (ct_discord + ct_concord), "%", "\n")
}
```
Pheno5-scaled vs Pheno5-fastMNN_case_id: 84 %
Pheno5-scaled vs Leiden5-scaled: 88 %
Pheno5-scaled vs Leiden5-fastMNN_case_id: 86 %
Pheno5-fastMNN_case_id vs Leiden5-scaled: 84 %
Pheno5-fastMNN_case_id vs Leiden5-fastMNN_case_id: 91 %
Leiden5-scaled vs Leiden5-fastMNN_case_id: 84 %

## Attribute consensus cell categories

### Majority vote

Cells for which 3 out of 4 methods agree on the cell categories are attributed this category, other cells as considered as "ambiguous".

```{r consensus-cell-categories}
# Majority vote
Mode <- function(x) {
  ux <- unique(x)
  ux[which.max(tabulate(match(x, ux)))]
}

cell_types[, cell_type := apply(cell_types, 1, Mode)]
cell_types[, Paste := do.call(paste0, .SD),
           .SDcols = grep("CellType_", colnames(cell_types))]
cell_types[, count := stringr::str_count(Paste, cell_type)]
cell_types[count < 3, cell_type := "Ambiguous"]
cell_types[, `:=`(Paste = NULL, count = NULL)]

cat("Percentage of cells with an ambiguous cell type: ", 100 * nrow(
  cell_types[cell_type == "Ambiguous"]) / nrow(cell_types), "%")
```

### Compare cell type attributions

```{r cellcat-alluvial, warning=FALSE}
celltype_cols <- c("cell_type", colnames(colData(sce_exo))[grepl(paste0(
  "CellType_", methods_sel, collapse = "|"), colnames(colData(sce_exo)))])

p <- cell_types[, .(number = .N), by = celltype_cols] |>
  ggplot(aes(axis1 = get(celltype_cols[1]),
             axis2 = get(celltype_cols[2]),
             axis3 = get(celltype_cols[3]),
             axis4 = get(celltype_cols[4]),
             axis5 = get(celltype_cols[5]),
             y = number)) +
  geom_stratum() +
  geom_text(stat = "stratum", infer.label = TRUE) +
  geom_alluvium(aes(fill = get(celltype_cols[1]))) +
  scale_x_discrete(limits = gsub("cell_type_", "", celltype_cols)) +
  scale_fill_manual(values = palettes$colors[c(TRUE, FALSE)],
                    name = "Cell type") +
  mytheme$standard() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
if (do_print) print(p)

fn <- paste0(paste(today, "CellType", "alluvial", sep = "_"), ".png")
do.call(ggsave, c(list(fn, p), plotsave_param))
```

### Number of cells by cell category

```{r cells-per-category}
as_tibble(cell_types) |>
  select(cell_type) |>
  group_by(cell_type) |>
  add_count(name = "Number of cells") |>
  distinct()
```



# **Finalize and save **

## Add cell types to the main SCE

Add cell types to the non-immune SCE object

```{r add-celltypes-sce-1}
cell_types <- cell_types[order(match(cell_types$cell_id, colnames(sce_exo))), ] 
colData(sce_exo)$cell_type <- cell_types$CellType_Leiden5_scaled
# colData(sce_exo)$cell_type <- cell_types$cell_type
```

Make sure that cell categories are aligned with cell types

```{r update-cell-categories}
sce_exo[, sce_exo$cell_type %in% c("Acinar", "Ductal")]$cell_category <- "exocrine"

sce_exo[, sce_exo$cell_type %in% c("Beta", "Alpha", "Delta", "IsletOther")]$cell_category <- "islet"

# FIMXE: add nerves?
sce_exo[, sce_exo$cell_type %in% c("Endothelial", "Mesenchymal", "SmoothMuscle", "Nerves")]$cell_category <- "mesenchymal_endothelial"
```

Add cell types to the main SCE

```{r add-celltypes-sce-2}
# Make sure row order is preserved
row_order <- rownames(colData(spe)[spe$cell_category != "immune", ])
sce_exo <- sce_exo[, order(match(colnames(sce_exo), row_order))]

# Add cell types and updated cell categories
colData(spe)[row_order, ]$cell_category <- sce_exo$cell_category
colData(spe)[row_order, ]$cell_type <- sce_exo$cell_type
```


## Save the updated SPE

```{r save-spe}
print(spe)
fn_spe <- file.path(paths$folder_script, paste0(paths$object_type, "_", paths$panel_type, ".rds"))
saveRDS(spe, fn_spe)
```
