---
title: "09_CellTypeCompo_cells_Immune"
author: "Nicolas Damond, Nathan Steenbuck"
date: "Created: 25 May, 2022; Compiled: `r format(Sys.time(), '%d %b, %Y')`"
output: workflowr::wflow_html
editor_options:
  chunk_output_type: inline
---

```{r setup, include=FALSE}
if (rstudioapi::isAvailable()) {
  script_name <- basename(rstudioapi::getSourceEditorContext()$path)
} else {
  script_name <- knitr::current_input()
}

cur_user <- Sys.info()[["user"]]
script_name <- "09_CellTypeCompo_cells_Immune.Rmd"

if (cur_user == "ubuntu") {
  source(file.path("/", "mnt", "central_nas", "projects", 
    "type1_diabetes", "nathan", "T1D_Vol", "T1D_analysis", 
    "analysis", "helpers.R"))  #n_cores <- future::availableCores() - 1
  n_cores <- 4
  future::plan(future::multicore(workers = n_cores))
  paths <- getPaths(script_name)
  knitr::opts_knit$set(root_dir = paths$home_data)
} else {
  source(file.path("/", "home", "nsteen", "scripts", "T1D_analysis", "analysis", "helpers.R"))
  n_cores <- 8
  future::plan(future::multicore(workers = n_cores))
  paths <- getPaths(script_name)
  paths$folder_script <- paths$cluster_folder_script
  paths$folder_in <- paths$cluster_folder_in
  paths$folder_out <- paths$cluster_home 
  knitr::opts_knit$set(root_dir = paths$cluster_home)
}

do_print <- FALSE
seed <- 123456
set.seed(seed)
options <- furrr::furrr_options(seed = seed)
paths$prev <- paste("08_CellTypesIslet", paths$object_type, paths$panel_type, sep = "_")
knitr::opts_chunk$set(echo = TRUE)
```


# **A0: Goal**

Calculate and plot immune cell type and cluster composition across cases and 
stages.  

Analyses are applied only to immune cells, which should be defined as `immune` 
in the `spe$cell_category` column. Calculations and plots are performed at both
the immune cell types and at the immune cluster levels.

After calculating compositions, the following plots are generated:

+ **Individual image composition**: composition for every single image.
  - Absolute: total immune cell number per image.
  - Relative: relative immune cell number per image (0 to 1).

+ **Aggregated composition** - Absolute: boxplots showing absolute composition by 
patient.
  - Density: cell number / image area.
  - Fraction: cell number / total cell number on the image.

+ **Aggregated composition** - Relative: bar plots showing relative composition.
  - Normalized by cell type or by cluster.
  - Normalized by patient or by stage.
  
+ **Differential abundance tests!**

# A1: Settings

## A1.1: Load packages

```{r packages, results='hide'}
suppressPackageStartupMessages(c(
  library(dplyr),
  library(tidyr),
  library(data.table),
  library(SpatialExperiment),
  library(ggplot2),
  library(ggpubr),
  library(edgeR)
))
```

## A1.2: Paths and settings

```{r settings}
# Paths
if (!dir.exists(paths$folder_script)) dir.create(paths$folder_script)
plotsave_param$path <- paths$folder_script
plotsave_param_large$path <- paths$folder_script

# Misc settings
today <- gsub("-", "", Sys.Date())
```

## A1.3: Read in the SPE data

Load the SpatialExperiment (SPE) object saved at the previous step.

```{r load-data}
fn_spe <- file.path(paths$folder_out, paths$prev, paste0(paths$object_type, "_", paths$panel_type, ".rds"))
spe <- readRDS(fn_spe)
print(spe)
```


## A1.4: Calculate image area

```{r image-area}
spe$image_area <- spe$width_px * spe$height_px / 10^6
```

## A1.5: Subset immune cells

```{r subset-immune}
spe_imm <- spe[, spe$cell_category == "immune" & 
                 spe$cell_type != "Ambiguous"]
print(spe_imm)
```

## A1.6: Define the analysis parameters

```{r parameters}
# List of cell types
celltypes <- c("T_CD4", "T_CD8", "B", "NK", "Myeloid", "Neutrophil", "CD303_VIM", "T_DN", "Other")
names(celltypes) <- celltypes
cat("Immune cell types\n", celltypes)

# Color palette for plotting
palette_ct <- c(palettes$colors[seq_len(length(celltypes))])
names(palette_ct) <- celltypes

# List of clusters and associated color palettes
cluster_cols <- c("cell_cluster_scaled") #, "cell_cluster_fastMNN_case_id")
names(cluster_cols) <- c("scaled") #, "fastMNN_case_id")

clusters <- palettes_clust <- vector(mode = "list")
for (i in seq_along(cluster_cols)) {
  clusters[[i]] <- sort(unique(spe_imm[[cluster_cols[i]]]))
  palettes_clust[[i]] <- c(palettes$colors50[seq_len(length(clusters[[i]]))])
}
names(clusters) <- names(palettes_clust) <- cluster_cols
print(c("Immune cell clusters", clusters))

# Variables to plot
plot_variables <- c("donor_type", "case_id")
names(plot_variables) <- c("Stage", "Donor")
```



# **A2: Calculate cell type composition**

Here, we calculate the immune cell type (`image_compo_ct`) and immune clusters 
(`image_compo_clust`) composition per image.


## A2.1: Define IMMUNE cell type composition per image

The following values are computed, first by cell type, then by cluster:
- Number of immune cells per image.
- Immune cell fraction, over the total number of **immune** cells on the image.
- Immune cell density: number of immune cells / image size.

### A2.1.1: List image IDs

Calculations of cell numbers and cell fractions are performed at the image 
level rather than at the islet level. This means that all islets from the same 
image are considered as a single islet. Most images contain only one islet and
the composition of neighboring islets should be similar.

```{r image-ids}
image_ids <- unique(spe_imm$image_id)
```

### A2.1.2: Calculate image size and cell number per image

The image area is calculated in square millimeters.

```{r calculate-image-stats}
# Total number of cells per image
total_cells <- as_tibble(colData(spe)) |>
  filter(image_id %in% image_ids) |>
  dplyr::add_count(image_id, name = "CellsPerImage") |>
  select(c("image_id", "CellsPerImage")) |>
  distinct()

# Number of immune cells per image
total_immune_cells <- as_tibble(colData(spe_imm)) |>
  dplyr::add_count(image_id, name = "ImmuneCellsPerImage") |>
  select(c("image_id", "ImmuneCellsPerImage")) |>
  distinct()

# Image area (note: every image contains immune cells, so this works.)
image_area <- colData(spe_imm) |>
  as_tibble() |>
  select(c("image_id", "width_px", "height_px")) |>
  distinct() |>
  mutate(image_area = (width_px * height_px) / 1e6) |>
  select(c("image_id", "image_area"))

image_stats <- inner_join(total_cells, total_immune_cells, by = "image_id") |>
  inner_join(image_area, by = "image_id")
```

### A2.1.3: Calculate immune cell type composition (image)

```{r composition-celltype}
# Number of cells of each cell type
image_compo_ct <- as_tibble(colData(spe_imm)) |>
  group_by(image_id, cell_type) |>
  dplyr::count(name = "CellNb") |>
  ungroup() |>
  complete(image_id, cell_type,
           fill = list(CellNb = 0))

# Merge with image stats and calculate fractions and densities
image_compo_ct <- image_compo_ct |>
  inner_join(image_stats, by = "image_id") |>
  mutate(fraction_total = CellNb / CellsPerImage,
         fraction_immune = CellNb / ImmuneCellsPerImage,
         density = CellNb / image_area)

# Import relevant metadata (e.g., disease stage and patient ids)
cn_keep <- c("image_id", plot_variables)

image_compo_ct <- as_tibble(colData(spe_imm))[, cn_keep] |>
  distinct() |>
  inner_join(image_compo_ct, by = "image_id") |>
  mutate(cell_type = factor(cell_type, levels = celltypes),
         image_id = factor(image_id, levels = image_ids)) |>
  arrange(image_id, cell_type)

# Reorder the dataset by CD4 T cell fraction (for plotting)
image_compo_ct_wide <- image_compo_ct |>
  pivot_wider(
    id_cols = image_id,
    values_from = c(CellNb, fraction_total, fraction_immune, density),
    names_from = c(cell_type))

ordcols <- c(paste0("fraction_total_", celltypes))
image_compo_ct <- image_compo_ct_wide |>
  select(all_of(c("image_id", ordcols))) |>
  inner_join(image_compo_ct, by = "image_id") |>
  arrange(fraction_total_T_CD4) |>
  select(-all_of(ordcols)) |>
  mutate(ord = nrow(image_compo_ct):1)

image_compo_ct_wide <- DataFrame(image_compo_ct_wide)
```

### A2.1.4: Calculate immune cluster composition

```{r composition-cluster}
image_compo_clust <- image_compo_clust_wide <- vector(mode = "list")
  
# Number of cells of each cluster
for (i in seq_along(cluster_cols)) {
  spe_imm[["immune_cluster"]] <- spe_imm[[cluster_cols[[i]]]]
  
  image_compo_clust[[i]] <- as_tibble(colData(spe_imm)) |>
    group_by(image_id, immune_cluster) |>
    dplyr::count(name = "CellNb") |>
    ungroup() |>
    complete(image_id, immune_cluster,
             fill = list(CellNb = 0)) 
  
  # Merge with image stats and calculate fractions and densities
  image_compo_clust[[i]] <- image_compo_clust[[i]] |>
    inner_join(image_stats, by = "image_id") |>
    mutate(fraction_total = CellNb / CellsPerImage,
           fraction_immune = CellNb / ImmuneCellsPerImage,
           density = CellNb / image_area)

  # Import relevant metadata and re-order the dataset
  cn_keep <- c("image_id", plot_variables)
  
  image_compo_clust[[i]] <- as_tibble(colData(spe_imm))[, cn_keep] |>
    distinct() |>
    inner_join(image_compo_clust[[i]], by = "image_id") |>
    mutate(immune_cluster = factor(immune_cluster,
                                   levels = clusters[[i]]),
           image_id = factor(image_id, levels = image_ids)) |>
    arrange(image_id, immune_cluster)

  # Reorder the dataset by fraction of the first T CD4 cluster (for plotting)
  image_compo_clust_wide[[i]] <- image_compo_clust[[i]] |>
    pivot_wider(
      id_cols = image_id,
      values_from = c(CellNb, fraction_total, fraction_immune, density),
      names_from = c(immune_cluster))
  
  ordcols <- c(paste0("fraction_total_", clusters[[i]]))
  image_compo_clust[[i]] <- image_compo_clust_wide[[i]] |>
    select(all_of(c("image_id", ordcols))) |>
    inner_join(image_compo_clust[[i]], by = "image_id") |>
    arrange(fraction_total_T_CD4_1) |>
    select(-all_of(ordcols)) |>
    mutate(ord = nrow(image_compo_clust[[i]]):1)
  
  image_compo_clust_wide[[i]] <- DataFrame(image_compo_clust_wide[[i]])
  spe_imm[["cell_cluster"]] <- NULL
}
names(image_compo_clust) <- names(image_compo_clust_wide) <- cluster_cols
```

## A2.2: Define TOTAL cell type composition per image

### A2.2.1: List all cell types
```{r celltype-composition}
celltypes2 <- c("Beta", "Alpha", "Delta", "IsletOther", # Islet
                "Ductal", "Acinar", # Exo
                "Endothelial", "Mesenchymal", "SmoothMuscle", "Nerves", # Stroma.
                "Other", "Ambiguous", # Other
                "Neutrophil", "T_CD4", "T_CD8", "B", "NK", "Myeloid", "CD303_VIM", "T_DN") # Immune)

# Number of cells of each cell type
image_compo_ct_all <- as_tibble(colData(spe)) |>
  filter(cell_type %in% celltypes2, donor_type != "Pancreatitis") |> 
  group_by(image_id, cell_type) |>
  dplyr::count(name = "CellNb") |>
  ungroup() |>
  complete(image_id, cell_type,
           fill = list(CellNb = 0))

# Merge with image stats and calculate fractions and densities
image_compo_ct_all <- image_compo_ct_all |>
  inner_join(image_stats, by = "image_id") |>
  mutate(fraction_total = CellNb / CellsPerImage,
         fraction_immune = CellNb / ImmuneCellsPerImage,
         density = CellNb / image_area)

# Import relevant metadata (e.g., disease stage and patient ids)
cn_keep <- c("image_id", plot_variables)

image_compo_ct_all <- as_tibble(colData(spe))[, cn_keep] |>
  filter(donor_type != "Pancreatitis") |> 
  distinct() |>
  inner_join(image_compo_ct_all, by = "image_id") |>
  mutate(cell_type = factor(cell_type, levels = celltypes2),
         image_id = factor(image_id, levels = image_ids)) |>
  arrange(image_id, cell_type)

# Reorder the dataset by CD4 T cell fraction (for plotting)
image_compo_ct_all_wide <- image_compo_ct_all |>
  pivot_wider(
    id_cols = image_id,
    values_from = c(CellNb, fraction_total, fraction_immune, density),
    names_from = c(cell_type))

ordcols <- c(paste0("fraction_total_", celltypes2))
image_compo_ct_all <- image_compo_ct_all_wide |>
  select(all_of(c("image_id", ordcols))) |>
  inner_join(image_compo_ct_all, by = "image_id") |>
  arrange(fraction_total_T_CD4) |>
  select(-all_of(ordcols)) |>
  mutate(ord = nrow(image_compo_ct_all):1)

image_compo_ct_all_wide <- DataFrame(image_compo_ct_all_wide)
```

```{r case-all}
case_compo_all <- image_compo_ct_all |>
  dplyr::group_by(donor_type, case_id, cell_type) |> 
  dplyr::summarize(across(c(CellNb, CellsPerImage), sum)) |>
  dplyr::rename(CellsPerCase = "CellsPerImage") |>
  dplyr::mutate(fraction = CellNb / CellsPerCase) |>
  dplyr::arrange(donor_type, case_id, cell_type)
```

## A2.3: Joint composition of Immune CC + ALL cell types.
### A2.3.1: By ROI.
```{r combine-compo}
sub_compo_clust <- image_compo_clust[[1]] |> 
  filter(donor_type != "Pancreatitis") |>
  select(image_id, image_area, donor_type, case_id, CellNb, immune_cluster, density, CellsPerImage) |> 
  mutate(cell_type = stringr::str_remove(immune_cluster, "_[0-9]+$")) |>
  mutate(cell_type = ifelse(immune_cluster == "Treg", "T_CD4", cell_type)) |>
  mutate(cell_type = ifelse(immune_cluster %in% c("T-naive", "T-EMRA"), "T_CD8", cell_type)) |>
  filter(cell_type %in% c("Myeloid", "Neutrophil", "T_CD8", "T_CD4", "NK"))

sub_image_compo_all <- image_compo_ct_all |> 
  filter(!cell_type %in% c("Myeloid", "Neutrophil", "T_CD8", "T_CD4", "NK")) |> 
  select(image_id, donor_type, image_area, case_id, CellNb, cell_type, density, CellsPerImage) |> 
  mutate(immune_cluster = cell_type)

joint_all <- bind_rows(sub_image_compo_all, sub_compo_clust)

## Delete clusters with 0-counts.
clusters_to_remove <- joint_all |> group_by(immune_cluster) |> summarise(mu = mean(density)) |> filter(mu == 0) |> pull(immune_cluster)
clusters[[1]] <- clusters[[1]][!clusters[[1]] %in% clusters_to_remove]
```

### A2.3.2: By CASE.

```{r}
case_joint_all <- joint_all |>
  group_by(donor_type, case_id, immune_cluster, cell_type) |>
  summarize(cell_nb = sum(CellNb),
            total_area = sum(image_area), .groups = "keep") |>
  dplyr::rename("CellsPerCase" = cell_nb) |>
  mutate(mean_density = CellsPerCase / total_area) |>
  mutate(donor_type = factor(donor_type, levels = metadata(spe)$stages),
         case_id = factor(case_id, levels = metadata(spe)$cases)) |>
  arrange(case_id, donor_type) |> 
  ungroup()
```

# A3: Add cell type compo to SCE and save

The cell type and cluster composition are added to the metadata of the SCE

```{r add-compo-spe}
if (is.null(metadata(spe)$celltype_compo) ||
    is.null(metadata(spe)$cluster_compo)) {
  metadata(spe)$celltype_compo <- image_compo_ct_wide
  metadata(spe)$cluster_compo  <- image_compo_clust_wide
  print(spe)
  saveRDS(spe, fn_spe)
}
```

## A3.1: Remove pancreatitis cases

*OPTIONAL* Should pancreatitis cases be removed before plotting?

```{r remove-pancreatitis}
remove_pancreatitis <- TRUE

if (remove_pancreatitis) {
  spe_imm <- spe_imm[, spe_imm$donor_type != "Pancreatitis"]
  
  metadata(spe)$stages <- metadata(spe)$stages[
    metadata(spe)$stages != "Pancreatitis"]
  palettes$stages <- palettes$stages[names(palettes$stages) != "Pancreatitis"]

  metadata(spe)$cases <- metadata(spe)$cases[!metadata(spe)$cases %in% c("6036", "6043", "6061", "6150", "6225", "6506", "6522", "8005", "8008")]
  palettes$cases <- palettes$cases[!names(palettes$cases) %in% c("6036", "6043", "6061", "6150", "6225", "6506", "6522", "8005", "8008")]
  
  image_compo_ct <- image_compo_ct[
    image_compo_ct$donor_type != "Pancreatitis", ]
  
  for (i in seq_along(cluster_cols)) {
    image_compo_clust[[i]] <- image_compo_clust[[i]][
      image_compo_clust[[i]]$donor_type != "Pancreatitis", ]
  }
}
```

# A4: Plot cell type composition
The absolute immune composition is plotted for every single image.
-> Fractions and Absolute numbers of immune cells. 

## A4.1: Add cell type to the cluster composition
```{r}
image_compo_clust[["cell_cluster_scaled"]] <- image_compo_clust[["cell_cluster_scaled"]] |> 
  mutate(cell_type = stringr::str_remove(immune_cluster, "_[0-9]+$")) |> 
  mutate(cell_type = ifelse(immune_cluster == "Treg", "T_CD4", cell_type)) |> 
  mutate(cell_type = ifelse(immune_cluster == "T-naive", "T_CD8", cell_type)) |> 
  mutate(cell_type = ifelse(immune_cluster == "T-EMRA", "T_CD8", cell_type))
```

## A4.2: Stacked Barplot
Overview of immune cell type composition across all images.

```{r stacked-barplot}
temp <- scuttle::makePerCellDF(spe_imm, use_dimred = FALSE) |>
  filter(!cell_type %in% c("Ambiguous", "IsletLike")) |>
  dplyr::count(cell_type)
# 
p <- temp |> 
  # mutate(cell_type = factor(cell_type, levels = names(palettes$celltype))) |>
  ggplot(aes(x = "", y = n, fill = cell_type)) +
  geom_bar(stat = "identity", width = 0.8) +
  labs(x = "", y = "Cell number") +
  mytheme$large() +
  scale_fill_manual("Cell Type",
                    values = palettes$colors) +
  theme(axis.text.x = element_text(hjust = 1, vjust = 1, angle = 45))
print(p)
fn <- paste0(paste(today, "CellType", "StackedBarplot", "byStage",
                   sep = "_"), ".png")
do.call(ggsave, c(list(fn, p), plotsave_param))
```


## A4.3: Absolute composition

### A4.3.1: By immune cell type

```{r individual-absolute-celltype}
# All cells
p <- image_compo_ct |>
  ggplot(aes(x = reorder(image_id, ImmuneCellsPerImage),
             y = CellNb)) +
  geom_bar(stat = "identity", aes(fill = cell_type), width = 1) +
  scale_fill_manual("Cell type",
                    labels = names(celltypes),
                    values = palette_ct) +
  labs(title = "Absolute composition - All islets",
       x = "Individual islet sections",
       y = "Number of cells per islet section") +
  mytheme$standard() +
  theme(
    axis.title.x = element_text(margin = margin(t = -6)),
    axis.text.x = element_blank(),
    axis.line.x = element_blank(),
    axis.ticks.x = element_blank(),
  )

if (do_print) print(p)
fn <- paste0(paste(today, "Individual", "Absolute", "CellType", "Ungrouped",
                   sep = "_"), ".png")
do.call(ggsave, c(list(fn, p), plotsave_param))

# By stage
p1 <- p + facet_grid(~factor(donor_type, levels = metadata(spe)$stages),
                     scales = "free_x")
if (do_print) p1

fn <- paste0(paste(today, "Individual", "Absolute", "CellType", "byStage",
                   sep = "_"), ".png")
do.call(ggsave, c(list(fn, p1), plotsave_param))

# By case
p2 <- p + facet_grid(~factor(case_id, levels = metadata(spe)$cases),
                     scales = "free_x") +
  theme(strip.text.x = element_text(angle = 90))
if (do_print) p2

fn <- paste0(paste(today, "Individual", "Absolute", "CellType", "byCase",
                   sep = "_"), ".png")
do.call(ggsave, c(list(fn, p2), plotsave_param))
```

### A4.3.1: By immune cluster

```{r individual-absolute-cluster}
for (i in seq_along(cluster_cols)) {
  
  # All cells
  p <- image_compo_clust[[i]] |>
    filter(!immune_cluster %in% clusters_to_remove) |>
    ggplot(aes(x = reorder(image_id, ImmuneCellsPerImage),
               y = CellNb)) +
    geom_bar(stat = "identity", aes(fill = immune_cluster), width = 1) +
    scale_fill_manual("Immune cluster",
                      labels = clusters[[i]],
                      values = palettes_clust[[i]]) +
    labs(title = "Absolute composition - All islets",
         x = "Individual islet sections",
         y = "Number of cells per islet section") +
    mytheme$standard() +
    theme(
      axis.title.x = element_text(margin = margin(t = -6)),
      axis.text.x = element_blank(),
      axis.line.x = element_blank(),
      axis.ticks.x = element_blank(),
    ) + facet_wrap(~cell_type, scale = "free")
  if (do_print) print(p)
  fn <- paste0(paste(today, "Individual", "Absolute", "Cluster", "Ungrouped",
                     names(cluster_cols)[i], sep = "_"), ".png")
  do.call(ggsave, c(list(fn, p), plotsave_param_large))
  
  # By stage
  p1 <- p + facet_grid(~factor(donor_type, levels = metadata(spe)$stages),
                       scales = "free_x")
  if (do_print) print(p1)
  
  fn <- paste0(paste(today, "Individual", "Absolute", "Cluster", "byStage",
                     names(cluster_cols)[i], sep = "_"), ".png")
  do.call(ggsave, c(list(fn, p1), plotsave_param))
  
  # By case
  p2 <- p + facet_grid(~factor(case_id, levels = metadata(spe)$cases),
                       scales = "free_x") +
    theme(strip.text.x = element_text(angle = 90))
  if (do_print) print(p2)
  
  fn <- paste0(paste(today, "Individual", "Absolute", "Cluster", "byCase",
                     names(cluster_cols)[i], sep = "_"), ".png")
  do.call(ggsave, c(list(fn, p2), plotsave_param_large))
}
```


## A4.4: Relative composition

### A4.4.1: By cell type

```{r individual-relative-celltype}
p <- image_compo_ct |>
  ggplot(aes(x = reorder(image_id, rev(ord)),
             y = fraction_immune)) +
  geom_bar(stat = "identity", aes(fill = cell_type), width = 1) +
  scale_fill_manual("Cell Type",
                    labels = names(celltypes),
                    values = palette_ct) +
  labs(title = "Relative composition - All islets",
       x = "Individual islet sections",
       y = "Cell type fraction") +
  mytheme$standard() +
  theme(
    axis.title.x = element_text(margin = margin(t = -6)),
    axis.text.x = element_blank(),
    axis.line.x = element_blank(),
    axis.ticks.x = element_blank(),
  )
if (do_print) print(p)

fn <- paste0(paste(today, "Individual", "Relative", "CellType", "Ungrouped",
                   sep = "_"), ".png")
do.call(ggsave, c(list(fn, p), plotsave_param))

# By stage
p1 <- p + facet_grid(~factor(donor_type, levels = metadata(spe)$stages),
                     scales = "free_x")
if (do_print) print(p1)

fn <- paste0(paste(today, "Individual", "Relative", "CellType", "byStage",
                   sep = "_"), ".png")
do.call(ggsave, c(list(fn, p1), plotsave_param))

# By case
p2 <- p + facet_grid(~factor(case_id, levels = metadata(spe)$cases),
                     scales = "free_x") +
  theme(strip.text.x = element_text(angle = 90))
if (do_print) print(p2)

fn <- paste0(paste(today, "Individual", "Relative", "CellType", "byCase",
                   sep = "_"), ".png")
do.call(ggsave, c(list(fn, p2), plotsave_param))
```

### A4.4.2: By cluster

```{r individual-relative-cluster}
for (i in seq_along(cluster_cols)) {
  
  p <- image_compo_clust[[i]] |>
    filter(!immune_cluster %in% clusters_to_remove) |>
    ggplot(aes(x = reorder(image_id, rev(ord)),
               y = fraction_immune)) +
    geom_bar(stat = "identity", aes(fill = immune_cluster), width = 1) +
    scale_fill_manual("Immune cluster",
                      labels = clusters[[i]],
                      values = palettes_clust[[i]]) +
    labs(title = "Relative composition - All islets",
         x = "Individual islet sections",
         y = "Immune cluster fraction") +
    mytheme$standard() +
    theme(
      axis.title.x = element_text(margin = margin(t = -6)),
      axis.text.x = element_blank(),
      axis.line.x = element_blank(),
      axis.ticks.x = element_blank(),
    ) + facet_wrap(~ cell_type, scale = "free")
  if (do_print) print(p)
  
  fn <- paste0(paste(today, "Individual", "Relative", "Cluster", "Ungrouped",
                     names(cluster_cols)[i], sep = "_"), ".png")
  do.call(ggsave, c(list(fn, p), plotsave_param))
  
  # By stage
  p1 <- p + facet_grid(~factor(donor_type, levels = metadata(spe)$stages),
                       scales = "free_x")
  if (do_print) print(p1)
  
  fn <- paste0(paste(today, "Individual", "Relative", "Cluster", "byStage",
                     names(cluster_cols)[i], sep = "_"), ".png")
  do.call(ggsave, c(list(fn, p1), plotsave_param))
  
  # By case
  p2 <- p + facet_grid(~factor(case_id, levels = metadata(spe)$cases),
                       scales = "free_x") +
    theme(strip.text.x = element_text(angle = 90))
  if (do_print) print(p2)
  
  fn <- paste0(paste(today, "Individual", "Relative", "Cluster", "byCase",
                     names(cluster_cols)[i], sep = "_"), ".png")
  do.call(ggsave, c(list(fn, p2), plotsave_param))
}
```




# **A5: Aggregated composition - Absolute**

The immune cell type / immune cluster composition is aggregated by patient.

## A5.1: Immune cell density

Number of cells per square millimeter.

### A5.1.1: All immune cells (by case)
Boxplots of (log)-total densities by case + stage.
```{r aggregated-density-all}
# Calculate immune densities by donor
cur_dat <- image_compo_ct |>
  group_by(donor_type, case_id) |>
  summarize(cell_nb = sum(CellNb),
            total_area = sum(image_area), .groups = "keep") |>
  mutate(mean_density = cell_nb / total_area) |>
  mutate(donor_type = factor(donor_type, levels = metadata(spe)$stages),
         case_id = factor(case_id, levels = metadata(spe)$cases)) |>
  arrange(case_id, donor_type) |>
  as.data.frame()

# Plot density
p <- cur_dat |>
  ggplot(aes(x = donor_type, y = mean_density, fill = donor_type)) +
  geom_boxplot(outlier.shape = NA) +
  geom_jitter(width = 0.1) +
  scale_fill_manual("Cell Type",
                    labels = metadata(spe)$stages,
                    values = palettes$stages) +
  labs(title = paste("Mean immune cell density"),
       x = "Stage",
       y = "Cell density (cells / mm^2)") +
  mytheme$large() +
  theme(axis.text.x = element_text(hjust = 1, vjust = 1, angle = 45))
if (do_print) print(p)

fn <- paste0(paste(today, "Density", "byCase", "AllCells", 
                   sep = "_"), ".png")
do.call(ggsave, c(list(fn, p), plotsave_param))

# Plot log of density
p <- cur_dat |>
  ggplot(aes(x = donor_type, y = log(mean_density), fill = donor_type)) +
  geom_boxplot(outlier.shape = NA) +
  geom_jitter(width = 0.1) +
  scale_fill_manual("Cell Type",
                    labels = metadata(spe)$stages,
                    values = palettes$stages) +
  labs(title = paste("Mean immune cell density (log)"),
       x = "Stage",
       y = "log of cell density (cells / mm^2)") +
  mytheme$large() +
  theme(axis.text.x = element_text(hjust = 1, vjust = 1, angle = 45))

if (do_print) print(p)

fn <- paste0(paste(today, "Density", "byCase", "AllCells", "Log",
                   sep = "_"), ".png")
do.call(ggsave, c(list(fn, p), plotsave_param))
```


## Immune cell fraction

Number of immune cells divided by total cell number on the image.

### All immune cells

```{r aggregated-fraction-all}
# Calculate immune densities by donor
cur_dat <- image_compo_ct |>
  group_by(donor_type, case_id) |>
  summarize(cell_nb = sum(CellNb),
            total_cell_nb = sum(CellsPerImage), .groups = "keep") |>
  mutate(mean_fraction = cell_nb / total_cell_nb) |>
  mutate(donor_type = factor(donor_type, levels = metadata(spe)$stages),
         case_id = factor(case_id, levels = metadata(spe)$cases)) |>
  arrange(case_id, donor_type) |>
  as.data.frame()

# Plot density
p <- cur_dat |>
  ggplot(aes(x = donor_type, y = mean_fraction, fill = donor_type)) +
  geom_boxplot(outlier.shape = NA) +
  geom_jitter(width = 0.1) +
  scale_fill_manual("Cell Type",
                    labels = metadata(spe)$stages,
                    values = palettes$stages) +
  labs(title = paste("Mean cell fraction"),
       x = "Stage",
       y = "Cell fraction (of total cell number)") +
  mytheme$large() +
  theme(axis.text.x = element_text(hjust = 1, vjust = 1, angle = 45))
if (do_print) print(p)

fn <- paste0(paste(today, "Fraction", "byCase", "AllCells", 
                   sep = "_"), ".png")
do.call(ggsave, c(list(fn, p), plotsave_param))

# Log of density
p <- cur_dat |>
  ggplot(aes(x = donor_type, y = log(mean_fraction), fill = donor_type)) +
  geom_boxplot(outlier.shape = NA) +
  geom_jitter(width = 0.1) +
  scale_fill_manual("Cell Type",
                    labels = metadata(spe)$stages,
                    values = palettes$stages) +
  labs(title = paste("Mean cell fraction (log) per patient"),
       x = "Stage",
       y = "Cell fraction (of total cell number)") +
  mytheme$large() +
  theme(axis.text.x = element_text(hjust = 1, vjust = 1, angle = 45))

if (do_print) print(p)

fn <- paste0(paste(today, "Fraction", "byCase", "AllCells", "Log",
                   sep = "_"), ".png")
do.call(ggsave, c(list(fn, p), plotsave_param))
```

### By cell type

```{r aggregated-fraction-celltype}
# Calculate immune densities by donor and cell type
cur_dat <- image_compo_ct |>
  group_by(donor_type, case_id, cell_type) |>
  summarize(cell_nb = sum(CellNb),
            total_cell_nb = sum(CellsPerImage), .groups = "keep") |>
  mutate(mean_fraction = cell_nb / total_cell_nb) |>
  mutate(cell_type = factor(cell_type, levels = celltypes),
         donor_type = factor(donor_type, levels = metadata(spe)$stages),
         case_id = factor(case_id, levels = metadata(spe)$cases)) |>
  arrange(cell_type, case_id, donor_type) |>
  as.data.frame()

# Density
for (j in seq_along(celltypes)) {
  p <- cur_dat[cur_dat$cell_type == celltypes[j], ] |>
    ggplot(aes(x = donor_type, y = mean_fraction, fill = donor_type)) +
    geom_boxplot(outlier.shape = NA) +
    geom_jitter(width = 0.1) +
    scale_fill_manual("Cell Type",
                      labels = metadata(spe)$stages,
                      values = palettes$stages) +
    labs(title = paste("Mean cell fraction per patient -", celltypes[j]),
         x = "Stage",
         y = "Cell fraction (of total cell number)") +
    mytheme$large() +
    theme(axis.text.x = element_text(hjust = 1, vjust = 1, angle = 45))
  if (do_print) print(p)
  
  fn <- paste0(paste(today, "Fraction", "byCase", "CellType", celltypes[j],
                     sep = "_"), ".png")
  do.call(ggsave, c(list(fn, p), plotsave_param))
}
```

### By cluster.  

Not run as this yields too many plots, these data will be easier to visualize 
with plotting approaches taken below.

```{r aggregated-fraction-cluster}
# for (i in seq_along(cluster_cols)) {
#   # Calculate immune densities by donor and cell type
#   cur_dat <- image_compo_clust[[i]] |>
#     group_by(donor_type, case_id, immune_cluster) |>
#     summarize(cell_nb = sum(CellNb),
#               total_cell_nb = sum(CellsPerImage), .groups = "keep") |>
#     mutate(mean_density = cell_nb / total_cell_nb) |>
#     mutate(immune_cluster = factor(immune_cluster, levels = clusters[[i]]),
#            donor_type = factor(donor_type, levels = metadata(spe)$stages),
#            case_id = factor(case_id, levels = metadata(spe)$cases)) |>
#     arrange(immune_cluster, case_id, donor_type) |>
#     as.data.frame()
# 
#   # Density
#   for (j in seq_along(clusters[[i]])) {
#     p <- cur_dat[cur_dat$immune_cluster == clusters[[i]][j], ] |>
#       ggplot(aes(x = donor_type, y = mean_fraction, fill = donor_type)) +
#       geom_boxplot(outlier.shape = NA) +
#       geom_jitter(width = 0.1) +
#       scale_fill_manual("Cluster",
#                         labels = metadata(spe)$stages,
#                         values = palettes$stages) +
#       labs(title = paste("Mean cell density per patient -", clusters[[i]][j]),
#            x = "Stage",
#            y = "Cell fraction (of total cell number)") +
#       mytheme$large() +
#       theme(axis.text.x = element_text(hjust = 1, vjust = 1, angle = 45))
#     print(p)
# 
#     fn <- paste0(paste(today, "Fraction", "byCase", "Cluster",
#                        cluster_cols[j], clusters[[i]][j], sep = "_"), ".png")
#     do.call(ggsave, c(list(fn, p), plotsave_param))
#   }
# }
```



# A6: **Aggregated composition - Relative**

Here, the cell type composition is aggregated and normalized from 0 to 1 to 
compare cell type compositions at different stages of the disease and in 
different patients. 

## A6.1: Aggregating function

The `aggregate_compo` function calculates the number of cells (`CellNb`) by
cluster and group, and then computes the relative value for each cluster.

For instance, to calculate relative cell type densities by case and stage:
- `aggreg_val`: value to aggregate by (`image_area`).
- `cluster`: cell type (`cell_type`).
- `group1`: disease stage (`donor_type`).
- `orderclust` and `order1` are used to order the dataset.

=> The function will first calculate the average cell type density, which is 
the  number of cells of a given cell type in the selected groups (here, 
`donor_type`) divided by the total aggregating value (here, `image_area`).  

It will then compute the total density (sum of cell type densities for all 
groups). Finally it will normalize the cell type densities, i.e., 
divide individual cell type densities by the total density (so that the sum of 
relative densities for one cell type is equal to 1).

```{r aggregate_compo-function}
aggregate_compo <- function(x, aggreg_val, cluster, orderclust, group1, order1) {
  
  # Average value by group1
  compo <- as_tibble(x) |>
    group_by(Group1 = get(group1), Cluster = get(cluster)) |>
    summarize(cell_nb = sum(CellNb),
              total_aggreg_val = sum(get(aggreg_val)),
              .groups = "keep") |>
    mutate(mean_value = cell_nb / total_aggreg_val) |>
    
    # Calculate relative value
    group_by(Cluster) |>
    mutate(total_value = sum(mean_value)) |>
    mutate(relative_value = mean_value / total_value) |>
    
    # Arrange
    mutate(Cluster = factor(Cluster, levels = orderclust),
           Group1 = factor(Group1, levels = order1)) |>
    arrange(Cluster, Group1)
  
  return(compo |>
           filter(!is.na(Cluster) & !is.na(Group1)))
}
```


## A6.2: Normalized by cell type/cluster

Cell type fractions and densities by case or patient, normalized by cell type.

### A6.2.1: Cell types

Cell type fractions and cell type densities.

```{r aggregated-normalized-celltype}
# Cell type fraction X stage
p <- aggregate_compo(
  image_compo_ct, aggreg_val = "CellsPerImage",
  cluster = "cell_type", orderclust = celltypes,
  group1 = "donor_type", order1 = rev(metadata(spe)$stages)) |>
  ggplot(aes(x = Cluster, y = relative_value, fill = as.factor(Group1))) +
  geom_bar(stat = "identity") +
  scale_fill_manual(values = palettes$stages) +
  labs(x = "Cell type", y = "Relative cell type fraction",
       title = "Relative cell type fraction by stage", fill = "Stage") +
  mytheme$large() +
  theme(axis.text.x = element_text(hjust = 1, vjust = 1, angle = 90))
if (do_print) print(p)

fn <- paste0(paste(today, "Aggregated", "byCellType", "Fraction", "Stage",
                   sep = "_"), ".png")
do.call(ggsave, c(list(fn, p), plotsave_param))

# Cell type density X stage
p <- aggregate_compo(
  image_compo_ct, aggreg_val = "image_area",
  cluster = "cell_type", orderclust = celltypes,
  group1 = "donor_type", order1 = rev(metadata(spe)$stages)) |>
  
  ggplot(aes(x = Cluster, y = relative_value, fill = as.factor(Group1))) +
  geom_bar(stat = "identity") +
  scale_fill_manual(values = palettes$stages) +
  labs(x = "Cell type", y = "Relative cell type density",
       title = "Relative cell type density by stage", fill = "Stage") +
  mytheme$large() +
  theme(axis.text.x = element_text(hjust = 1, vjust = 1, angle = 90))
if (do_print) print(p)

fn <- paste0(paste(today, "Aggregated", "byCellType", "Density",  "Stage",
                   sep = "_"), ".png")
do.call(ggsave, c(list(fn, p), plotsave_param))

# Cell type fraction X case
p <- aggregate_compo(
  image_compo_ct, aggreg_val = "CellsPerImage",
  cluster = "cell_type", orderclust = celltypes,
  group1 = "case_id", order1 = rev(metadata(spe)$cases))  |>
  
  ggplot(aes(x = Cluster, y = relative_value, fill = as.factor(Group1))) +
  geom_bar(stat = "identity") +
  scale_fill_manual(values = palettes$cases) +
  labs(x = "Cell type", y = "Relative cell type fraction",
       title = "Relative cell type fraction by case", fill = "Case") +
  mytheme$large() +
  theme(axis.text.x = element_text(hjust = 1, vjust = 1, angle = 90))
if (do_print) print(p)

fn <- paste0(paste(today, "Aggregated", "byCellType", "Fraction", "Case",
                   sep = "_"), ".png")
do.call(ggsave, c(list(fn, p), plotsave_param_large))

# Cell type density X case
p <- aggregate_compo(
  image_compo_ct, aggreg_val = "image_area",
  cluster = "cell_type", orderclust = celltypes,
  group1 = "case_id", order1 = rev(metadata(spe)$cases))  |>
  
  ggplot(aes(x = Cluster, y = relative_value, fill = as.factor(Group1))) +
  geom_bar(stat = "identity") +
  scale_fill_manual(values = palettes$cases) +
  labs(x = "Cell type", y = "Relative cell type density",
       title = "Relative cell type density by case", fill = "Case") +
  mytheme$large() +
  theme(axis.text.x = element_text(hjust = 1, vjust = 1, angle = 90))
if (do_print) print(p)

fn <- paste0(paste(today, "Aggregated", "byCellType", "Density", "Case",
                   sep = "_"), ".png")
do.call(ggsave, c(list(fn, p), plotsave_param_large))
```

### A6.2.2: Cluster

Immune cluster fractions and densities.

```{r aggregated-normalized-cluster}
for (i in seq_along(cluster_cols)) {
  # Cell type fraction X stage
  p <- aggregate_compo(
    image_compo_clust[[i]], aggreg_val = "CellsPerImage",
    cluster = "immune_cluster", orderclust = clusters[[i]],
    group1 = "donor_type", order1 = rev(metadata(spe)$stages)) |>
    
    ggplot(aes(x = Cluster, y = relative_value, fill = as.factor(Group1))) +
    geom_bar(stat = "identity") +
    scale_fill_manual(values = palettes$stages) +
    labs(x = "Immune cluster", y = "Relative cluster fraction",
         title = "Relative cluster fraction by stage", fill = "Stage") +
    mytheme$large() +
    theme(axis.text.x = element_text(hjust = 1, vjust = 1, angle = 90))
  if (do_print) print(p)
  
  fn <- paste0(paste(today, "Aggregated", "byCluster", "Fraction", "Stage",
                     cluster_cols[[i]], sep = "_"), ".png")
  do.call(ggsave, c(list(fn, p), plotsave_param))
  
  # Cell type density X stage
  p <- aggregate_compo(
    image_compo_clust[[i]], aggreg_val = "image_area",
    cluster = "immune_cluster", orderclust = clusters[[i]],
    group1 = "donor_type", order1 = rev(metadata(spe)$stages)) |>
    
    ggplot(aes(x = Cluster, y = relative_value, fill = as.factor(Group1))) +
    geom_bar(stat = "identity") +
    scale_fill_manual(values = palettes$stages) +
    labs(x = "Immune cluster", y = "Relative cluster fraction",
         title = "Relative cluster density by stage", fill = "Stage") +
    mytheme$large() +
    theme(axis.text.x = element_text(hjust = 1, vjust = 1, angle = 90))
  if (do_print) print(p)
  
  fn <- paste0(paste(today, "Aggregated", "byCluster", "Density", "Stage",
                     cluster_cols[[i]], sep = "_"), ".png")
  do.call(ggsave, c(list(fn, p), plotsave_param))
  
  # Cell type fraction X case
  p <- aggregate_compo(
    image_compo_clust[[i]], aggreg_val = "CellsPerImage",
    cluster = "immune_cluster", orderclust = clusters[[i]],
    group1 = "case_id", order1 = rev(metadata(spe)$cases))  |>
    
    ggplot(aes(x = Cluster, y = relative_value, fill = as.factor(Group1))) +
    geom_bar(stat = "identity") +
    scale_fill_manual(values = palettes$cases) +
    labs(x = "Immune cluster", y = "Relative cluster fraction",
         title = "Relative cluster fraction by case", fill = "Case") +
    mytheme$large() +
    theme(axis.text.x = element_text(hjust = 1, vjust = 1, angle = 90))
  if (do_print) print(p)
  
  fn <- paste0(paste(today, "Aggregated", "byCluster", "Fraction", "Case",
                     cluster_cols[[i]], sep = "_"), ".png")
  do.call(ggsave, c(list(fn, p), plotsave_param_large))
  
  # Cell type density X case
  p <- aggregate_compo(
    image_compo_clust[[i]], aggreg_val = "image_area",
    cluster = "immune_cluster", orderclust = clusters[[i]],
    group1 = "case_id", order1 = rev(metadata(spe)$cases))  |>
    ungroup() |> 
    ggplot(aes(x = Cluster, y = relative_value, fill = as.factor(Group1))) +
    geom_bar(stat = "identity") +
    scale_fill_manual(values = palettes$cases) +
    labs(x = "Immune cluster", y = "Relative cluster fraction",
         title = "Relative cluster density by case", fill = "Case") +
    mytheme$large() +
    theme(axis.text.x = element_text(hjust = 1, vjust = 1, angle = 90))
  if (do_print) print(p)
  
  fn <- paste0(paste(today, "Aggregated", "byCluster", "Density", "Case",
                     cluster_cols[[i]], sep = "_"), ".png")
  do.call(ggsave, c(list(fn, p), plotsave_param_large))
}
```


## A6.3: Normalized by stage or patient

Cell type fractions and densities normalized by stage or by patient.

### A6.3.1: Cell types

Cell type fractions and cell type densities.

```{r aggregated-normalized-group-celltype}
# Cell type fraction X stage
p <- aggregate_compo(
  image_compo_ct, aggreg_val = "CellsPerImage",
  cluster = "donor_type", orderclust = metadata(spe)$stages,
  group1 = "cell_type", order1 = rev(celltypes)) |>
  
  ggplot(aes(x = Cluster, y = relative_value, fill = as.factor(Group1))) +
  geom_bar(stat = "identity") +
  scale_fill_manual(values = palette_ct) +
  labs(x = "Cell type", y = "Relative cell type fraction",
       title = "Relative cell type fraction by stage", fill = "Stage") +
  mytheme$large() +
  theme(axis.text.x = element_text(hjust = 1, vjust = 1, angle = 90))
if (do_print) print(p)

fn <- paste0(paste(today, "Aggregated", "byStage", "Fraction", "CellType",
                   sep = "_"), ".png")
do.call(ggsave, c(list(fn, p), plotsave_param))

# Cell type density X stage
p <- aggregate_compo(
  image_compo_ct, aggreg_val = "image_area",
  cluster = "donor_type", orderclust = metadata(spe)$stages,
  group1 = "cell_type", order1 = rev(celltypes)) |>
  
  ggplot(aes(x = Cluster, y = relative_value, fill = as.factor(Group1))) +
  geom_bar(stat = "identity") +
  scale_fill_manual(values = palette_ct) +
  labs(x = "Cell type", y = "Relative cell type density",
       title = "Relative cell type density by stage", fill = "Stage") +
  mytheme$large() +
  theme(axis.text.x = element_text(hjust = 1, vjust = 1, angle = 90))
if (do_print) print(p)

fn <- paste0(paste(today, "Aggregated", "byStage", "Density",  "CellType",
                   sep = "_"), ".png")
do.call(ggsave, c(list(fn, p), plotsave_param))

# Cell type fraction X case
p <- aggregate_compo(
  image_compo_ct, aggreg_val = "CellsPerImage",
  cluster = "case_id", orderclust = metadata(spe)$cases,
  group1 = "cell_type", order1 = rev(celltypes)) |>

  ggplot(aes(x = Cluster, y = relative_value, fill = as.factor(Group1))) +
  geom_bar(stat = "identity") +
  scale_fill_manual(values = palette_ct) +
  labs(x = "Cell type", y = "Relative cell type fraction",
       title = "Relative cell type fraction by case", fill = "Case") +
  mytheme$large() +
  theme(axis.text.x = element_text(hjust = 1, vjust = 1, angle = 90))
if (do_print) print(p)

fn <- paste0(paste(today, "Aggregated", "byCase", "Fraction", "CellType",
                   sep = "_"), ".png")
do.call(ggsave, c(list(fn, p), plotsave_param_large))

# Cell type density X case
p <- aggregate_compo(
  image_compo_ct, aggreg_val = "image_area",
  cluster = "case_id", orderclust = metadata(spe)$cases,
  group1 = "cell_type", order1 = rev(celltypes)) |>
  
  ggplot(aes(x = Cluster, y = relative_value, fill = as.factor(Group1))) +
  geom_bar(stat = "identity") +
  scale_fill_manual(values = palette_ct) +
  labs(x = "Cell type", y = "Relative cell type density",
       title = "Relative cell type density by case", fill = "Case") +
  mytheme$large() +
  theme(axis.text.x = element_text(hjust = 1, vjust = 1, angle = 90))
if (do_print) print(p)

fn <- paste0(paste(today, "Aggregated", "byCase", "Density", "CellType",
                   sep = "_"), ".png")
do.call(ggsave, c(list(fn, p), plotsave_param_large))
```

### A6.3.2: Cluster

Immune cluster fractions and densities.

```{r aggregated-normalized-group-cluster}
for (i in seq_along(cluster_cols)) {

    # Cell type fraction X case
    p <- aggregate_compo(
      image_compo_clust[[i]], aggreg_val = "CellsPerImage",
      cluster = "case_id", orderclust = metadata(spe)$cases,
      group1 = "immune_cluster", order1 = rev(clusters[[i]])) |>
      
      ggplot(aes(x = Cluster, y = relative_value, fill = as.factor(Group1))) +
      geom_bar(stat = "identity") +
      scale_fill_manual(values = palettes_clust[[i]]) +
      labs(x = "Immune cluster", y = "Relative cluster fraction",
          title = "Relative cluster fraction by case", fill = "Case") +
      mytheme$large() +
      theme(axis.text.x = element_text(hjust = 1, vjust = 1, angle = 90))
    if (do_print) print(p)
    
    fn <- paste0(paste(today, "Aggregated", "byCase", "Fraction", "Cluster",
                      cluster_cols[[i]], sep = "_"), ".png")
    do.call(ggsave, c(list(fn, p), plotsave_param_large))

      # Cell type density X case
    p <- aggregate_compo(
      image_compo_clust[[i]], aggreg_val = "image_area",
      cluster = "case_id", orderclust = metadata(spe)$cases,
      group1 = "immune_cluster", order1 = rev(clusters[[i]])) |>
      
      ggplot(aes(x = Cluster, y = relative_value, fill = as.factor(Group1))) +
      geom_bar(stat = "identity") +
      scale_fill_manual(values = palettes_clust[[i]]) +
      labs(x = "Immune cluster", y = "Relative cluster density",
          title = "Relative cluster density by case", fill = "Case") +
      mytheme$large() +
      theme(axis.text.x = element_text(hjust = 1, vjust = 1, angle = 90))
    if (do_print) print(p)
    
    fn <- paste0(paste(today, "Aggregated", "byCase", "Density", "Cluster",
                      cluster_cols[[i]], sep = "_"), ".png")
    do.call(ggsave, c(list(fn, p), plotsave_param_large))
  
  ## Do Stage by Cell Types
  for (j in seq_along(celltypes)) {
      data <- image_compo_clust[[i]] |> filter(cell_type == celltypes[j])
      # Cluster fraction X stage
      p <- aggregate_compo(
        data, aggreg_val = "CellsPerImage",
        cluster = "donor_type", orderclust = metadata(spe)$stages,
        group1 = "immune_cluster", order1 = rev(clusters[[i]])) |>
        mutate(cell_type = stringr::str_remove(Cluster, "_[0-9]+$")) |> 
        ggplot(aes(x = Cluster, y = relative_value, fill = as.factor(Group1))) +
        geom_bar(stat = "identity") +
        scale_fill_manual(values = palettes_clust[[i]]) +
        labs(x = "Immune cluster", y = "Relative cluster fraction",
            title = "Relative cluster fraction by stage", fill = "Stage") +
        mytheme$large() +
        theme(axis.text.x = element_text(hjust = 1, vjust = 1, angle = 90))
        if (do_print) print(p)
  
    fn <- paste0(paste(today, "Aggregated", "byStage", "Fraction", "Cluster",
                      cluster_cols[[i]], celltypes[j], sep = "_"), ".png")
    do.call(ggsave, c(list(fn, p), plotsave_param))

      # Cell type density X stage
    p <- aggregate_compo(
      data, aggreg_val = "image_area",
      cluster = "donor_type", orderclust = metadata(spe)$stages,
      group1 = "immune_cluster", order1 = rev(clusters[[i]])) |>
      mutate(cell_type = stringr::str_remove(Cluster, "_[0-9]+$")) |> 
      ggplot(aes(x = Cluster, y = relative_value, fill = as.factor(Group1))) +
      geom_bar(stat = "identity") +
      scale_fill_manual(values = palettes_clust[[i]]) +
      labs(x = "Immune cluster", y = "Relative cluster density",
          title = "Relative cell type density by stage", fill = "Stage") +
      mytheme$large() +
      theme(axis.text.x = element_text(hjust = 1, vjust = 1, angle = 90))
    if (do_print) print(p)
    
    fn <- paste0(paste(today, "Aggregated", "byStage", "Density",  "Cluster",
                      cluster_cols[[i]], celltypes[j], sep = "_"), ".png")
    do.call(ggsave, c(list(fn, p), plotsave_param))
  }
}
```

# A7: edgeR with age-adjustment (**CELL TYPES**)

## A7.1: Prepare Data.

EXPLORATIVE ANALYSIS.
```{r aggregated-density-celltype}
t1d_duration <- colData(spe_imm)[, c("case_id", "donor_type", "age", "diabetes_duration", "C_peptide")] |> as_tibble()
temp_dur <- t1d_duration |> distinct(case_id, donor_type, age, diabetes_duration, C_peptide)

# By cell type
cur_dat <- image_compo_ct |>
  # factor to character
  mutate(cell_type = as.character(cell_type)) |> 
  group_by(donor_type, case_id, image_id, cell_type) |>
  summarize(cell_nb = CellNb,
            total_area = image_area, .groups = "keep") |>
  mutate(mean_density = cell_nb / total_area) |>
  mutate(case_id = factor(case_id, levels = metadata(spe)$cases)) |>
  arrange(cell_type, case_id, donor_type) |> 
  mutate(donor_type = factor(donor_type, levels = metadata(spe)$stages)) |>
  left_join(temp_dur, by = c("case_id", "donor_type")) |>
  ungroup()
cur_dat

# By immune cell cluster.
cur_dat_cl <- image_compo_clust[[1]] |>
  # factor to character
  mutate(immune_cluster = as.character(immune_cluster)) |> 
  group_by(donor_type, case_id, image_id, immune_cluster) |>
  summarize(cell_nb = CellNb,
            total_area = image_area, .groups = "keep") |>
  mutate(mean_density = cell_nb / total_area) |>
  mutate(case_id = factor(case_id, levels = metadata(spe)$cases)) |>
  arrange(immune_cluster, case_id, donor_type) |> 
  mutate(donor_type = factor(donor_type, levels = metadata(spe)$stages)) |>
  left_join(temp_dur, by = c("case_id", "donor_type")) |>
  ungroup()
cur_dat_cl

# Calculate immune densities by donor and cell type
cur_dat_case <- image_compo_ct |>
  group_by(donor_type, case_id, cell_type) |>
  summarize(cell_nb = sum(CellNb),
            total_area = sum(image_area), .groups = "keep") |>
  mutate(mean_density = cell_nb / total_area) |>
  mutate(cell_type = factor(cell_type, levels = celltypes),
         donor_type = factor(donor_type, levels = metadata(spe)$stages),
         case_id = factor(case_id, levels = metadata(spe)$cases)) |>
  arrange(cell_type, case_id, donor_type) |>
  left_join(temp_dur, by = c("case_id", "donor_type")) |>
  as.data.frame()
cur_dat_case
gc()
```

```{r prepare-edgeR}
case_stages <- colData(spe)[, c("donor_type", "case_id", "age")] |> 
  as_tibble() |>
  distinct() |> 
  mutate(age_group = ifelse(age < 13, "young", "older")) |>
  mutate(age_group = factor(age_group, levels = c("young", "older"))) |> 
  filter(donor_type != "Pancreatitis") |> 
  mutate(donor_type = stringr::str_remove(donor_type, "\\+"))

fn_cases_full <- file.path(paths$folder_in, "cases_full.xlsx")
cases_full <- readxl::read_excel(fn_cases_full) |> 
  mutate(case_id = as.character(case_id)) |> 
  select(case_id, ICU_time_days) |> 
  mutate(ICU_time_days = abs(as.numeric(stringr::str_remove(ICU_time_days, " days"))))

df_input <- case_compo_all |> 
  dplyr::ungroup() |> 
  inner_join(case_stages, by = c("case_id")) |> 
  left_join(cases_full, by = c("case_id"))

df_input <- df_input |> dplyr::rename("donor_type" = donor_type.y)
```

## A7.2: Define edgeR input
```{r edgeR-input}
# COUNT MATRIX.
edge <- df_input |>
  dplyr::select(c(case_id, cell_type, CellNb)) |>
  tidyr::pivot_wider(id_cols = cell_type,
                     names_from = case_id, values_from = CellNb) |>
  as.data.frame()

rownames(edge) <- edge$cell_type
edge$cell_type <- NULL
edge <- as.matrix(edge)

# GROUP INFORMATION
coldata <- unique(df_input[, c("donor_type", "case_id", "age_group", "ICU_time_days")]) |> as.data.frame()
coldata$ICU_time_days <- log1p(coldata$ICU_time_days)
coldata$donor_type = factor(coldata$donor_type, levels = c("NoDiabetes", "sAAb", "mAAb", "RecentOnset", "LongDuration"))
coldata$age_group = factor(coldata$age_group, levels = c("young", "older"))
```

Filter Cell types + Create Design Matrix.
```{r edge-design-matrix}
# Create DGEList (main object)
y_ab <- edgeR::DGEList(edge, samples = coldata)
y_ab

# Filter cell types. Add group information (donor-type).
keep <- edgeR::filterByExpr(y_ab, group = y_ab$samples$donor_type)
y_ab <- y_ab[keep, ]
summary(keep)

# Create design matrix. Adjust now for age + ICU-time.
y_ab$samples$ICU_time_days[is.na(y_ab$samples$ICU_time_days)] <-
  mean(y_ab$samples$ICU_time_days, na.rm = TRUE)
  
design <- model.matrix(~ donor_type + age_group + ICU_time_days, y_ab$samples) 
```

Estimate NB-dispersion. Common-dispersion.
```{r estimate-dispersion}
# y_ab <- edgeR::estimateDisp(y_ab, design, trend = "none")
y_ab <- edgeR::estimateDisp(y_ab, design)

summary(y_ab$common.dispersion)
edgeR::plotBCV(y_ab, cex=1)
```

Fit the NB-model. Turn off trend. 
```{r fit-ql-model}
fit_ab <- edgeR::glmQLFit(y_ab, design, robust = TRUE, abundance.trend = TRUE, legacy = FALSE)
summary(fit_ab$var.prior)
summary(fit_ab$df.prior)
edgeR::plotQLDisp(fit_ab, cex = 1)
```

## A7.3: Run edgeR tests and extract contrast of interest.

Run Tests between ND and distinct biological groups.
```{r test-diff-abundance}
my_contrasts <- limma::makeContrasts(
  NoDiabetes_sAAb = -donor_typesAAb,
  NoDiabetes_mAAb = -donor_typemAAb,
  NoDiabetes_RecentOnset = -donor_typeRecentOnset,
  NoDiabetes_LongDuration = -donor_typeLongDuration,
  sAAb_mAAb = donor_typesAAb - donor_typemAAb,
  mAAb_RecentOnset = donor_typemAAb - donor_typeRecentOnset,
  RecentOnset_LongDuration = donor_typeRecentOnset - donor_typeLongDuration,
  age_group = -age_groupolder,
  ICU_time_days = -ICU_time_days,
  levels = design
)

## Extract contrasts of interest. Either vs baseline; or between consecutive stages.
qlf_list <- lapply(colnames(my_contrasts), function(cn) {
  topTags(glmQLFTest(fit_ab, contrast = my_contrasts[, cn]), n = "all")$table |> 
    as_tibble(rownames = "cell_type") |> 
    mutate(contrast = cn)
}) |> 
  bind_rows() |> 
  # select Immune-cell types only!
  filter(cell_type %in% c("T_CD8", "T_CD4", "Myeloid", "Neutrophil", 
                          "T_DN", "B", "NK", "CD303_VIM"))

## Clearning.
edgeR_output <- qlf_list |> 
  filter(contrast != "ICU_time_days") |> 
  tidyr::separate_wider_delim(contrast, names = c("group1", "group2"), delim = "_") |> 
  mutate(group1 = ifelse(group1 == "sAAb", "sAAb+", 
                         ifelse(group1 == "mAAb", "mAAb+", group1))) |> 
  mutate(group2 = ifelse(group2 == "sAAb", "sAAb+", 
                         ifelse(group2 == "mAAb", "mAAb+", group2)))

case_stages2 <- case_stages |> 
  mutate(donor_type = ifelse(donor_type == "sAAb", "sAAb+",
                         ifelse(donor_type == "mAAb", "mAAb+", donor_type))) |> 
  mutate(donor_type = factor(donor_type, levels = metadata(spe)$stages))
```

## A7.4: Plot results.

Plot stats.
This is Fig3C and ExtFig5b.

```{r fig2c_suppfig6b}
library(ggpubr); library(rstatix)

edgeR_non_age <- edgeR_output |> filter(group1 != "age")

### Major immune cell types.
p <- plot_abundance_stages(stats = edgeR_non_age, 
          dat = cur_dat_case, 
          cur_celltypes = c("Myeloid", "Neutrophil", "T_CD4", "T_CD8"),
          x = "donor_type", y = "mean_density", facet_by = "cell_type", 
          fill_by = "donor_type", ct_type = "cell_type",
          title = "") + 
    facet_wrap(~cell_type, scales = "free_y", axes = "all") + 
  theme(axis.text.x = element_text(angle = 45, hjust = 0.5, vjust = 0.5)) + 
  theme(panel.spacing = unit(1, "cm"))
        
print(p)
fn <- paste0(paste(today, "Density", "byCase", "CellType", "Major",
                    sep = "_"), ".png")
do.call(ggsave, c(list(fn, p), plotsave_param_large))
saveRDS(p, file.path(paths$home_git, "figures", "fig3c.rds"))

### Minor immune cell types.
p <- plot_abundance_stages(stats = edgeR_non_age, 
          dat = cur_dat_case, 
          cur_celltypes = c("B", "NK", "CD303_VIM", "T_DN"),
          x = "donor_type", y = "mean_density", facet_by = "cell_type", 
          fill_by = "donor_type", 
          title = "")

print(p)
fn <- paste0(paste(today, "Density", "byCase", "CellType", "Minor",
                    sep = "_"), ".png")
do.call(ggsave, c(list(fn, p), plotsave_param_large))
saveRDS(p, file.path(paths$home_git, "figures", "extfig5b.rds"))
```

# A8: Immune Cluster; across disease stages

## A8.1: edgeR for Clusters.

Input: again all Immune Clusters + other Cell types to get correct fractions.
Output: Cell types of interest: only Myeloid, T-CD4, T-CD8 clusters.

- Prepare data.

```{r edge-r-clusters}
case_join2 <- case_joint_all |>
  tidyr::drop_na()  |> 
  mutate(donor_type = stringr::str_remove(donor_type, "\\+")) |> 
  inner_join(case_stages, by = c("case_id", "donor_type"))  |> 
  left_join(cases_full, by = c("case_id"))

edge <- case_join2 |> 
  dplyr::filter(!immune_cluster %in% c("Myeloid", "Neutrophil", "T_CD4", "T_CD8", "NK")) |> 
  dplyr::select(c(case_id, immune_cluster, CellsPerCase)) |>
  tidyr::pivot_wider(id_cols = immune_cluster,
                     names_from = case_id, values_from = CellsPerCase) |>
  as.data.frame()

rownames(edge) <- edge$immune_cluster
edge$immune_cluster <- NULL
edge <- as.matrix(edge)

coldata <- unique(case_join2[, c("donor_type", "case_id", "age_group", "ICU_time_days")])
coldata$ICU_time_days <- log1p(coldata$ICU_time_days)
coldata$donor_type = factor(coldata$donor_type, levels = c("NoDiabetes", "sAAb", "mAAb", "RecentOnset", "LongDuration"))
coldata$age_group = factor(coldata$age_group, levels = c("young", "older"))
```

Create DGEList (main object) + Filter Cell types + Create Design Matrix.
Removes T-CD4-12, T-CD4-9, T-CD8-5 (which do not exist) and T-regs.
```{r edge-design-matrix-clusters}
# Create DGEList (main object)
y_ab <- edgeR::DGEList(edge, samples = coldata)
y_ab

# remove very rare cell types.
keep <- edgeR::filterByExpr(y_ab, group = y_ab$samples$donor_type,
                            min.count = 5, min.total.count = 10)
y_ab <- y_ab[keep, ]
summary(keep)
print(keep[!keep])
# Create design matrix
y_ab$samples$ICU_time_days[is.na(y_ab$samples$ICU_time_days)] <-
  mean(y_ab$samples$ICU_time_days, na.rm = TRUE)
design <- model.matrix(~donor_type + age_group + ICU_time_days, y_ab$samples)
```

Estimate NB-dispersion.
```{r estimate-dispersion-clusters}
y_ab <- edgeR::estimateDisp(y_ab, design, trend = "none")
summary(y_ab$common.dispersion)
plotBCV(y_ab, cex=1)
```

Fit the NB-model. Turn off trend. 
```{r fit-ql-model-clusters}
fit_ab <- edgeR::glmQLFit(y_ab, design, robust = TRUE, abundance.trend = TRUE, legacy = FALSE)

summary(fit_ab$var.prior)
summary(fit_ab$df.prior)
edgeR::plotQLDisp(fit_ab, cex = 1)
```

## A8.2: Run Tests between ND and distinct biological groups.

Run Tests between ND and distinct biological groups.
```{r test-diff-abundance}
my_contrasts <- limma::makeContrasts(
  NoDiabetes_sAAb = -donor_typesAAb,
  NoDiabetes_mAAb = -donor_typemAAb,
  NoDiabetes_RecentOnset = -donor_typeRecentOnset,
  NoDiabetes_LongDuration = -donor_typeLongDuration,
  sAAb_mAAb = donor_typesAAb - donor_typemAAb,
  mAAb_RecentOnset = donor_typemAAb - donor_typeRecentOnset,
  RecentOnset_LongDuration = donor_typeRecentOnset - donor_typeLongDuration,
  age_group = -age_groupolder,
  ICU_time_days = -ICU_time_days,
  levels = design
)

## Extract contrasts of interest. Either vs baseline; or between consecutive stages.
qlf_list <- lapply(colnames(my_contrasts), function(cn) {
  topTags(glmQLFTest(fit_ab, contrast = my_contrasts[, cn]), n = "all")$table |> 
    as_tibble(rownames = "cell_type") |> 
    mutate(contrast = cn)
}) |> 
  bind_rows() |> 
  filter(grepl(("Myeloid|T_CD4|T_CD8|NK|Neutrophil|T-naive|T-EMRA"), cell_type))

## Cleaning.
edgeR_output <- qlf_list |> 
  filter(contrast != "ICU_time_days") |>
  tidyr::separate_wider_delim(contrast, names = c("group1", "group2"), delim = "_") |> 
  mutate(group1 = ifelse(group1 == "sAAb", "sAAb+", 
                         ifelse(group1 == "mAAb", "mAAb+", group1))) |> 
  mutate(group2 = ifelse(group2 == "sAAb", "sAAb+", 
                         ifelse(group2 == "mAAb", "mAAb+", group2)))
```

## A8.3: Plot results.

```{r myeloid-subtypes}
library(ggpubr); library(rstatix)

edgeR_non_age <- edgeR_output |> filter(group1 != "age")
input_df <- case_join2 |> mutate(donor_type = ifelse(donor_type == "sAAb", "sAAb+", 
                         ifelse(donor_type == "mAAb", "mAAb+", donor_type)))

### Myeloid subtypes
p <- plot_abundance_stages(stats = edgeR_non_age, 
          dat = input_df, 
          ct_type = "cluster",
          cur_celltypes = c("Myeloid_1", "Myeloid_2", "Myeloid_3", "Myeloid_4",
                             "Myeloid_5", "Myeloid_6", "Myeloid_7", "Myeloid_8"),
          x = "donor_type", y = "mean_density", facet_by = "cell_type", 
          fill_by = "donor_type", 
          title = "")
print(p)
fn <- paste0(paste(today, "Density", "byCase", "Cluster", "Myeloid", 
                   sep = "_"), ".png")
do.call(ggsave, c(list(fn, p), plotsave_param_large))

### NK-cell subtypes
p <- plot_abundance_stages(stats = edgeR_non_age, 
          dat = input_df, 
          ct_type = "cluster",
            cur_celltypes = c("NK_1", "NK_2", "NK_3"),
          x = "donor_type", y = "mean_density", facet_by = "cell_type", 
          fill_by = "donor_type", 
          title = "")

fn <- paste0(paste(today, "Density", "byCase", "Cluster", "NK", 
                   sep = "_"), ".png")
do.call(ggsave, c(list(fn, p), plotsave_param_large))

### T-CD4 subtypes
p <- plot_abundance_stages(stats = edgeR_non_age,
          dat = input_df, 
          ct_type = "cluster",
          cur_celltypes = c("T_CD4_1", "T_CD4_2", "T_CD4_3", "T_CD4_4",
                             "T_CD4_5", "T_CD4_6", "T_CD4_7", "T_CD4_8", 
                             "T_CD4_9", "T_CD4_10"),
          x = "donor_type", y = "mean_density", facet_by = "cell_type", 
          fill_by = "donor_type", 
          title = "")
fn <- paste0(paste(today, "Density", "byCase", "Cluster", "T_CD4", 
                   sep = "_"), ".png")
do.call(ggsave, c(list(fn, p), plotsave_param_large))

### T-CD8 subtypes
p <- plot_abundance_stages(stats = edgeR_non_age,
          dat = input_df, 
          ct_type = "cluster",
          cur_celltypes = c("T_CD8_1", "T_CD8_2", "T_CD8_3", "T_CD8_4",
                             "T-naive", "T-EMRA", "T_CD8_6", "T_CD8_7"),
          x = "donor_type", y = "mean_density", facet_by = "cell_type",
          fill_by = "donor_type",
          title = "")
fn <- paste0(paste(today, "Density", "byCase", "Cluster", "T_CD8", 
                   sep = "_"), ".png")
do.call(ggsave, c(list(fn, p), plotsave_param_large))
```

# A9: ICIs CellTypes.

## A9.1: Prepare Data.

```{r prepare-edgeR}
library(edgeR)
## Use ICIs only.
icis <- image_compo_ct_all |> filter(cell_type == "Beta") |> filter(density > 0) |> pull(image_id)

ici_all <- image_compo_ct_all |> 
  filter(image_id %in% icis) |> 
  group_by(donor_type, case_id, cell_type) |>
  summarize(CellNb = sum(CellNb), image_area = sum(image_area), .groups = "keep") |>
  mutate(density = CellNb / image_area) |> 
  mutate(donor_type = factor(donor_type, levels = metadata(spe)$stages)) |>
  mutate(case_id = factor(case_id, levels = metadata(spe)$cases)) |> 
  filter(donor_type != "LongDuration")

df_input <- ici_all |> 
  dplyr::ungroup() |> 
  inner_join(case_stages, by = c("case_id")) |> 
  left_join(cases_full, by = c("case_id"))

df_input <- df_input |> dplyr::rename("donor_type" = donor_type.y)
```

## A9.2: Define edgeR input
```{r edgeR-input}
# COUNT MATRIX.
edge <- df_input |>
  dplyr::select(c(case_id, cell_type, CellNb)) |>
  tidyr::pivot_wider(id_cols = cell_type,
                     names_from = case_id, values_from = CellNb) |>
  as.data.frame()

rownames(edge) <- edge$cell_type
edge$cell_type <- NULL
edge <- as.matrix(edge)

# GROUP INFORMATION
coldata <- unique(df_input[, c("donor_type", "case_id", "age_group", "ICU_time_days")])
coldata$ICU_time_days <- log1p(coldata$ICU_time_days)
coldata$donor_type = factor(coldata$donor_type, levels = c("NoDiabetes", "sAAb", "mAAb", "RecentOnset"))
coldata$age_group = factor(coldata$age_group, levels = c("young", "older"))
```

Filter Cell types + Create Design Matrix.
```{r edge-design-matrix}
# Create DGEList (main object)
y_ab <- edgeR::DGEList(edge, samples = coldata)
y_ab

# Filter cell types. Add group information (donor-type).
keep <- edgeR::filterByExpr(y_ab, group = y_ab$samples$donor_type)
y_ab <- y_ab[keep, ]
summary(keep)

# Create design matrix. Adjust now for age.
y_ab$samples$ICU_time_days[is.na(y_ab$samples$ICU_time_days)] <-
  mean(y_ab$samples$ICU_time_days, na.rm = TRUE)
design <- model.matrix(~ donor_type + age_group + ICU_time_days, y_ab$samples)
```

Estimate NB-dispersion. Common-dispersion.
```{r estimate-dispersion}
# y_ab <- edgeR::estimateDisp(y_ab, design, trend = "none")
y_ab <- edgeR::estimateDisp(y_ab, design)

summary(y_ab$common.dispersion)
edgeR::plotBCV(y_ab, cex=1)
```

Fit the NB-model. 
```{r fit-ql-model}
fit_ab <- edgeR::glmQLFit(y_ab, design, robust = TRUE, abundance.trend = TRUE, legacy = FALSE)
summary(fit_ab$var.prior)
summary(fit_ab$df.prior)
edgeR::plotQLDisp(fit_ab, cex = 1)
```

## A9.3: Run edgeR tests and extract contrast of interest.

Run Tests between ND and distinct biological groups.
```{r test-diff-abundance}
my_contrasts <- limma::makeContrasts(
  NoDiabetes_sAAb = -donor_typesAAb,
  NoDiabetes_mAAb = -donor_typemAAb,
  NoDiabetes_RecentOnset = -donor_typeRecentOnset,
  sAAb_mAAb = donor_typesAAb - donor_typemAAb,
  mAAb_RecentOnset = donor_typemAAb - donor_typeRecentOnset,
  age_group = -age_groupolder,
  ICU_time_days = -ICU_time_days,
  levels = design
)

## Extract contrasts of itnerest. Either vs baseline; or between consecutive stages.
qlf_list <- lapply(colnames(my_contrasts), function(cn) {
  topTags(glmQLFTest(fit_ab, contrast = my_contrasts[, cn]), n = "all")$table |> 
    as_tibble(rownames = "cell_type") |> 
    mutate(contrast = cn)
}) |> 
  bind_rows() |> 
  # select Immune-cell types only!
  filter(cell_type %in% c("T_CD8", "T_CD4", "Myeloid", "Neutrophil", 
                          "T_DN", "B", "NK", "CD303_VIM"))

## Clearning.
edgeR_output <- qlf_list |> 
  filter(contrast != "ICU_time_days") |>
  tidyr::separate_wider_delim(contrast, names = c("group1", "group2"), delim = "_") |> 
  mutate(group1 = ifelse(group1 == "sAAb", "sAAb+", 
                         ifelse(group1 == "mAAb", "mAAb+", group1))) |> 
  mutate(group2 = ifelse(group2 == "sAAb", "sAAb+", 
                         ifelse(group2 == "mAAb", "mAAb+", group2)))
```

## A9.4: Plot results.
```{r ici-celltypes}
library(ggpubr); library(rstatix)
edgeR_non_age <- edgeR_output |> filter(group1 != "age") |> 
  mutate(group1 = factor(group1, levels = metadata(spe)$stages[1:4])) |> 
  mutate(group2 = factor(group2, levels = metadata(spe)$stages[1:4]))

plot_df <- df_input |> 
  mutate(donor_type = ifelse(donor_type == "sAAb", "sAAb+", 
                         ifelse(donor_type == "mAAb", "mAAb+", donor_type))) |> 
  dplyr::rename("mean_density" = density) |> 
  mutate(donor_type = factor(donor_type, levels = metadata(spe)$stages[1:4]))

### Major immune cell types.
p <- plot_abundance_stages(stats = edgeR_non_age,
          dat = plot_df, 
          cur_celltypes = c("Myeloid", "Neutrophil", "T_CD4", "T_CD8"),
          x = "donor_type", y = "mean_density", facet_by = "cell_type", 
          fill_by = "donor_type", ct_type = "cell_type",
          title = "") + 
  facet_wrap(~cell_type, scales = "free_y", axes = "all") + 
  scale_x_discrete(breaks = c("NoDiabetes", "sAAb+", "mAAb+", "RecentOnset"), 
                   labels = c("Control", "sAAb+", "mAAb+", "Onset")) +
    scale_fill_manual(name = "Stage", 
                     values = palettes$stages[1:4], 
                     breaks = c("NoDiabetes", "sAAb+", "mAAb+", "RecentOnset"),
                    labels = c("Control", "sAAb+", "mAAb+", "Onset")) + 
  theme(axis.text.x = element_text(angle = 45, hjust = 0.5, vjust = 0.5)) + 
  theme(panel.spacing = unit(1, "cm"))
    
print(p)
fn <- paste0(paste(today, "Density", "byCase", "CellType", "Major", "ICI",
                    sep = "_"), ".png")
do.call(ggsave, c(list(fn, p), plotsave_param_large))

### Minor immune cell types.
p <- plot_abundance_stages(stats = edgeR_non_age,
          dat = plot_df, 
          cur_celltypes = c("B", "NK", "CD303_VIM", "T_DN"),
          x = "donor_type", y = "mean_density", facet_by = "cell_type", 
          fill_by = "donor_type",
        title = "")
print(p)
fn <- paste0(paste(today, "Density", "byCase", "CellType", "Minor", "ICI",
                    sep = "_"), ".png")
do.call(ggsave, c(list(fn, p), plotsave_param_large))
```

## A9.5: AGE!

```{r ici-age}
edgeR_age <- edgeR_output |> 
  filter(group1 == "age")

# get max-y-value PER cell type.
cur_celltypes <- c("Myeloid", "T_CD4", "T_CD8", "Neutrophil", "B", "NK", "CD303_VIM", "T_DN")
p_vals <- edgeR_age
```


```{r}
# Violin-Plot of AGE-association.
df <- plot_df |> 
  filter(cell_type %in% cur_celltypes) |> 
  mutate(age_groups2 = ifelse(age_group == "young", "< 13", ">= 13")) |> 
  mutate(mean_density = log1p(mean_density)) |> 
  mutate(age_groups2 = factor(age_groups2, levels = c("< 13", ">= 13")))

p_vals <- df |> 
  group_by(cell_type) |> 
  summarise(max = max(mean_density), .groups = "drop") |>
  dplyr::right_join(p_vals, by = c("cell_type")) |> 
  mutate(.y. = "mean_density",
          group1 = "< 13", group2 = ">= 13",
          p = PValue,
          p.adj = FDR,
           p.signif = cut(p.adj, breaks = c(-Inf, 0.001, 0.01, 0.05, 0.1, Inf), labels = c("***", "**", "*", "•", ""))
          )

cur_formula2 <- formula(paste("mean_density ~ age_groups2"))
stat.test <- df |> 
  group_by(cell_type) |> 
  t_test(cur_formula2, ref.group = "< 13") |> 
  add_xy_position(x = "age_groups2", step.increase = 0.07)
stat.test

# Combine formatting + P-value dataframe.
stat.test.test <- p_vals |> 
    left_join(stat.test, by = c("cell_type", ".y.", "group1", "group2"))

stat.test.test

p <- ggviolin(df, x = "age_groups2", y = "mean_density", fill = "age_groups2",
          facet.by = "cell_type",
          draw_quantiles = c(0.25, 0.5, 0.75), palette = NULL) + 
    geom_jitter(width = 0.2, size = 5, alpha = 0.8, aes(color = donor_type)) +
    stat_pvalue_manual(stat.test.test, label = "p.signif", tip.length = 0.1,
                      size = 10, label.size = 10) + 
    facet_wrap(~cell_type, scales = "free_y") +
    scale_fill_manual("Age Groups", values = palettes$colors50[c(9, 12)])  + 
    scale_color_manual("Donor Type",
                     values = palettes$stages) +
    labs(x = "Age Groups (years)", y = "log1p: Cell density (cells / mm^2)") + 
    mytheme$large_new() + 
    theme(axis.text.x = element_text(angle = 45, hjust = 1)) + 
    guides(fill = "none")

print(p)
fn <- paste0(paste(today, "CellTypes", "Age", sep = "_"), ".png")
do.call(ggsave, c(list(fn, p), plotsave_param_large))

extfig5c <- p
saveRDS(extfig5c, file.path(paths$home_git, "figures", "extfig5c.rds"))
```

# A10: ICI Clusters.

## A10.1: edgeR for Clusters.
```{r edge-r-clusters-ici}
ici_all <- joint_all |> 
  filter(image_id %in% icis) |> 
  group_by(donor_type, case_id, immune_cluster) |> 
  summarize(CellsPerCase = sum(CellNb), image_area = sum(image_area), .groups = "keep") |> 
  mutate(density = CellsPerCase / image_area) |> 
  filter(donor_type != "LongDuration") |> 
  mutate(donor_type = factor(donor_type, levels = metadata(spe)$stages[1:4])) |> 
  ungroup()

case_join2 <- ici_all |>
  mutate(donor_type = stringr::str_remove(donor_type, "\\+")) |> 
  inner_join(case_stages, by = c("case_id", "donor_type")) |> 
  left_join(cases_full, by = c("case_id"))
```

```{r edge-r-clusters}
edge <- case_join2 |> 
  dplyr::filter(!immune_cluster %in% c("Myeloid", "Neutrophil", "T_CD4", "T_CD8", "NK")) |> 
  dplyr::select(c(case_id, immune_cluster, CellsPerCase)) |>
  tidyr::pivot_wider(id_cols = immune_cluster,
                     names_from = case_id, values_from = CellsPerCase) |>
  as.data.frame()

rownames(edge) <- edge$immune_cluster
edge$immune_cluster <- NULL
edge <- as.matrix(edge)

coldata <- unique(case_join2[, c("donor_type", "case_id", "age_group", "ICU_time_days")])
coldata$ICU_time_days <- log1p(coldata$ICU_time_days)
coldata$donor_type = factor(coldata$donor_type, levels = c("NoDiabetes", "sAAb", "mAAb", "RecentOnset"))
coldata$age_group = factor(coldata$age_group, levels = c("young", "older"))
```

Create DGEList (main object) + Filter Cell types + Create Design Matrix.
Removes T-CD4-12, T-CD4-9, T-CD8-5 (which do not exist) and T-regs.
```{r edge-design-matrix-clusters}
# Create DGEList (main object)
y_ab <- edgeR::DGEList(edge, samples = coldata)
y_ab

# remove very rare cell types.
keep <- edgeR::filterByExpr(y_ab, group = y_ab$samples$donor_type,
                            min.count = 5, min.total.count = 10)
y_ab <- y_ab[keep, ]
summary(keep)
print(keep[!keep])
# Create design matrix
y_ab$samples$ICU_time_days[is.na(y_ab$samples$ICU_time_days)] <-
  mean(y_ab$samples$ICU_time_days, na.rm = TRUE)
design <- model.matrix(~donor_type + age_group + ICU_time_days, y_ab$samples)
```

Estimate NB-dispersion.
```{r estimate-dispersion-clusters}
y_ab <- edgeR::estimateDisp(y_ab, design, trend = "none")
summary(y_ab$common.dispersion)
plotBCV(y_ab, cex=1)
```

Fit the NB-model. Turn off trend. 
```{r fit-ql-model-clusters}
fit_ab <- edgeR::glmQLFit(y_ab, design, robust = TRUE, abundance.trend = TRUE, legacy = FALSE)

summary(fit_ab$var.prior)
summary(fit_ab$df.prior)
edgeR::plotQLDisp(fit_ab, cex = 1)
```

## A10.2: Run Tests between disease stages.

Run Tests between disease stages. 
```{r test-diff-abundance}
my_contrasts <- limma::makeContrasts(
  NoDiabetes_sAAb = -donor_typesAAb,
  NoDiabetes_mAAb = -donor_typemAAb,
  NoDiabetes_RecentOnset = -donor_typeRecentOnset,
  sAAb_mAAb = donor_typesAAb - donor_typemAAb,
  mAAb_RecentOnset = donor_typemAAb - donor_typeRecentOnset,
  age_group = -age_groupolder,
  ICU_time_days = -ICU_time_days,
  levels = design
)

## Extract contrasts of interest. Either vs baseline; or between consecutive stages.
qlf_list <- lapply(colnames(my_contrasts), function(cn) {
  topTags(glmQLFTest(fit_ab, contrast = my_contrasts[, cn]), n = "all")$table |> 
    as_tibble(rownames = "cell_type") |> 
    mutate(contrast = cn)
}) |> 
  bind_rows() |> 
  filter(grepl(("Myeloid|T_CD4|T_CD8|NK|Neutrophil|T-naive|T-EMRA"), cell_type))

## Cleaning.
edgeR_output <- qlf_list |> 
  filter(contrast != "ICU_time_days") |>
  tidyr::separate_wider_delim(contrast, names = c("group1", "group2"), delim = "_") |> 
  mutate(group1 = ifelse(group1 == "sAAb", "sAAb+", 
                         ifelse(group1 == "mAAb", "mAAb+", group1))) |> 
  mutate(group2 = ifelse(group2 == "sAAb", "sAAb+", 
                         ifelse(group2 == "mAAb", "mAAb+", group2)))
```

## A10.3: Plot results.
```{r ici-clusters-plot}
edgeR_non_age <- edgeR_output |> filter(group1 != "age")
input_df <- case_join2 |> mutate(donor_type = ifelse(donor_type == "sAAb", "sAAb+", 
                         ifelse(donor_type == "mAAb", "mAAb+", donor_type))) |> 
                  mutate(mean_density = density) |> 
                  mutate(donor_type = factor(donor_type, levels = metadata(spe)$stages[1:4]))

### Myeloid subtypes
p <- plot_abundance_stages(stats = edgeR_non_age, 
          dat = input_df, 
          ct_type = "cluster",
          cur_celltypes = c("Myeloid_1", "Myeloid_2", "Myeloid_3", "Myeloid_4",
                             "Myeloid_5", "Myeloid_6", "Myeloid_7", "Myeloid_8"),
          x = "donor_type", y = "mean_density", facet_by = "cell_type", 
          fill_by = "donor_type", 
          title = "") + 
  facet_wrap(~cell_type, labeller = as_labeller(myeloid_labels), scales = "free_y", axes = "all") + 
  scale_x_discrete(breaks = c("NoDiabetes", "sAAb+", "mAAb+", "RecentOnset"), 
                   labels = c("Control", "sAAb+", "mAAb+", "Onset")) +
    scale_fill_manual(name = "Stage", 
                     values = palettes$stages[1:4], 
                     breaks = c("NoDiabetes", "sAAb+", "mAAb+", "RecentOnset"),
                    labels = c("Control", "sAAb+", "mAAb+", "Onset")) + 
  theme(axis.text.x = element_text(angle = 45, hjust = 0.5, vjust = 0.5)) + 
  theme(panel.spacing = unit(1, "cm"))

print(p)
fn <- paste0(paste(today, "Density", "byCase", "Cluster", "Myeloid", "ICI",
                   sep = "_"), ".png")
plotsave_param_large2 <- plotsave_param_large
plotsave_param_large2$height <- 600
do.call(ggsave, c(list(fn, p), plotsave_param_large2))

### NK-cell subtypes
p <- plot_abundance_stages(stats = edgeR_non_age, 
          dat = input_df, 
          ct_type = "cluster",
            cur_celltypes = c("NK_1", "NK_2", "NK_3"),
          x = "donor_type", y = "mean_density", facet_by = "cell_type", 
          fill_by = "donor_type", 
          title = "") + 
          mytheme$standard_new()
print(p)
fn <- paste0(paste(today, "Density", "byCase", "Cluster", "NK", "ICI",
                   sep = "_"), ".png")
do.call(ggsave, c(list(fn, p), plotsave_param_large))

### T-CD4 subtypes
p <- plot_abundance_stages(stats = edgeR_non_age,
      dat = input_df |> filter(!immune_cluster %in% c("T_CD4_Other")), 
      ct_type = "cluster",
      cur_celltypes = c("T_CD4_1", "T_CD4_2", "T_CD4_3", "T_CD4_4",
                        "T_CD4_5", "T_CD4_6", "T_CD4_7", "T_CD4_8", 
                        "T_CD4_9", "T_CD4_10"),
      x = "donor_type", y = "mean_density", facet_by = "cell_type", 
      fill_by = "donor_type", 
      title = "") + 
  facet_wrap(~cell_type, labeller = as_labeller(tcd4_labels3), scales = "free_y", axes = "all") + 
  mytheme$standard_new()
print(p)
fn <- paste0(paste(today, "Density", "byCase", "Cluster", "T_CD4", "ICI",
                   sep = "_"), ".png")
do.call(ggsave, c(list(fn, p), plotsave_param_large))

### T-CD8 subtypes
p <- plot_abundance_stages(stats = edgeR_non_age,
          dat = input_df, 
          ct_type = "cluster",
          cur_celltypes = c("T_CD8_1", "T_CD8_2", "T_CD8_3", "T_CD8_4",
                             "T-naive", "T-EMRA", "T_CD8_6", "T_CD8_7"),
          x = "donor_type", y = "mean_density", facet_by = "cell_type",
          fill_by = "donor_type",
          title = "") + mytheme$standard_new()

fn <- paste0(paste(today, "Density", "byCase", "Cluster", "T_CD8", "ICI",
                   sep = "_"), ".png")
do.call(ggsave, c(list(fn, p), plotsave_param_large))

### Neutrophil subtypes.
p <- plot_abundance_stages(stats = edgeR_non_age,
          dat = input_df, 
          ct_type = "cluster",
          cur_celltypes = c("Neutrophil_1", "Neutrophil_2", "Neutrophil_3", "Neutrophil_4"),
          x = "donor_type", y = "mean_density", facet_by = "cell_type",
          fill_by = "donor_type",
          title = "") + mytheme$standard_new()
fn <- paste0(paste(today, "Density", "byCase", "Cluster", "Neutrophil", "ICI",
                   sep = "_"), ".png")
do.call(ggsave, c(list(fn, p), plotsave_param_large))
```


**Myeloid-cells**

Plot Myeloid-3 and Myeloid-4 jointly. Main Figure on Figure 5.
```{r}
library(patchwork)
edgeR_na <- edgeR_non_age |> 
  mutate(cell_type = recode(cell_type, !!!myeloid_labels))

input_df_x <- input_df  |> 
  mutate(immune_cluster = recode(immune_cluster, !!!myeloid_labels))

p <- plot_abundance_stages(stats = edgeR_na |> mutate(cell_type = factor(cell_type, levels = c("M1/M2-like act.", "cDC"))),
          dat = input_df_x |> mutate(immune_cluster = factor(immune_cluster, levels = c("M1/M2-like act.", "cDC"))),
          ct_type = "cluster",
          cur_celltypes = c("M1/M2-like act.", "cDC"),
          x = "donor_type", y = "mean_density", facet_by = "cell_type",
          fill_by = "donor_type", nrow = 2,
          title = "")  + 
          mytheme$standard_new()
print(p)
fn <- paste0(paste(today, "Density", "byCase", "Cluster", "Myeloid_3_4", "Vertical",
                    sep = "_"), ".png")
plotsave_param2 <- plotsave_param
plotsave_param2$height <- 400
do.call(ggsave, c(list(fn, p), plotsave_param2))

fig5d <- p
saveRDS(fig5d, file.path(paths$home_git, "figures", "fig5d.rds"))
```

Myeloid-Cells: Plot all besides 3 and 4.
```{r myeloid-states}
p <- plot_abundance_stages(stats = edgeR_na  |> 
    mutate(cell_type = factor(cell_type, levels = myeloid_labels[c(1, 2, 5, 6, 7, 8)])),
          dat = input_df_x  |> mutate(immune_cluster = factor(immune_cluster, levels = myeloid_labels[c(1, 2, 5, 6, 7, 8)])),
          ct_type = "cluster",
          cur_celltypes = myeloid_labels[c(1, 2, 5, 6, 7, 8)],
          x = "donor_type", y = "mean_density", facet_by = "cell_type",
          fill_by = "donor_type", nrow = 2,
          title = "") + 
          mytheme$large_new()
print(p)
fn <- paste0(paste(today, "Density", "byCase", "Cluster", "Myeloid_1_2_5_6_7_8",
                    sep = "_"), ".png")
do.call(ggsave, c(list(fn, p), plotsave_param_large))

# EDIT NAMES!
extfig8a <- p
saveRDS(extfig8a, file.path(paths$home_git, "figures", "extfig8a.rds"))
```

**T-CD4-cells**
Plot T-CD4-5
```{r t-cd4-5-differential-abundance}
edgeR_na <- edgeR_non_age |> 
  mutate(cell_type = recode(cell_type, !!!tcd4_labels))

input_df_x <- input_df  |> 
  mutate(immune_cluster = recode(immune_cluster, !!!tcd4_labels))

p <- plot_abundance_stages(stats = edgeR_na,
          dat = input_df_x, 
          ct_type = "cluster",
          cur_celltypes = tcd4_labels["T_CD4_5"],
          x = "donor_type", y = "mean_density", facet_by = "cell_type",
          fill_by = "donor_type", nrow = 1, step.increase = 0.07,
          title = "") + mytheme$standard_new()

fn <- paste0(paste(today, "Density", "byCase", "Cluster", "TCD4_5_v2",
                    sep = "_"), ".png")
do.call(ggsave, c(list(fn, p), plotsave_param))

fig4e <- p
saveRDS(fig4e, file.path(paths$home_git, "figures", "fig4e.rds"))
```


Plot T-CD4-1, 2, 3, 4, 6, 7, 8, 9, 10
```{r differential-abundance-tcd4-states}
edgeR_na <- edgeR_non_age |> 
  mutate(cell_type = recode(cell_type, !!!tcd4_labels))

input_df_x <- input_df  |> 
  mutate(immune_cluster = recode(immune_cluster, !!!tcd4_labels))

p <- plot_abundance_stages(stats = edgeR_na,
          dat = input_df_x,
          ct_type = "cluster",
          cur_celltypes = tcd4_labels[c(1,2,3,4,6,7,8,9,10)],
          x = "donor_type", 
          y = "mean_density", 
          facet_by = "cell_type",
          fill_by = "donor_type", 
          nrow = 3, 
          parse_labels = FALSE,  
          step.increase = 0.15,
          title = "") + mytheme$large_new()
print(p)
fn <- paste0(paste(today, "Density", "byCase", "Cluster", "TCD4_1_2_3_4_6_7_8_9_10",
                    sep = "_"), ".png")
do.call(ggsave, c(list(fn, p), plotsave_param_large))

# Edit names:
extfig6a <- p
saveRDS(extfig6a, file.path(paths$home_git, "figures", "extfig6a.rds"))
```

Plot T-CD8-1, 3, 4, 6, 7, 8, 
```{r tcd8-states-da}
edgeR_na <- edgeR_non_age |> 
  mutate(cell_type = recode(cell_type, !!!tcd8_labels))
input_df_x <- input_df  |>
  mutate(immune_cluster = recode(immune_cluster, !!!tcd8_labels))

p <- plot_abundance_stages(stats = edgeR_na,
          dat = input_df_x,
          ct_type = "cluster",
          cur_celltypes = tcd8_labels[c(1,2, 3, 4, 5, 6, 8)],
          x = "donor_type", 
          y = "mean_density", 
          facet_by = "cell_type",
          fill_by = "donor_type", 
          nrow = 3, 
          parse_labels = FALSE,  
          step.increase = 0.15,
          title = "") + mytheme$large_new()

fn <- paste0(paste(today, "Density", "byCase", "Cluster", "TCD8_1_2_3_naive_6_emra",
                    sep = "_"), ".png")
do.call(ggsave, c(list(fn, p), plotsave_param_large))

# EDIT NAMES.
extfig6b <- p
saveRDS(extfig6b, file.path(paths$home_git, "figures", "extfig6b.rds"))
```

Plot T-CD8-7.

```{r t-cd8-7-differential-abundance}
p <- plot_abundance_stages(stats = edgeR_non_age,
          dat = input_df,
          ct_type = "cluster",
          cur_celltypes = c("T_CD8_7"),
          x = "donor_type", y = "mean_density", facet_by = "cell_type",
          fill_by = "donor_type",
          title = "") + mytheme$standard_new()

p1 <- p + facet_wrap(~ cell_type, labeller = as_labeller(c("T_CD8_7" = 
  "PD1+,CD27+")), scales = "free_y")

fn <- paste0(paste(today, "Density", "byCase", "Cluster", "TCD8_7_v2",
                    sep = "_"), ".png")
do.call(ggsave, c(list(fn, p1), plotsave_param))

## 
fig4f <- p1
saveRDS(fig4f, file.path(paths$home_git, "figures", "fig4f.rds"))
```


## A10.5: Association with AGE.

### 10.5.1: Prepare data.
```{r ici-age}
edgeR_age <- edgeR_output |> filter(group1 == "age")

# get max-y-value PER cell type.
cur_celltypes <- unique(edgeR_age$cell_type)
cur_celltypes <- cur_celltypes[!grepl("NK|Neutrophil", cur_celltypes)]

plot_df <- case_join2 |> 
  mutate(donor_type = ifelse(donor_type == "sAAb", "sAAb+", 
                         ifelse(donor_type == "mAAb", "mAAb+", donor_type))) |> 
  dplyr::rename("mean_density" = density) |> 
  mutate(donor_type = factor(donor_type, levels = metadata(spe)$stages[1:4]))

p_vals <- edgeR_age
```

### 10.5.2: Formatting for Barplot.
```{r}
df <- plot_df |> 
  dplyr::rename("cell_type" = immune_cluster) |>
  filter(cell_type %in% cur_celltypes) |> 
  mutate(age_groups2 = ifelse(age_group == "young", "< 13", ">= 13")) |> 
  mutate(age_groups2 = factor(age_groups2, levels = c("< 13", ">= 13")))

p_vals <- p_vals |> 
  filter(cell_type %in% cur_celltypes) |>
  mutate(.y. = "mean_density",
          group1 = "< 13", group2 = ">= 13",
          p = PValue,
          p.adj = FDR,
           p.signif = cut(p.adj, breaks = c(-Inf, 0.001, 0.01, 0.05, 0.1, Inf), labels = c("***", "**", "*", "•", ""))
          )
```

### 10.5.3: Barplot of significances.

```{r}
## Barplot overview.
p <- p_vals |>
  mutate(estimate = ifelse(logFC > 0, ">= 13 years", "< 13 years")) |>
  mutate(significance = ifelse(p.adj < 0.05, "signif", "non_signif")) |>
  filter(p.adj < 0.1) |> 
  ggplot(aes(x = reorder(cell_type, -log10(p.adj)), y = -log10(p.adj), fill = estimate)) +
  geom_bar(stat = "identity") +
  geom_hline(yintercept = -log10(0.05), linetype = "dashed") +
  labs(title = paste("Differential abundance"),
        x = "Cell Type",
        y = "-log10(P-value)") +
  mytheme$large() +
  theme(axis.text.x = element_text(hjust = 1, vjust = 1, angle = 45)) + 
  theme(legend.position = "none")

print(p)
fn <- paste0(paste(today, "Density", "byCase", "Cluster", "Age", "Stages",
                    sep = "_"), ".png")
do.call(ggsave, c(list(fn, p), plotsave_param))

extfig10a <- p 
saveRDS(extfig10a, file.path(paths$home_git, "figures", "extfig10a.rds"))
```


```{r}
## Determine which cell clusters or cell clusters are AGE, STAGE, AGE+STAGE
age_stage <- edgeR_output |> 
  mutate(contrast = paste(group1, group2, sep = "_")) |> 
  select(cell_type, contrast, FDR) |> 
  pivot_wider(names_from = contrast, values_from = FDR) |> 
  mutate(AGE = ifelse(age_group < 0.1, TRUE, FALSE)) |> 
  mutate(STAGE = ifelse(`NoDiabetes_sAAb+` < 0.1 | 
                        `NoDiabetes_mAAb+` < 0.1 |
                        `NoDiabetes_RecentOnset` < 0.1 |
                        `sAAb+_mAAb+` < 0.1 |
                        `mAAb+_RecentOnset` < 0.1, TRUE, FALSE))

not_age_stage <- age_stage |> 
  filter(STAGE == FALSE & AGE == FALSE)
not_age_stage

age_only <- age_stage |> 
  filter(STAGE == FALSE & AGE == TRUE)
age_only

stage_only <- age_stage |> 
  filter(STAGE == TRUE & AGE == FALSE)
stage_only

age_stage_both <- age_stage |> 
  filter(STAGE == TRUE & AGE == TRUE) |> arrange(age_group)
age_stage_both

not_ins_relevant <- c(not_age_stage |> pull(cell_type), age_only |> pull(cell_type))

## These are saved for script 19.
saveRDS(not_ins_relevant[not_ins_relevant != "Myeloid_4"], file.path(paths$folder_script, "not_ins_relevant_clusters.rds"))
```

Age-ONLY: NK-1 + Myeloid-4
For Myeloid-4 we have shown that it shows islet attraction and plays role in motifs.
-> Just hard to detect.

Neither Age + Stage; multiple non-activated clusters (e.g. Myeloid-1).

### A10.6: LMM ALL STAGES: Boxplot for CD11c+ Macrophages, T-CD8-2 and T-CD8-7.

```{r}
# Boxplot for CD11c+ Macrophages.
df_age_plot <- plot_df |> 
  dplyr::rename("cell_type" = immune_cluster) |>
  # clusters oi.
  filter(cell_type %in% c("Myeloid_5", "T_CD8_2")) |>
  mutate(cell_type = ifelse(cell_type == "Myeloid_5", "CD11c+ Macrophages", "CD8+ T-RM low act.")) |>
  mutate(log_mean = log1p(mean_density)) |> 
  mutate(age_groups2 = ifelse(age_group == "young", "< 13", ">= 13")) |> 
  mutate(age_groups2 = factor(age_groups2, levels = c("< 13", ">= 13"))) |> 
  mutate(mean_density = log1p(mean_density))

# Already performed P-val adjustment above.
stats <- p_vals |> 
  filter(cell_type %in% c("Myeloid_5", "T_CD8_2")) |>
  mutate(cell_type = ifelse(cell_type == "Myeloid_5", "CD11c+ Macrophages", "CD8+ T-RM low act.")) |>
  select(.y., group1, group2,p, p.adj, p.signif, logFC, cell_type)

# get max-y-value PER cell type.
p_vals_max <- df_age_plot  |> 
  group_by(cell_type) |>
  summarise(max = max(mean_density), .groups = "drop")

# Get for the layout. 
cur_formula2 <- formula(paste("mean_density ~ age_groups2"))
library(rstatix); library(ggpubr)
stat.test <- df_age_plot |> 
  group_by(cell_type) |>
  t_test(cur_formula2, ref.group = "< 13") |> 
  adjust_pvalue()  |> 
  add_significance("p.adj") |> 
  add_xy_position(x = "age_groups2", step.increase = 0.07) |> 
  select(-c(statistic, p, p.adj))
stat.test

# Combine formatting + P-value dataframe.
stat.test.test <- stats |> 
    left_join(stat.test, by = c("cell_type", ".y.", "group1", "group2"))  |> 
    left_join(p_vals_max, by = c("cell_type")) |> 
    mutate(y.position = y.position * 1.2)
stat.test.test

p <- ggviolin(df_age_plot, x = "age_groups2", y = "mean_density", fill = "age_groups2",
              palette = NULL, facet.by = "cell_type", 
              draw_quantiles = c(0.25, 0.5, 0.75)) + 
    geom_jitter(width = 0.2, size = 3, alpha = 0.8, aes(color = donor_type)) +
    scale_fill_manual("Age Groups",
                  values = palettes$colors50[c(9, 12)])  + 
    scale_color_manual("Stage", values = palettes$stages[1:4]) +
    facet_wrap(~cell_type, scales = "free_y") +
    labs(x = "Age Groups (years)", y = "log1p: Cell Density (cells / mm^2)") + 
    mytheme$standard_new() + 
    theme(axis.text.x = element_text(angle = 45, hjust = 1)) + 
    guides(fill = "none") + 
  stat_pvalue_manual(stat.test.test, label = "p.signif", tip.length = 0.1,
                  size = 10, label.size = 10)

print(p)

fn <- paste0(paste(today, "ImmuneCluster", "AgeAssociation", "Myeloid_5",
                   sep = "_"), ".png")
do.call(ggsave, c(list(fn, p), plotsave_param))

fig6b <- p
saveRDS(fig6b, file.path(paths$home_git, "figures", "fig6b.rds"))
```

#-----------------------------------------------------------

# A13: Cell-Clusters - PERI_ISLET.

## A13.1: Cell-Counts peri-islet islet region.

Get Cell-Counts JUST within + immediate surrounding the Islet-mask ~ - 20 µm radius.

```{r spe_isl}
spe_isl <- spe[, spe$distance_to_islet >= -20 & spe$donor_type != "Pancreatitis"]

# ICIs only.
spe_isl <- spe_isl[, spe_isl$image_id %in% icis]
print(spe_isl)
gc()
```

### A13.1.2: Islet-level cell type composition.

```{r islet-level}
celltypes2 <- c("Beta", "Alpha", "Delta", # Islet
                "Ductal", "Acinar", # Exocrine.
                "Endothelial", "Mesenchymal", "SmoothMuscle", "Nerves", # Stroma.
                "Neutrophil", "T_CD4", "T_CD8", "B", "NK", "Myeloid", "CD303_VIM", "T_DN", # Immune.
                "Other", "IsletOther", "Ambiguous") # Other.

# Number of cells per islet
total_cells_all <- tibble::as_tibble(colData(spe_isl)) |>
  dplyr::add_count(image_id, name = "CellsPerIslet") |>
  dplyr::select(c("image_id", "CellsPerIslet")) |>
  dplyr::distinct()

# Number of cells of each cell type
image_compo_all <- as_tibble(colData(spe_isl)) |>
  filter(cell_type %in% celltypes2) |>
  dplyr::group_by(image_id, cell_type) |>
  dplyr::count(name = "CellNb") |>
  dplyr::ungroup() |>
  tidyr::complete(image_id, cell_type,
                  fill = list(CellNb = 0)) 

# Merge both datasets and calculate the fraction of each cell type per islet
image_compo_all <- image_compo_all |>
  dplyr::inner_join(total_cells_all, by = "image_id") |>
  dplyr::filter(CellsPerIslet > 0) |>
  dplyr::mutate(fraction = CellNb / CellsPerIslet)

image_compo_all <- as_tibble(colData(spe_isl)) |> 
  filter(cell_type %in% celltypes2) |>
  dplyr::select(all_of(cn_keep)) |>
  dplyr::distinct() |>
  dplyr::inner_join(image_compo_all, by = "image_id") |>
  dplyr::mutate(cell_type = factor(cell_type, levels = celltypes2),
                image_id = factor(image_id, levels = image_ids)) |>
  dplyr::arrange(image_id, cell_type)

# Reorder the dataset by beta-cell fraction
image_compo_wide_all <- image_compo_all |>
  tidyr::pivot_wider(id_cols = image_id,
                     values_from = c(CellNb, fraction),
                     names_from = c(cell_type))

ordcols2 <- c(paste0("fraction_", celltypes2))
image_compo_all <- image_compo_wide_all |>
  dplyr::select(all_of(c("image_id", ordcols2))) |>
  dplyr::inner_join(image_compo_all, by = "image_id") |>
  dplyr::arrange(fraction_Beta) |>
  # dplyr::select(-all_of(ordcols)) |>
  dplyr::mutate(ord = seq_len(nrow(image_compo_all))) |> 
  dplyr::rename(case_id = "Donor", donor_type = "Stage") |> 
  dplyr::mutate(cell_type = factor(cell_type, levels = celltypes2),
                case_id = factor(case_id, levels = metadata(spe)$cases),
                donor_type = factor(donor_type, levels = metadata(spe)$stages))
```

```{r case-all2}
case_compo_all_islet <- image_compo_all |>
  dplyr::group_by(donor_type, case_id, cell_type) |> 
  dplyr::summarize(across(c(CellNb, CellsPerIslet), sum)) |>
  dplyr::rename(CellsPerCase = "CellsPerIslet") |>
  dplyr::mutate(fraction = CellNb / CellsPerCase) |>
  dplyr::arrange(donor_type, case_id, cell_type)
```

```{r remove-pancreatitis2}
remove_pancreatitis <- TRUE
if (remove_pancreatitis) {
  case_compo_all_islet <- case_compo_all_islet |> dplyr::filter(donor_type != "Pancreatitis")
}
```

### A13.1.2: Islet-level cell CLUSTER composition.

Cell Cluster!
```{r edgeR-cluster-islet1}
# Number of cells per islet
spe_isl$immune_cluster <- ifelse(spe_isl$cell_cluster_scaled == "0", spe_isl$cell_type, spe_isl$cell_cluster_scaled)
immune_cl <- unique(spe_isl$immune_cluster)
image_ids <- unique(spe_isl$image_id)

total_cells_all <- tibble::as_tibble(colData(spe_isl)) |>
  dplyr::add_count(image_id, name = "CellsPerIslet") |>
  dplyr::select(c("image_id", "CellsPerIslet")) |>
  dplyr::distinct()

# Number of cells of each cell type
image_compo_all_cl <- as_tibble(colData(spe_isl)) |>
  filter(immune_cluster %in% immune_cl) |>
  dplyr::group_by(image_id, immune_cluster) |>
  dplyr::count(name = "CellNb") |>
  dplyr::ungroup() |>
  tidyr::complete(image_id, immune_cluster,
                  fill = list(CellNb = 0)) 

# Merge both datasets and calculate the fraction of each cell type per islet
image_compo_all_cl <- image_compo_all_cl |>
  dplyr::inner_join(total_cells_all, by = "image_id") |>
  dplyr::filter(CellsPerIslet > 0) |>
  dplyr::mutate(fraction = CellNb / CellsPerIslet)

# Add Meta-data
image_compo_all_cl <- as_tibble(colData(spe_isl)) |> 
  filter(immune_cluster%in% immune_cl) |>
  dplyr::select(all_of(cn_keep)) |>
  dplyr::distinct() |>
  dplyr::inner_join(image_compo_all_cl, by = "image_id") |>
  dplyr::mutate(image_id = factor(image_id, levels = image_ids)) |>
  dplyr::arrange(image_id, immune_cluster)

# Reorder the dataset by beta-cell fraction
image_compo_wide_all_cl <- image_compo_all_cl |>
  tidyr::pivot_wider(id_cols = image_id,
                     values_from = c(CellNb, fraction),
                     names_from = c(immune_cluster))

ordcols2 <- c(paste0("fraction_", immune_cl))
image_compo_all_cl <- image_compo_wide_all_cl |>
  dplyr::select(all_of(c("image_id", ordcols2))) |>
  dplyr::inner_join(image_compo_all_cl, by = "image_id") |>
  dplyr::arrange(fraction_Beta) |>
  # dplyr::select(-all_of(ordcols)) |>
  dplyr::mutate(ord = seq_len(nrow(image_compo_all_cl))) |> 
  dplyr::rename(case_id = "Donor", donor_type = "Stage") |> 
  dplyr::mutate(case_id = factor(case_id, levels = metadata(spe_isl)$cases),
                donor_type = factor(donor_type, levels = metadata(spe_isl)$stages))
```

```{r case-all3}
case_compo_all_cl_islet <- image_compo_all_cl |>
  dplyr::group_by(donor_type, case_id, immune_cluster) |> 
  dplyr::summarize(across(c(CellNb, CellsPerIslet), sum)) |>
  dplyr::rename(CellsPerCase = "CellsPerIslet") |>
  dplyr::mutate(fraction = CellNb / CellsPerCase) |>
  dplyr::arrange(donor_type, case_id, immune_cluster)
```

### A13.1.3: Get Insulitic Islet Information.
```{r}
## Get Ins. Islets
fn <- file.path(paths$folder_out, "13_MarkerExpression_cells_Immune", "insulitic_images.rds")

if (file.exists(fn)) {
  image_ids <- readRDS(fn)
  image_compo_all_cl <- image_compo_all_cl |> 
    mutate(insulitic = if_else(image_id %in% image_ids, "insulitis", "non-insulitis"))
}
```

### A13.1.4: Input for age-association of immune clusters.

```{r}
cur_dat_cl <- image_compo_all_cl |>
  select(-starts_with("fraction_")) |> 
  # factor to character
  left_join(temp_dur, by = c("case_id", "donor_type")) |>
  ungroup() |> 
  # remove _1, _2, _10, 
  mutate(cell_type = stringr::str_remove(immune_cluster, pattern = "_[0-9]+$")) |> 
  mutate(cell_type = ifelse(cell_type %in% c("`T-naive`","`T-EMRA`"), "T_CD8", cell_type))

## Age and immune cluster.
df_age <- cur_dat_cl |> 
  mutate(age_groups2 = ifelse(age < 13, "young", "high")) |>
  mutate(age_groups2 = factor(age_groups2, levels = c("high", "young"))) |> 
  mutate(immune_cluster = ifelse(immune_cluster %in% c("B_1", "B_2"), "B", immune_cluster)) |>
  dplyr::filter(stringr::str_detect(immune_cluster, pattern = "(^Myeloid|^T|^B)")) |> 
  dplyr::filter(!immune_cluster %in% c("T_DN_1", "T_DN_2", "CD303_VIM_1", "CD303_VIM_2", "Lympho_Other", "T_CD4_Other", "Myeloid_Other")) |> 
  group_by(image_id, donor_type, case_id, insulitic, age, cell_type,  
           age_groups2, immune_cluster, diabetes_duration) |>
  summarize(mean_density = mean(CellNb, , na.rm = TRUE), 
            fraction = mean(fraction, na.rm = TRUE), 
            CellNb = sum(CellNb), .groups = "drop")
```


## A13.2: Run edgeR for immune clusters in insulitic-islets!

Select only image-ids (mAAb+/RECENT-ONSET) with Insulitis.
The contrast here is **between age-groups**!

edgeR: CellNb(insulitic) ~ age_group

Aim is to detect differential immune-compositions between age-groups in insulitic islets!
-> We identified B-cells/T-Naive 

```{r}
# Select only donors with at least 1 insulitic islet and < 1 y since onset.
case_ids <-  cur_dat_cl |> 
  filter(donor_type == "mAAb+" | donor_type == "RecentOnset" & diabetes_duration < 1) |>
  filter(insulitic == "insulitis") |> 
  distinct(case_id) |> 
  pull(case_id)

## Select image-ids with insulitis only.
insulitic_df <- cur_dat_cl |> 
  filter(case_id %in% case_ids, insulitic == "insulitis") |> 
  mutate(age_group2 = ifelse(age < 13, "younger", "older")) |>
  mutate(age_group2 = factor(age_group2, levels = c("younger", "older"))) |>
  mutate(immune_cluster = ifelse(immune_cluster %in% c("B_1", "B_2"), "B", immune_cluster)) |>
  group_by(case_id, donor_type, CellsPerIslet, immune_cluster, age_group2) |>
  summarize(CellNb = sum(CellNb), .groups = "drop")

insulitic_df
```

### A13.2.1: Prepare edgeR input for insulitic islets.
  ```{r}
## Aggregate by Case-ID!
case_compo_all_islet <- insulitic_df |>
  dplyr::group_by(donor_type, case_id, immune_cluster, age_group2) |> 
  dplyr::summarize(across(c(CellNb, CellsPerIslet), sum)) |>
  ungroup() |>
  dplyr::rename(CellsPerCase = "CellsPerIslet") |>
  dplyr::mutate(fraction = CellNb / CellsPerCase) |>
  dplyr::arrange(donor_type, case_id, immune_cluster) |> 
  dplyr::left_join(cases_full, by = "case_id")

edge <- case_compo_all_islet |>
  dplyr::ungroup() |>
  dplyr::select(c(case_id, immune_cluster, CellNb, age_group2)) |>
  tidyr::pivot_wider(id_cols = immune_cluster,
                     names_from = case_id, values_from = CellNb) |>
  as.data.frame()

rownames(edge) <- edge$immune_cluster
edge$immune_cluster <- NULL
edge <- as.matrix(edge)

# coldata <- unique(case_compo_all_islet[, c("donor_type", "case_id")])
coldata <- unique(case_compo_all_islet[, c("age_group2", "case_id", "ICU_time_days")]) |> as.data.frame()
coldata$ICU_time_days <- log1p(coldata$ICU_time_days)
coldata$age_group = factor(coldata$age_group2, levels = c("younger", "older"))
```

Filter Cell types + Create Design Matrix.
```{r edge-design-matrix-islet}
# Create DGEList (main object)
y_ab <- edgeR::DGEList(edge, samples = coldata)
y_ab

# Filter cell types.  Remove NK, B, T-DN and CD303-VIM. Too rare in islets.
keep <- edgeR::filterByExpr(y_ab, group = y_ab$samples$age_group2)
keep["Myeloid_4"] <- TRUE
y_ab <- y_ab[keep, ]
summary(keep)
keep[!keep]

# Create design matrix
y_ab$samples$ICU_time_days[is.na(y_ab$samples$ICU_time_days)] <- mean(y_ab$samples$ICU_time_days, na.rm = TRUE)
design <- model.matrix(~age_group2 + ICU_time_days, y_ab$samples)
```

Estimate NB-dispersion.
```{r estimate-dispersion-islet}
y_ab <- edgeR::estimateDisp(y_ab, design)
summary(y_ab$common.dispersion)
plotBCV(y_ab, cex=1)
```

Fit the NB-model. 

```{r fit-ql-model-islet}
fit_ab <- edgeR::glmQLFit(y_ab, design, robust = TRUE, abundance.trend = TRUE, legacy = FALSE)
summary(fit_ab$var.prior)
summary(fit_ab$df.prior)
plotQLDisp(fit_ab, cex = 1)
```

### A13.2.2: Run edgeR tests and extract contrast of interest.
```{r}
my_contrasts <- limma::makeContrasts(
  age_group = -`age_group2older`,
  ICU_time_days = -ICU_time_days,
  levels = design
)

## Extract contrasts of interest. Either vs baseline; or between consecutive stages.
qlf_list <- lapply(colnames(my_contrasts), function(cn) {
  topTags(glmQLFTest(fit_ab, contrast = my_contrasts[, cn]), n = "all")$table |> 
    as_tibble(rownames = "cell_type") |> 
    mutate(contrast = cn)
}) |> 
  bind_rows() |> 
  filter(grepl(("Myeloid|T_CD4|T_CD8|NK|Neutrophil|T-naive|T-EMRA|B"), cell_type))

full_results <- tibble::as_tibble(edgeR::glmQLFTest(fit_ab)$table, rownames = "immune_cluster")

p_vals <- qlf_list |> 
  filter(cell_type %in% 
    c("Myeloid_3", "Myeloid_4", "Myeloid_5", "T-naive", 
      "T_CD8_7", "T_CD8_2", "B", "T_CD4_5"), contrast == "age_group") |> 
  mutate(p.adj = FDR)
```

## A13.3: Plot results.
```{r}
# Peri-insulitic images. B/T_CD4 ratio.
p <- df_age |> 
  select(-c(fraction))   |> 
  #mutate(peri_insulitic = ifelse(image_id %in% peri_insulitic, "peri-insulitis", "non-peri-insulitis")) |>
  filter(insulitic == "insulitis") |> 
  # filter(donor_type %in% c("RecentOnset")) |>
  filter(donor_type %in% c("RecentOnset", "mAAb+")) |>
  pivot_wider(id_cols = c(image_id, donor_type, case_id, insulitic, age, age_groups2), 
                names_from = immune_cluster, values_from = CellNb) |>
  #group_by(case_id, age, age_groups, donor_type) |>
  # summarize(across(where(is.numeric), ~ mean(.x)), .groups = "drop") |>
  mutate(T_CD4 = T_CD4_1 + T_CD4_2 + T_CD4_3 + T_CD4_4 + T_CD4_5 + T_CD4_6 + T_CD4_7 + T_CD4_8 + T_CD4_9 + T_CD4_10) |>
  mutate(frac = B / T_CD4) |>
  filter(T_CD4 > 0) |>
#   pivot_longer(cols = c(B, T_CD4, T_CD8_1), names_to = "cell_type", values_to = "CellNb") |>
  ggplot(aes(x = age, y = frac, color = age_groups2)) +
  geom_point() + 
  geom_hline(yintercept = 1, linetype = "dashed") + 
  scale_y_log10() + 
  ylab("B / T_CD4 ratio") + 
  mytheme$standard_new()

fn <- paste(today, "B_TCD4_ratio", "byCase", "Insulitis", "ICI.png", sep = "_")
do.call(ggsave, c(list(fn, p), plotsave_param_large))
```

```{r}
subtypes_oi <- c("T-naive", "B", 
                "Myeloid_3", "T_CD8_7", 
                "Myeloid_5","T_CD8_2", 
                 "Myeloid_4", "T_CD4_5")

# By Case-ID:
tt <- df_age |> 
  select(-c(fraction))   |> 
  filter(insulitic == "insulitis") |>
  mutate(age_groups2 = ifelse(age < 13, "< 13", ">= 13")) |>
  mutate(age_groups2 = factor(age_groups2, levels = c("< 13", ">= 13"))) |>
  # filter(donor_type %in% c("RecentOnset")) |>
  filter(donor_type %in% c("mAAb+", "RecentOnset")) |>
  filter(donor_type == "mAAb+" | diabetes_duration < 1) |>
  filter(immune_cluster %in% subtypes_oi) |> 
  mutate(immune_cluster = factor(immune_cluster, levels = subtypes_oi)) |> 
  group_by(case_id, age, age_groups2, donor_type, 
           immune_cluster, insulitic) |>
  summarize(across(where(is.numeric), ~ mean(.x)), .groups = "drop") |> 
  mutate(CellNb = log1p(CellNb))

tt
# Already performed P-val adjusted above.
stats <- p_vals |> 
  mutate(.y. = "CellNb",
          immune_cluster = cell_type,
          group1 = "< 13", group2 = ">= 13",
           p.signif = cut(p.adj, breaks = c(-Inf, 0.001, 0.01, 0.05, 0.1, Inf), labels = c("***", "**", "*", "•", ""))
          ) |> 
  select(.y., group1, group2,
          p.adj, p.signif, immune_cluster)

# get max-y-value PER cell type.
p_vals_max <- tt  |> 
  group_by(immune_cluster) |>
  summarise(max = max(CellNb), .groups = "drop")

# Get for the layout. 
cur_formula2 <- formula(paste("CellNb ~ age_groups2"))
library(rstatix); library(ggpubr)
stat.test <- tt |> 
  group_by(immune_cluster) |>
  t_test(cur_formula2, ref.group = "< 13") |> 
  add_xy_position(x = "age_groups2", step.increase = 0.07)
stat.test

# Combine formatting + P-value dataframe.
stat.test.test <- stats |> 
    left_join(stat.test, by = c("immune_cluster", ".y.", "group1", "group2"))  |> 
    left_join(p_vals_max, by = c("immune_cluster")) |>
    mutate(y.position = max*1.5) |> 
    mutate(immune_cluster = factor(immune_cluster, levels = subtypes_oi))

stat.test.test

# mAAb+ and RecentOnset (< 1 year duration)
p <- ggpubr::ggviolin(tt, x = "age_groups2", y = "CellNb", fill = "age_groups2",
                  palette = palettes$stages, facet.by = "immune_cluster",
                  draw_quantiles = c(0.25, 0.5, 0.75)) +
  geom_jitter(width = 0.2, size = 8, alpha = 0.8, 
              aes(color = donor_type)) +
  scale_fill_manual("Age Groups", values = palettes$colors50[c(9, 12)])  + 
  scale_color_manual("Stage", values = palettes$stages, 
                      breaks = c("mAAb+", "RecentOnset"),
                      labels = c("mAAb+", "Onset")) +
  facet_wrap(~immune_cluster, scales = "free_y", axes = "all") +
  labs(x = "Age Groups (years)", y = "log1p: CellNb") + 
  mytheme$large_new() + 
  stat_pvalue_manual(stat.test.test, label = "p.signif", tip.length = 0.2,
                    size = 20, label.size = 10) +  # Format as 0.1, 1, 10, ...
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) + 
  guides(fill = "none") + 
  geom_hline(yintercept = 1, linetype = "dashed") + 
  # extend y-axis a bit.
  scale_y_continuous(expand = expansion(mult = c(0, 0.1)))

print(p)
fn <- paste0(paste(today, "Density", "byCase", "CellType", "T_CD8", "B", "T_CD4", "Myeloid", "T_CD8_7", "T_CD8_2", "T-naive",
                   sep = "_"), ".png")
do.call(ggsave, c(list(fn, p), plotsave_param_large))

fig6f <- p
saveRDS(fig6f, file.path(paths$home_git, "figures", "fig6f.rds"))
```