---
title: "13_MarkerExpression_cells_Immune"
author: "Nathan Steenbuck, Nicolas Damond"
date: "Created: 05 November, 2022; Compiled: `r format(Sys.time(), '%d %b, %Y')`"
output: workflowr::wflow_html
editor_options:
  chunk_output_type: inline
---

```{r setup, include=FALSE}
if (rstudioapi::isAvailable()) {
  script_name <- basename(rstudioapi::getSourceEditorContext()$path)
} else {
  script_name <- knitr::current_input()
}

cur_user <- Sys.info()[["user"]]
script_name <- "13_MarkerExpression_cells_Immune.Rmd"

if (cur_user == "ubuntu") {
    source(file.path("/", "mnt", "central_nas", "projects", 
    "type1_diabetes", "nathan", "T1D_Vol", "T1D_analysis", 
    "analysis", "helpers.R"))  
  #n_cores <- future::availableCores() - 1
  n_cores <- 4
  future::plan(future::multicore(workers = n_cores))
  paths <- getPaths(script_name)
  knitr::opts_knit$set(root_dir = paths$home_data)
  do_print <- FALSE
} else {
  source(file.path("/", "home", "nsteen", "scripts", "T1D_analysis", "analysis", "helpers.R"))
  n_cores <- 8
  future::plan(future::multicore(workers = n_cores))
  paths <- getPaths(script_name)
  paths$folder_script <- paths$cluster_folder_script
  paths$folder_in <- paths$cluster_folder_in
  paths$folder_out <- paths$cluster_home 
  knitr::opts_knit$set(root_dir = paths$cluster_home)
  do_print <- TRUE
}

seed <- 123456
set.seed(seed)
options <- furrr::furrr_options(seed = seed)
paths$prev <- paste("08_CellTypesIslet", paths$object_type, paths$panel_type, sep = "_")
knitr::opts_chunk$set(echo = TRUE)
```

Rscript -e "rmarkdown::render('T1D_analysis/analysis/13_MarkerExpression_cells_Immune.Rmd')"


# **Goal**

Examine protein abundance. The marker expression levels are plotted in function of cell types, donors, T1D stages and distance to islet edge.  

Current data representations:
* Heatmaps (annotated by cell type, donor and stage).
* Cell type comparison (boxplots, violin plots, density plots).
* Case and stage comparison (boxplots, violin plots, density plots).
* Distance to islet edge (scatter plots).

Additionally, immune cell composition and expression is compared between ICIs and IDIs,
as well as between Insulitic & Non-Insulitic cases.

# A0: **Settings**

## A0.1: Load packages

```{r packages, results='hide'}
suppressPackageStartupMessages(c(
  library(data.table),
  library(dplyr),
  library(SpatialExperiment),
  library(scuttle),
  library(parallel),
  library(patchwork),
  library(tidyr),
  library(RSpectra)
))
```

## A0.2: Paths and settings

```{r settings}
# Paths
if (!dir.exists(paths$folder_script)) dir.create(paths$folder_script)
plotsave_param$path <- paths$folder_script
plotsave_param_large$path <- paths$folder_script

# Misc settings
today <- gsub("-", "", Sys.Date())
```

##  A0.3: Read in the data

Load the SpatialExperiment (SPE) object saved at the previous step.

```{r load-data}
fn_spe <- file.path(paths$folder_out, paths$prev, paste0(paths$object_type, "_", paths$panel_type, ".rds"))
spe <- readRDS(fn_spe)
print(spe)
```

## A0.4: Read in Immune Subset only

```{r load_sce_imm}
sce_imm <- spe[, spe$cell_category == "immune"]
# fn_sceimm <- file.path(paths$folder_script, paste0("sce_imm", "_", paths$object_type, "_", paths$panel_type, ".rds"))
# saveRDS(sce_imm, fn_sceimm)
# sce_imm <- readRDS(fn_sceimm)
```


## A0.5: Parameters

**Note: this step has to be manually adapted**  

- Define the assay(s) of interest.  
- Define the markers to keep for marker expression analysis.  

```{r parameters}
# Censor value for counts scaling
censor_val <- 0.9999

# Select assays and reduced dimensions
assay_sel <- c("exprs", "scaled")
writeLines(c("Assay:", assay_sel[assay_sel %in% assayNames(spe)]))

# Variables to plot
plot_variables <- c("donor_type", "case_id")
names(plot_variables) <- c("stages", "cases")
cat("\nVariables to plot:", plot_variables)
plot_variables_stages <- plot_variables["stages"]

# List of cell types to plot
# celltypes <- names(metadata(spe)$channels)
# names(celltypes) <- celltypes
## Add other cell types as well!
celltypes <- c("T_CD4", "T_CD8", "B", "NK", "Neutrophil", 
               "Myeloid", "Alpha", "Ductal", "Acinar", "Endothelial", "Beta")
names(celltypes) <- rep("cell_type", length(celltypes))

# Color palette
palette_ct <- palettes$colors[seq_len(length(celltypes))]
names(palette_ct) <- celltypes

# All channels to display
channels <- unique(unlist(metadata(spe)$channels))
```

## Subset the SCE

```{r subset-spe}
sce_sub <- as(spe, "SingleCellExperiment")
sce_sub <- sce_sub[channels, sce_sub$cell_type %in% celltypes]
```

# **A1: Prepare the data**

## A1.1: Functions

### A1.1.1: Aggregate SPE objects.

```{r function-aggregate-SPE}
aggregate_val <- function(x, cluster_col, aggreg_col, .channels,
                          stats = "mean", assay.type = "counts") {
  x <- aggregateAcrossCells(
    x, ids = DataFrame(
      SampleId = x[[cluster_col]], x[[aggreg_col]]),
    subset.row = .channels, statistics = stats,
    use.assay.type = assay.type, use.dimred = FALSE
  )
  colnames(x) <- paste(x[[cluster_col]], x[[aggreg_col]], sep = "_")
  return(x)
}
```

### A1.1.2: Transform and scale counts

```{r function-transform-counts}
transf_counts <- function(x, censor_val = 0.99) {
  assay(x, "exprs") <- asinh(assay(x, "counts"))
  
  quant <- apply(assay(x, "counts"), 1,
                 quantile, probs = censor_val)
  assay(x, "scaled") <- apply(assay(x, "counts"), 2, function(x) x / quant)
  assay(x, "scaled")[assay(x, "scaled") > 1] <- 1
  assay(x, "scaled")[assay(x, "scaled") < 0] <- 0
  return(x)
}
```

## A1.2: Summarize 

### A1.2.1: By cell type and stage

Average marker expression per cell type and stage.

```{r summarize-celltype-stage}
sce_ct_stage <- aggregate_val(
  sce_sub, "cell_type", "donor_type", channels
)
sce_ct_stage <- transf_counts(sce_ct_stage, censor_val)
```

### A1.2.2: By cell type and case

Average marker expression per cell type and case.

```{r summarize-celltype-case}
sce_ct_case <- aggregate_val(
  sce_sub, "cell_type", "case_id", channels
)
sce_ct_case <- transf_counts(sce_ct_case, censor_val)
```

### A1.2.3: By cell type and Islet

Average marker expression per cell type and Islet.
For this we take the average of all cells across the **image_id**.
One image may contain multiple islets, but this corrects for small islets, which skews averages.

```{r summarize-celltype-case-islet}
sce_ct_islet <- aggregate_val(
  sce_sub, "cell_type", "image_id", channels)
sce_ct_islet <- transf_counts(sce_ct_islet, censor_val)
```

### A1.2.4: Save the SCEs

```{r save_rds, results='hide'}
# save sce_ct_case
fn_sce_ct_case <- file.path(paths$folder_script, paste0("sce_ct_case", "_", paths$object_type, "_", paths$panel_type, ".rds"))
saveRDS(sce_ct_case, fn_sce_ct_case)
# sce_ct_case <- readRDS(fn_sce_ct_case)

# save sce_ct_stage
fn_sce_ct_stage <- file.path(paths$folder_script, paste0("sce_ct_stage", "_", paths$object_type, "_", paths$panel_type, ".rds"))
saveRDS(sce_ct_stage, fn_sce_ct_stage)
# sce_ct_stage <- readRDS(fn_sce_ct_stage)

# save sce_ct_islet:
fn_sce_ct_islet <- file.path(paths$folder_script, paste0("sce_ct_islet", "_", paths$object_type, "_", paths$panel_type, ".rds"))
saveRDS(sce_ct_islet, fn_sce_ct_islet)
# sce_ct_islet <- readRDS(fn_sce_ct_islet)
```

# **A2: Visualize marker expression on heatmaps**

```{r packages2, results='hide'}
suppressPackageStartupMessages(c(
  library(ggplot2),
  library(viridis),
  library(heatmaply),
  library(dittoSeq),
  library(ggridges),
  library(mclust)
))
```

## A2.1: dittoHeatmap

### A2.1.1: Heatmaps by stage
                     
```{r dittoheatmap-stages}
purrr::map(assay_sel, \(cur_assay) {
  # Non-scaled
  fn1 <- paste0(paste(today, "HeatmapDitto", "CellType", "Stage", cur_assay, 
                      "NotScaled", sep = "_"), ".png")
  
  h1 <- list(object = sce_ct_stage, assay = cur_assay,
             annot.by = c("donor_type", "cell_type"),
             annotation_colors = list(donor_type = palettes$stages,
                                      cell_type = palette_ct),
             scale = "none", heatmap.colors = viridis(100),
             cluster_cols = TRUE,
             main = paste("Cell type - Not scaled -", cur_assay),
             plot_method = "plotly")
  do.call(dittoHeatmap, h1)
  do.call(dittoHeatmap, c(h1, filename = file.path(paths$folder_script, fn1)))
  
  # Centered and scaled
  fn2 <- paste0(paste(today, "HeatmapDitto", "CellType", "Stage", cur_assay, 
                      "CenteredScaled", sep = "_"), ".png")

  h2 <- list(object = sce_ct_stage, assay = cur_assay,
             annot.by = c("donor_type", "cell_type"),
             annotation_colors = list(donor_type = palettes$stages,
                                      cell_type = palette_ct),
             cluster_cols = TRUE,
             heatmap.colors = colorRampPalette(c("dark blue", "white",
                                                 "dark red"))(100),
             breaks = seq(-3, 3, length.out = 101),
             main = paste("Cell type - Centered and Scaled -", cur_assay))
  do.call(dittoHeatmap, h2)
  do.call(dittoHeatmap, c(h2, filename = file.path(paths$folder_script, fn2)))
  remove(h1, h2)
})
```

### A2.1.2: Heatmaps by case

```{r dittoheatmap-cases}
purrr::map(assay_sel, \(cur_assay) {
  # Non-scaled
  fn1 <- paste0(paste(today, "HeatmapDitto", cur_assay, "CellType", "Case",
                      "NotScaled", sep = "_"), ".png")
  
  h1 <- list(object = sce_ct_case, assay = cur_assay,
             annot.by = c("case_id", "donor_type", "cell_type"),
             annotation_colors = list(case_id = palettes$cases,
                                      donor_type = palettes$stages,
                                      cell_type = palette_ct),
             scale = "none", heatmap.colors = viridis(100),
             cluster_cols = TRUE,
             main = paste("Cell type - Not scaled -", cur_assay),
             plot_method = "plotly")
  do.call(dittoHeatmap, h1)
  do.call(dittoHeatmap, c(h1, filename = file.path(paths$folder_script, fn1)))
  
  # Centered and scaled
  fn2 <- paste0(paste(today, "HeatmapDitto", cur_assay, "CellType", "Case",
                      "CenteredScaled", sep = "_"), ".png")

  h2 <- list(object = sce_ct_case, assay = cur_assay,
             annot.by = c("case_id", "donor_type", "cell_type"),
             annotation_colors = list(case_id = palettes$cases,
                                      donor_type = palettes$stages,
                                      cell_type = palette_ct),
             cluster_cols = TRUE,
             heatmap.colors = colorRampPalette(c("dark blue", "white",
                                                 "dark red"))(100),
             breaks = seq(-3, 3, length.out = 101),
             main = paste("Cell type - Centered and Scaled -", cur_assay))
  do.call(dittoHeatmap, h2)
  do.call(dittoHeatmap, c(h2, filename = file.path(paths$folder_script, fn2)))
  remove(h1, h2)
})
```



# **A3: Marker expression across cell types**

Marker expression is only plot by stage (by case is unreadable if too many cases).

## A3.1: Prepare the data

Convert SCE to melted data frame.

```{r convert-sce-datatable}
dat <- vector(mode = "list", length = length(assay_sel))
names(dat) <- assay_sel

for (i in seq_along(assay_sel)) {
  dat[[i]] <- makePerCellDF(
    sce_ct_case,
    exprs_values = assay_sel[i],
    features = channels,
    use.coldata = c("cell_type", plot_variables),
    use_dimred = FALSE) %>%
    mutate(case_id = factor(case_id, levels = metadata(spe)$cases),
           donor_type = factor(donor_type, levels = metadata(spe)$stages),
           cell_type = factor(cell_type, levels = celltypes)) %>%
    arrange(cell_type, case_id, donor_type) %>%
    as.data.table()
  
  dat[[i]] <- melt.data.table(
    dat[[i]],
    id.vars = c("cell_type", plot_variables),
    measure.vars = channels,
    variable.name = "channel",
    value.name = assay_sel[i]
  )
  dat[[i]][, case_id := as.factor(case_id)]
}
```

## A3.2: Boxplots

```{r boxplots-celltype, fig.width=15, fig.height=10}
purrr::iwalk(assay_sel, \(cur_assay, idx) {
  purrr::iwalk(plot_variables_stages, \(cur_variable, k) {
    cur_dat <- dat[[idx]]
    cur_dat[, MeanExprs := mean(get(cur_assay)),
            by = c(plot_variables[k], "cell_type", "channel")]
    
    p <- plot_boxes(cur_dat, x = "cell_type", y = cur_assay, 
                    fill_by = "MeanExprs", color_by = plot_variables_stages[k],
                    facet_by = "channel", ncol = 10,
                    title = paste(plot_variables_stages[k], cur_assay, "Cell types")) +
      scale_color_manual(values = palettes[[names(plot_variables_stages[k])]])
    
    fn <- paste0(paste(today, "Boxplot", "CellTypes", cur_assay,
                       names(plot_variables[k]), sep = "_"), ".png")
    do.call(ggsave, c(list(fn, p), plotsave_param_large))
    if (do_print) print(p)
  })
})
```

## A3.3: Violin plots

```{r violin-celltype, message=FALSE, warning=FALSE, fig.width=15, fig.height=10}
purrr::iwalk(assay_sel, \(cur_assay, idx) {
  purrr::iwalk(plot_variables_stages, \(cur_variable, k) {
    cur_dat <- dat[[idx]]
    cur_dat[, MeanExprs := mean(get(cur_assay)),
            by = c(plot_variables_stages[k], "cell_type", "channel")]
    
    p <- plot_violin(cur_dat, x = "cell_type", y = cur_assay, 
                     fill_by = "MeanExprs", color_by = plot_variables_stages[k],
                     facet_by = "channel", ncol = 10,
                     title = paste(plot_variables_stages[k], cur_assay,
                                   "Cell types")) +
      scale_color_manual(values = palettes[[names(plot_variables_stages[k])]])
    
    fn <- paste0(paste(today, "Violin", "CellTypes", cur_assay,
                       names(plot_variables_stages[k]), sep = "_"), ".png")
    do.call(ggsave, c(list(fn, p), plotsave_param_large))
    suppressMessages(if (do_print) print(p))
  })
})
```


## A3.4: Violin Plots, without LongDuration
```{r violin-celltype-woLD}
purrr::iwalk(assay_sel, \(cur_assay, idx) {
  purrr::iwalk(plot_variables_stages, \(cur_variable, k) {
    
    cur_dat <- dat[[idx]] |> filter(!(cell_type == "Beta" & donor_type == "LongDuration"))
    cur_dat[, MeanExprs := mean(get(cur_assay)),
            by = c(plot_variables_stages[k], "cell_type", "channel")]
    
    p <- plot_violin(cur_dat, x = "cell_type", y = cur_assay, 
                     fill_by = "MeanExprs", color_by = plot_variables_stages[k],
                     facet_by = "channel", ncol = 10,
                     title = paste(plot_variables_stages[k], cur_assay,
                                   "Cell types")) +
      scale_color_manual(values = palettes[[names(plot_variables_stages[k])]])
    
    fn <- paste0(paste(today, "Violin", "CellTypes_noLD", cur_assay,
                       names(plot_variables_stages[k]), sep = "_"), ".png")
    do.call(ggsave, c(list(fn, p), plotsave_param_large))
    if (do_print) suppressMessages(print(p))
  })
})
```

## A3.5: Density plots

```{r density-celltypes, message=FALSE, warning=FALSE, fig.width=15, fig.height=10}
purrr::iwalk(assay_sel, \(cur_assay, i) {
  purrr::iwalk(plot_variables_stages, \(cur_variable, k) {
    cur_dat <- dat[[i]]
    cur_dat <- subset(cur_dat, !is.infinite(cur_assay))
    cur_dat[, MeanExprs := mean(get(cur_assay)),
            by = c(plot_variables[k], "cell_type", "channel")]
    
    p <- plot_density(cur_dat, x = cur_assay, y = "cell_type", 
                      fill_by = "MeanExprs", color_by = plot_variables[k],
                      facet_by = "channel", ncol = 4,
                      title = paste(plot_variables[k], cur_assay, "Cell types")) +
      scale_color_manual(values = palettes[[names(plot_variables[k])]])
    
    fn <- paste0(paste(today, "Density", "CellTypes", cur_assay,
                       names(plot_variables[k]), sep = "_"), ".png")
    do.call(ggsave, c(list(fn, p), plotsave_param_large))
    if (do_print) suppressMessages(print(p))
  })
})
```

## A3.6: Remove unneeded objects

```{r clean-env}
remove(sce_sub, dat); gc()
```



# A4: **Marker expression by cell type**

## A4.1: Read in the data

Load the list containing one SPE object per cell type.

```{r load-data2}
# Load the SCE list
prev <- "12_CellTypeSCEs_cells_Immune"
fn_sces <- file.path(paths$folder_out, prev, paste0("celltypes", "_", paths$panel_type, ".rds"))
sces <- readRDS(fn_sces)
```

Convert SPEs to SCEs

```{r convert-to-SCEs}
sces <- lapply(sces, as, "SingleCellExperiment")
print(sces)
```

## A4.2: Aggregate the data

Aggregate by cell type and case.

```{r aggregate-sces-celltype-case}
sces_ct_cases <- lapply(sces, aggregate_val, "cell_type", "case_id", NULL)
sces_ct_cases <- lapply(sces_ct_cases, transf_counts, censor_val)
```

Convert SCEs to melted data frame.

```{r prepare-data-celltype-case}
prepare_data <- function(cur_sce, assay, variables) {
  cur_channels <- metadata(cur_sce)$channels
  
  cur_sce <- makePerCellDF(
    cur_sce,
    features = cur_channels,
    assay.type = assay,
    use.coldata = variables,
    use_dimred = FALSE) %>%
    mutate(case_id = factor(case_id, levels = metadata(cur_sce)$cases),
           donor_type = factor(donor_type, levels = metadata(cur_sce)$stages)) %>%
    arrange(case_id, donor_type) %>%
    as.data.table()
  
  cur_sce <- cur_sce %>%
    melt.data.table(
      id.vars = variables,
      measure.vars = cur_channels,
      variable.name = "channel",
      value.name = assay
    )
  
  # COMMENT OUT TO INCLUDE PANCREATITIS CASES
  cur_sce <- cur_sce[cur_sce$donor_type != "Pancreatitis", ]
  
  return(cur_sce)
}

# Aggregate by cell type and case
dat_ct_cases <- vector(mode = "list", length = length(assay_sel))
names(dat_ct_cases) <- assay_sel

for (i in seq_along(assay_sel)) {
  dat_ct_cases[[assay_sel[i]]] <- lapply(
    sces_ct_cases, prepare_data, assay_sel[i], plot_variables)
}

# # Single cell data
# dat <- vector(mode = "list", length = length(assay_sel))
# names(dat) <- assay_sel
# 
# purrr::map(assay_sel, \(cur_assay) {
#   dat[[cur_assay] <- lapply(
#     sces, prepare_data, cur_assay, plot_variables)
# }
```

## A4.3: Boxplots

```{r boxplots-celltype-case, fig.width=15, fig.height=10}
purrr::map(assay_sel, \(cur_assay) {
  purrr::map(names(dat_ct_cases[[1]]), \(cur_celltype) {
    # cur_assay <- assay_sel[1]; cur_celltype <- celltypes[1]
    purrr::iwalk(plot_variables_stages, \(cur_variable, k) {
      
      cur_dat <- dat_ct_cases[[cur_assay]][[cur_celltype]]
      cur_dat[, MeanExprs := mean(get(cur_assay)),
              by = c(plot_variables_stages[k], "channel")]
      
      p <- plot_boxes(cur_dat, x = plot_variables_stages[k], y = cur_assay, 
                      fill_by = "MeanExprs",
                      color_by = plot_variables_stages[k],
                      facet_by = "channel", ncol = 10,
                      title = paste(plot_variables_stages[k], cur_assay,
                                    cur_celltype)) +
        theme(axis.text.x = element_text(angle = 90, hjust = 1)) +
        scale_color_manual(values = palettes[[names(plot_variables_stages[k])]])
      
      fn <- paste0(paste(today, "Boxplot", cur_celltype, cur_assay, 
                         names(plot_variables_stages[k]), sep = "_"), ".png")
      do.call(ggsave, c(list(fn, p), plotsave_param_large))
      if (do_print) print(p)
    })
  })
})
```

## A4.4: Violin plots

```{r violin-casestage, message=FALSE, warning=FALSE, fig.width=15, fig.height=10}
purrr::map(assay_sel, \(cur_assay) {
  purrr::map(names(dat_ct_cases[[1]]), \(cur_celltype) {
    purrr::iwalk(plot_variables_stages, \(cur_variable, k) {
      cur_dat <- dat_ct_cases[[cur_assay]][[cur_celltype]]
      cur_dat[, MeanExprs := mean(get(cur_assay)),
              by = c(plot_variables_stages[k], "channel")]
      ## Calculate Wilcox-Test between ND and other stages.
      p_vals <- cur_dat  |> 
        as_tibble()  |> 
                # rename scaled to exprs if needed.
        rename_with(~ stringr::str_replace(., "scaled", "exprs")) |>  
        nest(data = -c(channel)) |> 
        # dplyr::mutate(test = purrr::map(data, ~ broom.mixed::tidy(lmerTest::lmer(data = ., MeanExprs ~ donor_type + (1 | case_id))))) |>
        mutate(test = purrr::map(data, ~ broom::tidy(lm(data = ., exprs ~ donor_type)))) |>
        tidyr::unnest(test) |> 
        filter(term != "(Intercept)") |> 
        mutate(donor_type = stringr::str_remove(term, "donor_type"))
      
      p_vals$significance <- cut(p_vals$p.value,
                                breaks = c(-Inf, 0.001, 0.01, 0.05, Inf),
                                labels = c("***", "**", "*", ""))
                               
      max_y_value <- max(cur_dat[[cur_assay]]) * 1.2
      pos_pvals <- max_y_value * 0.85
      p <- plot_violin(cur_dat, x = plot_variables_stages[k],
                       y = cur_assay, fill_by = "MeanExprs",
                       color_by = plot_variables_stages[k],
                       facet_by = "channel", ncol = 6,
                       title = paste(plot_variables_stages[k], cur_assay,
                                     cur_celltype)) +
        ggplot2::ylim(c(0, max_y_value)) + 
        theme(axis.text.x = element_text(angle = 90, hjust = 1)) +
        scale_color_manual(values = palettes[[names(plot_variables_stages[k])]])
        # add the p.values. Plot p-value per channel AND per donor_type
      # Adding significance stars
      p <- p + geom_text(data = p_vals, aes(x = donor_type, y = pos_pvals, label = significance), 
                        inherit.aes = FALSE, vjust = -0.5, color = "black") +
              facet_wrap(~ channel, ncol = 6)
      print(p)
      
      fn <- paste0(paste(today, "Violin", cur_celltype, cur_assay,
                         names(plot_variables_stages[k]), sep = "_"), ".png")
      do.call(ggsave, c(list(fn, p), plotsave_param))
      suppressMessages(if (do_print) print(p))
    })
  })
})
```

## A4.5: Density plots

```{r density-casestage, message=FALSE, warning=FALSE, fig.width=15, fig.height=10}
purrr::map(assay_sel, \(cur_assay) {
  purrr::map(names(dat_ct_cases[[1]]), \(cur_celltype) {
    purrr::iwalk(plot_variables_stages, \(cur_variable, k) {
      
      cur_dat <- dat_ct_cases[[cur_assay]][[cur_celltype]]
      cur_dat <- subset(cur_dat, !is.infinite(cur_assay))
      cur_dat[, MeanExprs := mean(get(cur_assay)),
              by = c(plot_variables_stages[k], "channel")]
      
      p <- plot_density(cur_dat, x = cur_assay, y = plot_variables_stages[k], 
                        fill_by = "MeanExprs",
                        color_by = plot_variables_stages[k],
                        facet_by = "channel", ncol = 4,
                        title = paste(plot_variables_stages[k], cur_assay,
                                      cur_celltype)) +
        scale_color_manual(values = palettes[[names(plot_variables_stages[k])]])
      
      fn <- paste0(paste(today, "Density", cur_celltype, cur_assay,
                         names(plot_variables_stages[k]), sep = "_"), ".png")
      do.call(ggsave, c(list(fn, p), plotsave_param))
      suppressMessages(if (do_print) print(p))
    })
  })
})
```


# A5: Downstream analysis

## A5.1: Prepare the data

### A5.1.1: Define ICI/IDIs.

Define ICI and IDIs, which immune cells are located.
```{r ici-idi}
image_ids <- sce_imm$image_id |> unique()

## Define ICI and IDIs. DF: 1 row per image.
composition_islet <- colData(spe) |> 
  as_tibble() |> 
  filter(image_id %in% image_ids) |> 
  group_by(image_id, donor_type, case_id) |> 
  summarize(fraction_Beta = sum(cell_type == "Beta") / sum(cell_category == "islet") * 100) |> 
  ungroup() |> 
  mutate(islet_type = if_else(fraction_Beta > 0, "ICI", "IDI")) |> 
  select(-fraction_Beta)
```

### A5.1.2: Define Insulitic islets.

Define Insulitis. 
An image is defined as insulitic, if it contains >= 6 CD3+ cells in at least one islet. 

```{r label-insulitic}
insulitic_ids <- colData(sce_imm) |> 
  as_tibble() |> 
  filter(donor_type != "Pancreatitis") |>
  select(image_id, cell_type, donor_type, islet_id, islet_closest, distance_to_islet) |>
  filter(image_id %in% image_ids) |> 
  filter(cell_type %in% c("T_DN", "T_CD8", "T_CD4")) |> 
  filter(distance_to_islet >= -5) |>
  dplyr::count(donor_type, image_id, islet_closest) |> 
  arrange(desc(n)) |> 
  filter(n >= 6) |> 
  distinct(image_id) |> 
  pull(image_id)

message("Number of insulitic ROIs: ", length(insulitic_ids))
saveRDS(insulitic_ids, file.path(paths$folder_script, "insulitic_images.rds"))
```

Add Composition Data **(cell Types)** to Data Frame. Each row one Image.

```{r add-composition-cts}
# Immune Cell Type Image Composition.
image_compo_ct_wide <- metadata(spe)$celltype_compo |> as_tibble()

## Join composition data with metadata. 
cols_islet <- image_compo_ct_wide |> 
  inner_join(composition_islet, by = "image_id") |>
  tidyr::drop_na() |> 
  mutate(donor_type = factor(donor_type, levels = metadata(spe)$stages),
         case_id = factor(case_id, levels = metadata(spe)$cases),
         islet_type = factor(islet_type, levels = c("ICI", "IDI"))) |> 
  select(-starts_with("fraction"), -starts_with("CellNb")) |> 
  rename_with(~ stringr::str_remove(., "density_")) |> 
  filter(donor_type != "Pancreatitis") |> 
  mutate(donor_type = factor(donor_type, levels = c("NoDiabetes", "sAAb+", "mAAb+", "RecentOnset", "LongDuration")))
```

Equal procedure for the **cluster** data. Join composition with metadata.
Own Dataframe for Cluster Data. Each row one Image.

```{r cluster_data-ici}
image_compo_clust_wide <- metadata(spe)$cluster_compo$cell_cluster_scaled |> as_tibble()
cols_islet_cluster <- image_compo_clust_wide |>
  as_tibble() |> 
  filter(image_id %in% image_ids) |> 
  inner_join(composition_islet, by = "image_id") |>
  tidyr::drop_na() |> 
  mutate(donor_type = factor(donor_type, levels = metadata(spe)$stages),
         case_id = factor(case_id, levels = metadata(spe)$cases),
         islet_type = factor(islet_type, levels = c("ICI", "IDI"))) |> 
  select(-starts_with("fraction"), -starts_with("CellNb"))

rm(composition_islet); gc()
```


## A5.2: Expression Values & DistScore

```{r get-full-exprs}
# Define channels of interest.
channels <- metadata(spe)$channels
channels_imm <- unique(unlist(channels))

## Prepare the data (average expression per Cell Type) per Image.
exprs_vals_full <- 
  as_tibble(scuttle::makePerCellDF(spe, features = channels_imm, 
                                  assay.type = "exprs", use.dimred = FALSE)) |> 
  select(all_of(channels_imm), case_id, donor_type, image_id, cell_type, cell_cluster_scaled) |>
  # average expression per cell type, case, donor type, image.
  group_by(image_id, case_id, donor_type, cell_type) |> 
  summarize(across(all_of(channels_imm), \(x) mean(x, na.rm = TRUE)), .groups= "drop")

## Add Distscore
dist_score <- as_tibble(metadata(spe)$celltype_compo_score) |> 
  select(image_id, donor_type, case_id, starts_with("NormScoreArea")) |> 
  pivot_longer(
    cols = starts_with("NormScoreArea"),
    names_to = "cell_type",
    values_to = "NormScoreArea"
  ) |> 
  mutate(cell_type = stringr::str_remove(cell_type, "NormScoreArea_")) |> 
  ungroup()

## Left-joint, to keep expression values for all images.
exprs_vals_full <- exprs_vals_full |> 
  left_join(dist_score, by = c("image_id", "donor_type", "case_id", "cell_type")) |> 
  mutate(donor_type = factor(donor_type, levels = metadata(spe)$stages)) |> 
  arrange(case_id, donor_type, image_id)
gc()
```

# A6: MHC-II analysis
Check expression of MHC-II in cell types and islets.

## A6.1: Prepare the data

Here-we still have one with grouped data per case.
```{r dat-case}
dat_case <- vector(mode = "list", length = length(assay_sel))
names(dat_case) <- assay_sel

for (i in seq_along(assay_sel)) {
  dat_case[[i]] <- scuttle::makePerCellDF(
    sce_ct_case,
    exprs_values = assay_sel[i],
    features = channels_imm,
    use.coldata = c("cell_type", plot_variables, "image_id", "distance_to_islet"),
    use_dimred = FALSE) |>
    mutate(case_id = factor(case_id, levels = metadata(spe)$cases),
           donor_type = factor(donor_type, levels = metadata(spe)$stages),
           cell_type = factor(cell_type, levels = celltypes)) |>
    arrange(cell_type, case_id, donor_type) |> 
    as.data.table()
  
  dat_case[[i]] <- melt.data.table(
    dat_case[[i]],
    id.vars = c("cell_type", plot_variables, "image_id", "distance_to_islet"),
    measure.vars = channels_imm,
    variable.name = "channel",
    value.name = assay_sel[i]
  )
  dat_case[[i]][, case_id := as.factor(case_id)]
}
```

```{r islet-case-dat}
dat <- vector(mode = "list", length = length(assay_sel))
names(dat) <- assay_sel

for (i in seq_along(assay_sel)) {
  dat[[i]] <- scuttle::makePerCellDF(
    sce_ct_islet,
    # sce_ct_case,
    exprs_values = assay_sel[i],
    features = channels_imm,
    use.coldata = c("cell_type", plot_variables, "image_id", "distance_to_islet"),
    use_dimred = FALSE) |>
    mutate(case_id = factor(case_id, levels = metadata(spe)$cases),
           donor_type = factor(donor_type, levels = metadata(spe)$stages),
           cell_type = factor(cell_type, levels = celltypes)) |>
    arrange(cell_type, case_id, donor_type) |>
    as.data.table()
  
  dat[[i]] <- melt.data.table(
    dat[[i]],
    id.vars = c("cell_type", plot_variables, "image_id", "distance_to_islet"),
    measure.vars = channels_imm,
    variable.name = "channel",
    value.name = assay_sel[i]
  )
  dat[[i]][, case_id := as.factor(case_id)]
}
```

## A6.2: MHC-II hyperexpression in Cell Types

Akin, to the results in MHC-I we perform differential expression analysis for MHC-II.
```{r mhc-ii}
library(emmeans)
cur_assay <- "exprs"
exprs_dat <- dat[["exprs"]]

exprs_islet <- exprs_dat |> 
  filter(donor_type %in% c("NoDiabetes", "sAAb+", "mAAb+", "RecentOnset", "LongDuration")) |>
  filter(cell_type %in% c("Alpha", "Beta", "Acinar", "Ductal")) |> 
  mutate(cell_type = factor(cell_type, levels = c("Beta", "Alpha", "Acinar", "Ductal")))

icis <- colData(spe) |> 
  as_tibble() |> 
  filter(cell_type == "Beta") |> 
  distinct(image_id) |>
  pull(image_id)

to_del <- colData(spe) |> 
  as_tibble() |> 
  filter(cell_type == "Beta") |> 
  dplyr::count(case_id) |> 
  filter(n < 100) |> pull(case_id)

exprs_islet <- exprs_islet |> 
  left_join(colData(spe) |> as_tibble() |> distinct(case_id, age_group.y, ICU_time_days.y), by = "case_id") |> 
  as_tibble() |> 
  filter(image_id %in% icis) |>
  filter(!(case_id %in% to_del)) |>
  mutate(ICU_time_days.y = ifelse(is.na(ICU_time_days.y), mean(ICU_time_days.y, na.rm = TRUE), ICU_time_days.y))

library(ggpubr); library(rstatix)
p <- plot_expression_violin(exprs_islet, 
                            "HLA_DR", 
                            cell_types = c("Alpha", "Beta", "Acinar", "Ductal"),
                            facet_by = "cell_type",
                            immune_adjustment = TRUE,
                            y_label = "HLA_DR expression")
print(p)

fn <- paste0(paste(today, "Boxplot", "MHC-II", "CellTypes", "HLA_ABC", sep = "_"), ".png")
do.call(ggsave, c(list(fn, p), plotsave_param_large))

fig2b <- p
saveRDS(fig2b, file.path(paths$home_git, "figures", "fig2b.rds"))
```

## A6.3 Islet-wide upregulation?

Check Correlations between cell types.

```{r mhc-ii-islet}
## MHC-II -> Hypothesize for islet-wide upregulation.
ab_hla <- exprs_vals_full  |> 
  pivot_longer(all_of(c(channels_imm, "NormScoreArea")), names_to = "channel", values_to = "exprs") |>
  dplyr::filter(channel == "HLA_DR") |> 
  dplyr::filter(donor_type %in% c("NoDiabetes", "sAAb+", "mAAb+", "RecentOnset")) |>
  mutate(donor_type = factor(donor_type, levels = c("NoDiabetes", "sAAb+", "mAAb+", "RecentOnset"))) |>
  tidyr::pivot_wider(names_from = cell_type, values_from = exprs)

# Compute correlation between alpha and beta expression in same islet.
# 0.7 - 0.77
ab_hla |> dplyr::group_by(donor_type) |>
  dplyr::summarise(cors = cor(Beta, Alpha, use = "complete.obs", method = "spearman"))

dd_test <- ab_hla |> filter(donor_type == "RecentOnset")
lmerTest::lmer(Beta ~ Alpha + (1 | case_id), data = dd_test) |> broom.mixed::tidy() |> filter(term == "Alpha")
cor.test(dd_test$Beta, dd_test$Alpha, method = "spearman")

# Beta/Delta: 0.7/0.8
ab_hla |> dplyr::group_by(donor_type) |>
  dplyr::summarise(cors = cor(Beta, Delta, use = "complete.obs", method = "spearman"))

# Beta-Ductal: 0.3 - 0.6
ab_hla |> dplyr::group_by(donor_type) |> 
  dplyr::summarise(cors = cor(Beta, Ductal, use = "complete.obs", method = "spearman"))

# 0.4 - 0.6. Beta + Endothelial (surprisingly; but icnrease along stage).
ab_hla  |> dplyr::group_by(donor_type) |> 
  dplyr::summarise(cors = cor(Beta, Endothelial, use = "complete.obs", method = "spearman"))

p <- ab_hla |> 
  select(donor_type, case_id, image_id, channel, 
         Alpha, Beta, Delta, Acinar, Ductal, Endothelial) |> 
  pivot_longer(c("Alpha", "Delta", "Acinar", "Ductal"), names_to = "cell_type", values_to = "exprs") |> 
  mutate(cell_type = factor(cell_type, levels = c("Alpha", "Delta", "Acinar", "Ductal"))) |>
  ggplot(aes(x = Beta, y = exprs, color = donor_type)) +
  geom_point(size = 0.5, alpha = 0.5) +
  geom_smooth(method = "lm") +
  theme(axis.text.x = element_text(angle = 90, hjust = 1)) +
  labs(x = expression(beta ~ "-cell HLA_DR expression"), 
       y = "HLA-DR expression", color = "Stage") +
  mytheme$large_new() +
  theme(legend.position = "none") + 
  facet_wrap(~ cell_type, ncol = 2, scales = "free_y") + 
  stat_cor(method = "spearman", aes(color = donor_type, label = ..r.label..),
           label.x = 0, size = 12, show.legend = FALSE) + 
  scale_color_manual(values = palettes[["stages"]]) + 
  theme(legend.position = "top",                 # Move legend above the plot
        legend.justification = "center",         # Center the legend
        legend.box = "horizontal") 

print(p)
# Define 2x2 grid
fn <- paste0(paste(today, "Correlation_Islet_HLA_DR", sep = "_"), ".png")
do.call(ggsave, c(list(fn, p), plotsave_param_large))

extfig4b <- p
saveRDS(extfig4b, file.path(paths$home_git, "figures", "extfig4b.rds"))

p <- ab_hla |> 
  select(donor_type, case_id, image_id, channel, 
         Alpha, Beta, Delta, Acinar, Ductal, Endothelial) |> 
  pivot_longer(c("Alpha", "Delta", "Acinar", "Ductal"), names_to = "cell_type", values_to = "exprs") |> 
  mutate(cell_type = factor(cell_type, levels = c("Alpha", "Delta", "Acinar", "Ductal"))) |>
  group_by(donor_type, case_id, cell_type) |> 
  summarise(Beta = mean(Beta, na.rm = TRUE), exprs = mean(exprs, na.rm = TRUE)) |> 
  ungroup() |>
  ggplot(aes(x = Beta, y = exprs, color = donor_type)) +
  geom_point(size = 3, alpha = 0.5) +
  geom_smooth(method = "lm") +
  theme(axis.text.x = element_text(angle = 90, hjust = 1)) +
  labs(x = expression(beta ~ "-cell HLA_DR expression"), 
       y = "HLA-DR expression", color = "Stage") +
  mytheme$large_new() +
  theme(legend.position = "none") + 
  facet_wrap(~ cell_type, ncol = 2, scales = "free_y") + 
  stat_cor(method = "spearman", aes(color = donor_type, label = ..r.label..),
           label.x = 0, size = 12, show.legend = FALSE) + 
  scale_color_manual(values = palettes[["stages"]]) + 
  theme(legend.position = "top",                 # Move legend above the plot
        legend.justification = "center",         # Center the legend
        legend.box = "horizontal") 

print(p)
# Define 2x2 grid
fn <- paste0(paste(today, "Correlation_Islet_HLA_DR_DONOR", sep = "_"), ".png")
do.call(ggsave, c(list(fn, p), plotsave_param_large))
```

sd vs mean. Potentially HLA-DR less uniform than HLA-ABC?
-> Check this.
```{r}
exprs_dat  |> 
  filter(cell_type %in% c("Beta"), channel == "HLA_DR") |> 
  group_by(cell_type, donor_type) |> 
  summarise(cv = sd(exprs) / mean(exprs), mm = mean(exprs))
```

## A6.4: Statistics MHC-II upregulation

### Correlation MHC-II Beta-cells vs Immune Cells Infiltration DistScore

```{r mhc-ii-infiltrationscore}
hla_distscore <- exprs_vals_full |> filter(cell_type %in% c("Beta", "Myeloid", "T_CD4", "T_CD8", "Neutrophil")) |> 
  tidyr::pivot_wider(names_from = c(cell_type), values_from = where(is.numeric)) |> 
  dplyr::select(image_id, case_id, donor_type, starts_with("NormScoreArea"), starts_with("HLA_DR")) 

data_tt <- hla_distscore  |> 
  pivot_longer(starts_with("NormScoreArea"), names_to = "cell_type", values_to = "NormScoreArea") |>
  pivot_longer(starts_with("HLA_DR"), names_to = "channel", values_to = "HLA_DR_Beta") |> 
  filter(channel == "HLA_DR_Beta") |> 
  filter(!cell_type %in% c("NormScoreArea_Beta")) |> 
  dplyr::filter(donor_type %in% c("NoDiabetes", "sAAb+", "mAAb+", "RecentOnset")) |> 
  mutate(cell_type = stringr::str_remove(cell_type, "NormScoreArea_")) |>
  mutate(donor_type = factor(donor_type, levels = c("NoDiabetes", "sAAb+", "mAAb+", "RecentOnset")))

p <- data_tt |> 
  group_by(cell_type, donor_type, case_id) |>
  summarise(HLA_DR_Beta = mean(HLA_DR_Beta, na.rm = TRUE),
            NormScoreArea = mean(NormScoreArea, na.rm = TRUE)) |>
  ungroup() |>
  ggplot(aes(x = HLA_DR_Beta, y = log1p(NormScoreArea), color = donor_type)) +
  geom_point(size = 3) +
  scale_color_manual(values = palettes[["stages"]], breaks = c("NoDiabetes", "sAAb+", "mAAb+", "RecentOnset")) + 
  geom_smooth(method = "lm", alpha = 0.3) +
  theme(axis.text.x = element_text(angle = 90, hjust = 1)) +
  labs(x = expression(beta ~ "-cell HLA-DR expression"), y = "log: Infiltration Score", color = "Stage") +
  mytheme$large_new() + 
  facet_wrap(~ cell_type, scales = "free_y", axes = "all") + 
  theme(legend.position = "none") + 
  stat_cor(method = "spearman", aes(color = donor_type, label = ..r.label..),
           label.x = 0.2, label.y.npc = "top", size = 12)

print(p)
fn <- paste0(paste(today, "Correlation_HLA_DR_DistScore_DONOR", sep = "_"), ".png")
do.call(ggsave, c(list(fn, p), plotsave_param_large))

## By Islet.
p <- data_tt |> 
  ggplot(aes(x = HLA_DR_Beta, y = log1p(NormScoreArea), color = donor_type)) +
  geom_point(size = 1) +
  scale_color_manual(values = palettes[["stages"]], breaks = c("NoDiabetes", "sAAb+", "mAAb+", "RecentOnset")) + 
  geom_smooth(method = "lm", alpha = 0.3) +
  theme(axis.text.x = element_text(angle = 90, hjust = 1)) +
  labs(x = expression(beta ~ "-cell HLA-DR expression"), y = "log: Infiltration Score", color = "Stage") +
  mytheme$large_new() + 
  facet_wrap(~ cell_type, scales = "free_y", axes = "all") + 
  theme(legend.position = "none") + 
  stat_cor(method = "spearman", aes(color = donor_type, label = ..r.label..),
           label.x = 2, r.accuracy = 0.01, label.y.npc = "bottom", size = 12)
print(p)
fn <- paste0(paste(today, "Correlation_HLA_DR_DistScore", sep = "_"), ".png")
do.call(ggsave, c(list(fn, p), plotsave_param_large))

extfig4d <- p
saveRDS(extfig4d, file.path(paths$home_git, "figures", "extfig4d.rds"))

## Statistics:
## Strong significance. 
summary(lmerTest::lmer(data = hla_distscore |> filter(donor_type == "RecentOnset"), HLA_DR_Beta ~ NormScoreArea_Neutrophil + (1 | case_id)))
summary(lmerTest::lmer(data = hla_distscore |> filter(donor_type == "RecentOnset"), HLA_DR_Beta ~ NormScoreArea_Myeloid + (1 | case_id)))
summary(lmerTest::lmer(data = hla_distscore |> filter(donor_type == "RecentOnset"), HLA_DR_Beta ~ NormScoreArea_T_CD8+ (1 | case_id)))
summary(lmerTest::lmer(data = hla_distscore |> filter(donor_type == "RecentOnset"), HLA_DR_Beta ~ NormScoreArea_T_CD4 + (1 | case_id)))

## Pictures: 
hla_distscore |> slice_max(HLA_DR_Beta, n = 5) |> select(image_id)
```

## A6.5: Cells neighboring MHC-II+ and MHC-II- beta-cells.

Step 1: Define MHC-II+ and MHC-II- Beta-cells (Otsu-threshold.) Only for mAAb+/RecentOnsets.
Step 2: Aggregate the neighboring cells of MHC-II+ and MHC-II- Beta-cells.
Step 3: Compare the neighboring cells of MHC-II+ and MHC-II- Beta-cells.

### A6.5.1: Define MHC-II+ and MHC-II- Beta-cells
Step 1: Define MHC-II+ and MHC-II- Beta-cells (Otsu-threshold.) Only for mAAb+/RecentOnsets.

```{r}
spe_sub <- spe[, spe$donor_type %in% c("mAAb+", "RecentOnset")]
exprs_hla <- makePerCellDF(spe_sub, features = "HLA_DR", 
                          use.dimred = FALSE, assay = "exprs",
                          use.coldata = c("cell_type", "donor_type", "case_id", "image_id", "cell_id")) |>
  as_tibble() |> 
  filter(cell_type == "Beta")

beta_cells <- exprs_hla |> 
  select(image_id, case_id, donor_type, cell_type, HLA_DR) |> 
  mutate(HLA_DR = as.numeric(HLA_DR)) |> 
  arrange(image_id, case_id, donor_type)

## Calculate Otsu threshold for HLA_DR in Beta-cells.
beta_cells_otsu <- pliman::otsu(beta_cells$HLA_DR)
message("Otsu Threshold for HLA_DR in Beta-cells: ", beta_cells_otsu)

p <- exprs_hla |> 
  ggplot(aes(x = HLA_DR, fill = donor_type)) +
  geom_histogram(bins = 100) + 
  geom_vline(xintercept = beta_cells_otsu, linetype = "dashed", color = "red") +
  scale_fill_manual(values = palettes[["stages"]]) +
  labs(x = "HLA_DR expression", y = "Number of cells", fill = "Stage") +
  mytheme$standard()

print(p)
fn <- paste0(paste(today, "HLA_DR_Histogram", sep = "_"), ".png")
do.call(ggsave, c(list(fn, p), plotsave_param_large))
```

```{r hla-dr-beta-cells}
## HLA-DR+ beta-cells.
hla_pos_beta <- exprs_hla |> 
  mutate(hla_pos = ifelse(HLA_DR >= beta_cells_otsu, "HLA_DR+", "HLA_DR-")) |> 
  filter(hla_pos == "HLA_DR+") |>
  pull(cell_id) |> unique()

spe_sub$hla_pos <- ifelse(spe_sub$cell_id %in% hla_pos_beta, "HLA_DR+", "HLA_DR-")
```

### A6.5.2: Aggregate neighboring cells of MHC-II+ and MHC-II- Beta-cells
Step 2: Aggregate the neighboring cells of MHC-II+ and MHC-II- Beta-cells.

```{r aggregate-neighbors-hla}
## Aggregate the neighboring cells of MHC-II+ and MHC-II- Beta-cells.
tt <- imcRtools::aggregateNeighbors(
  spe_sub,
  colPairName = "neighborhood",
  aggregate_by = "metadata",
  count_by = "cell_type",
  proportions = FALSE,
  assay_type = NULL,
  subset_row = NULL,
  statistic = c("mean", "median", "sd", "var"),
  name = NULL
)

## When aggregating across all images, see that enriched next to practically all immune cells.
input_df <- colData(tt)[, c("hla_pos", "cell_type", "islet_id", "image_id", "case_id", "donor_type", "aggregatedNeighbors", "cell_category")] |> 
  as_tibble() |> 
  dplyr::rename_with(~ stringr::str_remove(., "aggregatedNeighbors.")) |>
  filter(cell_type == "Beta")

## Select only Islets with >= 3 HLA_DR+ beta-cells.
hla_pos_islets <- input_df  |> 
  filter(hla_pos == "HLA_DR+") |> 
  dplyr::count(islet_id)  |> filter(n >= 3) |> pull(islet_id) |> unique() 
```

### A6.5.3: Plot: Compare the neighboring cells of MHC-II+ and MHC-II- Beta-cells
Step 3: Compare the neighboring cells of MHC-II+ and MHC-II- Beta-cells.
```{r plot-neighbors-hla}
## Plot these results.
plot_df <- input_df  |> 
  filter(islet_id %in% hla_pos_islets) |> 
  group_by(hla_pos, image_id, case_id) |> 
  summarise(across(where(is.numeric), ~ mean(.x, na.rm = TRUE)))  |> 
  pivot_longer(
    cols = where(is.numeric),
    names_to = "cell_type",
    values_to = "density"
  ) |> 
  filter(cell_type %in% c("Alpha", "Beta", "Delta", "IsletOther", 
                          "Acinar", "Ductal", "Endothelial", "T_CD8", 
                          "T_CD4", "T_DN", "Myeloid", "NK", "B", "Neutrophil")) |> 
  ungroup() |> 
  mutate(cell_category = case_when(cell_type %in% c("Beta", "Alpha", "Delta", "IsletOther") ~ "islet",
            cell_type %in% c("Endothelial", "Acinar", "Ductal") ~ "exo_stroma",
            cell_type %in% c("T_CD8", "T_CD4", "T_DN", "Myeloid", "B", "Neutrophil", "NK") ~ "immune",
            TRUE ~ cell_type))
```

### A6.5.4: Volcano Plot: HLA-DR+ vs HLA-DR- Beta-cells

```{r volcano-hla-dr-beta}
p_vals <- plot_df |> 
  tidyr::nest(data = -cell_type) |> 
  dplyr::mutate(test = purrr::map(data, ~ broom.mixed::tidy(lmerTest::lmer(data = ., 
          density ~ hla_pos + (1 | case_id))))) |> 
  unnest(test) |> 
  filter(term == "hla_posHLA_DR+") |> 
  mutate(p.adj = p.adjust(p.value, method = "fdr")) |> 
  mutate(p.signif = cut(p.adj, 
          breaks = c(-Inf, 0.001, 0.01, 0.05, 0.1, Inf), 
          labels = c("***", "**", "*",   ".", "ns")))

library('EnhancedVolcano')

input <- plot_df |> 
  group_by(cell_type, hla_pos) |> 
  summarise(density = mean(density, na.rm = TRUE)) |> 
  pivot_wider(names_from = hla_pos, values_from = density) |> 
  mutate(log2FC = log2(`HLA_DR+` / `HLA_DR-`)) |> 
  arrange(desc(log2FC)) |> 
  mutate(direction = ifelse(log2FC > 0, "Enriched", "Depleted")) |>
  inner_join(p_vals, by = "cell_type") |> 
  mutate(p.adj = ifelse(p.adj < 1e-26, 1e-26, p.adj))

## volcano plot
p <-  EnhancedVolcano(input,
    lab = input$cell_type,
    title = NULL,
    subtitle = NULL,
    caption = NULL,
    x = 'log2FC',
    y = 'p.adj', 
    xlab = "log2 Fold Change",
    ylab = "-log10(padj)",
    axisLabSize = 8,
    xlim = c(-0.5, 2), 
    ylim = c(0, 28),
    selectLab = input$cell_type,  # force all labels
    FCcutoff = 0.2,
    pCutoff = 0.05,
    pointSize = 20,
    boxedLabels = TRUE,
    labSize = 10.0,
    drawConnectors = TRUE, 
    legendPosition = "top") + 
    mytheme$large_new() + 
    ylab("-log10(p.adj)") + 
    theme(axis.text.x = element_text(size=30)) + 
    theme(axis.text.y = element_text(size=30))

print(p)
fn <- paste0(paste(today, "HLA_DR_CellType_Volcano", sep = "_"), ".png")
do.call(ggsave, c(list(fn, p), plotsave_param_large))

revfigvolcano <- p
saveRDS(p, file.path(paths$home_git, "figures", "rev_supplfig6b.rds"))
```

```{r}
library(rstatix); library(ggpubr)
max_vals <- plot_df |> 
  filter(cell_category %in% c("exo_stroma", "immune", "islet")) |>
  group_by(cell_type, cell_category, case_id, hla_pos) |> 
  summarise(density = mean(density, na.rm = TRUE))  |> 
  group_by(cell_type) |>
  summarise(max_val = max(density, na.rm = TRUE))

stat_test <- plot_df |> 
  mutate(cell_type = factor(cell_type, levels = c("Alpha", "Beta", "Delta", "IsletOther", 
                                                  "Acinar", "Ductal", "Endothelial", 
                                                  "T_CD8", "T_CD4", "T_DN", "Myeloid", 
                                                  "NK", "B", "Neutrophil"))) |>
  group_by(cell_type) |> 
  rstatix::t_test(density ~ hla_pos) |> 
  add_xy_position(x = "cell_type") |> 
  inner_join(p_vals, by = "cell_type") |> 
  select(cell_type, .y., group1, group2, y.position, groups, x, xmin, xmax, p.signif) |> 
  inner_join(max_vals, by = "cell_type") |> 
  mutate(y.position = log1p(max_val) * 1.1)

p3 <- plot_df |> 
  filter(cell_category %in% c("exo_stroma", "immune", "islet")) |>
  group_by(cell_type, cell_category, case_id, hla_pos) |> 
  summarise(density = mean(density, na.rm = TRUE)) |>
  mutate(cell_type = factor(cell_type, levels = c("Alpha", "Beta", "Delta", "IsletOther", 
                                                  "Acinar", "Ductal", "Endothelial", 
                                                  "T_CD8", "T_CD4", "T_DN", "Myeloid", 
                                                  "NK", "B", "Neutrophil"))) |>
  ggplot(aes(x = cell_type, y = log1p(density), fill = hla_pos)) + 
  geom_boxplot(outlier.shape = NA) +
  geom_point(position = position_dodge(width = 0.75)) +
  scale_fill_manual(values = c("HLA_DR+" = "orange", "HLA_DR-" = "red")) +
  labs(x = "Cell type", y = expression("log1p: # of average " * beta * " -cell neighbors"), fill = "HLA-DR status") +
  mytheme$large_new() + 
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) + 
  theme(legend.position = "top")  + 
  stat_pvalue_manual(stat_test, label = "p.signif", tip.length = 0.01, size = 8)

print(p3)

fn <- paste0(paste(today, "HLA_DR_CellType_Density", sep = "_"), ".png")
do.call(ggsave, c(list(fn, p3), plotsave_param_large))

revfigboxplot <- p3
saveRDS(revfigboxplot, file.path(paths$home_git, "figures", "rev_supplfig6a.rds"))
```


### A6.6: Volcano Plot: HLA-DR+ vs HLA-DR- Exocrine Cells.

### A6.6.1: Define MHC-II+ and MHC-II- Beta-cells
Step 1: Define MHC-II+ and MHC-II- Beta-cells (Otsu-threshold.) Only for mAAb+/RecentOnsets.

```{r}
spe_sub <- spe[, spe$donor_type %in% c("mAAb+", "RecentOnset")]
exprs_hla <- makePerCellDF(spe_sub, features = "HLA_DR", 
                          use.dimred = FALSE, assay = "exprs",
                          use.coldata = c("cell_type", "donor_type", "case_id", "image_id", "cell_id")) |>
  as_tibble() |> 
  filter(cell_type %in% c("Acinar", "Ductal"))

exo_cells <- exprs_hla |> 
  select(image_id, case_id, donor_type, cell_type, HLA_DR) |> 
  mutate(HLA_DR = as.numeric(HLA_DR)) |> 
  arrange(image_id, case_id, donor_type)

## Calculate Otsu threshold for HLA_DR in Exocrine cells.
exo_cells_otsu <- pliman::otsu(exo_cells$HLA_DR)
message("Otsu Threshold for HLA_DR in Exocrine cells: ", exo_cells_otsu)

# Choose a more stringent threshold for exocrine cells.
# This is the beta-cells threshold; it is larger than the computed exocrine threshold.
p <- exprs_hla |> 
  ggplot(aes(x = HLA_DR, fill = donor_type)) +
  geom_histogram(bins = 100) + 
  geom_vline(xintercept = beta_cells_otsu, linetype = "dashed", color = "red") +
  scale_fill_manual(values = palettes[["stages"]]) +
  labs(x = "HLA_DR expression", y = "Number of cells", fill = "Stage") +
  mytheme$standard()

print(p)
fn <- paste0(paste(today, "HLA_DR_Histogram_Exo", sep = "_"), ".png")
do.call(ggsave, c(list(fn, p), plotsave_param_large))
```

```{r hla-dr-beta-cells}
## HLA-DR+ exocrine cells.
hla_pos_exo <- exprs_hla |> 
  mutate(hla_pos = ifelse(HLA_DR >= beta_cells_otsu, "HLA_DR+", "HLA_DR-")) |> 
  filter(hla_pos == "HLA_DR+") |>
  pull(cell_id) |> unique()

spe_sub$hla_pos <- ifelse(spe_sub$cell_id %in% hla_pos_exo, "HLA_DR+", "HLA_DR-")
```

### A6.5.2: Aggregate neighboring cells of MHC-II+ and MHC-II- Exocrine cells
Step 2: Aggregate the neighboring cells of MHC-II+ and MHC-II- Exocrine cells.

```{r aggregate-neighbors-hla}
## Aggregate the neighboring cells of MHC-II+ and MHC-II- Exocrine cells.
tt <- imcRtools::aggregateNeighbors(
  spe_sub,
  colPairName = "neighborhood",
  aggregate_by = "metadata",
  count_by = "cell_type",
  proportions = FALSE,
  assay_type = NULL,
  subset_row = NULL,
  statistic = c("mean", "median", "sd", "var"),
  name = NULL
)

## When aggregating across all images, see that enriched next to practically all immune cells.
input_df <- colData(tt)[, c("hla_pos", "cell_type", "islet_id", "image_id", "case_id", "donor_type", "aggregatedNeighbors", "cell_category")] |> 
  as_tibble() |> 
  dplyr::rename_with(~ stringr::str_remove(., "aggregatedNeighbors.")) |>
  filter(cell_type %in% c("Acinar", "Ductal"))

## Select only Islets with >= 3 HLA_DR+ exo-cells.
hla_pos_rois <- input_df  |> 
  filter(hla_pos == "HLA_DR+") |> 
  dplyr::count(image_id)  |> filter(n >= 3) |> pull(image_id) |> unique() 
```


### A6.5.3: Plot: Compare the neighboring cells of MHC-II+ and MHC-II- Beta-cells
Step 3: Compare the neighboring cells of MHC-II+ and MHC-II- Beta-cells.
```{r plot-neighbors-hla}
## Plot these results.
plot_df <- input_df  |> 
  filter(image_id %in% hla_pos_rois) |> 
  group_by(hla_pos, image_id, case_id) |> 
  summarise(across(where(is.numeric), ~ mean(.x, na.rm = TRUE)))  |> 
  pivot_longer(
    cols = where(is.numeric),
    names_to = "cell_type",
    values_to = "density"
  ) |> 
  filter(cell_type %in% c("Alpha", "Beta", "Delta", "IsletOther", 
                          "Acinar", "Ductal", "Endothelial", "T_CD8", 
                          "T_CD4", "T_DN", "Myeloid", "NK", "B", "Neutrophil")) |> 
  ungroup() |> 
  mutate(cell_category = case_when(cell_type %in% c("Beta", "Alpha", "Delta", "IsletOther") ~ "islet",
            cell_type %in% c("Endothelial", "Acinar", "Ductal") ~ "exo_stroma",
            cell_type %in% c("T_CD8", "T_CD4", "T_DN", "Myeloid", "B", "Neutrophil", "NK") ~ "immune",
            TRUE ~ cell_type))

case_ids <- plot_df |> distinct(hla_pos, image_id, case_id) |> dplyr::count(hla_pos, case_id) |> 
  pivot_wider(names_from = hla_pos, values_from = n, values_fill = 0) |> 
  filter(`HLA_DR+` >= 3 & `HLA_DR-` >= 3) |> 
  pull(case_id)
```            

```{r volcano-hla-dr-ductal}
p_vals <- plot_df |> 
  filter(case_id %in% case_ids) |>
  tidyr::nest(data = -cell_type) |> 
  dplyr::mutate(test = purrr::map(data, ~ broom.mixed::tidy(lmerTest::lmer(data = ., 
          density ~ hla_pos + (1 | case_id))))) |> 
  unnest(test) |> 
  filter(term == "hla_posHLA_DR+") |> 
  mutate(p.adj = p.adjust(p.value, method = "fdr")) |> 
  mutate(p.signif = cut(p.adj, 
          breaks = c(-Inf, 0.001, 0.01, 0.05, 0.1, Inf), 
          labels = c("***", "**", "*",   ".", "ns")))
p_vals
library('EnhancedVolcano')

input <- plot_df |> 
  group_by(cell_type, hla_pos) |> 
  summarise(density = mean(density, na.rm = TRUE)) |> 
  pivot_wider(names_from = hla_pos, values_from = density) |> 
  mutate(log2FC = log2(`HLA_DR+` / `HLA_DR-`)) |> 
  arrange(desc(log2FC)) |> 
  mutate(direction = ifelse(log2FC > 0, "Enriched", "Depleted")) |>
  inner_join(p_vals, by = "cell_type") |>
  mutate(p.adj = ifelse(p.adj < 1e-50, 1e-50, p.adj))

input

## volcano plot
p <-  EnhancedVolcano(input,
    lab = input$cell_type,
    title = "HLA-DR+ vs HLA-DR- Exocrine-cells",
    subtitle = NULL,
    caption = NULL,
    x = 'log2FC',
    y = 'p.adj', 
    xlim = c(-0.5, 2), 
    ylim = c(0, 52),
    pCutoff = 0.05,
    selectLab = input$cell_type,  # force all labels
    FCcutoff = 0.2,
    pointSize = 20,
    boxedLabels = TRUE,
    labSize = 10.0,
    drawConnectors = TRUE) + 
    mytheme$large_new() + 
    ylab("-log10(p.adj)") + 
    theme(axis.text.x = element_text(size=30)) + 
    theme(axis.text.y = element_text(size=30))

print(p)
fn <- paste0(paste(today, "HLA_DR_CellType_Volcano_Exocrine", sep = "_"), ".png")
do.call(ggsave, c(list(fn, p), plotsave_param_large))

revfigvolcano_exo <- p
saveRDS(revfigvolcano_exo, file.path(paths$home_git, "figures", "revfigvolcano_exo.rds"))
```


```{r}
library(rstatix); library(ggpubr)
max_vals <- plot_df |> 
  filter(case_id %in% case_ids) |>
  filter(cell_category %in% c("exo_stroma", "immune", "islet")) |>
  group_by(cell_type, cell_category, case_id, hla_pos) |> 
  summarise(density = mean(density, na.rm = TRUE), .groups = "drop")  |> 
  group_by(cell_type) |>
  summarise(max_val = max(density, na.rm = TRUE))

stat_test <- plot_df |> 
  filter(case_id %in% case_ids) |>
  mutate(cell_type = factor(cell_type, levels = c("Alpha", "Beta", "Delta", "IsletOther", 
                                                  "Acinar", "Ductal", "Endothelial", 
                                                  "T_CD8", "T_CD4", "T_DN", "Myeloid", 
                                                  "NK", "B", "Neutrophil"))) |>
  group_by(cell_type) |> 
  rstatix::t_test(density ~ hla_pos) |> 
  add_xy_position(x = "cell_type") |> 
  inner_join(p_vals, by = "cell_type") |> 
  select(cell_type, .y., group1, group2, y.position, groups, x, xmin, xmax, p.signif) |> 
  inner_join(max_vals, by = "cell_type") |> 
  mutate(y.position = log1p(max_val) * 1.1)

p3 <- plot_df |> 
  filter(case_id %in% case_ids) |>
  filter(cell_category %in% c("exo_stroma", "immune", "islet")) |>
  group_by(cell_type, cell_category, case_id, hla_pos) |> 
  summarise(density = mean(density, na.rm = TRUE)) |>
  mutate(cell_type = factor(cell_type, levels = c("Alpha", "Beta", "Delta", "IsletOther", 
                                                  "Acinar", "Ductal", "Endothelial", 
                                                  "T_CD8", "T_CD4", "T_DN", "Myeloid", 
                                                  "NK", "B", "Neutrophil"))) |>
  ggplot(aes(x = cell_type, y = log1p(density), fill = hla_pos)) + 
  geom_boxplot(outlier.shape = NA) +
  geom_point(position = position_dodge(width = 0.75)) +
  scale_fill_manual(values = c("HLA_DR+" = "orange", "HLA_DR-" = "red")) +
  labs(x = "Cell type", y = "log1p: # of average exocrine cell neighbors", fill = "HLA-DR status") +
  mytheme$large_new() + 
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) + 
  theme(legend.position = "top")  + 
  stat_pvalue_manual(stat_test, label = "p.signif", tip.length = 0.01, size = 8)

print(p3)

fn <- paste0(paste(today, "HLA_DR_CellType_Density_Exocrine", sep = "_"), ".png")
do.call(ggsave, c(list(fn, p3), plotsave_param_large))

saveRDS(p3, file.path(paths$home_git, "figures", "rev_supplfig6c.rds"))
```

# **A7: ICI vs IDI:**

## **A7.1: Number of ICIs per Stage**
```{r number-of-icis}
p <- cols_islet  |> 
  dplyr::count(islet_type, donor_type) |> 
  dplyr::group_by(donor_type) |>
  dplyr::mutate(total = sum(n),
                frac = n / total) |> 
  # Barplot:
  ggplot(aes(x = islet_type, y = frac, fill = donor_type)) +
  geom_bar(stat = "identity", position = "dodge") +
  scale_fill_manual(values = palettes[["stages"]]) +
  labs(x = "Islet type", y = "Fraction of islets", fill = "Stage") +
  mytheme$standard()

fn <- paste0(paste(today, "IsletType", sep = "_"), ".png")
do.call(ggsave, c(list(fn, p), plotsave_param))
print(p)
```

Number of Insulitic Images per Stage. 
Separated by ICI vs IDI. 
```{r plot-insulitic-images}
## Plot Barplot for Insulitic vs Non-Insulitic.
p <- cols_islet |> 
  mutate(insulitic = ifelse(image_id %in% insulitic_ids, "Insulitic", "Non-Insulitic")) |>
  dplyr::count(donor_type, islet_type, insulitic) |>
  filter(insulitic == "Insulitic") |>
  ggplot(aes(x = donor_type, y = n, fill = donor_type)) +
  geom_bar(stat = "identity", position = "dodge", 
           aes(alpha = islet_type), show.legend = c(color = TRUE, fill = FALSE)) + 
  scale_alpha_manual(values = c("ICI" = 1, "IDI" = 0.5), name = "Islet type")  +
  scale_fill_manual(values = palettes$stages, name = "Stage") +
  labs(x = "Stage", y = "Number of insulitic images", fill = "Insulitic") +
  mytheme$standard_new() + 
  theme(axis.text.x = element_text(angle = 45, hjust = 1, size = 22))

print(p)
fn <- paste0(paste(today, "Insulitic", "IsletType", sep = "_"), ".png")
do.call(ggsave, c(list(fn, p), plotsave_param))
```

## A7.2: Compare the Immune cell compositions for ICI vs IDI

Compare Cell Type Composition between ICI vs IDIs.
No major differences.
```{r ici-idi-composition-ct}
## Compare the Immune cell compositions for ICI vs IDI
ici_idi <- cols_islet |> 
  select(where(is.numeric), islet_type, case_id, donor_type, image_id) |>
  tidyr::pivot_longer(cols = where(is.numeric), names_to = "cell_type", values_to = "density")

## Grouped barplots
p <- ici_idi |> 
  filter(donor_type %in% c("RecentOnset")) |> 
  group_by(case_id, donor_type, cell_type, islet_type) |>
  summarize(density = mean(density))  |> 
  ggplot(aes(x = cell_type, y = log1p(density), fill = islet_type)) +
  geom_boxplot(outlier.shape = NA) +
  scale_fill_manual(values = c("ICI" = "orange", "IDI" = "red")) +
  labs(x = "Cell type", y = "Density - RecenOnset", fill = "Islet type") +
  mytheme$standard() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

print(p)
fn <- paste0(paste(today, "ImmuneCellComposition_ici_idi", sep = "_"), ".png")
do.call(ggsave, c(list(fn, p), plotsave_param_large))

## Paired plot for Myeloid:
## Average per Donor.
p <- ici_idi |> 
  filter(donor_type %in% c("mAAb+", "RecentOnset")) |> 
  filter(cell_type == "Myeloid") |> 
  # Group by Donor + CT.
  group_by(case_id, donor_type, cell_type, islet_type) |>
  summarize(density = mean(density))  |> 
  ggplot(aes(x = islet_type, y = density, fill = islet_type)) +
  geom_boxplot(outlier.shape = NA) +
  geom_line(aes(group = case_id, color = as.factor(case_id)), 
                position = position_dodge(width = 0.9)) +
  geom_point(aes(group = case_id, color = as.factor(case_id)), 
             position = position_dodge(width = 0.9)) +
  scale_fill_manual(values = c("ICI" = "blue", "IDI" = "red")) +
  labs(x = "Islet type", y = "Fraction of cells", fill = "Islet type") +
  mytheme$standard() + 
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) + 
  facet_wrap(~donor_type, scales = "free_y")

print(p)
fn <- paste0(paste(today, "ImmuneCellComposition_ici_idi_myeloid", sep = "_"), ".png")
do.call(ggsave, c(list(fn, p), plotsave_param_large))

## MM-model between ICI and IDI.
p <- ici_idi  |> 
  filter(donor_type %in% c("mAAb+", "RecentOnset")) |> 
  filter(cell_type != "Other") |> 
  tidyr::nest(data = -c(cell_type, donor_type)) |> 
  dplyr::mutate(test = purrr::map(data, ~ broom.mixed::tidy(lme4::lmer(data = ., density ~ islet_type + (1 | case_id))))) |> 
  tidyr::unnest(test) |> 
  filter(term %in% c("islet_typeIDI")) |>
  dplyr::select(donor_type, statistic, cell_type, term) |> 
  ggplot(aes(x = cell_type, y = statistic, fill = donor_type)) +
  geom_bar(stat = "identity", position = "dodge") +
  scale_fill_manual(values = palettes[["stages"]]) +
  labs(x = "Cell type", y = "Statistic", fill = "Stage") +
  mytheme$standard() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) + 
  # horizontal-line at 2 and -2
  geom_hline(yintercept = 2, linetype = "dashed", color = "black") + 
  geom_hline(yintercept = -2, linetype = "dashed", color = "black")
print(p)

fn <- paste0(paste(today, "ImmuneCellComposition_ici_idi_mixed_effects", sep = "_"), ".png")
do.call(ggsave, c(list(fn, p), plotsave_param))
```

## A7.3: Immune Cell Cluster - ICI vs IDI
```{r ici-idi-composition-clust}
ici_idi_clust <- cols_islet_cluster |> 
  select(starts_with("density"), islet_type, case_id, image_id, donor_type) |>
  tidyr::pivot_longer(cols = starts_with("density_"), 
                      names_to = "cell_type", values_to = "density") |> 
  # remove the "fraction_total_" prefix & numeric suffix
  mutate(cell_clust = stringr::str_remove(cell_type, "density_")) |> 
  mutate(cell_type = stringr::str_remove(cell_clust, "_[0-9]$")) |> 
  mutate(cell_type = ifelse(cell_clust == "T.naive", "T_CD8", cell_clust),
         cell_type = ifelse(cell_clust == "T.EMRA", "T_CD8", cell_clust),
         cell_type = ifelse(cell_clust == "Treg", "T_CD4", cell_clust))

## Grouped barplots
p <- ici_idi_clust |> 
  filter(donor_type %in% c("mAAb+", "RecentOnset")) |> 
  group_by(case_id, donor_type, cell_clust, islet_type) |>
  summarize(density = mean(density)) |>
  ggplot(aes(x = cell_clust, y = density, fill = islet_type)) +
  geom_boxplot(outlier.shape = NA) +
  scale_fill_manual(values = c("ICI" = "blue", "IDI" = "red")) +
  labs(x = "Cell type", y = "Fraction of cells", fill = "Islet type") +
  mytheme$standard() + 
  # rotate x-axis labels
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  facet_wrap(~donor_type, scales = "free_y")

print(p)
fn <- paste0(paste(today, "ImmuneCellComposition_ici_idi_clust", sep = "_"), ".png")
do.call(ggsave, c(list(fn, p), plotsave_param_large))
```

-> These comparisons are not very useful.

# A8: Expression differences between ICI and IDIs

## A8.1: Prepare Data.
Here, Cell Types.
```{r idi-ici-expression-diffs}
# Define channels of interest.
channels <- metadata(spe)$channels
channels_imm <- unique(unlist(channels))
 
## Join with the islet type.
ici_idi <- cols_islet |> select(image_id, donor_type, case_id, islet_type) |> 
  distinct(image_id, donor_type, case_id, islet_type)

## Prepare the data (average expression per Cell Type)
exprs_vals_full <- as_tibble(scuttle::makePerCellDF(spe, features = channels_imm, 
                                                    assay.type = "exprs", use.dimred = FALSE)) |> 
  select(all_of(channels_imm), case_id, age, donor_type, image_id, cell_type, cell_cluster_scaled) |>
  filter(cell_type %in% c("T_CD4", "T_CD8", "B", "NK", "Neutrophil", "Myeloid", 
                          "Beta", "Alpha", "Ductal", "Endothelial"))

exprs_vals_full <- exprs_vals_full |> 
  left_join(dist_score, by = c("image_id", "donor_type", "case_id", "cell_type")) |> 
  # by ROI.
  group_by(image_id, case_id, donor_type, cell_type) |> 
  summarize(across(where(is.numeric), \(x) mean(x, na.rm = TRUE)), .groups = "drop")

## Join with Expression data. 1 row per image and cell type.
ici_idi_df <- exprs_vals_full |> 
  left_join(ici_idi, by = c("image_id", "donor_type", "case_id")) |> 
  filter(!is.na(islet_type)) |> 
  mutate(islet_type = factor(islet_type, levels = c("ICI", "IDI"))) |> 
  mutate(donor_type = factor(donor_type, levels = c("NoDiabetes", "sAAb+", "mAAb+", "RecentOnset")))
gc()
```

## A8.2: MM-models

### A8.2.1: Cell Type
Run MM-effect models per Cell Type, Channel and Donor Type (mAAb+, RecentOnset).
```{r mm-model}
## Run MM-models for each cell type, channel and donor_type (mAAb+, RecentOnset).
input_stats <- ici_idi_df  |> 
  filter(donor_type %in% c("mAAb+", "RecentOnset")) |>
  filter(cell_type != "Beta") |> 
  # channels are the columns
  pivot_longer(cols = all_of(channels_imm), names_to = "channel", values_to = "MeanExprs")

results <- input_stats  |> 
  tidyr::nest(data = -c(donor_type, channel, cell_type)) |> 
  dplyr::mutate(test = purrr::map(data, ~ broom.mixed::tidy(lmerTest::lmer(data = ., MeanExprs ~ islet_type + (1 | case_id))))) |> 
  tidyr::unnest(test)  |> 
  filter(term %in% c("islet_typeIDI")) |>
  dplyr::select(channel, donor_type, estimate, statistic, cell_type, term, p.value)
results
```

### A8.2.2: Analyze + plot linear mixed-effects models per Cell Type & stage
Analyze + plot linear mixed-effects models per Cell Type & stage.

```{r ici-ici-lmm-ct}
celltypes_oi <- c("T_CD8", "T_CD4", "B", "NK", "Myeloid", "Neutrophil", 
                  "Alpha", "Ductal", "Endothelial")
channels_ct <- metadata(spe)$channels[celltypes_oi]

## Myeloid: Before M1 afterwards M2 phenotype.
purrr::map2(celltypes_oi, channels_ct, \(cur_celltype, cur_channels) {
  p <- results |>
  filter(cell_type == !!cur_celltype) |>
  filter(channel %in% !!cur_channels) |> 
  filter(abs(statistic) > 2) |>
  ## Plot:
  ggplot(aes(x = channel, y = statistic, fill = donor_type)) +
  # vertical lines at + 2 and - 2
  geom_hline(yintercept = 2, linetype = "dashed") +
  geom_hline(yintercept = -2, linetype = "dashed") +
  geom_bar(stat = "identity", position = "dodge") +
  scale_fill_manual(values = palettes[["stages"]]) +
  labs(x = "Channel", y = "Fixed-effects Statistics", fill = "Stage") +
  theme(axis.text.x = element_text(angle = 90, hjust = 1)) +
  mytheme$standard() + 
  labs(title = cur_celltype)
  print(p)
  fn <- paste0(paste(today, "IsletType", cur_celltype, sep = "_"), ".png")
  do.call(ggsave, c(list(fn, p), plotsave_param))
})
```

-> Biggest changes:
Exhaustion/Activation/Survival in T-Cells. Lower CD73 in ICIs.
-> CD73 to restore immunity afterwards likely.

M1 -> M2 phenotype in myeloid cells. (also higher CD73) in IDIs.
Few differences in e.g. Neutrophils, NK-cells.

Endothelial: CD73 higher in IDIs.
Ductal: CD54/HLA-DR higher ICIs.
Alpha: only non-significant changes.

### Plot Myeloid Cell Differences.

```{r myeloid-ici-vs-idis}
## Expression as Violin Plots for CD163, CD206, CD16, CS, CD54, HLA-DR, 
channels_oi <- c("CD206", "CD16", "CS", "HLA_DR", "CD54", "CD163")
## NOTE: applicable to both mAAb+ and RecentOnset.
donor_type <- "RecentOnset"
library(ggpubr); library(rstatix)

input_myeloid <- input_stats |> 
  filter(cell_type == "Myeloid") |>
  #filter(cell_cluster_scaled == "Myeloid_3") |>
  filter(donor_type == !!donor_type)

input_myeloid |> 
  dplyr::count(islet_type, case_id) |> 
  arrange(case_id) |> 
  dplyr::count(case_id) |> filter(n == 2) |> pull(case_id) |> unique() -> case_ids

input_myeloid <- input_myeloid |> 
  filter(case_id %in% case_ids) |>
  pivot_wider(names_from = channel, values_from = MeanExprs)

p <- plot_ici_insulitic_expression(df = input_myeloid, features_oi = channels_oi,
                                   cell_types = "Myeloid",
                                   cur_var = "islet_type", x_var = "islet_type",
                                   group1 = "ICI", group2 = "IDI", facet_by = "channel", 
                                   y_label = "Macrophage expression", 
                                   cur_palette = palettes$colors[3:4]) + 
                                  geom_line(aes(group = case_id), alpha = 0.5) + 
                                  theme(legend.position = "none")
print(p)

fn <- paste0(paste(today, "MyeloidExpression_ICI_IDI_Violin", sep = "_"), ".png")
do.call(ggsave, c(list(fn, p), plotsave_param))
```


## A8.6 Expression differences between ICI and IDIs (My-3)

Just for Myeloid-3 (activated M1/M2 phenotype).
```{r icisidis-myeloid3-expression-diffs}
## Prepare the data (average expression for My3 per ImageID)
gc()
exprs_vals_my3 <- as_tibble(scuttle::makePerCellDF(spe, features = channels_imm, 
                                                   assay.type = "exprs", use.dimred = FALSE)) |> 
  select(all_of(channels_imm), case_id, donor_type, image_id, age, cell_type, cell_cluster_scaled) |>
  filter(cell_cluster_scaled == "Myeloid_3") |> 
  group_by(image_id, case_id, donor_type, cell_cluster_scaled) |> 
  summarize(across(all_of(channels_imm), \(x) mean(x, na.rm = TRUE))) |> 
  ungroup()

## Join with Expression data. 1 row per image and cell type.
ici_idi_my3 <- exprs_vals_my3 |> 
  left_join(ici_idi, by = c("image_id", "donor_type", "case_id")) |> 
  filter(!is.na(islet_type)) |>
  mutate(islet_type = factor(islet_type, levels = c("ICI", "IDI")))
```

Run MM-effect models per Cell Type, Channel and Donor Type (mAAb+, RecentOnset).
```{r mm-model-my3}
## Run MM-models for each cell type, channel and donor_type (mAAb+, RecentOnset).
input_stats <- ici_idi_my3  |> 
  filter(donor_type %in% c("mAAb+", "RecentOnset")) |>
  # channels are the columns
  pivot_longer(cols = all_of(channels_imm), names_to = "channel", values_to = "MeanExprs")

results <- input_stats  |> 
  tidyr::nest(data = -c(donor_type, channel, cell_cluster_scaled)) |> 
  dplyr::mutate(test = purrr::map(data, ~ broom.mixed::tidy(lmerTest::lmer(data = ., MeanExprs ~ islet_type + (1 | case_id))))) |> 
  tidyr::unnest(test)  |> 
  filter(term %in% c("islet_typeIDI")) |>
  dplyr::select(channel, donor_type, estimate, statistic, term, p.value)
```

### Plot Myeloid Cell Differences.

```{r my3-ici-vs-idis}
## Expression as Violin Plots for CD163, CD206, CD16, CS, CD54, HLA-DR, 
channels_oi <- c("CD206", "CD16", "CS", "HLA_DR", "CD54", "CD163")

## NOTE: applicable to both mAAb+ and RecentOnset.
donor_type <- "RecentOnset"
case_ids <- input_stats |> 
  dplyr::count(islet_type, case_id) |> 
  arrange(case_id) |> 
  dplyr::count(case_id) |> filter(n == 2) |> pull(case_id) |> unique()

cols_islet <- input_stats |> 
  filter(cell_cluster_scaled == "Myeloid_3") |>
  filter(donor_type == !!donor_type) |> 
  filter(case_id %in% case_ids) |>
  pivot_wider(names_from = channel, values_from = MeanExprs) |>  
  dplyr::rename("cell_type" = "cell_cluster_scaled")

p <- plot_ici_insulitic_expression(df = cols_islet, features_oi = channels_oi,
                                   cell_types = "Myeloid_3",
                                   cur_var = "islet_type", x_var = "islet_type",
                                   group1 = "ICI", group2 = "IDI", facet_by = "channel", 
                                   y_label = "Act. M1/M2-like expression", 
                                   cur_palette = palettes$colors[3:4]) + 
                                  geom_line(aes(group = case_id), alpha = 0.5) + 
                                  theme(legend.position = "none")
print(p)

fn <- paste0(paste(today, "MyeloidExpression_ICI_IDI_Violin_Myeloid_3", sep = "_"), ".png")
do.call(ggsave, c(list(fn, p), plotsave_param))

extfig9b <- p
saveRDS(extfig9b, file.path(paths$home_git, "figures", "extfig9b.rds"))
```


# A9: Stats EXPRESSIOn with PROGRESSION

## A9.1: Expression progression for ICIs (CELL_TYPES)
-> Partly already did in the upper part of the script.
-> Difference: use only ICIs!!

-> This is to get a global overview.

First, cell types.
```{r lmms-progression}
## Run MM-models for each cell type, channel and donor_type 
## Run only for mAAb+ and RecentOnset ICIs
full_res_exp <- ici_idi_df |> 
  filter(islet_type == "ICI") |>
  filter(donor_type %in% c("NoDiabetes", "sAAb+", "mAAb+", "RecentOnset")) |>
  mutate(age_group = ifelse(age < 13, "<13", ">= 13")) |> 
  mutate(donor_type = factor(donor_type, levels = c("NoDiabetes", "sAAb+", "mAAb+", "RecentOnset"))) |>
  tidyr::pivot_longer(cols = all_of(channels_imm), names_to = "channel", values_to = "MeanExprs") |>
  tidyr::nest(data = -c(channel, cell_type)) |>
  dplyr::mutate(test = purrr::map(data, ~ broom.mixed::tidy(lmerTest::lmer(data = ., MeanExprs ~ donor_type + age_group + (1 | case_id))))) |>
  tidyr::unnest(test) |> 
  filter(effect == "fixed", term != "(Intercept)") |> 
  # mutate donor_type (remove the "donor_type" prefix)
  mutate(donor_type = stringr::str_remove(term, "donor_type"))  |> 
  mutate(donor_type = factor(donor_type, levels = c("sAAb+", "mAAb+", "RecentOnset"))) |> 
  select(-c(df, group, effect))

## Separate per Cell Type.
celltypes_oi <- unique(full_res_exp$cell_type)
purrr::map(celltypes_oi, \(cur_celltype) {
  p <- full_res_exp  |> 
    filter(cell_type == !!cur_celltype) |> 
    arrange(desc(statistic)) |> 
    filter(p.value < 0.1)  |> 
    filter(channel %in% metadata(spe)$channels[[cur_celltype]]) |>
    filter(grepl("donor_type", term)) |> 
    ggplot(aes(x = channel, y = statistic, fill = donor_type)) +
    geom_bar(stat = "identity", position = "dodge") +
    labs(x = "Donor Type", y = "Statistic", fill = "Stage", title = paste(cur_celltype)) +
    scale_fill_manual(values = palettes[["stages"]]) +
    mytheme$standard() +
    theme(axis.text.x = element_text(angle = 45, hjust = 1))
  print(p)
  fn <- paste0(paste(today, "Expression_progression", cur_celltype, sep = "_"), ".png")
  do.call(ggsave, c(list(fn, p), plotsave_param))
})
```

Checks for beta-cells.
```{r}
full_res_exp |> 
  filter(cell_type == "Beta") |> 
  filter(channel %in% metadata(spe)$channels[["Beta"]]) |> 
  print(n = "all")

## LIneage down: SYP, NKX6.1, INS, PDX1 down; IFN up: HLA-DR + CD54 up.
## All metabolic markers0 (non-significantly) upregulated.
tt2 <- ici_idi_df |> 
  filter(islet_type == "ICI") |>
  filter(cell_type == "Beta") |>
  filter(donor_type %in% c("NoDiabetes", "sAAb+", "mAAb+", "RecentOnset")) |>
  mutate(donor_type = factor(donor_type, levels = c("NoDiabetes", "sAAb+", "mAAb+", "RecentOnset"))) |>
  tidyr::pivot_longer(cols = all_of(channels_imm), names_to = "channel", values_to = "MeanExprs") |>
  tidyr::nest(data = -c(channel, cell_type)) |>
  dplyr::mutate(test = purrr::map(data, \(.x) {
    tt <- lmerTest::lmer(data = .x, MeanExprs ~ donor_type + (1 | case_id))
    emmeans::emmeans(tt, specs = "donor_type", pbkrtest.limit = 10000) |>
      emmeans::contrast("pairwise") |>
      broom.mixed::tidy()
  }))  |> 
  tidyr::unnest(test)

tt2 |> 
  filter(channel %in% c("SYP", "HLA_DR", "ATP5A", "LDHA", "CD73", "HK1", "CS", "INS")) |> 
  mutate(estimate = estimate*-1) |> 
  print(n = "all")
```

```{r}
## CD54.
exprs_vals_full |> 
  filter(cell_type == "Beta") |> 
  select(image_id, case_id, donor_type, CD54) |>
  arrange(desc(CD54))
```

## A9.2: Cell Cluster Expression Differences

Join with ICI data for other cluster data.
```{r ici-cl-exprs}
## by cluster
gc()
exprs_vals_full <- as_tibble(scuttle::makePerCellDF(spe, features = channels_imm, 
                                                    assay.type = "exprs", use.dimred = FALSE)) |> 
  select(all_of(channels_imm), case_id, donor_type, image_id, cell_type, cell_cluster_scaled) |>
  filter(cell_type %in% c("T_CD4", "T_CD8", "B", "NK", "Neutrophil", "Myeloid", 
                          "Beta", "Alpha", "Ductal", "Endothelial"))

exprs_vals_cl <- exprs_vals_full |> 
  group_by(image_id, case_id, donor_type, cell_cluster_scaled) |> 
  summarize(across(all_of(channels_imm), \(x) mean(x, na.rm = TRUE)))

ici_idi_df_cl <- exprs_vals_cl |> 
  left_join(ici_idi, by = c("image_id", "donor_type", "case_id")) |> 
  tidyr::drop_na() |> 
  mutate(islet_type = factor(islet_type, levels = c("ICI", "IDI")))
```

## A9.3: Linear Mixed-Effects Models for Cell Clusters
Now, cell-clusters. Only between ICIs along progression. 
used for overview of changes in expression along progression.

```{r lmmsprogression-ct}
full_res_exp_cl <- ici_idi_df_cl |> 
  filter(islet_type == "ICI") |>
  filter(donor_type %in% c("NoDiabetes", "sAAb+", "mAAb+", "RecentOnset")) |>
  mutate(donor_type = factor(donor_type, levels = c("NoDiabetes", "sAAb+", "mAAb+", "RecentOnset"))) |>
  tidyr::pivot_longer(cols = all_of(channels_imm), names_to = "channel", values_to = "MeanExprs") |>
  tidyr::nest(data = -c(channel, cell_cluster_scaled)) |>
  dplyr::mutate(test = purrr::map(data, ~ broom.mixed::tidy(lmerTest::lmer(data = ., MeanExprs ~ donor_type + (1 | case_id))))) |>
  tidyr::unnest(test) |> 
  filter(effect == "fixed", term != "(Intercept)") |> 
  # mutate donor_type (remove the "donor_type" prefix)
  mutate(donor_type = stringr::str_remove(term, "donor_type"))  |> 
  mutate(donor_type = factor(donor_type, levels = c("sAAb+", "mAAb+", "RecentOnset"))) |> 
  select(-c(df, group, effect)) |> 
  filter(!cell_cluster_scaled %in% c("0", "B1", "B2"))

## Separate per Cell Type.
clusters_oi <- unique(full_res_exp_cl$cell_cluster_scaled)
purrr::map(clusters_oi, \(cur_cluster) {
  cur_celltype <- stringr::str_remove(cur_cluster, "_[0-9]+$")
  ifelse(cur_celltype == "T-naive", cur_celltype <- "T_CD8", cur_celltype)
  ifelse(cur_celltype == "T-EMRA", cur_celltype <- "T_CD8", cur_celltype)
  p <- full_res_exp_cl  |> 
    filter(cell_cluster_scaled == !!cur_cluster) |> 
    arrange(desc(statistic)) |> 
    filter(p.value < 0.1)  |> 
    filter(channel %in% metadata(spe)$channels[[cur_celltype]]) |>
    ggplot(aes(x = channel, y = statistic, fill = donor_type)) +
    geom_bar(stat = "identity", position = "dodge") +
    labs(x = "Donor Type", y = "Statistic", fill = "Stage", title = paste(cur_cluster)) +
    scale_fill_manual(values = palettes[["stages"]]) +
    mytheme$standard() +
    theme(axis.text.x = element_text(angle = 45, hjust = 1))
  print(p)
  fn <- paste0(paste(today, "Expression_progression", cur_cluster, sep = "_"), ".png")
  do.call(ggsave, c(list(fn, p), plotsave_param))
})
```

## A9.3: Plot changes for specific cell clusters across progression
### A9.3.1: T-CD4-5
Plot changes for T-CD4-5 across progression.
```{r tcd4-5-progression}
library(emmeans)
channels_to_plot <- c("GranB", "PD1", "TIM3", "CD103")

t_cd4_5_df <- ici_idi_df_cl |> 
  filter(islet_type == "ICI") |>
  filter(donor_type %in% c("NoDiabetes", "sAAb+", "mAAb+", "RecentOnset")) |>
  mutate(donor_type = factor(donor_type, levels = c("NoDiabetes", "sAAb+", "mAAb+", "RecentOnset"))) |> 
  filter(cell_cluster_scaled == "T_CD4_5") |> 
  mutate(cell_type = "T_CD4") |> 
  pivot_longer(cols = all_of(channels_imm), names_to = "channel", values_to = "exprs")

t_cd4_5_df <- t_cd4_5_df |>
  left_join(colData(spe) |> as_tibble() |> distinct(case_id, age_group.y, ICU_time_days.y), by = "case_id") |> 
  as_tibble() |> 
  filter(image_id %in% icis) |>
  filter(!(case_id %in% to_del))

p <- plot_expression_violin(exprs_dat = t_cd4_5_df, 
                            channels_oi = channels_to_plot, 
                            cell_types = "T_CD4", 
                            facet_by = "channel", 
                            y_label = "CD4+ PD1+ act. expression",
                            immune_adjustment = TRUE)
print(p)
  
fn <- paste0(paste(today, "T_CD4_5", "Progression", sep = "_"), ".png")
do.call(ggsave, c(list(fn, p), plotsave_param_large))

extfig6f <- p
saveRDS(extfig6f, file.path(paths$home_git, "figures", "extfig6f.rds"))
```

### A9.3.2: T-CD8-7
Plot changes for T-CD8-7 across progression.
```{r tcd87-progression}
channels_to_plot <- c("GranB", "TIM3", "PD1", "CD103")

t_cd8_7_df <- ici_idi_df_cl |> 
  filter(islet_type == "ICI") |>
  filter(donor_type %in% c("NoDiabetes", "sAAb+", "mAAb+", "RecentOnset")) |>
  mutate(donor_type = factor(donor_type, levels = c("NoDiabetes", "sAAb+", "mAAb+", "RecentOnset"))) |> 
  filter(cell_cluster_scaled == "T_CD8_7") |> 
  mutate(cell_type = "T_CD8") |> 
  pivot_longer(cols = all_of(channels_imm), names_to = "channel", values_to = "exprs")

t_cd8_7_df <- t_cd8_7_df |>
  left_join(colData(spe) |> as_tibble() |> distinct(case_id, age_group.y, ICU_time_days.y), by = "case_id") |> 
  as_tibble() |> 
  filter(image_id %in% icis) |>
  filter(!(case_id %in% to_del))

p <- plot_expression_violin(exprs_dat = t_cd8_7_df, 
                            channels_oi = channels_to_plot, 
                            cell_types = "T_CD8", 
                            facet_by = "channel", 
                            y_label = "PD1+,CD27+ expression",
                            immune_adjustment = TRUE) + 
          facet_wrap(~channel, scales = "free_y", nrow = 3, ncol = 2)
print(p)
  
fn <- paste0(paste(today, "T_CD8_7", "Progression", sep = "_"), ".png")
plotsave_param_large2 <- plotsave_param_large
plotsave_param_large2$width <- 400
plotsave_param_large2$height <- 600
do.call(ggsave, c(list(fn, p), plotsave_param_large2))

extfig6g <- p
saveRDS(extfig6g, file.path(paths$home_git, "figures", "extfig6g.rds"))
```

### A9.3.3: T-regs
Plot changes for T-regs across progression.
```{r tregs-progression}
channels_to_plot <- c("TIM3", "FoxP3", "PD1", 
                      "CS", "HK1", "CD45RO")

treg_df <- ici_idi_df_cl |> 
  filter(islet_type == "ICI") |>
  filter(donor_type %in% c("NoDiabetes", "sAAb+", "mAAb+", "RecentOnset")) |>
  mutate(donor_type = factor(donor_type, levels = c("NoDiabetes", "sAAb+", "mAAb+", "RecentOnset"))) |> 
  filter(cell_cluster_scaled == "T_CD4_10") |> 
  mutate(cell_type = "T_CD4") |> 
  pivot_longer(cols = all_of(channels_imm), names_to = "channel", values_to = "exprs")

treg_df <- treg_df |>
  left_join(colData(spe) |> as_tibble() |> distinct(case_id, age_group.y, ICU_time_days.y), by = "case_id") |> 
  as_tibble() |> 
  filter(image_id %in% icis) |>
  filter(!(case_id %in% to_del))
  
p <- plot_expression_violin(exprs_dat = treg_df, 
                            channels_oi = channels_to_plot, 
                            cell_types = "T_CD4", 
                            facet_by = "channel", 
                            y_label = "Treg expression",
                            immune_adjustment = TRUE)
print(p)
  
fn <- paste0(paste(today, "T_regs", "Progression", sep = "_"), ".png")
do.call(ggsave, c(list(fn, p), plotsave_param_large))
```

# A10: Visualization of Cell Clusters and Cell Types
## Load images and masks

```{r viz-cDC-clust-select}
all_channels <- rownames(sce_imm)

viz_clust <- NULL # c(22, 21, 9) # c(21, 9, 11, 12, 2, 20, 1, 5, 3, 4)
viz_method <- "cell_type"
viz_assay <- "scaled"
```

```{r viz-lympho-cluster-load}
viz_clust <- TRUE
if (!is.null(viz_clust)) {
  nb_images <- 14
  image_extension <- ".tiff"
  clust_name <- viz_method
  
  # Subset the SCE
  sce_viz <- spe

  set.seed(seed)

  image_sub <- sort(sample(
    unique(sce_viz$image_fullname),
    min(length(unique(sce_viz$image_fullname)), nb_images)))
  
  image_sub <- unique(c(image_sub, images <- 
    c("6228_Immune_ROI_082.tiff", "6551_Immune_ROI_069.tiff", 
      "6362_Immune_ROI_019.tiff", "6178_Immune_ROI_029.tiff",
      # CD54:
      "6550_Immune_ROI_024.tiff", "6534_Immune_ROI_040.tiff",
      # Myeloid-4:
      "6509_Immune_ROI_026.tiff", "6447_Immune_ROI_009.tiff", "6533_Immune_ROI_051.tiff", "6551_Immune_ROI_069.tiff", 
      "6551_Immune_ROI_050.tiff",
      # Myeloid-11:
      "6458_Immune_ROI_077.tiff", "6458_Immune_ROI_046.tiff", "6044_Immune_ROI_049.tiff", "6367_Immune_ROI_076.tiff",
      # Myeloid-12:
      "6429_Immune_ROI_008.tiff", "6449_Immune_ROI_072.tiff", "6534_Immune_ROI_066.tiff",
      # Myeloid-7:
      "6520_Immune_ROI_067.tiff", "6324_Immune_ROI_016.tiff", "6526_Immune_ROI_009.tiff",
      # Myeloid-10:
      "6228_Immune_ROI_023.tiff", "6534_Immune_ROI_076.tiff"
      )))

  # Folders
  folder_images <- file.path(paths$folder_in, "img", paths$panel_type)
  folder_masks <- file.path(paths$folder_in, "masks_cells",
                            paths$panel_type, "whole-cell")
  
  images <- cytomapper::loadImages(file.path(folder_images, image_sub))
  cytomapper::channelNames(images) <- all_channels
  # Add image names to metadata
  mcols(images)$ImageName <- names(images)
  mcols(images)$ImageName <- paste0(mcols(images)$ImageName)

  # Load images and masks
  #images <- yoloader(
  #  x = sce_viz,
  #  image_dir = folder_images,
  #  image_names = image_sub,
  #  type = "stacks"
  #)

  masks <- yoloader(
    x = sce_viz,
    image_dir = folder_masks,
    image_names = image_sub,
    as.is = TRUE,
    type = "masks"
  )
  
  sce_viz <- sce_viz[, sce_viz$image_fullname %in% image_sub]
  sce_viz$ImageName <- gsub(image_extension, "", sce_viz$image_fullname)
}
```

### Cytoviewer app.
```{r cytoviewer}
library(cytoviewer)
sce_viz[[clust_name]] <- as.factor(sce_viz[[clust_name]])
app <- cytoviewer(image = images, 
                  mask = masks, 
                  object = sce_viz,
                  img_id = "ImageName", 
                  cell_id = "cell_number")

if (interactive()) {
  shiny::runApp(app)
}
```




# A11: Umap and Heatmap for Panel Immune.
## A11.1: UMAP full

Run UMAP for all cells in the immune panel.

```{r umap_full}
channels <- rownames(spe)
spe_sub <- spe[, spe$cell_id %in% metadata(spe)[["subset"]]]
cur_dat <- scuttle::makePerCellDF(spe_sub, features = channels, 
                                  assay.type = "exprs", use.dimred = TRUE) |> 
  as_tibble() |> 
  select(all_of(channels), starts_with("UMAP"), cell_type, cell_category, donor_type, case_id, image_id) |> 
  mutate(cell_type_cat = if_else(cell_category == "immune", "Immune", cell_type)) |> 
  filter(!cell_type_cat %in% c("Other", "Ambiguous"))

palettes$celltype <- (c(
  Beta = palettes$colors[1], Alpha = palettes$colors[2],
  Delta = palettes$colors[3], IsletOther = palettes$colors[4],
  Ductal = palettes$colors[5], Acinar = palettes$colors[6],
  Endothelial = palettes$colors[7], Mesenchymal = palettes$colors[8],
  SmoothMuscle = palettes$colors[9], Nerves = palettes$colors[10],
  Immune = palettes$colors[11]
))

palette_ct <- c(palettes$colors[seq_len(length(celltypes))])
p <- plot_dim_red(cur_dat, "UMAP_exprs", "cell_type_cat", palette = palettes$celltype) + 
  labs(title = "", x = "", y = "") + 
  theme(
    legend.title = element_text(size = 25),  # Adjust the legend title font size
    legend.text = element_text(size = 22)    # Adjust the legend text font size
  ) + 
  guides(color = guide_legend(title = "Cell Type", override.aes = list(size = 5)))
  
fn <- paste0(paste(today, "UMAP", "full", sep = "_"), ".png")
do.call(ggsave, c(list(fn, p), plotsave_param))
```

## A11.2: Full Heatmap
All non-immune Cell Types.
```{r heatmap_full}
hmlist <- list()
channels <- c("SYP", "INS", "SST", "GCG", "CAV1", "SMA", "LDHA", "VIM", "CD73", 
              "PDX1", "NKX6_1", "CD57", "CD56", "CD54", "TMEM173")

spe_sub2 <- spe_sub[, (spe_sub$cell_type %in% c("Beta", "Alpha", "Delta", "Endothelial", 
                                                "Ductal", "Acinar", "Mesenchymal",
                                                "SmoothMuscle", "Nerves"))]
# Summarize the data
hm <- summarize_heatmap(spe_sub2,
                        expr_values = "scaled",
                        cluster_by = "cell_type",
                        channels = channels)

# Counts
ct_anno <- as.data.frame(table(spe_sub$cell_type))
rownames(ct_anno) <- ct_anno$Var1
ct_anno <- ct_anno[rownames(hm), ]
cell_type_counts <- ct_anno$Freq
names(cell_type_counts) <- ct_anno$Var1

# Create row annotation for cell type counts
breaks <- c(0, 0.5, 0.7, 1)  # Define custom breaks for the color scale
colors <- viridis::viridis(4) 
# Create a color function with more emphasis on the lower values
col_fun <- circlize::colorRamp2(breaks, colors)

# Create row annotation for cell type counts
row_anno <- ComplexHeatmap::rowAnnotation(
  Counts = ComplexHeatmap::anno_barplot(cell_type_counts, 
                         gp = grid::gpar(fill = "skyblue", col = "black"),
                         border = TRUE,
                         width = grid::unit(4, "cm"),
      axis_param = list(gp = grid::gpar(fontsize = 25)),
),       show_annotation_name = FALSE
)

# Display the heatmap
fn <- file.path(paths$folder_script, paste0(paste(today, "cell_type_full", "Heatmap",
                   sep = "_"), ".png"))
png(fn, width = 1200, height = 800)
p <- ComplexHeatmap::Heatmap(heatmaply::normalize(hm), name = "Scaled expression", 
                              col= col_fun,
                              show_row_names = TRUE, show_column_names = TRUE,
                              cluster_rows = TRUE, cluster_columns = TRUE,
                              row_names_gp = grid::gpar(fontsize = 30, fontfamily = "Arial"),
                              column_names_gp = grid::gpar(fontsize = 30, fontfamily = "Arial"), 
                              heatmap_legend_param = list(legend_direction = "horizontal",
                                                          title_position = "topcenter",
                                                          title_gp = grid::gpar(fontsize = 40, fontfamily = "Arial"),
                                                          labels_gp = grid::gpar(fontsize = 40, fontfamily = "Arial")),                             
                            row_dend_gp = grid::gpar(lwd = 3),  # Increase dendrogram line width for rows
                            column_dend_gp = grid::gpar(lwd = 3),  # Increase dendrogram line width for columns
                            row_names_side = "left",
                            row_dend_side = "right") + 
      row_anno
ComplexHeatmap::draw(p, heatmap_legend_side = "top", 
                      padding = unit(c(12, 20, 2, 2), "mm"))   # right and bottom padding

dev.off()
```