---
title: "14_DimensionalityReduction_cells_Immune"
author: "Nicolas Damond, Nathan Steenbuck"
date: "Created: 05 November, 2022; Compiled: `r format(Sys.time(), '%d %b, %Y')`"
output: workflowr::wflow_html
editor_options:
  chunk_output_type: inline
---

```{r setup, include=FALSE}

cur_user <- Sys.info()[["user"]]
script_name <- "14_DimensionalityReduction_cells_Immune.Rmd"

if (cur_user == "ubuntu") {
  source(file.path("/", "mnt", "T1D_Vol", "T1D_analysis", "analysis", "helpers.R"))
  #n_cores <- future::availableCores() - 1
  n_cores <- 2
  future::plan(future::multicore(workers = n_cores))
  paths <- getPaths(script_name)
  knitr::opts_knit$set(root_dir = paths$home_data)
  do_print <- FALSE
} else {
  source(file.path("/", "home", "nsteen", "scripts", "T1D_analysis", "analysis", "helpers.R"))
  n_cores <- 8
  future::plan(future::multicore(workers = n_cores))
  paths <- getPaths(script_name)
  paths$folder_script <- paths$cluster_folder_script
  paths$folder_in <- paths$cluster_folder_in
  paths$folder_out <- paths$cluster_home 
  knitr::opts_knit$set(root_dir = paths$cluster_home)
  do_print <- TRUE
}

seed <- 123456
set.seed(seed)
options <- furrr::furrr_options(seed = seed)
paths$prev <- paste("08_CellTypesIslet", paths$object_type, paths$panel_type, sep = "_")
knitr::opts_chunk$set(echo = TRUE)
options(future.globals.maxSize = 60 * 1024 * 1024 * 1024)
```


# **Goal**

Run dimensionality reduction and plot reduced dimensions for individual cell types.  


# **Settings**  

## Load packages

```{r packages, results='hide'}
suppressPackageStartupMessages(c(
  library(data.table),
  library(dplyr),
  library(SpatialExperiment),
  library(parallel)
))
```

## Paths and settings

```{r settings}
# Paths
if (!dir.exists(paths$folder_script)) dir.create(paths$folder_script)
plotsave_param$path <- paths$folder_script
plotsave_param_large$path <- paths$folder_script

# Misc settings
today <- gsub("-", "", Sys.Date())
```

##  Read in the data

Load the SpatialExperiment (SPE) object saved at the previous step.

```{r load-data}
fn_spe <- file.path(paths$folder_out, paths$prev, paste0(paths$object_type, "_", paths$panel_type, ".rds"))
spe <- readRDS(fn_spe)
print(spe)
```

##  Read in the data

Load the list containing one SCE object per cell type.  

```{r load-data-celltypes}
folder_spes <- file.path(paths$folder_out, paste0("12_CellTypeSCEs_", paths$object_type, "_", paths$panel_type))
fn_spes <- file.path(folder_spes, "celltypes_Immune.rds")

spes <- readRDS(fn_spes)
print(spes)
```



# **Reduce dimensions** 

Dimensionality reduction is performed on each SCE object in the `spes` list.

Dimensionality reduction algorithms currently used:  
- **PCA**
- **UMAP**
- **tSNE**
- **Diffusion map**

## Packages

```{r packages2, results='hide'}
suppressPackageStartupMessages(c(
  library(uwot),
  library(destiny),
  library(viridis),
  library(scater),
  library(foreach),
  library(doParallel)
))
```

## Select parameters

Select the dimensionality reductions algorithms, assays and channels to use and the variables to plot on reduced dimensions.

```{r select-assay2}
# Select the dimensionality reduction algorithms to use
dimred_sel <- c("UMAP", "DiffMap") # "tSNE", 
writeLines(c("Dimensionality reduction:", dimred_sel))

# Select the assay(s) of interest
assay_sel <- c("scaled", "fastMNN_case_id") # "exprs", 
writeLines(c("\nAssays:", assay_sel[assay_sel %in% assayNames(spes[[1]])]))

# Variables to plot
plot_variables <- c("stages", "cases", "colors", "colors50")
names(plot_variables) <- c("donor_type", "case_id",
                           "cell_type", "immune_cluster")
writeLines("\nVariables to plot:")
print(plot_variables)

# Cell state channels (not used for dimensionality reduction)
channels_state <- c("Ki67", "DNA1", "DNA2")
```

## Run PCA

```{r run-PCA}
# Function to calculate PCA
calc_PCA <- function(x, exprs_values, channels = NULL) {
  if (is.null(channels))
    cur_channels <- rownames(x)[!rownames(x) %in% channels_state]
  else
    cur_channels <- channels
  
  purrr::map(exprs_values, \(cur_exprs) {
    dimred_name <- paste("PCA", cur_exprs, sep = "_")
    
    if (!dimred_name %in% SingleCellExperiment::reducedDimNames(x)) {
      x <- scater::runPCA(
        x, exprs_values = cur_exprs, subset_row = cur_channels,
        ncomponents = 5, ntop = (length(cur_channels) / 2), name = dimred_name
      )
    }
  })
  return(x)
}

# Run PCA
spes <- unlist(purrr::map(spes, \(spes_x) {
  calc_PCA(spes_x, exprs_values = assay_sel)}))
```

## Run UMAP

```{r run-UMAP}
# Function to calculate UMAP
calc_UMAP <- function(x, exprs_values, channels = NULL) {
  if (is.null(channels))
    cur_channels <- rownames(x)[!rownames(x) %in% channels_state]
  else
    cur_channels <- channels
  
  purrr::map(exprs_values, \(cur_exprs) {
    dimred_name <- paste("UMAP", cur_exprs, sep = "_")
    
    if (!dimred_name %in% SingleCellExperiment::reducedDimNames(x)) {
      cur_umap <- umap(t(SummarizedExperiment::assay(x, cur_exprs)[cur_channels, ]))
      colnames(cur_umap) <- c("UMAP1", "UMAP2")
      rownames(cur_umap) <- colnames(SummarizedExperiment::assay(x, cur_exprs))
      SingleCellExperiment::reducedDim(x, dimred_name) <<- cur_umap
      remove(cur_umap)
    }
  })
  return(x)
}

# Run UMAP
spes <- unlist(purrr::map(spes, \(spe_x) {
    calc_UMAP(spe_x, exprs_values = assay_sel) 
  }))
```

## Run Diffusion Map

DiffMap is run on the principal components of each assay, rather than on the counts.

```{r run-diffmap}
# Function to calculate Diffusion Map
calc_diff_map <- function(x, exprs_values, channels = NULL) {
  if (is.null(channels))
    cur_channels <- rownames(x)[!rownames(x) %in% channels_state]
  else
    cur_channels <- channels
    
  purrr::map(exprs_values, \(cur_exprs) {
    dimred_name <- paste("DiffMap", cur_exprs, sep = "_")
    
    if (!dimred_name %in% SingleCellExperiment::reducedDimNames(x)) {
      pca_name <- paste("PCA", cur_exprs, sep = "_")
      cur_pca <- SingleCellExperiment::reducedDim(x, pca_name)
      
      cur_diffmap <- DiffusionMap(cur_pca)
      SingleCellExperiment::reducedDim(x, dimred_name) <- cur_diffmap@eigenvectors[, c("DC1", "DC2"), drop = FALSE]
      colnames(SingleCellExperiment::reducedDim(x, dimred_name)) <- c("DiffMap1", "DiffMap2")
    }
  })
  return(x)
}

# Run Diffusion Map
spes <- unlist(purrr::map(spes, \(spe_x) {
    calc_diff_map(spe_x, exprs_values = assay_sel) 
  }))
```

## Save the updated SCE

Contains the new reduced dimensions.

```{r save-spes}
print(spes)
saveRDS(spes, fn_spes)
```


# **Plot reduced dimensions**

Because many plots are generated, the functions are run in parallel to save time. The plots are saved but not displayed in the notebook.

## Plot variables

```{r plot-dimred-variables}
lplot_dim_red <- function(x, exprs_values, dimred, variables) {
  cur_dat <- makePerCellDF(x, use_dimred = TRUE) |>
    mutate(case_id = factor(case_id, levels = metadata(spe)$cases),
           donor_type = factor(donor_type, levels = metadata(spe)$stages)) |>
    arrange(case_id, donor_type)
  
  purrr::map(exprs_values, \(cur_exprs) {
    purrr::map(dimred_sel, \(cur_dimred) {
      purrr::map2(plot_variables, names(plot_variables), \(cur_var, name_cur_var) {
        dimred_name <- paste(cur_dimred, cur_exprs, sep = "_")

        if (dimred_name %in% SingleCellExperiment::reducedDimNames(x)) {
          p <- plot_dim_red(cur_dat, dimred_name, name_cur_var,
                            size = 0.1, palette = palettes[[cur_var]])
          fn <- paste0(paste(today, mainExpName(x), dimred_name,
                             cur_var, sep = "_"), ".png")
          do.call(ggsave, c(list(fn, p), plotsave_param))
        }
      })
    })
  })
  remove(cur_dat)
}

# Plot
purrr::map(spes, \(spe_x) {
  lplot_dim_red(spe_x, exprs_values = assay_sel, dimred = dimred_sel,
                variables = plot_variables)
})
```

## Plot marker expression

```{r plot-dimred-channels}
# Function for plotting marker expression on reduced dimensions
lplot_dim_red_channels <- function(x, exprs_values, dimred, channels = NULL) {
  if (is.null(channels))
    cur_channels <- rownames(x)[!rownames(x) %in% channels_state]
  else
    cur_channels <- channels
  
  purrr::map(exprs_values, \(cur_exprs) {
    cur_dat <- as.data.table(
      makePerCellDF(x,
                    features = cur_channels,
                    exprs_values = cur_exprs))
    
    for (j in seq_along(dimred)) {
      dimred_name <- paste(dimred[j], cur_exprs, sep = "_")
      
      if (dimred_name %in% SingleCellExperiment::reducedDimNames(x)) {
        cur_dat_long <- melt.data.table(
          cur_dat,  measure.vars = cur_channels, variable.name = "channel",
          id.vars = c(paste0(dimred_name, ".1"), paste0(dimred_name, ".2")),
          value.name = cur_exprs)
        
        # Plot marker expression
        p <- plot_dim_red_channels(cur_dat_long, dimred_name, cur_exprs,
                                   cur_channels)
        fn <- paste0(paste(today, mainExpName(x), dimred_name, "Channels",
                           sep = "_"), ".png")
        do.call(ggsave, c(list(fn, p), plotsave_param_large))
      }
    }
  })
  remove(cur_dat, cur_dat_long)
}

# Plot
purrr::map(spes, \(spe_x) {
  lplot_dim_red_channels(spe_x, exprs_values = assay_sel, dimred = dimred_sel)
})
```
