---
title: "19_Insulitic_cells_Islet"
author: "Nathan Steenbuck"
date: "Created: 24 May, 2025; Compiled: `r format(Sys.time(), '%d %b, %Y')`"
output: workflowr::wflow_html
editor_options:
  chunk_output_type: inline
---

```{r setup, include=FALSE}
if (rstudioapi::isAvailable()) {
  script_name <- basename(rstudioapi::getSourceEditorContext()$path)
} else {
  script_name <- knitr::current_input()
}

cur_user <- Sys.info()[["user"]]
script_name <- "19_Insulitic_cells_Islet.Rmd"

if (cur_user == "ubuntu") {
  source(file.path("/", "mnt", "central_nas", "projects", "type1_diabetes", "nathan", 
                   "T1D_Vol", "T1D_analysis", "analysis", "helpers.R"))
  #n_cores <- future::availableCores() - 1
  n_cores <- 4
  future::plan(future::multicore(workers = n_cores))
  paths <- getPaths(script_name)
  knitr::opts_knit$set(root_dir = paths$home_data)
  do_print <- FALSE
} else {
  source(file.path("/", "home", "nsteen", "scripts", "T1D_analysis", "analysis", "helpers.R"))
  n_cores <- 8
  future::plan(future::multicore(workers = n_cores))
  paths <- getPaths(script_name)
  paths$folder_script <- paths$cluster_folder_script
  paths$folder_in <- paths$cluster_folder_in
  paths$folder_out <- paths$cluster_home 
  knitr::opts_knit$set(root_dir = paths$cluster_home)
  do_print <- TRUE
}

seed <- 123456
set.seed(seed)
options <- furrr::furrr_options(seed = seed)
paths$prev <- paste("08_SubclusteringBeta", paths$object_type, paths$panel_type, sep = "_")
knitr::opts_chunk$set(echo = TRUE)
```



# **Goal**

Examine protein abundance. 
Compare ICIs (Insulin-containing islets) vs IDIs (Insulin-deficient islets).
Compare Insulitic ICI vs ICIs.

# **A0: Settings and Data**

## Load packages

```{r packages, results='hide'}
suppressPackageStartupMessages(c(
  library(data.table),
  library(dplyr),
  library(SpatialExperiment),
  library(scuttle),
  library(parallel),
  library(patchwork),
  library(tidyr),
  library(emmeans)
))
```


```{r packages2, results='hide'}
suppressPackageStartupMessages(c(
  library(ggplot2),
  library(viridis),
  library(heatmaply),
  library(dittoSeq),
  library(ggridges),
  library(rstatix),
  library(ggpubr)
))
```
## Paths and settings

```{r settings}
# Paths
if (!dir.exists(paths$folder_script)) dir.create(paths$folder_script)
plotsave_param$path <- paths$folder_script
plotsave_param_large$path <- paths$folder_script

# Misc settings
today <- gsub("-", "", Sys.Date())
emm_options(lmerTest.limit = 10000)
```

##  Read in the data

Load the SpatialExperiment (SPE) object saved at the previous step.

```{r load-data}
fn_spe <- file.path(paths$folder_out, paths$prev, paste0(paths$object_type, "_", paths$panel_type, ".rds"))
spe <- readRDS(fn_spe)
print(spe)
```

## Parameters

**Note: this step has to be manually adapted**  

- Define the assay(s) of interest.  
- Define the markers to keep for marker expression analysis.  

```{r parameters}
# Censor value for counts scaling
censor_val <- 0.999

# Select assays and reduced dimensions
assay_sel <- c("exprs", "scaled")
writeLines(c("Assay:", assay_sel[assay_sel %in% assayNames(spe)]))

# Variables to plot
plot_variables <- c("donor_type", "case_id")
names(plot_variables) <- c("stages", "cases")
cat("\nVariables to plot:", plot_variables)
plot_variables_stages <- plot_variables["stages"]

# List of cell types to plot
celltypes <- names(metadata(spe)$channels)
names(celltypes) <- celltypes

# Color palette
palette_ct <- palettes$colors[seq_along(celltypes)]
names(palette_ct) <- celltypes

# All channels to display
channels <- unique(unlist(metadata(spe)$channels))
```

## Subset the SCE

Retain only celltypes of interest.
That is, Beta, Alpha, Delta, Acinar, Ductal, Endothelial and islet.
```{r subset-spe}
sce_sub <- as(spe, "SingleCellExperiment")
sce_sub <- sce_sub[channels, sce_sub$cell_type %in% celltypes]
remove_panreatitis <- TRUE
if (remove_panreatitis) {
  sce_sub <- sce_sub[, sce_sub$donor_type != "Pancreatitis"]
}
sce_sub$donor_type <- factor(sce_sub$donor_type, levels = c("NoDiabetes", "sAAb+", "mAAb+", "RecentOnset", "LongDuration")) # , "Pancreatitis"))
```


# A1: ICIs vs IDIs:
- ICIs: Insulin-containing islets
- IDIs: Insulin-deficient islets
Differential Calculations make only sense for Alpha & Delta-cells; not Beta-cells.

## A1.1: Prepare Data.
Channels, Image, + Islet-composition.
```{r ICIs-vs-IDIs}
channels <- unique(c(metadata(spe)$channels$Beta,
                     metadata(spe)$channels$Alpha, 
                     metadata(spe)$channels$Delta))

image_ids <- colData(sce_sub)$image_id |> unique()

## Islet CT Composition.
compositions <- metadata(sce_sub)$islet_compo |> 
    as_tibble() |> 
    filter(image_id %in% image_ids) |> 
    mutate(islet_type = if_else(fraction_Beta > 0, "ICI", "IDI"))
```

Expression data by ROI + join with compositional data.
```{r expression-data-ICI-IDI}
## Get Expression Data for Alpha / Delta / Endothelial / T_cell / Macrophage
## Aggregate by Image.
cols_islet <- scuttle::makePerCellDF(spe, features = channels, assay.type = "exprs", use.coldata = TRUE, use.dimred = FALSE) |> 
  as_tibble() |> 
  filter(cell_type %in% c("Alpha", "Delta", "Endothelial", "Tcell", "Macrophage")) |> 
  select(case_id, image_id, cell_type, donor_type, all_of(channels)) |> 
  # group-by image_id, case_id, cell_type, donor_type
  group_by(image_id, case_id, cell_type, donor_type) |>
  summarise(across(all_of(channels), mean), .groups = "drop")

## Join with islet composition.  
cols_islet <- cols_islet |>
  left_join(compositions, by = "image_id") |> 
  filter(islet_type %in% c("ICI", "IDI")) |> 
  tidyr::drop_na()
```

## A1.2: ICI vs IDI: Overviews.
PLOT ISLET TYPE (ICI vs IDI)
```{r islet_type}
p <- cols_islet  |> 
  dplyr::count(islet_type, donor_type) |> 
  dplyr::group_by(donor_type) |>
  dplyr::mutate(total = sum(n),
                frac = n / total) |> 
  ggplot(aes(x = islet_type, y = frac, fill = donor_type)) +
  geom_bar(stat = "identity", position = "dodge") +
  scale_fill_manual(values = palettes[["stages"]]) +
  labs(x = "Islet type", y = "Fraction of islets", fill = "Stage") +
  mytheme$standard()

print(p)
fn <- paste0(paste(today, "IsletType", sep = "_"), ".png")
do.call(ggsave, c(list(fn, p), plotsave_param))


p <- cols_islet  |> 
  dplyr::count(islet_type, donor_type, case_id) |> 
  dplyr::group_by(donor_type, case_id) |>
  dplyr::mutate(total = sum(n),
                frac = n / total) |> 
  ggplot(aes(x = islet_type, y = frac, fill = donor_type)) +
  geom_boxplot() +
  geom_point(position = position_dodge(width = 0.75)) + 
  scale_fill_manual(values = palettes[["stages"]]) +
  labs(x = "Islet type", y = "Fraction of islets", fill = "Stage") +
  mytheme$standard()

print(p)
fn <- paste0(paste(today, "IsletType", "Boxplot", sep = "_"), ".png")
do.call(ggsave, c(list(fn, p), plotsave_param))
```

## A1.3: Compute ICIs vs IDIs via LMM

### A1.3.1: LMM across everything for explorative overview.
Differential expression between ICIs and IDIs by MM-model.
Note: few IDIs in mAAb+.

This is mostly explorative.
```{r diff_exprs-ici-idi}
## ICI images.
ici_images <- cols_islet |> 
  filter(islet_type == "ICI") |>
  pull(image_id) |> unique()

## Select mAAb+ and RecentOnset
cols_islet_oi <- cols_islet |> 
  filter(donor_type %in% c("mAAb+", "RecentOnset")) |> 
  tidyr::pivot_longer(cols = c(all_of(channels)), names_to = "channel", values_to = "MeanExprs") |>
  dplyr::mutate(islet_type = factor(islet_type, levels = c("ICI", "IDI"))) |> 
  dplyr::select(case_id, image_id, cell_type, donor_type, channel, islet_type, MeanExprs)

## Run MM-model (without random slopes)
results <- cols_islet_oi |>
  ## nest per donor_type
  tidyr::nest(data = -c(donor_type, channel, cell_type)) |> 
  dplyr::mutate(test = purrr::map(data, ~ broom.mixed::tidy(lmerTest::lmer(data = ., MeanExprs ~ islet_type + (1 | case_id))))) |> 
  tidyr::unnest(test)  |> 
  filter(term == "islet_typeIDI") |>
  # log10 transform p-values
  dplyr::select(channel, donor_type, p.value, statistic, estimate, cell_type)

## Compare random slope + random intercept vs random intercept only. 
results2 <- cols_islet_oi |>
  ## nest per donor_type
  tidyr::nest(data = -c(donor_type, channel, cell_type)) |> 
  dplyr::mutate(test = purrr::map(data, ~ 
    broom.mixed::tidy(lmerTest::lmer(data = ., MeanExprs ~ islet_type + (1 + islet_type | case_id))))) |> 
  tidyr::unnest(test)  |> 
  filter(term == "islet_typeIDI") |>
  # log10 transform p-values
  dplyr::select(channel, donor_type, p.value, statistic, estimate, cell_type)
```

-> random slopes should not be included. 
-> stay with random intercept only model for further analysis.


### A1.3.2: Plot results.

Overview.
Macrophages, Delta, Alpha, Endothelial cells.

```{r ici-idi-diffexp-plot}
channels_macro <- c("ADAR", "C3", "ATPIF1", "CD155", "CAIX", "CD9", 
                    "HLA_ABC", "B2M", "NFkB", "Ki67",
                    "XBP1", "ERN1", "TXNIP", "WFS1")
metadata(spe)$channels$Macrophage <- channels_macro

purrr::map(c("Macrophage", "Delta", "Alpha", "Endothelial"), \(ct) {
  p <- results |>
    filter(cell_type == !!ct) |> 
    filter(abs(statistic) > 1) |>
    filter(channel %in% metadata(spe)$channels[[!!ct]]) |> 
    ## Plot:
    ggplot(aes(x = channel, y = statistic, fill = donor_type)) +
    geom_hline(yintercept = 2, linetype = "dashed") +
    geom_hline(yintercept = -2, linetype = "dashed") +
    geom_bar(stat = "identity", position = "dodge") +
    scale_fill_manual(values = palettes[["stages"]]) +
    labs(x = "Channel", y = "t-statistic", fill = "Stage") +
    theme(axis.text.x = element_text(angle = 90, hjust = 1)) +
    mytheme$standard() + 
    labs(title = ct)
  print(p)
  fn <- paste0(paste(today, ct, "ici_idi_diffexp", sep = "_"), ".png")
  do.call(ggsave, c(list(fn, p), plotsave_param))
})

purrr::map(c("Macrophage", "Delta", "Alpha", "Endothelial"), \(ct) {
  p <- results2 |>
    filter(cell_type == !!ct) |> 
    filter(abs(statistic) > 1) |>
    filter(channel %in% metadata(spe)$channels[[!!ct]]) |> 
    ## Plot:
    ggplot(aes(x = channel, y = statistic, fill = donor_type)) +
    geom_hline(yintercept = 2, linetype = "dashed") +
    geom_hline(yintercept = -2, linetype = "dashed") +
    geom_bar(stat = "identity", position = "dodge") +
    scale_fill_manual(values = palettes[["stages"]]) +
    labs(x = "Channel", y = "t-statistic", fill = "Stage") +
    theme(axis.text.x = element_text(angle = 90, hjust = 1)) +
    mytheme$standard() + 
    labs(title = paste(ct, " (Random Slope)"))
  print(p)
  fn <- paste0(paste(today, ct, "ici_idi_diffexp_randomsl", sep = "_"), ".png")
  do.call(ggsave, c(list(fn, p), plotsave_param))
})
```

Results: 
Alpha-cells:
ICI: Inflammation (MHC-I, MX1, SYP).
IDI: Homeostasis (GCG (!) + ProGCG (!), Islet TFs / all other nuclear markers).

ER-stress: 
-> *Higher* rates of ERN1 (sig.), TXNIP (sig.); WFS1, XBP1 (ns)!
-> lower rates of NFkB (ns).

Delta-cells:
Similar results; but SST loss actually for IDI vs ICI. 

Macrophages:
- ICI: MHC-I upregulation (p < -26); 
- Inflammation/Stress: CAIX, CD155, NFkB, TXNIP, XBP1 upregulation (M1-like).
- IDI: C3 only. -> more homeostatic role?

Plot stress markers + ER-stress markers for Alpha/Delta.

```{r}
features_oi <- c("ERN1", "XBP1", "WFS1", "TXNIP", "CD155", "NFkB")

purrr::map(c("Macrophage", "Delta", "Alpha", "Endothelial"), \(ct) {
  p <- results2 |>
    filter(cell_type == !!ct) |> 
    filter(abs(statistic) > 1) |>
    filter(channel %in% metadata(spe)$channels[[!!ct]]) |> 
    ## Plot:
    ggplot(aes(x = channel, y = statistic, fill = donor_type)) +
    geom_hline(yintercept = 2, linetype = "dashed") +
    geom_hline(yintercept = -2, linetype = "dashed") +
    geom_bar(stat = "identity", position = "dodge") +
    scale_fill_manual(values = palettes[["stages"]]) +
    labs(x = "Channel", y = "t-statistic", fill = "Stage") +
    theme(axis.text.x = element_text(angle = 90, hjust = 1)) +
    mytheme$standard() + 
    labs(title = paste(ct, " (Random Slope)"))
  print(p)
  fn <- paste0(paste(today, ct, "ici_idi_diffexp_randomsl", sep = "_"), ".png")
  do.call(ggsave, c(list(fn, p), plotsave_param))
})
```



## A1.4: Plot ICI comparisons (celltype-specific)

### A1.4.1: Alpha-cells LMMs 
Compare ICI vs IDI for Alpha-cells.

**NOTE: automatically only computes for Recent-onset!**
-> too few IDIs for mAAb+ cases.

```{r ici-idi-alphas}
## Homeostatic markers.
p <- plot_ici_insulitic_expression(df = cols_islet, features_oi = c("HLA_ABC", "ECad", "GCG", "ProGCG"), 
                                   cell_types = "Alpha",
                                   cur_var = "islet_type", x_var = "islet_type",
                                   group1 = "ICI", group2 = "IDI", facet_by = "channel", 
                                   y_label = "Alpha-cell expression", 
                                   cur_palette = palettes$colors[3:4]) + 
                                   geom_line(aes(group = case_id), alpha = 0.5) + 
                                   theme(legend.position = "none")

fn <- paste0(paste(today, "Boxplot", "Alpha", "ICI_IDI", "HLA_ABC", sep = "_"), ".png")
do.call(ggsave, c(list(fn, p), plotsave_param))
print(p)

## ER-stress + functional markers.
cols_islet <- cols_islet |> dplyr::rename("IRE1a-P" = ERN1)

p <- plot_ici_insulitic_expression(df = cols_islet, 
                                  features_oi = c("IRE1a-P", "XBP1", "WFS1", "TXNIP", "CD155", "NFkB", "HLA_ABC", "GCG"), 
                                   cell_types = "Alpha",
                                   cur_var = "islet_type", x_var = "islet_type",
                                   group1 = "ICI", group2 = "IDI", facet_by = "channel", 
                                   y_label = expression(alpha ~ " -cell normalized counts"),
                                   cur_palette = palettes$colors[3:4]) + 
                                   geom_line(aes(group = case_id), alpha = 0.5) +
                                   theme(legend.position = "none")

fn <- paste0(paste(today, "Boxplot", "Alpha", "ICI_IDI", "ER_stress", sep = "_"), ".png")
do.call(ggsave, c(list(fn, p), plotsave_param))
print(p)

saveRDS(p, file.path(paths$home_git, "figures", "extfig4i.rds"))

q <- cols_islet |> 
  filter(cell_type == "Alpha") |> 
  select(image_id, case_id, cell_type, donor_type, islet_type, ProGCG) |> 
  group_by(donor_type, case_id, islet_type) |>
  summarise(ProGCG = mean(ProGCG), .groups = "drop") |>
  filter((donor_type == "NoDiabetes" & islet_type == "ICI") | 
         (donor_type == "sAAb+" & islet_type == "ICI") |
         (donor_type == "mAAb+" & islet_type == "ICI") |
         (donor_type == "RecentOnset" & islet_type == "ICI") |
         (donor_type == "RecentOnset" & islet_type == "IDI") |
         (donor_type == "LongDuration" & islet_type == "IDI")) |>
  mutate(donor_type = factor(donor_type, levels = c("NoDiabetes", "sAAb+", "mAAb+", "RecentOnset", "LongDuration"))) |> 

  ggplot(aes(x = donor_type, y = ProGCG, fill = islet_type)) +
  geom_boxplot(aes(fill = donor_type), position = position_dodge()) +
  geom_point() + 
  labs(x = "Stage", y = "ProGCG expression", color = "Islet type") +
  mytheme$standard()
print(q)
# SST very variable across stages.
```

### A1.4.2: Delta-cells LMMs.

Unexpectedly, SST higher in ICIs.
ER-stress markers show same pattern as for Alpha-cells.

```{r delta-cells-icis-lmms}
p <- plot_ici_insulitic_expression(df = cols_islet, features_oi = c("HLA_ABC", "ECad", "SST", "PCSK1"), 
                                   cell_types = "Delta",
                                   cur_var = "islet_type", x_var = "islet_type",
                                   group1 = "ICI", group2 = "IDI", facet_by = "channel", 
                                   y_label = "Delta-cell expression", 
                                   cur_palette = palettes$colors[3:4]) + 
                                   geom_line(aes(group = case_id), alpha = 0.5) + 
                                   theme(legend.position = "none")

fn <- paste0(paste(today, "Boxplot", "Delta", "ICI_IDI", "HLA_ABC", sep = "_"), ".png")
do.call(ggsave, c(list(fn, p), plotsave_param))
print(p)

p <- plot_ici_insulitic_expression(df = cols_islet, 
                                   features_oi = c("IRE1a-P", "XBP1", "WFS1", "TXNIP", "CD155", "NFkB"), 
                                   cell_types = "Delta",
                                   cur_var = "islet_type", x_var = "islet_type",
                                   group1 = "ICI", group2 = "IDI", facet_by = "channel", 
                                   y_label = "Delta-cell expression", 
                                   cur_palette = palettes$colors[3:4]) + 
                                   geom_line(aes(group = case_id), alpha = 0.5) + 
                                   theme(legend.position = "none")

fn <- paste0(paste(today, "Boxplot", "Delta", "ICI_IDI", "ER_stress", sep = "_"), ".png")
do.call(ggsave, c(list(fn, p), plotsave_param))
print(p)

q <- cols_islet |> 
  filter(cell_type == "Delta") |> 
  select(image_id, case_id, cell_type, donor_type, SST) |> 
  group_by(donor_type, case_id) |>
  summarise(SST = mean(SST)) |>
  mutate(donor_type = factor(donor_type, levels = c("NoDiabetes", "sAAb+", "mAAb+", "RecentOnset", "LongDuration"))) |> 
  ggplot(aes(x = donor_type, y = SST, fill = donor_type)) +
  geom_boxplot(aes(fill = donor_type)) +
  geom_point() + 
  scale_fill_manual(values = palettes[["stages"]][1:5]) +
  labs(x = "Stage", y = "SST expression", color = "Islet type") +
  mytheme$standard()
print(q)
# SST very variable across stages.
```

## A1.5: ICI: peri-islet only (Macrophages, Endothelial, T-cells)

### A1.5.1: Prepare Data

Select only intra-islet + peri-islet cells. (p < -20).
```{r islet-resident-cells}
channels <- c("B2M", "C3", "CAIX", "CD155", "CD9", "HLA_ABC", "Ki67", "MX1", "NFkB",
            "PDL1", "XBP1", "TXNIP", "WFS1", "ERN1", "ATPIF1", "ADAR", "VIM", "CD163",
            "GCG", "ProGCG", "SST", "PCSK1")

cols_islet <- scuttle::makePerCellDF(spe, features = channels, assay.type = "exprs", use.coldata = TRUE, use.dimred = FALSE) |> 
  as_tibble() |> 
  filter(cell_type %in% c("Endothelial", "Tcell", "Macrophage", "Alpha")) |> 
  select(case_id, image_id, cell_type, distance_to_islet, donor_type, all_of(channels)) |> 
  filter(distance_to_islet < -20) |> 
  group_by(image_id, case_id, cell_type, donor_type) |>
  summarise(across(all_of(channels), mean), .groups = "drop")

## Join with islet composition.  
cols_islet <- cols_islet |>
  left_join(compositions, by = "image_id") |> 
  filter(islet_type %in% c("ICI", "IDI")) |> tidyr::drop_na()

## Select mAAb+ and RecentOnset
test_cols_islet <- cols_islet |> 
  filter(donor_type %in% c("mAAb+", "RecentOnset")) |> 
  tidyr::pivot_longer(cols = c(all_of(channels)), names_to = "channel", values_to = "MeanExprs") |>
  dplyr::mutate(islet_type = factor(islet_type, levels = c("ICI", "IDI"))) |> 
  dplyr::select(case_id, image_id, cell_type, donor_type, channel, islet_type, MeanExprs)
```

### A1.5.2: Run LMMs
```{r lmm-peri-islet}
## Run MM-model
results <- test_cols_islet  |> 
  ## nest per donor_type
  tidyr::nest(data = -c(donor_type, channel, cell_type)) |> 
  dplyr::mutate(test = purrr::map(data, ~ broom.mixed::tidy(lmerTest::lmer(data = ., MeanExprs ~ islet_type + (1 | case_id))))) |> 
  tidyr::unnest(test)  |> 
  filter(term == "islet_typeIDI") |>
  dplyr::select(channel, donor_type, p.value, statistic, estimate, cell_type)
```

### A1.5.3: Plot LMM results

```{r mm-results-islet-resident} 
metadata(spe)$channels$Tcell <- c("CD3", 
                                  "B2M", "HLA_ABC", "NFkB", "ADAR", "MX1",
                                  "CD155", "CD9", "Ki67", "ATPIF1", 
                                  "XBP1", "TXNIP", "WFS1", "ERN1")

## Plot results
purrr::map(c("Macrophage", "Endothelial", "Tcell"), 
            \(ct){
  p <- results |> 
    filter(cell_type == !!ct) |> 
    filter(p.value < 0.05) |> 
    filter(channel %in% metadata(spe)$channels[[!!ct]]) |>
    ggplot(aes(x = channel, y = statistic, fill = donor_type)) +
    geom_hline(yintercept = 2, linetype = "dashed") +
    geom_hline(yintercept = -2, linetype = "dashed") +
    geom_bar(stat = "identity", position = "dodge") +
    scale_fill_manual(values = palettes[["stages"]]) +
    labs(x = "Channel", y = "t-statistic", fill = "Stage") +
    theme(axis.text.x = element_text(angle = 90, hjust = 1)) +
    mytheme$standard() + 
    labs(title = ct)
  print(p)
  fn <- paste0(paste(today, ct, "islet_resident", sep = "_"), ".png")
  do.call(ggsave, c(list(fn, p), plotsave_param))
})

## Results: 
## Islet-resident Endothelial ICI -> Upregulation MHC-I. 
## Islet-resident ENdothelial IDI (nothing jumps out).

## Peris-Islet-resident Macrophages: 
## Similarly multiple ER-sterss markers + inflammation markers 
## upregulated in IDI.

# -> Note MX1 much higher in ICI in Immune Cells as well. 
## BUT: again across progression no major differences.
cols_islet |> 
  filter(cell_type == "Macrophage") |>
  group_by(donor_type) |>
  summarise(mean(MX1), .groups = "drop")
```

### A1.5.4: Plot ICI vs IDI for Peri-islet Macrophages
Compare ICI vs IDI for Peri-islet Macorphages.
```{r peri-islet-macrophages}
cols_islet

p <- plot_ici_insulitic_expression(df = cols_islet, features_oi = c("HLA_ABC", "NFkB", "CD155", "CD163"), 
                                   cell_types = "Macrophage",
                                   cur_var = "islet_type", x_var = "islet_type",
                                   group1 = "ICI", group2 = "IDI", facet_by = "channel", 
                                   y_label = "Macrophage expression", 
                                   cur_palette = palettes$colors[3:4]) + 
                                   geom_line(aes(group = case_id), alpha = 0.5) + 
                                   theme(legend.position = "none")
print(p)
fn <- paste0(paste(today, "Boxplot", "Macrophages", "ICI_IDI", "HLA_ABC", sep = "_"), ".png")
do.call(ggsave, c(list(fn, p), plotsave_param))

extfig9c <- p
saveRDS(extfig9c, file.path(paths$home_git, "figures", "extfig9c.rds"))
```


# **A2: Insulitis**
- Insulitis is defined as the presence of 6 >= CD3+ (T-cells) immune cells within the islets.
- Progression: 
  - CTRL -> Inflammation ICI -> Insulitic ICI -> Insulitic IDI -> IDI

- Before we compared: Inflammation ICI (containing few Insulitic ICI) vs Inflammation IDI
- NOW: Inflammation ICI vs Insulitic ICI.

-> Now we hope to identify Insulitic ICI SPECIFIC markers.

## A2.1: Prepare Data.
```{r insulitic}
## Define Insulitic Islets.
channels_islet <- unique(c(metadata(spe)$channels$Endothelial, 
                          metadata(spe)$channels$Beta, metadata(spe)$channels$Alpha,
                          metadata(spe)$channels$Delta, metadata(spe)$channels$Macrophage))

## ROI-level Data including ICI information.
exprs_df <- as_tibble(scuttle::makePerCellDF(spe, features = channels_islet, assay.type = "exprs", use.dimred = FALSE)) |> 
  select(all_of(channels_islet), case_id, donor_type, image_id, cell_type, distance_to_islet) |> 
  filter(donor_type != "Pancreatitis") |> 
  group_by(image_id, case_id, donor_type) |>
  mutate(islet_type = if_else(sum(cell_type == "Beta") > 0, "ICI", "IDI")) |> 
  mutate(insulitic = if_else(sum(cell_type == "Tcell" & distance_to_islet >= 0) >= 6, "insulitis", "non-insulitis")) |> 
  ungroup() |> 
  mutate(donor_type = factor(donor_type, levels = c("NoDiabetes", "sAAb+", "mAAb+", "RecentOnset", "LongDuration")))

fn <- file.path(paths$folder_out, "13_MarkerExpression_cells_Immune", "insulitic_images.rds")
if (file.exists(fn)) {
  image_ids <- readRDS(fn)
  image_ids <- gsub("_Immune_", "_Islet_", image_ids)
  exprs_df <- exprs_df |> 
    mutate(insulitic = if_else(image_id %in% image_ids, "insulitis", "non-insulitis"))
}
```

## A2.2: Expression changes between Insulitic and Non-Insulitic Islets:

```{r insulitic-vs-non-insulitic}
### Prepare the data
immune_insulitic <- exprs_df |> 
  filter(cell_type %in% c("Alpha", "Delta", "Endothelial", "Macrophage", "Beta", "Tcell"))

# Take only PERI_ISLET + ICI + INSULITIC Ones.
stats <- immune_insulitic  |> 
  filter(donor_type == "RecentOnset", islet_type == "ICI") |>
  filter(distance_to_islet >= -20) |>
  group_by(image_id, insulitic, case_id, cell_type) |>
  summarize(across(channels_islet, \(x) mean(x, na.rm = TRUE)), .groups = "drop") |>
  tidyr::pivot_longer(cols = all_of(channels_islet), names_to = "channel", values_to = "value") |>
  tidyr::nest(data = -c(channel, cell_type)) |>
  dplyr::mutate(test = purrr::map(data, ~ broom.mixed::tidy(lmerTest::lmer(data = ., value ~ insulitic + (1 | case_id))))) |>
  tidyr::unnest(test) |>
  filter(term != "(Intercept)", effect == "fixed") |>
  arrange(p.value) |> 
  mutate(p.adj = p.adjust(p.value, method = "fdr"))
```

## A2.3: Plot overview: Insulitic and Non-Insulitic Islets.

```{r plot-overview-non-insulitic}
## Plot - Islet:
palettes$celltypes <- c("Alpha" = palettes$colors[1], "Beta" = palettes$colors[2], 
                       "Delta" = palettes$colors[4])
p <- stats |> 
  filter(p.value < 0.1) |>
  filter(!channel %in% c("ARX", "PDL1")) |> 
  filter(!cell_type %in% c("Endothelial", "Macrophage", "Tcell")) |> 
  ggplot(aes(x = channel, y = statistic, fill = cell_type)) +
  geom_bar(stat = "identity", position = "dodge") +
  labs(x = "Channel", y = "Statistic", color = "Channel", title = "random_Int") +
  mytheme$standard() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  scale_fill_manual(values = palettes$celltypes)
print(p)
fn <- paste0(paste(today, "Insulitic_vs_NonInsulitic", "ICI", "Islets", sep = "_"), ".png")
do.call(ggsave, c(list(fn, p), plotsave_param))

### Plot - Non-islet cells:
palettes$celltypes <- c("Macrophage" = palettes$colors[6], "Endothelial" = palettes$colors[7], "Tcell" = palettes$colors[8]) 
p <- stats |> 
  filter(p.value < 0.1) |>
  filter(!channel %in% c("ARX", "PDL1")) |>
  filter(cell_type %in% c("Endothelial", "Macrophage", "Tcell")) |> 
  ggplot(aes(x = channel, y = statistic, fill = cell_type)) +
  geom_bar(stat = "identity", position = "dodge") +
  labs(x = "Channel", y = "Statistic", color = "Channel") +
  mytheme$standard() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  scale_fill_manual(values = palettes$celltypes)

print(p)
fn <- paste0(paste(today, "Insulitic_vs_NonInsulitic", "ICI", "Non_Islets", sep = "_"), ".png")
do.call(ggsave, c(list(fn, p), plotsave_param))
```

## A2.4: Insulitic. Cell-Types.

### A2.4.1: Beta-cells Insulitic vs Non-Insulitic ROIs.
Beta-cells Insulitic vs non-Insulitic ROIs.

Figure 2f.
```{r plot-fig-3f-beta}
## Compare between Insulitic and Non-insulitic ICIs.
insulitic_df <- immune_insulitic |> 
  mutate(insulitic = ifelse(insulitic == "insulitis", "Insulitic", "Non-Insulitic")) |>
  mutate(insulitic = factor(insulitic, levels = c("Insulitic", "Non-Insulitic")))

## Cases that have Insulitic Islets (paired).
insulitic_df |> 
  filter(islet_type == "ICI") |>
  distinct(insulitic, donor_type, case_id) |> 
  filter(insulitic == "Insulitic") |> 
  filter(donor_type == "RecentOnset") |>
  pull(case_id) -> cases_insulitic

to_del <- c("6324", "6533")

## 
p <- plot_ici_insulitic_expression(df = insulitic_df |> filter(case_id %in% cases_insulitic, !case_id %in% to_del, image_id %in% ici_images),
                                   features_oi = c("HLA_ABC", "MX1", "ADAR", "ECad"), 
                                   cell_types = "Beta", cur_var = "insulitic", x_var = "insulitic",
                                   group1 = "Insulitic", group2 = "Non-Insulitic", 
                                   facet_by = "channel",
                                   y_label = expression(beta ~"-cell expression"), 
                                   cur_palette = palettes$colors[1:2]) + 
                                   mytheme$standard_new() + 
                                   geom_line(aes(group = case_id), alpha = 0.5, color = "black") + 
                                   theme(legend.position = "none")

print(p)
fn <- paste0(paste(today, "Boxplot", "BetaCells", "Insulitic_vs_NonInsulitic", "HLA_ABC", sep = "_"), ".png")
do.call(ggsave, c(list(fn, p), plotsave_param))

fig2f <- p
saveRDS(fig2f, file.path(paths$home_git, "figures", "fig2f.rds"))
```

### A2.4.2: Islet-cells Insulitic vs Non-Insulitic ROIs.

We observed islet-wide upregulation of MX1, HLA-ABC, B2M in islet-cells,
but more specific beta-cell upregulation of ADAR in beta-cells.
(Non-significant in alpha-cells; and reduced effect size in delta-cells).
Thus, plot them for beta, alpha and delta cells.

```{r insulitic-islet-cells}
## Compare between Insulitic and Non-insulitic ICIs.
insulitic_df <- immune_insulitic |> 
  mutate(insulitic = ifelse(insulitic == "insulitis", "Insulitic", "Non-Insulitic")) |>
  mutate(insulitic = factor(insulitic, levels = c("Insulitic", "Non-Insulitic"))) |> 
  filter(islet_type == "ICI")

insulitic_df <- insulitic_df |> 
  filter(case_id %in% cases_insulitic, image_id %in% ici_images, !case_id %in% to_del)

p <- plot_ici_insulitic_expression(df = insulitic_df, 
                                   features_oi = c("HLA_ABC", "MX1", "ADAR", "ECad"), 
                                   cell_types = c("Alpha", "Delta"),
                                   cur_var = "insulitic", x_var = "cell_type",
                                    group1 = "Insulitic", group2 = "Non-Insulitic", facet_by = "channel",
                                    y_label = "Islet-cell expression", 
                                    cur_palette = palettes$colors[1:2]) + 
                                    mytheme$standard_new() +
                                    theme(legend.position = "none")

fn <- paste0(paste(today, "Boxplot", "IsletCells", "Insulitic_vs_NonInsulitic", sep = "_"), ".png")
do.call(ggsave, c(list(fn, p), plotsave_param))
print(p)
extfig4f <- p
saveRDS(extfig4f, file.path(paths$home_git, "figures", "extfig4f.rds"))
```

### A2.4.3: Beta-cells Insulitic vs Non-Insulitic ROIs - ER-stress

ER-stress markers. Only ERN1 non-significantly higher.

```{r insulitic-beta-cells-er}
p <- plot_ici_insulitic_expression(df = insulitic_df |> dplyr::filter(!case_id %in% to_del) |> dplyr::rename("IRE1a-P" = ERN1), 
                                    features_oi = c("IRE1a-P", "WFS1", "XBP1", "ECad", "TXNIP", "NFkB"), 
                                    cell_types = "Beta", cur_var = "insulitic", x_var = "insulitic",
                                    group1 = "Insulitic", group2 = "Non-Insulitic", facet_by = "channel",
                                    y_label = expression(beta ~"-cell mean expression"), 
                                    cur_palette = palettes$colors[1:2]) + 
                                    mytheme$standard_new() + 
                                    geom_line(aes(group = case_id), alpha = 0.5, color = "black") + 
                                    theme(legend.position = "none")
                                    
fn <- paste0(paste(today, "Boxplot", "BetaCells", "Insulitic_vs_NonInsulitic", "ER_stress", sep = "_"), ".png")
do.call(ggsave, c(list(fn, p), plotsave_param))
print(p)
extfig4h <- p
saveRDS(extfig4h, file.path(paths$home_git, "figures", "extfig4h.rds"))
```

### A2.4.4: Islet-cells: Insulitic vs Non-Insulitic ROIs - ER-stress

ER-stress markers. Only ERN1 non-significantly higher.

```{r insulitic-islet-cells-er}
p <- plot_ici_insulitic_expression(df = insulitic_df |> filter(!case_id %in% to_del), 
                                   features_oi = c("ERN1", "WFS1", "XBP1", "ECad", "TXNIP", "NFkB"), 
                                   cell_types = c("Alpha", "Delta"),
                                   cur_var = "insulitic", x_var = "cell_type",
                                   group1 = "Insulitic", group2 = "Non-Insulitic", facet_by = "channel",
                                   y_label = "Islet-cell expression", 
                                   cur_palette = palettes$colors[1:2]) + 
                                   mytheme$standard_new() + 
                                   theme(legend.position = "none")
                                    
fn <- paste0(paste(today, "Boxplot", "Islet", "Insulitic_vs_NonInsulitic", "ER_stress", sep = "_"), ".png")
do.call(ggsave, c(list(fn, p), plotsave_param))
print(p)
```

### A2.4.5: Beta-cells: Lineage markers.
Lineage markers are higher in insulitic islets. 
But this is actually driven mainly by outliers; no consistent effect.

```{r insulitic-beta-cells-er}
p <- plot_ici_insulitic_expression(df = insulitic_df |> filter(!case_id %in% to_del), 
                                    features_oi = c("ProINS", "INS", "PDX1", "NKX6_1"), 
                                    cell_types = "Beta", cur_var = "insulitic", x_var = "insulitic",
                                    group1 = "Insulitic", group2 = "Non-Insulitic", facet_by = "channel",
                                    y_label = expression(beta ~"-cell mean expression"), 
                                    cur_palette = palettes$colors[1:2]) + 
                                    mytheme$standard_new() + 
                                    geom_line(aes(group = case_id), alpha = 0.5, color = "black") + 
                                    theme(legend.position = "none")
                                    
fn <- paste0(paste(today, "Boxplot", "BetaCells", "Insulitic_vs_NonInsulitic", "Lineage_Markers", sep = "_"), ".png")
do.call(ggsave, c(list(fn, p), plotsave_param))
print(p)
```