---
title: "19_TcellsInsulitis_cells_Immune"
author: "Nathan Steenbuck"
date: "Created: 15 May, 2024; Compiled: `r format(Sys.time(), '%d %b, %Y')`"
output: workflowr::wflow_html
editor_options:
  chunk_output_type: inline
---

```{r setup, include=FALSE}
if (rstudioapi::isAvailable()) {
  script_name <- basename(rstudioapi::getSourceEditorContext()$path)
} else {
  script_name <- knitr::current_input()
}

cur_user <- Sys.info()[["user"]]
script_name <- "19_TcellsInsulitis_cells_Immune.Rmd"

if (cur_user == "ubuntu") {
    source(file.path("/", "mnt", "central_nas", "projects", 
    "type1_diabetes", "nathan", "T1D_Vol", "T1D_analysis", 
    "analysis", "helpers.R"))  #n_cores <- future::availableCores() - 1
    #n_cores <- future::availableCores() - 1
  n_cores <- 4
  future::plan(future::multicore(workers = n_cores))
  paths <- getPaths(script_name)
  knitr::opts_knit$set(root_dir = paths$home_data)
  do_print <- FALSE
} else {
  source(file.path("/", "home", "nsteen", "scripts", "T1D_analysis", "analysis", "helpers.R"))
  n_cores <- 8
  future::plan(future::multicore(workers = n_cores))
  paths <- getPaths(script_name)
  paths$folder_script <- paths$cluster_folder_script
  paths$folder_in <- paths$cluster_folder_in
  paths$folder_out <- paths$cluster_home 
  knitr::opts_knit$set(root_dir = paths$cluster_home)
  do_print <- TRUE
}

seed <- 123456
set.seed(seed)
options <- furrr::furrr_options(seed = seed)
paths$prev <- paste("08_CellTypesIslet", paths$object_type, paths$panel_type, sep = "_")
knitr::opts_chunk$set(echo = TRUE)
```


# **Goal**

Comparisons of T-cells in Insulitc and non-insulitic islets.

# A1: **Settings**

## A1.1: Load packages

```{r packages, results='hide'}
suppressPackageStartupMessages(c(
  library(data.table),
  library(dplyr),
  library(SpatialExperiment),
  library(scuttle),
  library(parallel),
  library(patchwork),
  library(tidyr),
  library(RSpectra),
  library(rstatix),
  library(ggpubr),
  library(emmeans)
))
```

## A1.2: Paths and settings

```{r settings}
# Paths
if (!dir.exists(paths$folder_script)) dir.create(paths$folder_script)
plotsave_param$path <- paths$folder_script
plotsave_param_large$path <- paths$folder_script

# Misc settings
today <- gsub("-", "", Sys.Date())
```

## A1.3: Read in the data

Load the SpatialExperiment (SPE) object saved at the previous step.

```{r load-data}
fn_spe <- file.path(paths$folder_out, paths$prev, paste0(paths$object_type, "_", paths$panel_type, ".rds"))
spe <- readRDS(fn_spe)
print(spe)
```

## A1.4: Read in Immune Subset only

```{r load_sce_imm}
sce_imm <- spe[, spe$cell_category == "immune"]
# fn_sceimm <- file.path(paths$folder_script, paste0("sce_imm", "_", paths$object_type, "_", paths$panel_type, ".rds"))
# saveRDS(sce_imm, fn_sceimm)
# sce_imm <- readRDS(fn_sceimm)
```

# **A2: Insulitis**

## A2.1: Define insulitis.
Define Insulitis as >= 6 infiltrating T-cells. 
Add DistScore + Infiltration Bins.
```{r add-inf-score}
## Prepare the data
quantiles <- rev(quantile(spe$distance_to_islet[
  spe$distance_to_islet <= 0], prob = c(0, 0.25, 0.5, 0.75, 0.89, 1)))
print(quantiles)

bins <- unname(round(quantiles))
distances <- c("inside", "< 20 um", "20-40 um", "> 40 um")
names(distances) <- RColorBrewer::brewer.pal(8, "Spectral")[c(1, 3, 4, 6)]
print(distances)

spe$infiltration <- ifelse(
  spe$distance_to_islet > bins[1], distances[1],
  ifelse(
    spe$distance_to_islet < bins[1] & spe$distance_to_islet >= bins[2], distances[2],
    ifelse(
      spe$distance_to_islet < bins[2] & spe$distance_to_islet >= bins[3], distances[3], distances[4])
    ))
spe$infiltration <- factor(spe$infiltration, levels = distances)

spe$DistScore = ifelse(spe$distance_to_islet >= -1, 1, 1 / abs(spe$distance_to_islet))
spe$DistScoreSqrt <- ifelse(spe$distance_to_islet >= -1, 1, sqrt(abs(spe$distance_to_islet)))
gc()
```

## A2.2: Define Insulitic Islets + ICIs
Define Insulitic Islets + ICIs. Filter out Pancreatitis
```{r label-insulitic}
channels <- metadata(spe)$channels
channels_imm <- unique(unlist(channels))

## Define Insulitic Islets + ICIs.
## Filter Pancreatitis
exprs_df <- as_tibble(scuttle::makePerCellDF(spe, features = channels_imm, assay.type = "exprs", use.dimred = FALSE)) |> 
  select(all_of(channels_imm), case_id, age, donor_type, DistScoreSqrt, DistScore, image_id, infiltration, cell_type, distance_to_islet, cell_cluster_scaled) |> 
  filter(donor_type != "Pancreatitis") |> 
  group_by(image_id, case_id, donor_type) |>
  mutate(islet_type = if_else(sum(cell_type == "Beta") > 0, "ICI", "IDI")) |> 
  mutate(tcell = if_else(cell_type %in% c("T_CD4", "T_CD8"), "T-cell", "non-T-cell")) |>
  ungroup() |> 
  mutate(donor_type = factor(donor_type, levels = c("NoDiabetes", "sAAb+", "mAAb+", "RecentOnset", "LongDuration")))

icis <- exprs_df |> filter(islet_type == "ICI") |> pull(image_id) |> unique()

## Get Ins. Islets
fn <- file.path(paths$folder_out, "13_MarkerExpression_cells_Immune", "insulitic_images.rds")

if (file.exists(fn)) {
  image_ids <- readRDS(fn)
  exprs_df <- exprs_df |> 
    mutate(insulitic = if_else(image_id %in% image_ids, "insulitis", "non-insulitis"))
}
```

# A3.: Islet-Infiltration and Insulitic Comparisons

## A3.1: Cell-Type Level

Compare the Infiltration Scores of Cell Types between Insulitic Islet and non-insulitic islets.
Only ICIs.

```{r infiltration-scores-cell-type-insulitic}
dist_score <- metadata(spe)$celltype_compo_score |> as_tibble() |> 
  select(image_id, case_id, donor_type, starts_with("NormScoreArea_"), starts_with("AbsScore")) |> 
  pivot_longer(cols = starts_with("NormScoreArea_"), 
               names_to = "cell_type", values_to = "NormScoreArea") |>
  mutate(cell_type = stringr::str_remove(cell_type, "NormScoreArea_")) |>
  pivot_longer(cols = starts_with("AbsScore"), 
               names_to = "cell_type2", values_to = "DistScore") |> 
  mutate(cell_type2 = stringr::str_remove(cell_type2, "AbsScore_")) |> 
  filter(cell_type == cell_type2) |> 
  mutate(insulitic = ifelse(image_id %in% image_ids, "insulitis", "non-insulitis"))

## select only cases with both insulitic and non-insulitic islets. (PAIRED ANALYSIS)
case_ids <- dist_score |> 
  dplyr::distinct(case_id, insulitic) |>
  dplyr::count(case_id) |> filter(n == 2) |> pull(case_id)

## Calculate P-value
p_vals <- dist_score |> 
  filter(case_id %in% case_ids) |>
  filter(donor_type %in% c("mAAb+", "RecentOnset")) |>
  filter(image_id %in% icis) |>
  filter(cell_type %in% c("T_CD4", "T_CD8", "Myeloid", "T_DN", "NK", "B", "Neutrophil")) |>
  tidyr::nest(data = -c(cell_type)) |> 
  dplyr::mutate(test = purrr::map(data, ~ broom.mixed::tidy(lmerTest::lmer(data = ., DistScore ~ insulitic + (1 | case_id))))) |> 
  tidyr::unnest(test) |>
  filter(term != "(Intercept)", effect == "fixed") |> 
  mutate(p.adj = p.adjust(p.value, method = "fdr")) |> 
  arrange(p.adj) |> 
  mutate(insulitic = stringr::str_remove(term, "insulitic")) |> 
  mutate(significance = cut(p.adj, breaks = c(-Inf, 0.001, 0.01, 0.05, Inf),
                            labels = c("***", "**", "*", "")))

 # get max-y-value PER channel.
max_values <- dist_score|> 
  group_by(cell_type) |> 
  summarise(p_vals_pos = max(DistScore)*0.9)
 # Join with p-vals
p_vals <- p_vals |> left_join(max_values, by = "cell_type")

## Plot
p <- dist_score |> 
  filter(cell_type %in% c("T_CD4", "T_CD8", "Myeloid", "T_DN", "NK", "B", "Neutrophil")) |>
  ggplot(aes(x = insulitic, y = DistScore, fill = insulitic)) +
  geom_boxplot() +
  labs(x = "Cell Type", y = "DistScore", color = "Donor Type") +
  mytheme$standard() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) + 
  facet_wrap(~cell_type, scales = "free_y")

p <- p + geom_text(data = p_vals, 
    aes(x = insulitic, y = p_vals_pos, label = significance), 
     size = 10) + 
facet_wrap(~cell_type, scales = "free_y")

print(p)
fn <- paste0(paste(today, "Average_DistScore_CellType", sep = "_"), ".png")
do.call(ggsave, c(list(fn, p), plotsave_param))
```

-> T-cells are obviously highly significant; given that they were used to define insulitis.
-> Myeloid >> NK > B_cells (because rare) > Neutrophils (barely)

## A3.2: Cell Cluster Level - GLOBAL OVERVIEW.
Compare DistScore of Insulitic vs Non-Insulitic Islets between Cell Clusters.
```{r infiltration-scores-cell-clusters-insulitic}
## Get DistScore for Cell Clusters.
dist_score_cluster <- metadata(spe)$cluster_compo_score[[1]] |> as_tibble() |> 
  select(image_id, case_id, donor_type, starts_with("NormScoreArea_"), starts_with("AbsScore")) |> 
  pivot_longer(cols = starts_with("NormScoreArea_"), 
               names_to = "cell_cluster_scaled", values_to = "NormScoreArea") |>
  mutate(cell_cluster_scaled = stringr::str_remove(cell_cluster_scaled, "NormScoreArea_")) |>
  pivot_longer(cols = starts_with("AbsScore"), 
               names_to = "cell_cluster_scaled2", values_to = "DistScore") |> 
  mutate(cell_cluster_scaled2 = stringr::str_remove(cell_cluster_scaled2, "AbsScore_")) |> 
  filter(cell_cluster_scaled == cell_cluster_scaled2) |> 
  mutate(insulitic = ifelse(image_id %in% image_ids, "insulitis", "non-insulitis")) |> 
  mutate(cell_type = stringr::str_remove(cell_cluster_scaled, "_[0-9]+$")) |> 
  mutate(cell_type = ifelse(cell_cluster_scaled == "T-naive", "T_CD8", cell_type)) |> 
  mutate(cell_type = ifelse(cell_cluster_scaled == "T-EMRA", "T_CD8", cell_type))

dist_score_cluster |> distinct(cell_type)

## Calculate P-value (OVERVIEW).
p_full <- dist_score_cluster |> 
  filter(donor_type %in% c("mAAb+", "RecentOnset")) |>
  filter(image_id %in% icis) |>
  tidyr::nest(data = -c(cell_cluster_scaled)) |> 
  dplyr::mutate(test = purrr::map(data, ~ broom.mixed::tidy(lmerTest::lmer(data = ., NormScoreArea ~ insulitic + (1 | case_id))))) |> 
  tidyr::unnest(test) |> 
  mutate(p.adj = p.adjust(p.value, method = "fdr")) |> 
  arrange(p.adj) |> 
  filter(term != "(Intercept)", effect == "fixed") |> 
  mutate(insulitic = stringr::str_remove(term, "insulitic")) |> 
  mutate(cell_type = stringr::str_remove(cell_cluster_scaled, "_[0-9]+$")) |> 
  mutate(cell_type = ifelse(cell_cluster_scaled == "T-naive", "T_CD8", cell_type)) |> 
  mutate(cell_type = ifelse(cell_cluster_scaled == "T-EMRA", "T_CD8", cell_type))

## Iterate through Cell Types:
celltypes_oi <- c("T_CD4", "T_CD8", "Myeloid", "NK")
purrr::map(celltypes_oi, \(cur_ct){
  message("Plotting DistScore for Cell Type: ", cur_ct)
  p_vals <- p_full |> filter(cell_type == !!cur_ct)
  p_vals$significance <- cut(p_vals$p.adj,
                            breaks = c(-Inf, 0.001, 0.01, 0.05, Inf),
                            labels = c("***", "**", "*", ""))
  # get max-y-value PER channel.
  max_values <- dist_score_cluster |> filter(cell_type == !!cur_ct) |>
    group_by(cell_cluster_scaled) |> 
    summarise(p_vals_pos = max(DistScore)*0.9)
  # Join with p-vals 
  p_vals <- p_vals |> left_join(max_values, by = "cell_cluster_scaled")
  p <- dist_score_cluster |> 
    filter(cell_type %in% !!cur_ct) |>
    ggplot(aes(x = insulitic, y = DistScore, fill = insulitic)) +
    geom_boxplot() +
    labs(x = "Cell Cluster", y = "DistScore", color = "Donor Type") +
    mytheme$standard() +
    theme(axis.text.x = element_text(angle = 45, hjust = 1)) + 
    facet_wrap(~cell_cluster_scaled, scales = "free_y") + 
    scale_fill_manual(values = palettes$colors[1:2])

  p <- p + geom_text(data = p_vals, 
    aes(x = insulitic, y = p_vals_pos, label = significance), 
     size = 10) +
    facet_wrap(~cell_cluster_scaled, scales = "free_y")
  print(p)

  fn <- paste0(paste(today, "Average_DistScore_Insulitis", cur_ct, sep = "_"), ".png")
  do.call(ggsave, c(list(fn, p), plotsave_param))
})
```

## A3.3: Specific Myeloid Cell Types. 
### A3.4: Myeloid-3 and Myeloid-4

Plot this plot with Myeloid-3 and Myeloid-4.
**ONLY INCLUDES paired comparisons!**
Show increase in DistScore for Myeloid-3 and Myeloid-4 in Insulitic Islets.

```{r prepare_data}
input_df <- dist_score_cluster |> 
  filter(donor_type %in% c("mAAb+", "RecentOnset")) |>
  filter(cell_cluster_scaled %in% c("Myeloid_3", "Myeloid_4", "Myeloid_5")) |>
  filter(image_id %in% icis)

## My3 cases.
case_ids_my3 <- input_df |> 
  filter(cell_cluster_scaled == "Myeloid_3") |>
  filter(NormScoreArea > 0) |>
  dplyr::distinct(case_id, insulitic) |>
  dplyr::count(case_id) |> filter(n == 2) |> pull(case_id) |> unique()
length(case_ids_my3) ## N = 23

## My4 cases.
case_ids_my4 <- input_df |> 
  filter(cell_cluster_scaled == "Myeloid_4") |>
  filter(NormScoreArea > 0) |> 
  dplyr::distinct(case_id, insulitic) |>
  dplyr::count(case_id) |> filter(n == 2) |> pull(case_id) |> unique()

length(case_ids_my4) ## N = 11

## My5 cases.
case_ids_my5 <- input_df |> 
  filter(cell_cluster_scaled == "Myeloid_5") |>
  filter(NormScoreArea > 0) |> 
  dplyr::distinct(case_id, insulitic) |>
  dplyr::count(case_id) |> filter(n == 2) |> pull(case_id) |> unique()
length(case_ids_my5) ## N = 23
```

Plot Myeloid 3 + Myeloid-4.
```{r}
case_stages <- colData(spe)[, c("case_id", "age")] |> as_tibble() |> distinct() |> 
  mutate(age_group = ifelse(age < 13, "< 13", ">= 13"))

input_df2 <- input_df |> 
  inner_join(case_stages, by = "case_id") |>
  filter((cell_cluster_scaled == "Myeloid_3" & case_id %in% case_ids_my3) |
         (cell_cluster_scaled == "Myeloid_4" & case_id %in% case_ids_my4))  |> 
  filter(donor_type == "RecentOnset")

p_vals <- input_df2  |>
  tidyr::nest(data = -c(cell_cluster_scaled)) |>
  dplyr::mutate(test = purrr::map(data, 
    ~ broom.mixed::tidy(lmerTest::lmer(data = ., log1p(NormScoreArea) ~ insulitic + age_group +  (1 | case_id))))) |>
  tidyr::unnest(test) |> 
  filter(term != "(Intercept)", effect == "fixed", term != "age_group>= 13") |> 
  mutate(.y. = "NormScoreArea",
          p = p.value,
          p.adj = p.adjust(p, method = "fdr"),
          p.signif = cut(p.adj, breaks = c(-Inf, 0.001, 0.01, 0.05, Inf), labels = c("***", "**", "*", ""))
          ) |> 
  select(.y., p, p.adj, p.signif, cell_cluster_scaled)

# Final plot.
plot_df <- input_df2 |> 
  filter(cell_cluster_scaled %in% c("Myeloid_3", "Myeloid_4")) |>
  group_by(cell_cluster_scaled, case_id, insulitic, age_group) |> 
  summarize(NormScoreArea = mean(NormScoreArea, na.rm = TRUE), .groups = "drop") |> 
  mutate(NormScoreArea = log1p(NormScoreArea))

# get max-y-value PER channel.
p_vals <- plot_df |> 
  filter(cell_cluster_scaled %in% c("Myeloid_3", "Myeloid_4")) |>
  group_by(cell_cluster_scaled) |> 
  summarise(p_vals_pos = log1p(max(NormScoreArea))) |> 
  right_join(p_vals, by = "cell_cluster_scaled")

# Join with p-vals.
stat.test <- plot_df |> 
  group_by(cell_cluster_scaled) |> 
  rstatix::t_test(
    NormScoreArea ~ insulitic, paired = TRUE,
  ) |> 
    add_xy_position(x = "insulitic", step.increase = 0.07, dodge = 0.8) |> 
    distinct() |> 
    select(cell_cluster_scaled, group1, group2, y.position)

stat.test.test <- p_vals |> 
  left_join(stat.test, by = "cell_cluster_scaled")
stat.test.test

p <- 
  ggpubr::ggviolin(plot_df, x = "insulitic", y = "NormScoreArea", 
                   fill = "insulitic", add = "point", facet.by = "cell_cluster_scaled",
                   draw_quantiles = c(0.25, 0.5, 0.75), 
          add.params = list(alpha = 1, size = 1.5)) +
  stat_pvalue_manual(stat.test.test, label = "p.signif", tip.length = 0.05,
                      size = 10, label.size = 10, vjust = 0.5) + 
  labs(x = "", y = "Log: Infiltration Score", color = "Donor Type") +
  mytheme$standard_new() +
  facet_wrap(~cell_cluster_scaled, scales = "free_y",
             labeller = labeller(cell_cluster_scaled = c("Myeloid_3" = "Act. M1/M2-like",
                                                         "Myeloid_4" = "cDC"))) + 
  scale_fill_manual(values = palettes$colors[1:2]) + 
  scale_x_discrete(labels = c("insulitis" = "Insulitic", "non-insulitis" = "Non-Insulitic")) + 
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) + 
  theme(
        axis.text.x = element_text(angle = 45, hjust = 1, size = 22),
        axis.text.y = element_text(size = 22),
        axis.title.x = element_text(size = 25),
        axis.title.y = element_text(size = 25),
        strip.text = element_text(size = 25)) + 
  geom_line(aes(group = case_id), alpha = 0.5)
print(p)

fn <- paste0(paste(today, "Average_DistScore_Insulitis_Myeloid3_Myeloid4", sep = "_"), ".png")
do.call(ggsave, c(list(fn, p), plotsave_param))

extfig9d <- p
saveRDS(extfig9d, file.path(paths$home_git, "figures", "extfig9d.rds"))
```

Plot Myeloid 5. NOTE: AGE-GROUP is significant here as co-variate (*).
```{r}
input_df2 <- input_df |> 
  filter((cell_cluster_scaled == "Myeloid_5" & case_id %in% case_ids_my5)) |> 
  inner_join(case_stages, by = "case_id")  |> 
  mutate(insulitic = ifelse(insulitic == "insulitis", "Insulitic", "Non-Insulitic")) |> 
  mutate(cell_cluster_scaled = ifelse(cell_cluster_scaled == "Myeloid_5", "CD11c+ Macrophages", cell_cluster_scaled)) |> 
  filter(donor_type == "RecentOnset")

p_vals <- input_df2  |>
  tidyr::nest(data = -c(cell_cluster_scaled)) |>
  dplyr::mutate(test = purrr::map(data, 
    ~ broom.mixed::tidy(lmerTest::lmer(data = ., log1p(NormScoreArea) ~ insulitic + age_group + (1 | case_id))))) |>
  tidyr::unnest(test) |> 
  filter(term != "(Intercept)", effect == "fixed", term != "age_group>= 13") |> 
  mutate(.y. = "NormScoreArea",
          p = p.value,
          p.adj = p.adjust(p, method = "fdr"),
          p.signif = cut(p.adj, breaks = c(-Inf, 0.001, 0.01, 0.05, Inf), labels = c("***", "**", "*", ""))
          ) |> 
  select(.y., p, p.adj, p.signif, cell_cluster_scaled)

# Final plot.
plot_df <- input_df2 |> 
  filter(cell_cluster_scaled %in% c("CD11c+ Macrophages")) |>
  group_by(cell_cluster_scaled, case_id, insulitic) |> 
  summarize(NormScoreArea = mean(NormScoreArea, na.rm = TRUE), .groups = "drop") |> 
  mutate(NormScoreArea = log1p(NormScoreArea))

# get max-y-value PER channel.
p_vals <- plot_df |> 
  filter(cell_cluster_scaled %in% c("CD11c+ Macrophages")) |>
  group_by(cell_cluster_scaled) |> 
  summarise(p_vals_pos = log1p(max(NormScoreArea))) |> 
  right_join(p_vals, by = "cell_cluster_scaled")

# Join with p-vals. This just for easier formating. Other P-values are used.
stat.test <- plot_df |> 
  group_by(cell_cluster_scaled) |> 
  rstatix::t_test(
    NormScoreArea ~ insulitic, paired = TRUE,
  ) |> 
    add_xy_position(x = "insulitic", step.increase = 0.07, dodge = 0.8) |> 
    distinct() |> 
    select(cell_cluster_scaled, group1, group2, y.position)

stat.test.test <- p_vals |> 
  left_join(stat.test, by = "cell_cluster_scaled") |> 
  mutate(y.position = y.position *1.3)
stat.test.test

p <- 
  ggpubr::ggviolin(plot_df, x = "insulitic", y = "NormScoreArea", 
                   fill = "insulitic", add = "point", facet.by = "cell_cluster_scaled",
                   draw_quantiles = c(0.25, 0.5, 0.75), 
          add.params = list(alpha = 1, size = 1.5)) +
  stat_pvalue_manual(stat.test.test, label = "p.signif", tip.length = 0.05,
                      size = 10, label.size = 10, vjust = 0.5) + 
  labs(x = "", y = "log: Infiltration Score", color = "Donor Type") +
  mytheme$standard_new() +
  scale_fill_manual(values = palettes$insulitic) + 
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) + 
  theme(legend.position = "none",
        axis.text.x = element_text(angle = 45, hjust = 1, size = 22),
        axis.text.y = element_text(size = 22),
        axis.title.x = element_text(size = 25),
        axis.title.y = element_text(size = 25),
        strip.text = element_text(size = 25)) + 
  geom_line(aes(group = case_id), alpha = 0.5) 

print(p)

fn <- paste0(paste(today, "Average_DistScore_Insulitis_Myeloid5", sep = "_"), ".png")
do.call(ggsave, c(list(fn, p), plotsave_param))

# Save the plot
extfig10c <- p
saveRDS(extfig10c, file.path(paths$home_git, "figures", "extfig10c.rds"))
```


# A4: Cluster and Identify Peri-Insulitic Lesions

## A4.1: Prepare Data.
For REC-ICI-Insulitis; cluster NormScoreArea.
-> Use AbsScore!
-> Normalized Scores will be correlated, due to the normalization.

Identify co-regulated Cell Clusters associated to Insulitis.

```{r peri-insulitic-lesions}
not_rel_clusters <- readRDS(file.path(paths$folder_out, "09_CellTypeCompo_cells_Immune", "not_ins_relevant_clusters.rds"))

peri_insulitic <- exprs_df  |> 
  filter(cell_type %in% c("Myeloid", "T_CD4", "T_CD8", "B")) |>
  filter(!cell_cluster_scaled %in% not_rel_clusters) |>
  filter(distance_to_islet >= -20) |>
  mutate(cell_cluster_scaled = ifelse(cell_type == "B", "B", cell_cluster_scaled))  |>
  # -70 IDI insulitic images.
  filter(islet_type == "ICI") |>
  # Average for: Image, Insulitic, Cell Cluster
  group_by(donor_type, case_id, image_id, cell_cluster_scaled, insulitic) |>
  summarize(n = sum(DistScoreSqrt), .groups = "drop") |> 
  tidyr::complete(image_id, cell_cluster_scaled, fill = list(n = 0)) |>
  group_by(image_id) |>
  mutate(insulitic = insulitic %>% table() %>%  which.max() %>% names()) |>
  mutate(donor_type = donor_type %>% table() %>%  which.max() %>% names()) |>
  ungroup() |> 
  tidyr::drop_na()

## Subset to only mAA/RecentOnset + Insulitic 
peri_insulitic_df <- peri_insulitic |> 
  ungroup() |> 
  # Remove clusters ending in "_Other"
  filter(!grepl("_Other", cell_cluster_scaled)) |>
  tidyr::pivot_wider(names_from = cell_cluster_scaled, values_from = n, values_fill = 0) |>
  dplyr::filter(donor_type %in% c("mAAb+", "RecentOnset")) |> 
  filter(insulitic == "insulitis")

rec_per_islet <- peri_insulitic_df |>
  select(-c(image_id, case_id, donor_type, insulitic))
```

### A4.2: Correlate DistScores between Cell Clusters.
### A4.2.1: Pearson

PEARSON Correlation of DistScores between Cell Clusters.
```{r pearson-peri-insulitic}
## Pearson
recent_onset_islet_pearson <- rec_per_islet |> 
  cor(use = "pairwise.complete.obs", method = "pearson")

saveRDS(recent_onset_islet_pearson, file.path(paths$home_git, "figures", "ch_fig6e.rds"))

## Complex-heatmap
set.seed(222)
fn <- paste0(paste(today, "Correlations", "peri_islet_pearson", sep = "_"), ".png")
png(file.path(paths$folder_script, fn), width = 1000, height = 1000)
p <- ComplexHeatmap::Heatmap(recent_onset_islet_pearson, col = circlize::colorRamp2(c(-1, 0, 1), c("blue", "white", "red")), 
                             name = "Correlation", 
                             clustering_distance_rows = "euclidean", clustering_method_rows = "complete", 
                             clustering_distance_columns = "euclidean", clustering_method_columns = "complete",
                             show_row_names = TRUE, show_column_names = TRUE,
                             # increase row and column names font
                             row_names_gp = grid::gpar(fontsize = 25),
                             column_names_gp = grid::gpar(fontsize = 25),
                             # increase legend font size
                                                           heatmap_legend_param = list(legend_direction = "vertical",
                                                          title_gp = grid::gpar(fontsize = 20, fontfamily = "Arial"),
                                                          labels_gp = grid::gpar(fontsize = 20, fontfamily = "Arial")),)
ComplexHeatmap::draw(p)
dev.off()
```


### A4.2.2: Pearson + log1p

PEARSON Correlation of DistScores between Cell Clusters.
```{r pearson-peri-insulitic}
## Pearson
recent_onset_islet_pearson <- rec_per_islet |> 
  # log-transform
  mutate(across(everything(), ~ log1p(.))) |>
  cor(use = "pairwise.complete.obs", method = "pearson")

## Complex-heatmap
set.seed(222)
fn <- paste0(paste(today, "Correlations", "peri_islet_pearson_log1p", sep = "_"), ".png")
png(file.path(paths$folder_script, fn), width = 1000, height = 1000)
p <- ComplexHeatmap::Heatmap(recent_onset_islet_pearson, col = circlize::colorRamp2(c(-1, 0, 1), c("blue", "white", "red")), 
                             name = "Correlation", 
                             clustering_distance_rows = "euclidean", clustering_method_rows = "complete", 
                             clustering_distance_columns = "euclidean", clustering_method_columns = "complete",
                             show_row_names = TRUE, show_column_names = TRUE,
                             # increase row and column names font
                             row_names_gp = grid::gpar(fontsize = 25),
                             column_names_gp = grid::gpar(fontsize = 25),
                             # increase legend font size
                                                           heatmap_legend_param = list(legend_direction = "vertical",
                                                          title_gp = grid::gpar(fontsize = 20, fontfamily = "Arial"),
                                                          labels_gp = grid::gpar(fontsize = 20, fontfamily = "Arial")),)
ComplexHeatmap::draw(p)
dev.off()
```

## A4.3: Correlation of Cell Clusters in Peri-Insulitic Islets.

-> Correlation of specific cell clusters in the peri-insulitic islets.

Plotting the correlations of specific cell clusters in the peri-insulitic islets.
```{r}
## Plot correlation scores of T-CD4-5 + B
age_cases <- exprs_df |> 
  select(age, case_id, donor_type) |>
  distinct()

input_df <- peri_insulitic |> 
  filter(cell_cluster_scaled %in% c("T_CD4_5", "T_CD8_7", "Myeloid_3", "B", "Myeloid_4", "T-naive")) |> 
  mutate(case_id = stringr::str_sub(image_id, 1, 4)) |> 
  left_join(age_cases, by = c("case_id", "donor_type")) |> 
  filter(donor_type %in% c("mAAb+", "RecentOnset")) |> 
  filter(insulitic == "insulitis") |> 
  pivot_wider(names_from = cell_cluster_scaled, values_from = n, values_fill = 0) |> 
  mutate(age_group = ifelse(age < 13, "<13", ">=13")) |>
  select(c("case_id", "age_group", "Myeloid_4", "Myeloid_3", "T-naive", "T_CD4_5", "B", "T_CD8_7", "age"))

p <- input_df |> 
  filter(T_CD4_5 > 0, `T-naive` > 0) |>
  ggplot(aes(x = T_CD4_5, y = `T-naive`, color = age_group)) +
  geom_point(size = 3) +
  geom_smooth(method = "lm", se = TRUE) +
  mytheme$standard() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  ggpubr::stat_cor(method = "pearson", aes(color = age_group), size = 6) +  # Correlation per group  
  theme(axis.text.x = element_text(size = 22),
        axis.text.y = element_text(size = 22),
        axis.title.x = element_text(size = 25),
        axis.title.y = element_text(size = 25),
        strip.text = element_text(size = 25))  + 
  labs(x = "Infiltration score: PD1+ act. cells", 
       y = "Infiltration score: T-naive") + 
  scale_color_manual(name = "Age Group (years)", 
                     values = palettes$colors50[c(9, 12)]) + 
  theme(legend.position = "top", 
        legend.orientation = "horizontal")
print(p)
fn <- paste0(paste(today, "Correlations", "Insulitic_T_CD4_5_T_naive", sep = "_"), ".png")
do.call(ggsave, c(list(fn, p), plotsave_param))
# saveRDS(p, file.path(paths$home_git, "figures", "extfig10b.rds"))

## T-naive + B-cells.
p <- input_df |> 
  filter(B > 0, `T-naive` > 0) |>
  ggplot(aes(x = log1p(B), y = log1p(`T-naive`), color = age_group)) +
  geom_point(size = 3) +
  geom_smooth(method = "lm", se = TRUE) +
  mytheme$standard() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  ggpubr::stat_cor(method = "pearson",  aes(group = age_group), size = 6) +  # Correlation per group  
  theme(axis.text.x = element_text(size = 22),
        axis.text.y = element_text(size = 22),
        axis.title.x = element_text(size = 25),
        axis.title.y = element_text(size = 25),
        strip.text = element_text(size = 25))  + 
  labs(x = "Infiltration Score: B-cells", 
       y = "Infiltration Score: T-naive") + 
  scale_color_manual(name = "Age Group (years)", 
                     values = palettes$colors50[c(9, 12)]) + 
  theme(legend.position = "top", 
        legend.orientation = "horizontal")
print(p)
fn <- paste0(paste(today, "Correlations", "Insulitic_B_Tnaive", sep = "_"), ".png")
do.call(ggsave, c(list(fn, p), plotsave_param))

## T-CD8-7 + Myeloid-3
p <- input_df |> 
  filter(T_CD8_7 > 0, Myeloid_3 > 0) |>
  ggplot(aes(x = log1p(Myeloid_3), y = log1p(T_CD8_7), color = age_group)) +
  geom_point(size = 3) +
  geom_smooth(method = "lm", se = TRUE) +
  mytheme$standard() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  ggpubr::stat_cor(method = "pearson", aes(color = age_group), size = 6) +  # Correlation per group  
  theme(axis.text.x = element_text(size = 22),
        axis.text.y = element_text(size = 22),
        axis.title.x = element_text(size = 25),
        axis.title.y = element_text(size = 25),
        strip.text = element_text(size = 25))  + 
  labs(x = "log1p: Infiltration Score: T_CD8_7s", 
       y = "log1p: Infiltration Score: Myeloid_3") + 
  scale_color_manual(name = "Age Group (years)", 
                     values = palettes$colors50[c(9, 15)]) + 
  theme(legend.position = "top", 
        legend.orientation = "horizontal")
print(p)
fn <- paste0(paste(today, "Correlations", "Insulitic_T_CD8_7_Myeloid3_AGE", sep = "_"), ".png")
do.call(ggsave, c(list(fn, p), plotsave_param))

## T-CD8-7 + Myeloid-3
p <- input_df |> 
  filter(T_CD8_7 > 0, Myeloid_3 > 0) |>
  ggplot(aes(x = log1p(Myeloid_3), y = log1p(T_CD8_7))) +
  geom_point(size = 3) +
  geom_smooth(method = "lm", se = TRUE) +
  mytheme$standard() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  ggpubr::stat_cor(method = "pearson", size = 6) +  # Correlation per group  
  theme(axis.text.x = element_text(size = 22),
        axis.text.y = element_text(size = 22),
        axis.title.x = element_text(size = 25),
        axis.title.y = element_text(size = 25),
        strip.text = element_text(size = 25))  + 
  labs(x = "log1p: Infiltration Score: T_CD8_7s", 
       y = "log1p: Infiltration Score: Myeloid_3") + 
  scale_color_manual(name = "Age Group (years)", 
                     values = palettes$colors50[c(9, 15)]) + 
  theme(legend.position = "top", 
        legend.orientation = "horizontal")
print(p)
fn <- paste0(paste(today, "Correlations", "Insulitic_T_CD8_7_Myeloid3_NOAGE", sep = "_"), ".png")
do.call(ggsave, c(list(fn, p), plotsave_param))

## Note: also aggregating by donor is highly significant.
summary(lmerTest::lmer(log1p(T_CD8_7) ~ log1p(Myeloid_3) + (1|case_id), 
  data = input_df |> 
  filter(T_CD8_7 > 0, Myeloid_3 > 0)))
```


```{r}
peri_insulitic <- exprs_df  |> 
  filter(cell_type %in% c("Myeloid", "T_CD4", "T_CD8", "B")) |>
  filter(!cell_cluster_scaled %in% not_rel_clusters) |>
  filter(distance_to_islet >= -20) |>
  mutate(cell_cluster_scaled = ifelse(cell_type == "B", "B", cell_cluster_scaled))  |>
  # -70 IDI insulitic images.
  filter(islet_type == "ICI") |>
  # Average for: Image, Insulitic, Cell Cluster
  group_by(donor_type, case_id, image_id, cell_cluster_scaled, insulitic) |>
  summarize(n = sum(DistScore), .groups = "drop") |> 
  tidyr::complete(image_id, cell_cluster_scaled, fill = list(n = 0)) |>
  group_by(image_id) |>
  mutate(insulitic = insulitic %>% table() %>%  which.max() %>% names()) |>
  mutate(donor_type = donor_type %>% table() %>%  which.max() %>% names()) |>
  ungroup() |> 
  tidyr::drop_na()

## Correlation of DistScores.
data_tt <- peri_insulitic |> 
    filter(donor_type %in% c("RecentOnset", "mAAb+")) |>
    mutate(insulitic = if_else(image_id %in% image_ids, "insulitis", "non-insulitis")) |> 
    filter(cell_cluster_scaled %in% c("Myeloid_3", "T_CD8_7")) |> 
    pivot_wider(names_from = cell_cluster_scaled, values_from = n, values_fill = 0) |> 
    filter(insulitic == "insulitis")

cor_test <- cor.test(log1p(data_tt$Myeloid_3), log1p(data_tt$T_CD8_7), method = "pearson")
cor_test_sp <- cor.test(data_tt$Myeloid_3, data_tt$T_CD8_7, method = "spearman")

(round(cor_test_sp$estimate, 2))
(signif(cor_test_sp$p.value, 3))
# Extract correlation coefficient and p-value
(pearson_r <- round(cor_test$estimate, 2))
(p_value <- signif(cor_test$p.value, 3))

p <- data_tt |>
    ggplot(aes(x = log1p(Myeloid_3), y = log1p(T_CD8_7))) +
    geom_point() +
    geom_smooth(method = "lm", se = TRUE) +
    labs(x = "Infiltration Score: M1/M2-like act.", 
         y = "Infiltration Score: PD1+,CD27+ T-CD8") +
    mytheme$standard() +
    theme(axis.text.x = element_text(angle = 45, hjust = 1)) + 
    ggpubr::stat_cor(method = "spearman", label.x = 0, label.y = 3.5, size = 10) + 
    theme(axis.text.x = element_text(size = 22),
          axis.text.y = element_text(size = 22),
          axis.title.x = element_text(size = 25),
          axis.title.y = element_text(size = 25),
          strip.text = element_text(size = 25))

print(p)
fn <- paste0(paste(today, "DistScore_Myeloid3_vs_T_CD8_7", sep = "_"), ".png")
do.call(ggsave, c(list(fn, p), plotsave_param))

fig5i <- p
saveRDS(fig5i, file.path(paths$home_git, "figures", "fig5i.rds"))
```


# A6: Expression changes between Insulitic and Non-Insulitic Islets:

Filter Peri-insulitic Cells. In REC.
```{r}
### Prepare the data.
### ICI. Peri-insulitic.
ici_immune <- exprs_df |> 
  filter(islet_type == "ICI", 
         cell_type %in% c("T_CD4", "T_CD8", "Myeloid", "T_DN", "NK", "B", "Neutrophil", "Endothelial")) |> 
  filter(infiltration %in% c("inside", "< 20 um"))
```


## A6.3: Endothelial Insulitis EXPRESSION ANALYSIS.
Compare insulitic vs non-insulitic ICIs.
```{r endothelial-insulitic-changes}
## Endothelial cells:
## COmpare expression for Endothelial cells next:
features_oi <- unique(c("CD54", metadata(spe)$channels$Endothelial))

## 
endo_df <- ici_immune  |> 
    filter(cell_type == "Endothelial", donor_type %in% c("RecentOnset", "mAAb+")) |> 
    select(all_of(features_oi), cell_cluster_scaled, case_id, insulitic, image_id, donor_type)

# Select paired case-ids with insulitic + non-insulitic images.
case_ids <- endo_df |> 
  filter(donor_type %in% c("RecentOnset", "mAAb+")) |>
  distinct(case_id, insulitic) |>
  dplyr::count(case_id) |> 
  filter(n == 2) |> pull(case_id) |> 
  unique()

## Calculate Expression Differences.
stats <- endo_df |> 
  filter(case_id %in% case_ids) |>
  group_by(image_id, insulitic, case_id, cell_cluster_scaled) |>
  summarize(across(all_of(features_oi), \(x) mean(x, na.rm = TRUE)), .groups = "drop") |>
  pivot_longer(cols = all_of(features_oi), names_to = "channel", values_to = "value") |>
  tidyr::nest(data = -c(cell_cluster_scaled, channel)) |>
  dplyr::mutate(test = purrr::map(data, ~ broom.mixed::tidy(lmerTest::lmer(data = ., value ~ insulitic + (1 | case_id))))) |>
  tidyr::unnest(test) |>
  filter(term != "(Intercept)", effect == "fixed") |>
  arrange(p.value) |> 
  mutate(p.adj = p.adjust(p.value, method = "fdr"))

## Plot:
p <- stats |> 
  filter(p.adj < 0.1) |>
  ggplot(aes(x = channel, y = statistic, fill = cell_cluster_scaled)) +
  geom_bar(stat = "identity", position = "dodge") +
  labs(x = "Channel", y = "Statistic", color = "Channel") +
  mytheme$standard() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
print(p)

fn <- paste0(paste(today, "Expression_Endo_Insulitis", sep = "_"), ".png")
do.call(ggsave, c(list(fn, p), plotsave_param))
```
-> As expected, strong upregulated in HLA-DR and CD54.



# A8: Insulitis in endocrine Cells.

## A8.1: Prepare Data.
```{r}
channels <- metadata(spe)$channels
channels_imm <- unique(c("VIM", unlist(channels)))

## Define Insulitic Islets + ICIs.
## Filter Pancreatitis
exprs_df <- as_tibble(scuttle::makePerCellDF(spe, features = channels_imm, assay.type = "exprs", use.dimred = FALSE)) |> 
  select(all_of(channels_imm), case_id, DistScore, donor_type, image_id, infiltration, cell_type, distance_to_islet, cell_cluster_scaled) |> 
  filter(donor_type != "Pancreatitis") |> 
  group_by(image_id, case_id, donor_type) |>
  mutate(islet_type = if_else(sum(cell_type == "Beta") > 0, "ICI", "IDI")) |> 
  mutate(tcell = if_else(cell_type %in% c("T_CD4", "T_CD8"), "T-cell", "non-T-cell")) |>  ungroup() |> 
  mutate(donor_type = factor(donor_type, levels = c("NoDiabetes", "sAAb+", "mAAb+", "RecentOnset", "LongDuration")))

fn <- file.path(paths$folder_out, "13_MarkerExpression_cells_Immune", "insulitic_images.rds")

if (file.exists(fn)) {
  image_ids <- readRDS(fn)
  exprs_df <- exprs_df |> 
    mutate(insulitic = if_else(image_id %in% image_ids, "insulitis", "non-insulitis"))
}
```

## A8.2: Define Insulitic Endocrine Cells.
```{r insulitic-endocrine-cells}
endocrine_insulitic <- exprs_df |> 
  filter(islet_type == "ICI", cell_type %in% c("Beta", "Alpha", "Endothelial"))

endocrine_insulitic_df <- endocrine_insulitic |> 
  group_by(image_id, cell_type, insulitic, case_id, donor_type) |> 
  summarize(across(all_of(channels_imm), \(x) mean(x, na.rm = TRUE)), .groups = "drop")  |> 
  arrange(image_id)

beta_df <- endocrine_insulitic_df |> 
  filter(cell_type == "Beta") |> 
  pivot_longer(cols = all_of(channels_imm), names_to = "channel", values_to = "MeanExprs")
```


## A8.3: Expression Changes in Insulitic vs Non-Insulitic Beta Cells.
```{r insulitic-vs-non-insulitic-beta}
## Compare between Insulitic and Non-insulitic ICIs.
features_oi <- c("HLA_DR", "SYP", "VIM", "CD54")

## Select only case-ids with insulitic + non-insulitic images.
case_ids <- beta_df |> 
  filter(donor_type %in% c("RecentOnset", "mAAb+")) |>
  distinct(case_id, insulitic) |>
  dplyr::count(case_id) |> 
  filter(n == 2) |> 
  pull(case_id) |> 
  unique()

## Delete rare beta-cell instances (to_del).
data_tt <- beta_df |> 
  filter(donor_type %in% c("RecentOnset")) |> 
  filter(case_id %in% case_ids) |>
  filter(channel %in% features_oi) |> 
  mutate(insulitic = ifelse(insulitic == "insulitis", "Insulitic", "Non-Insulitic")) |>
  mutate(insulitic = factor(insulitic, levels = c("Insulitic", "Non-Insulitic")))

beta_df_plot <- data_tt |> 
  group_by(case_id, donor_type, channel, insulitic) |> 
  summarize(MeanExprs = mean(MeanExprs, na.rm = TRUE), .groups = "drop")

## Calculate difference between INsulitic and Non-insulitic
p_vals <- data_tt |> 
  tidyr::nest(data = -c(channel)) |>
  dplyr::mutate(test = purrr::map(data, ~ 
            broom.mixed::tidy(lmerTest::lmer(data = ., MeanExprs ~ insulitic + (1 | case_id))))) |>
  tidyr::unnest(test) |>
  filter(effect == "fixed", term != "(Intercept)") |>
  dplyr::select(channel, p.value, statistic, estimate, term) |> 
  arrange(statistic) |> mutate(p.adj = p.adjust(p.value, method = "fdr")) |> 
  mutate(significance = cut(p.adj, breaks = c(-Inf, 0.001, 0.01, 0.05, Inf), labels = c("***", "**", "*", "")) ) |> 
  mutate(insulitic = stringr::str_remove(term, "insulitic"))

# Use this for actual P-Values.
p_vals <- p_vals |> 
  mutate(.y. = "MeanExprs",
         donor_type = stringr::str_remove(term, "donor_type"),
         group1 = "Insulitic",
         group2 = insulitic,
         p = p.value,
         p.adj = p.adjust(p, method = "fdr"),
        p.signif = cut(p.adj, breaks = c(-Inf, 0.001, 0.01, 0.05, 0.1, Inf), labels = c("***", "**", "*", "•", ""))
         ) |> 
  select(channel, .y., group1, group2,
         statistic, p, p.adj, p.signif)

max_values <- beta_df_plot |> 
  group_by(channel) |> 
  summarize(max = max(MeanExprs, na.rm = TRUE) + 0.4, .groups = "drop")

p_vals <- p_vals  |> 
  left_join(max_values, by = "channel")

## Calculate p-values with ggpubr.
# Use this for right formatting of DF.
stat.test <- data_tt  |> 
  group_by(channel) |> 
  t_test(MeanExprs ~ insulitic, ref.group = "Insulitic") |> 
  adjust_pvalue()  |> 
  add_significance("p.adj") |> 
  add_xy_position(x = "insulitic", step.increase = 0.07, dodge = 0.8) |> 
  select(-c(statistic, p, p.adj))

# Combine formatting + P-value dataframe.
stat.test.test <- p_vals |> 
  left_join(stat.test, by = c("channel", ".y.", "group1", "group2")) |> 
  mutate(y.position = max)

# Plot.
p <- ggviolin(beta_df_plot, x = "insulitic", y = "MeanExprs", fill = "insulitic", 
        facet.by = "channel", palette = palettes$colors[1:2], add = "point",
        draw_quantiles = c(0.25, 0.5, 0.75)) + 
  stat_pvalue_manual(stat.test.test, label = "p.signif", tip.length = 0.02,
                    size = 10, label.size = 10, vjust = +0.1, inherit.aes = FALSE) + 
  facet_wrap(~channel, scales = "free_y") +
  mytheme$standard() + 
  theme(axis.text.x = element_text(angle = 45, hjust = 1, size = 22),
        axis.text.y = element_text(size = 22),
        strip.text = element_text(size = 25),
        axis.title.y = element_text(size = 25),
        legend.position = "none") + 
  labs(x = "", y = expression(beta ~ "-cell expression")) + 
  scale_y_continuous(expand = expansion(mult = c(0, 0.15))) + 
  geom_line(aes(group = case_id), alpha = 0.5)
print(p)

fn <- paste0(paste(today, "Insulitic_Beta", sep = "_"), ".png")
do.call(ggsave, c(list(fn, p), plotsave_param))

extfig4g <- p
saveRDS(extfig4g, file.path(paths$home_git, "figures", "extfig4g.rds"))
```
-> HLA-DR + CD54 are significantly upregulated in insulitic beta-cells.


```{r select-endothelial-cells}
endo_df <- endocrine_insulitic_df |> 
  filter(cell_type == "Endothelial") |> 
  pivot_longer(cols = all_of(channels_imm), names_to = "channel", values_to = "MeanExprs")
```


## A8.3: Expression Changes in Insulitic vs Non-Insulitic Endothelial Cells.
```{r insulitic-vs-non-insulitic-endothelial}
## Compare between Insulitic and Non-insulitic ICIs.
features_oi <- c("HLA_DR", "VIM", "CD54")

## Select only case-ids with insulitic + non-insulitic images.
case_ids <- endo_df |> 
  filter(donor_type %in% c("RecentOnset", "mAAb+")) |>
  distinct(case_id, insulitic) |>
  dplyr::count(case_id) |> 
  filter(n == 2) |> 
  pull(case_id) |> 
  unique()
  
## Delete rare endothelial-cell instances (to_del).
data_tt <- endo_df |> 
  filter(donor_type %in% c("RecentOnset")) |> 
  filter(case_id %in% case_ids) |>
  filter(channel %in% features_oi) |> 
  mutate(insulitic = ifelse(insulitic == "insulitis", "Insulitic", "Non-Insulitic")) |>
  mutate(insulitic = factor(insulitic, levels = c("Insulitic", "Non-Insulitic")))

endo_df_plot <- data_tt |> 
  group_by(case_id, donor_type, channel, insulitic) |> 
  summarize(MeanExprs = mean(MeanExprs, na.rm = TRUE), .groups = "drop")

## Calculate difference between INsulitic and Non-insulitic
p_vals <- data_tt |> 
  tidyr::nest(data = -c(channel)) |>
  dplyr::mutate(test = purrr::map(data, ~ 
            broom.mixed::tidy(lmerTest::lmer(data = ., MeanExprs ~ insulitic + (1 | case_id))))) |>
  tidyr::unnest(test) |>
  filter(effect == "fixed", term != "(Intercept)") |>
  dplyr::select(channel, p.value, statistic, estimate, term) |> 
  arrange(statistic) |> mutate(p.adj = p.adjust(p.value, method = "fdr")) |> 
  mutate(significance = cut(p.adj, breaks = c(-Inf, 0.001, 0.01, 0.05, Inf), labels = c("***", "**", "*", "")) ) |> 
  mutate(insulitic = stringr::str_remove(term, "insulitic"))

# Use this for actual P-Values.
p_vals <- p_vals |> 
  mutate(.y. = "MeanExprs",
         donor_type = stringr::str_remove(term, "donor_type"),
         group1 = "Insulitic",
         group2 = insulitic,
         p = p.value,
         p.adj = p.adjust(p, method = "fdr"),
        p.signif = cut(p.adj, breaks = c(-Inf, 0.001, 0.01, 0.05, 0.1, Inf), labels = c("***", "**", "*", "•", ""))
         ) |> 
  select(channel, .y., group1, group2,
         statistic, p, p.adj, p.signif)

max_values <- endo_df_plot |> 
  group_by(channel) |> 
  summarize(max = max(MeanExprs, na.rm = TRUE) + 0.4, .groups = "drop")

p_vals <- p_vals  |> 
  left_join(max_values, by = "channel")

## Calculate p-values with ggpubr.
# Use this for right formatting of DF.
stat.test <- data_tt  |> 
  group_by(channel) |> 
  t_test(MeanExprs ~ insulitic, ref.group = "Insulitic") |> 
  adjust_pvalue()  |> 
  add_significance("p.adj") |> 
  add_xy_position(x = "insulitic", step.increase = 0.07, dodge = 0.8) |> 
  select(-c(statistic, p, p.adj))

# Combine formatting + P-value dataframe.
stat.test.test <- p_vals |> 
  left_join(stat.test, by = c("channel", ".y.", "group1", "group2")) |> 
  mutate(y.position = max)

# Plot.
p <- ggviolin(endo_df_plot, x = "insulitic", y = "MeanExprs", fill = "insulitic", 
        facet.by = "channel", palette = palettes$colors[1:2], add = "point",
        draw_quantiles = c(0.25, 0.5, 0.75)) + 
  stat_pvalue_manual(stat.test.test, label = "p.signif", tip.length = 0.02,
                    size = 10, label.size = 10, vjust = +0.1, inherit.aes = FALSE) + 
  facet_wrap(~channel, scales = "free_y") +
  mytheme$standard() + 
  theme(axis.text.x = element_text(angle = 45, hjust = 1, size = 22),
        axis.text.y = element_text(size = 22),
        strip.text = element_text(size = 25),
        axis.title.y = element_text(size = 25),
        legend.position = "none") + 
  labs(x = "", y = "Endothelial-cell expression") + 
  scale_y_continuous(expand = expansion(mult = c(0, 0.15))) + 
  geom_line(aes(group = case_id), alpha = 0.5)
print(p)

fn <- paste0(paste(today, "Insulitic_Endothelial", sep = "_"), ".png")
do.call(ggsave, c(list(fn, p), plotsave_param))

suppfig9 <- p
saveRDS(suppfig9, file.path(paths$home_git, "figures", "suppfig9.rds"))
```
-> HLA-DR + CD54 are significantly upregulated in insulitic endothelial-cells.