---
title: "23_ImageLevelAnalysis_cells_Islet"
author: "Nathan Steenbuck"
date: "Created: 24 May, 2025; Compiled: `r format(Sys.time(), '%d %b, %Y')`"
output: workflowr::wflow_html
editor_options:
  chunk_output_type: inline
---

```{r setup, include=FALSE}
if (rstudioapi::isAvailable()) {
  script_name <- basename(rstudioapi::getSourceEditorContext()$path)
} else {
  script_name <- knitr::current_input()
}

cur_user <- Sys.info()[["user"]]
script_name <- "23_ImageLevelAnalysis_cells_Islet.Rmd"

if (cur_user == "ubuntu") {
  source(file.path("/", "mnt", "central_nas", "projects", "type1_diabetes", "nathan", 
                   "T1D_Vol", "T1D_analysis", "analysis", "helpers.R"))
  #n_cores <- future::availableCores() - 1
  n_cores <- 4
  future::plan(future::multicore(workers = n_cores))
  paths <- getPaths(script_name)
  knitr::opts_knit$set(root_dir = paths$home_data)
  do_print <- FALSE
} else {
  source(file.path("/", "home", "nsteen", "scripts", "T1D_analysis", "analysis", "helpers.R"))
  n_cores <- 8
  future::plan(future::multicore(workers = n_cores))
  paths <- getPaths(script_name)
  if (!dir.exists(cluster_folder_script)) dir.create(cluster_folder_script)
  paths$folder_script <- paths$cluster_folder_script
  paths$folder_in <- paths$cluster_folder_in
  paths$folder_out <- paths$cluster_home 
  knitr::opts_knit$set(root_dir = paths$cluster_home)
  do_print <- TRUE
}

seed <- 123456
set.seed(seed)
options <- furrr::furrr_options(seed = seed)
options(future.globals.maxSize = 50 * 1024 * 1024 * 1024)
paths$prev <- paste("08_SubclusteringBeta", paths$object_type, paths$panel_type, sep = "_")
knitr::opts_chunk$set(echo = TRUE)
```


# A0: **Introduction and Settings**

Both total protein marker levels and localization are proxies for protein activity.

Here, the aim is to detect potential shifts in localization across disease stages. 
This could indicate distinct activities for the given marker.

Here, we mainly look at distinct protein localization for XBP1. 

As sXBP1 is a transcription factor, it is expected to be localized in the nucleus, 
whereas unspliced XBP1 is expected to be localized in the cytoplasm.
-> Note: clone detects both isoforms.


Improved analysis:
**- when defining pixels, I could take quantiles of uncertainty. **
This could make the analysis more robust to noise. 
However, based on the images - I dont expect much changes.

## A0.1: Load packages

```{r packages, results='hide'}
suppressPackageStartupMessages(c(
  library(data.table),
  library(dplyr),
  library(SpatialExperiment),
  library(scuttle),
  library(parallel),
  library(patchwork),
  library(tidyr),
  library(cytomapper),
  library(EBImage),
  library(reshape2),
  library(mclust)
))
```

## A0.2: Paths and settings

```{r settings}
# Paths
if (!dir.exists(paths$folder_script)) dir.create(paths$folder_script)
plotsave_param$path <- paths$folder_script
plotsave_param_large$path <- paths$folder_script

# Misc settings
today <- gsub("-", "", Sys.Date())
```

##  A0.3: Read in the SPE data

Load the SpatialExperiment (SPE) object saved at the previous step.

```{r load-data}
fn_spe <- file.path(paths$folder_out, paths$prev, paste0(paths$object_type, "_", paths$panel_type, ".rds"))
spe <- readRDS(fn_spe)
print(spe)
```

## A0.4: Subset the data to beta-cells.
```{r}
beta_cells <- spe[, spe$cell_type == "Beta"]
beta_cells <- beta_cells[, beta_cells$donor_type %in% c("NoDiabetes", "sAAb+", "mAAb+", "RecentOnset")]
```

# **Aim 1: Estimate the correlation between channels (mainly DNA1) for each image.**

Goal is to detect shifts in the correlation to DNA1, which would be a proxy for its localization shift.

## A1.1: Prepare Data
```{r subset-data}
metadata_oi <- c("cell_id", "image_id", "case_id", "donor_type", "age")
features_oi <- rownames(beta_cells)
# Channels to estimate correlation for.
channels_oi <- metadata(beta_cells)$channels$Beta
channels_oi <- c("DNA1", "DNA3", channels_oi[!channels_oi %in% c("Ki67", "ARX", "PDL1")])

exprs_df <- scuttle::makePerCellDF(beta_cells, 
  features = features_oi,
  assay.type = "exprs", 
  use.coldata = metadata_oi, 
  use.dimred = FALSE)

image_ids <- unique(exprs_df$image_id)
case_stages <- exprs_df |> distinct(case_id, donor_type, age) |> mutate(age_group = ifelse(age < 13, "young", "old"))
set.seed(22222)
rm(spe, beta_cells); gc()
```

## A1.2: Run the correlation pipeline.

Steps:
- Load image + Masks. 
- Across Images, extract ALL beta-cell pixels.
- For each image, extract the pixel values for the channels of interest.
- Calculate the correlation between the channels - MAINLY TO DNA1- for each image.

-> This yields a correlation matrix for each image, 
which can be used to assess shifts in localization across disease stages.

```{r run-pipeline}
calc_corrs <- FALSE

# iterate through all images. 
if (calc_corrs){
  corr_results <- purrr::map(image_ids, \(image_id){
  # image_id <- "6433_Islet_008"
    mask_ids <- unique(exprs_df$cell_id[exprs_df$image_id %in% image_id])
    
    # remove image-id prefix from mask_ids.
    mask_ids <- stringr::str_remove(mask_ids, paste0(image_id, "_"))
    message("Image: ", image_id, " - Number of mask ids: ", length(mask_ids))

    ### Load Images:
    image_extension <- ".tiff"
    image_id_long <- stringr::str_replace(image_id, "Islet_", "Islet_ROI_")
    image_id_full <- paste(image_id_long, image_extension, sep = "")

    # Paths.
    folder_images <- file.path(paths$folder_in, "img", paths$panel_type)
    folder_masks <- file.path(paths$folder_in, "masks_cells", paths$panel_type, "whole-cell")

    # ------ Load the masks and identify pixels of interest (beta-cell pixels).
    masks <- cytomapper::loadImages(file.path(folder_masks, image_id_full), as.is = TRUE)
    mask_image <- imageData(cytomapper::getImages(masks, image_id_long)[[1]])
    nrow(mask_image); ncol(mask_image)
    mask_df <- as.data.frame(which(!is.na(mask_image), arr.ind = TRUE))

    # Get the value of the mask at these indices (= mask labels)
    mask_df$label <- mask_image[cbind(mask_df$row, mask_df$col)]

    # Filter by mask-ids.
    result <- subset(mask_df, label %in% mask_ids)
    mask_pixels <- result[, c("row", "col")]
    head(result); head(mask_pixels)
    
    # ------- Load the images for the channels of interest.
    images <- cytomapper::loadImages(file.path(folder_images, image_id_full))
    cytomapper::channelNames(images) <- features_oi

    ## Subset the tiff images to the channels of interest.  
    images <- cytomapper::getChannels(images, channels_oi)

    # Add image names to metadata
    mcols(images)$ImageName <- names(images)
    mcols(images)$ImageName <- paste0(mcols(images)$ImageName)

    # Subset the images to the mask pixels.
    image_data <- imageData(cytomapper::getImages(images, image_id_long)[[1]])

    # Prepare to write pixel values channel-wise into HDF5
    # Flatten spatial dimensions, keep channels
    dims <- dim(image_data)  # Should be [rows, cols, channels]
    n_channels <- dims[3]
    pixel_count <- nrow(mask_pixels)

    # Set a default temp folder.
    HDF5Array::setHDF5DumpDir(tempdir())

    # h5_file <- paste0(image_id, "_beta_image_by_channel.h5")

    # Preallocate HDF5-backed matrix: rows = pixels, cols = channels
    h5_beta <- HDF5Array::writeHDF5Array(
      DelayedArray::DelayedArray(array(NA_real_, dim = c(pixel_count, n_channels))),
      # filepath = h5_file,
      name = "beta_pixels"
    )

    # Get the pixel values for the mask pixels. Across all channels.
    # This ofc. destroys spatial structure of image.
    # --- Channel-wise extraction to keep memory low
    for (ch in seq_len(n_channels)) {
      slice <- image_data[, , ch]
      h5_beta[, ch] <- slice[cbind(mask_pixels$row, mask_pixels$col)]
    }

    # Function to compute correlation between one column and all others
    cor_column <- function(h5_matrix, col_index) {
      x <- h5_matrix[, col_index]
      # Compute correlation between x and all columns
      apply(h5_matrix, 2, function(y) cor(x, y, use = "pairwise.complete.obs", method = "spearman"))
    }

    # Preallocate correlation matrix
    cor_matrix <- matrix(NA, n_channels, n_channels)

    # Compute upper triangle (matrix is symmetric)
    for (i in seq_len(n_channels)) {
      cor_vals <- cor_column(h5_beta, i)
      cor_matrix[i, ] <- cor_vals
    }

    # Optional: Set dimnames
    channel_names <- dimnames(image_data)[[3]]
    rownames(cor_matrix) <- channel_names
    colnames(cor_matrix) <- channel_names

    # Melt for plotting or export
    library(reshape2)
    cor_matrix_melted <- melt(cor_matrix)
    colnames(cor_matrix_melted) <- c("Channel1", "Channel2", "Correlation") 

    if (image_id == "6433_Islet_001") {
      message("Plotting correlation matrix for image: ", image_id)
      p <- ggplot(cor_matrix_melted, aes(x = Channel1, y = Channel2, fill = Correlation)) +
        geom_tile() +
        scale_fill_gradient2(low = "blue", high = "red", mid = "white", 
                            midpoint = 0, limit = c(-1, 1), space = "Lab", 
                            name="Correlation") +
        mytheme$standard_new() +
        theme(axis.text.x = element_text(angle = 45, hjust = 1))

    fn <- file.path(paths$folder_script, "beta_cell_correlation_matrix.png")
    ggsave(fn, plot = p, width = 10, height = 8, dpi = 300)
    }

    # Return:
    as_tibble(cor_matrix_melted) |> 
      mutate(image_id = image_id) |> 
      mutate(nr_beta_cells = length(mask_ids))
  }) |> bind_rows()
  saveRDS(corr_results, file = file.path(paths$folder_script, "beta_cell_correlation_results.rds"))
}
```

## A1.3: Analyze the correlation results.
### A1.3.1 Load the correlation results and label insulitic images.
Now, subset the images to the mask pixels.
```{r load-corr-results}
if (!calc_corrs){
  corr_results <- readRDS(file.path(paths$folder_script, "beta_cell_correlation_results.rds"))## Label Insulitic images
}

fn <- file.path(paths$folder_out, "19_TcellsInsulitis_cells_Immune", "insulitic_images2.rds")
if (file.exists(fn)) {
  insulitic_images <- readRDS(fn) |> stringr::str_replace("Immune", "Islet")
}
corr_results <- corr_results |> 
    mutate(insulitic = ifelse(image_id %in% insulitic_images, "Insulitic", "NonInsulitic"))
```

### A1.3.2 Check for differences in correlation between channels.
Now, check if any correlations are associated with disease.
Use linear model to test for differences in correlation between groups.

```{r corr-differences} 
to_del <- corr_results |> 
  filter(Channel1 == "INS", Channel2 == "DNA1") |>
  mutate(case_id = stringr::str_extract(image_id, "^[0-9]+")) |> 
  group_by(case_id) |> 
  summarise(sum_beta = sum(nr_beta_cells)) |> filter(sum_beta <= 10) |> 
  pull(case_id)

lmm_input <- corr_results |> 
  filter(nr_beta_cells > 3) |> 
  filter(Channel1 != Channel2) |>
  mutate(case_id = stringr::str_extract(image_id, "^[0-9]+")) |>
  filter(!case_id %in% to_del) |>
  left_join(case_stages, by = "case_id")
  #mutate(donor_type = if_else(insulitic == "Insulitic" & donor_type %in% c("RecentOnset", "mAAb+"), "Insulitic", donor_type)) |>
  #mutate(donor_type = factor(donor_type, 
  #  levels = c("NoDiabetes", "sAAb+", "mAAb+", "RecentOnset", "Insulitic")))

res <- lmm_input |>
  tidyr::nest(data = -c(Channel1, Channel2)) |>
  mutate(test = purrr::map(data, ~  broom.mixed::tidy(lmerTest::lmer(data = ., 
    Correlation ~ donor_type + age_group + (1 | case_id))))
  ) |> 
  unnest(test) |> 
  filter(term != "(Intercept)") |>
  filter(effect == "fixed") |> 
  arrange(p.value) |> 
  mutate(donor_type = stringr::str_remove(term, "donor_type")) |>
  mutate(p.adj = p.adjust(p.value, method = "fdr")) |> 
  select(-data, -group, -data, -std.error, -effect)
```

```{r plot-corr-differences}
c("INS", "DNA1", "DNA3", "ECad", "HLA_ABC", "MX1") |> 
  purrr::walk(\(channel) {
    p <- res |>
      filter(Channel2 == !!channel) |>
      mutate(direction = ifelse(estimate > 0, "positive", "negative")) |>
      ggplot(aes(x = reorder(Channel1, p.adj), y = -log10(p.adj), fill = direction)) +
      geom_bar(stat = "identity") +
      labs(x = "Channel", y = "Adjusted p-value") +
      mytheme$standard_new() + 
      theme(axis.text.x = element_text(angle = 45, hjust = 1)) + 
      facet_wrap(~ donor_type, scales = "free_y")
    
    fn <- paste0(today, "_correlation_differences_", channel, ".png")
    do.call(ggsave, c(list(fn, p), plotsave_param_large))
  })

# Reflects shifts across stages. Less corr. with INS/ProINS and higher with HLA-ABC.
res |> 
  filter(Channel1 == "XBP1", Channel2 == "DNA1") |> 
  select(Channel2, estimate, p.value, p.adj, donor_type) |> 
  arrange(p.adj) |> 
  print(n = 20)

# Higher correlation with both nuclear/cytoplasmatic markers. 
# Lower with CAIX
res |> 
  filter(Channel1 == "TXNIP") |> 
  select(Channel2, estimate, p.value, p.adj, donor_type) |> 
  arrange(p.adj) |> 
  print(n = 20)

# Plot only for ER-stress related markers.
p <- res |> filter(Channel1 == "DNA3") |>
  filter(Channel2 %in% c("TXNIP", "XBP1", "ERN1", "NFkB", "WFS1", "CAIX")) |>
  mutate(direction = ifelse(estimate > 0, "positive", "negative")) |>
  ggplot(aes(x = reorder(Channel2, p.adj), y = -log10(p.adj), fill = direction)) +
  geom_bar(stat = "identity") +
  labs(x = "Channel", y = "logpadj") +
  mytheme$standard_new() + 
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) + 
  facet_wrap(~ donor_type, scales = "free_y")

print(p)
fn <- paste0(today, "_correlation_differences_ER_only", ".png")
do.call(ggsave, c(list(fn, p), plotsave_param_large))
```

Boxplot with DNA1 + ECad for XBP1. 
```{r}
xbp1_plot <- lmm_input |> 
  filter(Channel1 == "XBP1", Channel2 %in% c("DNA1", "ECad", "ProINS")) |> 
  group_by(donor_type, case_id, Channel2) |>
  summarize(mean_correlation = mean(Correlation, na.rm = TRUE),
            sd_correlation = sd(Correlation, na.rm = TRUE),
            .groups = "drop")
palettes$stages2 <- palettes$stages
names(palettes$stages2) <- rev(c("Onset", "mAAb+", "sAAb+", "Control"))

p <- xbp1_plot |>
  mutate(donor_type = case_when(
    donor_type == "RecentOnset" ~ "Onset",
    donor_type == "mAAb+" ~ "mAAb+",
    donor_type == "sAAb+" ~ "sAAb+",
    donor_type == "NoDiabetes" ~ "Control"
  )) |>
  mutate(donor_type = factor(donor_type, levels = c("Control", "sAAb+", "mAAb+", "Onset"))) |>
  ggplot(aes(x = Channel2, y = mean_correlation, fill = donor_type)) +
  geom_boxplot(position = "dodge", outliers = FALSE) +
  geom_jitter() +
  geom_hline(yintercept = 0, linetype = "dashed") +
  mytheme$standard_new() + 
  scale_fill_manual(values = palettes$stages2, name = "Stage") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) + 
  labs(x = expression(beta ~ "-cell marker"), y = expression(beta ~ "-cell Spearman correlation (XBP1)", fill = "Stage")) + 
  theme(legend.position = "top")

print(p)
fn <- paste0(today, "_xbp1_correlation_boxplot.png")
do.call(ggsave, c(list(fn, p), plotsave_param))
saveRDS(p, file.path(paths$home_git, "figures", "rev_supplfig4b.rds"))
```

Compare Insulitic vs non-insulitic:

```{r}
inputs <- lmm_input |> 
  filter(Channel1 == "XBP1", Channel2 == "DNA1") |> 
  filter(donor_type %in% c("mAAb+", "RecentOnset")) |> 
  filter(!case_id %in% to_del) |> 
  distinct(insulitic, case_id) |> 
  dplyr::count(case_id) |> 
  dplyr::filter(n == 2) |> pull(case_id)

res_insulitic1 <- lmm_input |> 
  filter(case_id %in% inputs) |> 
  tidyr::nest(data = -c(Channel1, Channel2)) |>
  mutate(test = purrr::map(data, ~  broom.mixed::tidy(lmerTest::lmer(data = ., 
    Correlation ~ insulitic + age_group + (1 | case_id))))
  ) |> 
  unnest(test) |> 
  filter(term != "(Intercept)", effect == "fixed")


## Higher correlation (non-s) with DNA1+3 in non-insulitic.
## HENCE NO nuclear translocation in insulitis.
res_insulitic1 |> filter(Channel1 == "XBP1", Channel2 %in% c("B2M", "HLA_ABC", "SYP", "DNA1", "DNA3"), term != "age_groupyoung")

## No clear direction.
res_insulitic1 |> filter(Channel1 == "ADAR", Channel2 %in% c("B2M", "HLA_ABC", "SYP", "DNA1", "DNA3"), term != "age_groupyoung")
```


# Aim 2: Estimate nuclear, cytoplasmatic and total expression for each beta-cell.

1) Get the nuclear, cytoplasmatic and total expression for each beta-cell.
2) For this, iterate through all images, load the masks and images, then extract the nuclear pixels using a GMM.
  -> Each cell has nucleus + cytoplasm. 
  -> there are 2, 3 edge cases (very rare), but these are handled.
3) Calculate the average expression for each channel in the nuclear pixels, total pixels and cytoplasmatic pixels.
4) Write them to a .csv.

## A2.1: Run the pipeline.
```{r write-cyt-nuc-tot-exp-per-cell}
# iterate through all images. 
set.seed(222)
calc_nuc_cyto_tot <- FALSE
  if (calc_nuc_cyto_tot) {

  purrr::walk(image_ids, \(image_id){
    # image_id <- "6034_Islet_001"
    mask_ids <- unique(exprs_df$cell_id[exprs_df$image_id %in% image_id])

    # remove image-id prefix from mask_ids.
    mask_ids <- stringr::str_remove(mask_ids, paste0(image_id, "_"))
    message("Image: ", image_id, " - Number of mask ids: ", length(mask_ids))

    ### Load Images:
    image_extension <- ".tiff"
    image_id_long <- stringr::str_replace(image_id, "Islet_", "Islet_ROI_")
    image_id_full <- paste(image_id_long, image_extension, sep = "")

    # Paths.
    folder_images <- file.path(paths$folder_in, "img", paths$panel_type)
    folder_masks <- file.path(paths$folder_in, "masks_cells", paths$panel_type, "whole-cell")

    # ------ Load the masks and identify pixels of interest (beta-cell pixels).
    masks <- cytomapper::loadImages(file.path(folder_masks, image_id_full), as.is = TRUE)
    mask_image <- imageData(cytomapper::getImages(masks, image_id_long)[[1]])
    nrow(mask_image); ncol(mask_image)
    mask_df <- as.data.frame(which(!is.na(mask_image), arr.ind = TRUE))
    head(mask_df)
    
    # Get the value of the mask at these indices (= mask labels)
    mask_df$label <- mask_image[cbind(mask_df$row, mask_df$col)]

    # Filter by mask-ids and split by label (cell-id)!
    result <- subset(mask_df, label %in% mask_ids)
    X <- split(result, result$label)
    
    # ------- Load the images for the channels of interest.
    images <- cytomapper::loadImages(file.path(folder_images, image_id_full))
    cytomapper::channelNames(images) <- features_oi

    ## Subset the tiff images to the channels of interest.  
    images <- cytomapper::getChannels(images, channels_oi)

    # Add image names to metadata
    mcols(images)$ImageName <- names(images)
    mcols(images)$ImageName <- paste0(mcols(images)$ImageName)

    # Get image data as 3D array and DNA1 image.
    image_data <- imageData(cytomapper::getImages(images, image_id_long)[[1]])
    DNA1_patch <- image_data[, , "DNA1"]

    ## Now, iterate through every mask-id (CELL_LEVEL!)
    ## 1) First, extract the DNA1 patch for each cell, then run a GMM to identify nuclear pixels.
    ## 2) Then, go to the other channels and take their average pixel values. 
    
    average_exp <- purrr::map2(X, names(X), function(x_p, label_X) {
      # x_p <- X[[2]]; label_X <- names(X)[2]
      # Extract all DNA1 pixels of the current cell; then subset to NUCLEAR pixels (pos. for DNA1.)
      patch <- purrr::map2_dbl(x_p$row, x_p$col, ~ image_data[.x, .y, "DNA1", drop = FALSE])

      # For edge-case that DNA1 is 0. (segmentation errors).
      if (sum(patch) == 0) {
        message("No nuclear pixels found for cell: ", label_X, ". Skipping...")
        nuc_pixel_coords <- tibble::tibble(row = x_p$row, 
                                          col = x_p$col, 
                                          label = label_X)
        tot_pixel_coords <- tibble::tibble(row = x_p$row, 
                                          col = x_p$col, 
                                          label = label_X)
        cyto_pixel_coords <- tibble::tibble(row = x_p$row, 
                                          col = x_p$col, 
                                          label = label_X)
      } else {
        gmm <- Mclust(patch, G = 2, verbose = FALSE)
        if (is.null(gmm)) {
          message("GMM failed for cell: ", label_X, ". Skipping...")
          nuc_pixel_coords <- tibble::tibble(row = x_p$row, 
                                            col = x_p$col, 
                                            label = label_X)
          tot_pixel_coords <- tibble::tibble(row = x_p$row, 
                                            col = x_p$col, 
                                            label = label_X)
          cyto_pixel_coords <- tibble::tibble(row = x_p$row, 
                                          col = x_p$col, 
                                          label = label_X)
        } else{
          nuc_pixels <- which(gmm$classification == 2)  
          cyto_pixels <- which(gmm$classification == 1)
          nuc_pixel_coords <- tibble::tibble(row = x_p$row[nuc_pixels], 
                                          col = x_p$col[nuc_pixels], 
                                          label = label_X)
          tot_pixel_coords <- tibble::tibble(row = x_p$row, 
                                            col = x_p$col, 
                                            label = label_X)
          cyto_pixel_coords <- tibble::tibble(row = x_p$row[cyto_pixels], 
                                          col = x_p$col[cyto_pixels], 
                                          label = label_X)
        }
      }

      ## Iterate across channels:
      ## Get average nuclear expression for markers.
      nuc_exp <- purrr::map(dimnames(image_data)[[3]], function(channel) {
        pixels_nuc <- purrr::map2_dbl(nuc_pixel_coords$row, nuc_pixel_coords$col, ~ image_data[.x, .y, channel])
      
        # Average the pixel values for the nuclear pixels.
        mean_value <- mean(pixels_nuc, na.rm = TRUE)
        return(mean_value)
      }) |> unlist() |> setNames(dimnames(image_data)[[3]])

      tot_exp <- purrr::map(dimnames(image_data)[[3]], function(channel) {
        pixels_tot <- purrr::map2_dbl(tot_pixel_coords$row, tot_pixel_coords$col, ~ image_data[.x, .y, channel])
      
        # Average the pixel values for the total pixels.
        mean_value <- mean(pixels_tot, na.rm = TRUE)
        return(mean_value)
      }) |> unlist() |> setNames(dimnames(image_data)[[3]])

      ## For cytoplasmatic expression, we can use the total expression values.
      cyto_exp <- purrr::map(dimnames(image_data)[[3]], function(channel) {
        pixels_cyto <- purrr::map2_dbl(cyto_pixel_coords$row, cyto_pixel_coords$col, ~ image_data[.x, .y, channel])
      
        # Average the pixel values for the cytoplasmatic pixels.
        mean_value <- mean(pixels_cyto, na.rm = TRUE)
        return(mean_value)
      }) |> unlist() |> setNames(dimnames(image_data)[[3]])
    

      ## Creat the return-tibble with the average expression values.
      nuc_df <- tibble(channel = names(nuc_exp), values = nuc_exp) |> 
        mutate(label = label_X)

      tot_df <- tibble(channel = names(tot_exp), values = tot_exp) |> 
        mutate(label = label_X)
      
      cyto_df <- tibble(channel = names(cyto_exp), values = cyto_exp) |> 
        mutate(label = label_X) 

      list(nuclear = nuc_df, total = tot_df, cyto = cyto_df)
    })

    # Combine all "nuclear" into a list
    nuclear_list <- purrr::map(average_exp, "nuclear") |> bind_rows()
    total_list <- purrr::map(average_exp, "total") |> bind_rows()
    cyto_list <- purrr::map(average_exp, "cyto") |> bind_rows()
    
    ## Write nuclear expression values to a .csv-file.
    nuclear_list <- nuclear_list |> 
      mutate(image_id = image_id) |> 
      pivot_wider(names_from = channel, values_from = values)

    fn <- file.path(paths$folder_in, "nuclear_data_cells", paste0(image_id, ".csv"))
    write.csv(nuclear_list, file = fn, row.names = FALSE)

    ## Write total expression values to a .csv-file.
    total_list <- total_list |> 
      mutate(image_id = image_id) |> 
      pivot_wider(names_from = channel, values_from = values)
    
    fn_tot <- file.path(paths$folder_in, "total_data_cells", paste0(image_id, ".csv"))
    write.csv(total_list, file = fn_tot, row.names = FALSE)
    ## Write cytoplasmatic expression values to a .csv-file.
    cyto_list <- cyto_list |> 
      mutate(image_id = image_id) |> 
      pivot_wider(names_from = channel, values_from = values)

    fn_cyto <- file.path(paths$folder_in, "cyto_data_cells", paste0(image_id, ".csv"))
    write.csv(cyto_list, file = fn_cyto, row.names = FALSE)
  }) 
# , .options = options)
}
```

## A2.2: Save and Load the full data.
Read in the nuclear expression data for each image.
```{r read-expression-values}
compartments <- c("nuclear", "cyto", "total")

if (calc_nuc_cyto_tot){
  df <- purrr::map(compartments, \(.x){
    # Paths.
    path <- file.path(paths$folder_in, paste0(.x, "_data_cells"))
    files <- list.files(path, pattern = "\\.csv$", full.names = TRUE)
    
    ## Read .csv files in parallel.
    df_list <- furrr::future_map(files, \(f) readr::read_csv(f), .options = options)

    ## Reshape. Add caseID + cellID.
    exprs_df_nuc <- bind_rows(df_list) |> 
      mutate(cell_id = paste(image_id, label, sep = "_")) |> 
      select(-label) |> 
      mutate(case_id = stringr::str_extract(image_id, "^[0-9]+")) |>
      dplyr::inner_join(case_stages, by = "case_id")
    
    # Save.
    fn <- file.path(paths$folder_script, paste0(.x, "_expression.rds"))
    saveRDS(exprs_df_nuc, file = fn)
    #}
    exprs_df_nuc
  }) |> setNames(compartments)
} else{
   df <- purrr::map(compartments, \(.x){
    fn <- file.path(paths$folder_script, paste0(.x, "_expression.rds"))
    readRDS(file = fn)
  }) |> setNames(compartments)
}

exprs_df_nuc <- df[["nuclear"]]
exprs_df_cyto <- df[["cyto"]]
exprs_df_tot <- df[["total"]]
```

## A2.3: Generate a joint dataframe with nuclear, cytoplasmatic and total expression values.
Join the nuclear, cytoplasmatic and total expression values for each channel
into one joint dataframe.
Each row = 1 cell.

```{r generate-joint-dataframe}
### Compare across stages
exprs_df_tot <- exprs_df_tot |> 
  rename_with(~ paste0(.x, "_tot"), -c(image_id, cell_id, case_id, donor_type))

joint_df <- exprs_df_nuc |>
    inner_join(exprs_df_cyto, by = c("image_id", "cell_id", "case_id", "donor_type"), suffix = c("_nuc", "_cyto")) |> 
    inner_join(exprs_df_tot, by = c("image_id", "cell_id", "case_id", "donor_type"))

final_df <- joint_df |> 
  select(-starts_with("age")) |>
  pivot_longer(c(ends_with("_nuc"), ends_with("_cyto"), ends_with("_tot")),
               names_to = c("channel"),
               values_to = "count")
final_df
gc()
```

## A2.4: ALL: Plot the subcellular expression for each channel.
Plot the subcellular expression for each channel.
Plot for nuclear, cytoplasmatic and total expression.

```{r plot-subcellular-expression}
p <- final_df |> 
  # remove the _nuc, _cyto, _tot suffixes.
  mutate(channel2 = stringr::str_remove(channel, "_nuc|_cyto|_tot")) |>
  # keep either _nuc, _cyto or _tot
  mutate(channel = stringr::str_extract(channel, "_nuc|_cyto|_tot")) |>
  # transform counts:
  mutate(count = asinh(count)) |> 
  # by image-id
  group_by(donor_type, channel2, image_id, case_id, channel) |> 
  summarise(count = mean(count, na.rm = TRUE), .groups = "drop") |>
  # by case-id
  group_by(donor_type, channel, channel2, case_id) |> 
  summarise(count = mean(count, na.rm = TRUE), .groups = "drop") |>
  # plot.
  ggplot(aes(x = channel, y = count, fill = donor_type)) +
  geom_boxplot() +
  labs(x = "Channel", y = "Mean Nuclear Expression") +
  mytheme$standard_new() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) + 
  facet_wrap(~ channel2, scales = "free_y")

fn <- paste0(today, "_subcellular_expression_boxplot.png")
do.call(ggsave, c(list(fn, p), plotsave_param_large))
print(p)
```

## A2.5: LMMs: subcellular expression. 

### A2.5.1: Calculate P-values for subcellular expression.
Test, if taking the cyto/nuclear expression and calculating P-values makes a difference. 
-> No major differences observed.
```{r}
res_lmm <- final_df |> 
  filter(!case_id %in% to_del) |>
  # take Islet-average.
  group_by(channel, image_id, case_id, donor_type) |> 
  summarise(count = mean(count, na.rm = TRUE), .groups = "drop") |> 
  tidyr::nest(data = -channel) |> 
  mutate(test = purrr::map(data, ~ broom.mixed::tidy(lmerTest::lmer(data = ., 
    count ~ donor_type + (1 | case_id))))
  ) |> 
  unnest(test) |> 
  filter(term != "(Intercept)", effect == "fixed")

res_lmm |> 
  arrange(p.value) |> 
  print(n = "all")
```

### A2.5.2: Plot LMMs P-values for subcellular expression.

```{r}
p <- res_lmm |> 
  mutate(donor_type = stringr::str_remove(term, "donor_type")) |>
  mutate(donor_type = factor(donor_type, 
    levels = c("NoDiabetes", "sAAb+", "mAAb+", "RecentOnset"))) |>
  mutate(direction = ifelse(estimate > 0, "positive", "negative")) |>
  mutate(channel2 = stringr::str_remove(channel, "_nuc|_cyto|_tot")) |> 
  mutate(channel3 = stringr::str_extract(channel, "_nuc|_cyto|_tot")) |>
  ggplot(aes(x = reorder(channel2, p.value), y = -log10(p.value), fill = channel3)) +
  geom_bar(stat = "identity", position = "dodge") +
  labs(x = "Channel", y = "-log10(p-value)", fill = "") +
  mytheme$standard_new() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) + 
  facet_wrap(~ donor_type, ncol = 2, scales = "free_y", axes = "all")
print(p)
fn <- paste0(today, "_subcellular_expression_pvalues.png")
do.call(ggsave, c(list(fn, p), plotsave_param_large))
```


## A2.6: Calculate vs numeric stages.

Barplot of the p-values for the nuclear vs cytoplasmatic expression.
There is basically no difference between nuclear and cytoplasmatic expression for most channels.

## A 2.7: Calculate the fraction in nuclear and cytoplasmatic expression.

```{r delta-expression-nuc-cyto}
delta_exp <- joint_df |>
  select(image_id, cell_id, case_id, donor_type, 
         ends_with("_nuc"), ends_with("_cyto"), 
         ends_with("_tot"), -starts_with("age")) |>
  mutate(across(
    .cols = ends_with("_nuc"),
    .fns = \(nuc_val, .cols = cur_column()) {
      marker <- stringr::str_remove(.cols, "_nuc")
      cyto_col <- paste0(marker, "_cyto")
      if (cyto_col %in% names(joint_df)) {
        return(nuc_val/(nuc_val + joint_df[[cyto_col]]))
      } else {
        return(NA_real_)
      }
    },
    .names = "{.col}_frac_cyto"
  )) |> 
  select(image_id, cell_id, case_id, donor_type, 
         ends_with("_frac_cyto"), ends_with("_nuc"), ends_with("_tot"))
```

```{r plot-delta-expression-nuc-cyto}
p <- delta_exp |> 
  pivot_longer(
    cols = ends_with("_nuc_frac_cyto"),
    names_to = "channel",
    values_to = "delta_expression"
  ) |>
  group_by(channel, image_id, case_id, donor_type) |> 
  summarise(mean_exp = mean(delta_expression, na.rm = TRUE), .groups = "drop") |>
  group_by(case_id, channel, donor_type) |> 
  summarise(mean_exp = mean(mean_exp, na.rm = TRUE), .groups = "drop") |>
  # transform counts:
  mutate(mean_exp = asinh(mean_exp)) |>
  mutate(channel = stringr::str_remove(channel, "_nuc_frac_cyto")) |>
  ggplot(aes(x = channel, y = mean_exp, fill = donor_type)) +
  geom_boxplot() +
  # geom_jitter(width = 0.2, alpha = 0.5, dodge = TRUE) +
  labs(x = "Channel", y = "Frac Nuclear Expression") +
  mytheme$standard_new() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

fn <- paste0(today, "_delta_expression_boxplot.png")
do.call(ggsave, c(list(fn, p), plotsave_param_large))
print(p)
```


```{r}
palettes$stages2 <- palettes$stages
names(palettes$stages2) <- rev(c("Onset", "mAAb+", "sAAb+", "Control"))

p <- delta_exp |>
  pivot_longer(
    cols = ends_with("_nuc_frac_cyto"),
    names_to = "channel",
    values_to = "delta_expression"
  ) |>
  filter(channel %in% c("XBP1_nuc_frac_cyto", "ProINS_nuc_frac_cyto", "DNA1_nuc_frac_cyto")) |>
  group_by(channel, image_id, case_id, donor_type) |> 
  summarise(mean_exp = mean(delta_expression, na.rm = TRUE), .groups = "drop") |>
  group_by(case_id, channel, donor_type) |> 
  summarise(mean_exp = mean(mean_exp, na.rm = TRUE), .groups = "drop") |>
  # transform counts:
  mutate(mean_exp = asinh(mean_exp)) |>
  mutate(channel = stringr::str_remove(channel, "_nuc_frac_cyto")) |>
  mutate(channel = factor(channel, levels = c("DNA1", "ProINS", "XBP1", "ECad"))) |>
  mutate(donor_type = case_when(donor_type == "NoDiabetes" ~ "Control",
                                donor_type == "sAAb+" ~ "sAAb+",
                                donor_type == "mAAb+" ~ "mAAb+",
                                donor_type == "RecentOnset" ~ "Onset")) |>
  mutate(donor_type = factor(donor_type, levels = c("Control", "sAAb+", "mAAb+", "Onset"))) |> 
  ggplot(aes(x = channel, y = mean_exp, fill = donor_type)) +
  geom_boxplot(outliers = FALSE) +
  geom_hline(yintercept = 0.5, linetype = "dashed", color = "black") +
  geom_jitter(position = position_jitterdodge(jitter.width = 0.5, dodge.width = 0.75)) +
  labs(x = expression(beta ~ "-cell marker"), y = "Nuclear fraction of normalized counts") +
  mytheme$standard_new() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) + 
  scale_fill_manual(values = palettes$stages2, name = "Stage") + 
  theme(legend.position = "top")

fn <- paste0(today, "_delta_expression_boxplot_selected.png")
do.call(ggsave, c(list(fn, p), plotsave_param))
print(p)

saveRDS(p, file.path(paths$home_git, "figures", "rev_supplfig4b.rds"))
```

-> Nothing really stands out.
Note: no change for XBP1.

Had calculated the p-values for cyto-nuc before -> nothing that stood out.