# A0: Aims and Summary:
Here, we show how to access, visualize and analyze the Steenbuck-Damond et al. 2026 IMC datasets. 

```{r}
library(SpatialExperiment)
library(cytomapper)
```

# A1: Data Access:

We suggest to use the `imcdatasets` package to access the SingleCellExperiment (SCE) and mask data.
Per panel, we provide one set of images, masks, and SCE objects.
And suggest to use the `Zenodo` entry (linked from the publication) to download all other data (images, models, raw data).

## Pancreatlas:

An informative subset of the data can be visualized in the Pancreatlas web portal (https://www.pancreatlas.org).
Pancreatlas provides a user-friendly interface for visualizing annotated images from the two IMC panels.
Pancreatlas allows the user to visualize the "Islet" and "Immune" panel side-by-side, which is useful for jointly visualizing and analyzing the 79 unique measured protein markers.  
Data can be filtered via distinct annotations, such as "Insulitic". 

Pancreatlas uses the same `image_id` as the SCE and image/mask data, which allows for easy linking of the data across the different platforms.

## Zenodo:

Full Data (masks, images, models, processed SCE objects, raw data) are available from:
- Zenodo: https://zenodo.org/record/14968076 and linked repositories.

## imcdatasets: 

The full data (masks, SCE object) or a subset (masks, images, SCE objects) can be accessed via the `imcdatasets` package.

NOTE: The data is currently being added to `imcdatasets` and uploaded to `ExperimentHub`. 
Until then, please pull the data from this Zenodo repository (https://zenodo.org/uploads/21371432).
This is the SAME data as will be available via `imcdatasets`.

Data does not differ strongly from the original Zenodo upload, 
but additional metadata columns are included and naming conventions have been updated. 

```{r}
# require(devtools)
# devtools::install_github("BodenmillerGroup/imcdatasets")
```

### Import function

Function to download and read RDS files.

```{r import-function}
importAndReadData <- function(url, output_dir, filename) {
  # Create full path
  file_path <- file.path(output_dir, filename)
  # Download the RDS file
  if (!file.exists(file_path)) {
    download.file(url, destfile = file_path, mode = "wb")
  }
  # Read and return the RDS object
  readRDS(file_path)
}
```

### Download SCEs


```{r dataset-version, echo=FALSE}
dataset_name <- "SteenbuckDamond_2026_Pancreas"
dataset_version <- "v1"
cat("Dataset version:", dataset_version)
```

Setting the working and output directories

```{r directories}
# Temporary directory to unzip files
# workdir <- tempdir()
workdir <- getwd()
Sys.setenv(workdir = workdir)

# Output directory
dataset_dir <- file.path(".", dataset_name)
if(!(dir.exists(dataset_dir))) dir.create(dataset_dir)

outdir <- file.path(dataset_dir, dataset_version)
if(!(dir.exists(outdir))) dir.create(outdir)

# Increase timeout period so that large files can be downloaded
timeout <- getOption('timeout')
options(timeout = 1000)
```

Note: this downloads the SUBSET of the SCEs.

```{r download-sce}
# Download SCE SUBSET.
url_sce_islet <- "https://zenodo.org/records/21371432/files/sce_islet.rds"
url_sce_immune <- "https://zenodo.org/records/21371432/files/sce_immune.rds"

# FULL SCEs:
# url_sce_islet <- "https://zenodo.org/records/21240077/files/sce_full_islet.rds"
# url_sce_immune <- "https://zenodo.org/records/21240077/files/sce_full_immune.rds"

url_sces <- c(url_sce_islet, url_sce_immune)
names(url_sces) <- c("islet", "immune")

sce_list <- purrr::imap(url_sces, function(url, name) importAndReadData(url, workdir, paste0("sce_", name, ".rds")))
```

# A2: Data 

## A2.1: Markers

Per panel, we measured 47 markers, including 2 DNA markers.
Note: PD-L1 was not specific and was not analyzed. 

```{r markers}
(rownames(sce_list[["islet"]]))
(rownames(sce_list[["immune"]]))
```

Some cell-type expressed markers are given in the SCE metadata (not exhaustive).

```{r channels}
metadata(sce_list[["islet"]])$channels
metadata(sce_list[["immune"]])$channels
```

## A2.2: Cell Categories, Cell Types and Cell States

Cell Type Categories are available via the `cell_category` column in the `colData` of the SingleCellExperiment object 
and distinguishes major cell categories like `Endocrine`, `Exocrine`, `Immune`, `Mesenchymal` and `Endothelial`.

Cell Type Annotation Information is available via the `cell_type` columns in the `colData` of the SingleCellExperiment object.

```{r celltype-categories}
(table(colData(sce_list[["islet"]])$cell_category))
(table(colData(sce_list[["immune"]])$cell_category))

(table(colData(sce_list[["islet"]])$cell_type))
(table(colData(sce_list[["immune"]])$cell_type))
```

Cell Subclustering for key Immune cells Information is available via the `cell_cluster_scaled` column in the `colData` of the SingleCellExperiment object
of the `Immune` panel. Names of the subclusters are available via the `myeloid_labels`, `tcd4_labels` and `tcd8_labels` objects.

```{r immune-cell-subclusters}
table(colData(sce_list[["immune"]])$cell_cluster_scaled)

# Myeloid cells: Macs + cDCs.
myeloid_labels <- c(
  "Myeloid_1" = "Exocrine low act.", "Myeloid_2" = "M1/M2-like low act.",
  "Myeloid_3" = "M1/M2-like act.", "Myeloid_4" = "cDC",
  "Myeloid_5" = "CD11c+", "Myeloid_6" = "Exocrine act.",
  "Myeloid_7" = "Exocrine MPO+", "Myeloid_8" = "CD54+", "Myeloid_Other" = "Ambiguous"
)

# T_CD4.
tcd4_labels <- c(
  "T_CD4_1" = "CD4+ Other",  "T_CD4_2" = "T-EM-like low act.",
  "T_CD4_3" = "PD1+ low act.", "T_CD4_4" = "T-CM-like act.",
  "T_CD4_5" = "PD1+ act.", "T_CD4_6" = "T-EM-like act.",
  "T_CD4_7" = "PD1low act.", "T_CD4_8" = "PD1low low act.", 
  "T_CD4_9" = "CD73+", "T_CD4_10" = "T-reg",
  "T_CD4_Other" = "Ambiguous"
)

# T_CD8.
tcd8_labels <- c(
  "T_CD8_1" = "T-CD8 Other", "T_CD8_2" = "T-RM low act.",
  "T_CD8_3" = "GranB+", "T_CD8_4" = "T-EM/CM-like",
  "T-naive" = "T-naive", "T_CD8_6" = "T-RM act.",
  "T_CD8_7" = "T-exeff", "T-EMRA" = "T-EMRA"
)
```

## A2.3: Cell-specific annotations

### A2.3.1: Distance to islet

The distance of a cell to the nearest islet is available as `cell_distance_to_islet` column in the `colData`.
A negative distance indicates that the cell is outside the islet, while a positive distance indicates that the cell is located outside of an islet.
A distance of `-20` indicates that the cell is located 20 µm **away** from the islet.
A distance of `20` indicates that the cell is located 20 µm **inside** the islet.

```{r distance-to-islet}
summary(colData(sce_list[["islet"]])$cell_distance_to_islet)
summary(colData(sce_list[["immune"]])$cell_distance_to_islet)
```

### A2.3.2: Infiltration Score.

The distance to the nearest islet is used to compute the the infiltration score. 
This weights abundance and distance to islet for each immuen cell. 
It is available both for Cell Types and Cell Subclusters in the `Immune Panel`.
The "NormScoreArea" is the islet-area normalized infiltration score, as shown in the manuscript.

```{r infiltration-score}
metadata(sce_list[["immune"]])$celltype_compo_score |> 
  tibble::as_tibble() |> 
  dplyr::select(dplyr::starts_with("NormScoreArea"))

metadata(sce_list[["immune"]])[["cluster_compo_score"]][[1]] |> 
  tibble::as_tibble() |> 
  dplyr::select(dplyr::starts_with("NormScoreArea"))
```

The `image_id` can be used to link the infiltration score to the respective expression data.

### A2.3.3: Pseudotime-Beta cells.
The pseudotime of a beta cell is available as the `cell_pseudotime_beta` column in the `colData`.
This is only derived for beta-cells on the Islet Panel.
The pseudotime was derived using slingshot and indicates the relative progression of beta-cells along predicted T1D progression.

```{r pseudotime_beta}
colData(sce_list[["islet"]])[, c("cell_type", "cell_pseudotime_beta")] |> 
    tibble::as_tibble() |> 
    dplyr::filter(cell_type == "Beta") |>
    dplyr::select(cell_pseudotime_beta)
```

## A2.4: Islet-level Information

If one is interested in analyzing islet-level information, for example to analyze structural integrity of islets, 
one can use the `islet_id` reflective of the islet masks available at `Zenodo`. 
There are more islet-ids on the Immune panel than on the Islet panel, as the Immune panel has an expanded FOVs (around 50 µm) and more islets are thus captured.

```{r islet-ids}
length(unique(colData(sce_list[["islet"]])$islet_id))
length(unique(colData(sce_list[["immune"]])$islet_id))
```



## A2.5: ROI types

Biologically important ROI types.
These can be used to refine disease staging of islets. 
IDIs are (if they experienced beta-cell loss) further in disease progression, than ICIs. 
Insulitic ROIs are usually further in disease progression than non-insulitic ROIs (if they are both ICIs).

### A2.5.1: Insulitic ROIs.
Insulitic ROI annotation are available via the 'image_insulitic' column in the `colData` of the SingleCellExperiment object.
They were defined using the `Immune Panel`.

ICI vs IDIs (Insulitic containing vs deficient Islets) were defined using the `Islet Panel` and are available via the 'image_ici_idi' column in the `colData` of the SingleCellExperiment object.

```{r insulitic-ici-rois}
colData(sce_list[["immune"]]) |> 
    tibble::as_tibble() |> 
    dplyr::group_by(image_id, patient_donor_type) |> 
    dplyr::summarise(insultic = unique(image_insulitic), 
                     n_cells = dplyr::n(),
                     ici_idi = unique(image_ici_idi), .groups = "drop")
```

## A2.6: Niche Information:

Spatial Tissue domains were only derived for the Immune panel. They are available via the 
`CN_cn_k1070` column in the `colData` of the SingleCellExperiment object.

```{r niche-information}
table(colData(sce_list[["immune"]])[["CN_cn_k1070"]])
```

## A2.7: Metadata and Patient-Information. 

General patient information and metadata is available in all columns with a "patient_" prefix. 
Metadata of main interest are:
`patient_case_id`, `patient_age`, `patient_donor_type`, `patient_diabetes_duration`.
```{r metadata-information}
colnames(colData(sce_list[["islet"]])) |> 
  stringr::str_subset("patient_")
```

# A3: Masks

Whole-cell segmentation masks are available via `imcdatasets` (a subset and all masks) or via `Zenodo`.
and can be read into a CytoImageList.

## Download and unzip cell segmentation masks.
```{r download-cell-masks}
url_masks <- "https://zenodo.org/records/21371432/files/masks_islet.rds"
dest_dir <- file.path(workdir, "masks_islet.rds")

if (!file.exists(dest_dir)) {
  download.file(url_masks, destfile = dest_dir, mode = "wb")
}

# IMMUNE PANEL
url_masks <- "https://zenodo.org/records/21371432/files/masks_immune.rds"
dest_dir <- file.path(workdir, "masks_immune.rds")

if (!file.exists(dest_dir)) {
  download.file(url_masks, destfile = dest_dir, mode = "wb")
}
````

```{r create-cytoimagelist-masks}
masks_islet <- readRDS(file.path(workdir, "masks_islet.rds"))
masks_immune <- readRDS(file.path(workdir, "masks_immune.rds"))
```


# Images:
Processed images (.tiffs):

Data Subset are available in the following repositories:

- Islet + Immune panel: https://zenodo.org/records/21371432
- Full Image Data (.tiffs) are available via linked repositories (https://zenodo.org/record/14968076).


```{r images-islet}
dest_dir <- file.path(workdir, "images_islet.rds")

if (!dir.exists(dest_dir)) {
  url_img <- "https://zenodo.org/records/21371432/files/images_islet.rds"
  download.file(url_img, destfile = dest_dir, mode = "wb")
}

images_islet <- readRDS(dest_dir)
```

```{r images-immune}
dest_dir <- file.path(workdir, "images_immune.rds")

if (!dir.exists(dest_dir)) {
  url_img <- "https://zenodo.org/records/21371432/files/images_immune.rds"
  download.file(url_img, destfile = dest_dir, mode = "wb")
}

images_immune <- readRDS(dest_dir)
```

Visualizing images. 
Images can be visualized using the `cytomapper` package.
Alternatively, images can be visualized using `cytoviewer` or `napari` (Python).

```{r cytomapper}
if(length(setdiff(names(images_islet), names(masks_islet))) > 0 ) {
  stop("Some images do not have a corresponding mask. Please check the data.")
}

idx <- 10

cytomapper::plotPixels(image = images_islet[idx], 
                       object = sce_list[["islet"]], 
                       mask = masks_islet[idx], 
                       img_id = "image_name",
                        colour_by = c("SYP", "KRT19", "CD163"),
                        cell_id = "cell_number",
                        outline_by = "cell_type", 
                        bcg = list("SYP" = c(0, 2, 1),
                                     "KRT19" = c(0, 5, 1),  
                                     "CD163" = c(0, 5, 1)),
                        save_plot = list(filename = file.path(workdir, "image_islet.png"),
                                             scale = 1)
)
```
